### 2017-06-25 docs/manuscript.tex
writing manuscript

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Towards the construction of a causal map of the human phenome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{amsmath}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Towards the construction of a causal map of the human phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{\texttt{r\ format(Sys.time(),\ \textquotesingle{}\%d\ \%B\ \%Y\textquotesingle{})}}

\begin{document}
\maketitle

\subsection{Introduction}\label{introduction}

Mendelian randomisation (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is known as an instrumental variable
for the exposure, and can be exploited to mimic a randomised controlled
trial, making causal inference by comparing the outcome phenotypes
between those individuals that have the exposure-increasing allele
against those who do not. Multiple independent genetic variants for a
particular exposure can be used jointly to improve causal inference, the
premise being that each variant is an independent natural experiment,
and an overall causal estimate can be obtained by meta-analysing the
single estimates from each instrument.

Genome-wide association studies (GWAS) have identified genetic
instrumental variables for thousands of phenotypes. Recent developments
in Mendelian randomisation have enabled knowledge of instrumental
variables to be applied using only summary level data (known as
two-sample MR, 2SMR). Here, in order to infer the causal effect of an
exposure on an outcome all that is required is an estimate of the
genetic effects of the instrumenting SNP on the exposure, and the
corresponding estimate of the effect on the outcome. This has two major
advantages. First, GWAS summary data is non-disclosive and often
publicly available. Second, causal inference can be made between
phenotypes even if they have not been measured in the same samples,
limiting the breadth of possible causal estimates only to the
availability of GWAS summary data for the traits in question.

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure. The
extent of this problem is not to be understated, and many methods have
been developed that attempt to reliably obtain unbiased causal estimates
under specific models of HP. It is considered best practice to report
estimates from all available methods as sensitivity analyses when
presenting causal estimates, however this strategy is not necessarily
optimal for several reasons. First, if different methods disagree it is
not possible to know which is correct because the appropriate model of
HP is not known. Second, though the IVW approach is most statistically
powerful under no HP, it can have high false negative or low true
positive rates in the presence of HP compared to other methods. Given
that pleiotropy has been hypothesised to be universal, defaulting to the
IVW method in the first instance and using other methods as sensitivity
analyses may not be appropriate. Third, the available methods do not
cover all possible models of HP, and therefore an automated method for
instrument selection may be necessary. Fourth, it could be of interest
to make causal effect estimates for thousands of traits, in which case a
discerning evaluation of each causal effect of interest may not be
possible or convenient.

In this paper we introduce new machine learning approaches that attempt
to automate both instrument and method selection. Using curated GWAS
summary data for thousands of phenotypes, we use these new methods to
construct a graph of millions of causal estimates.

\subsection{Methods}\label{methods}

\subsubsection{GWAS summary data and their use in
2SMR}\label{gwas-summary-data-and-their-use-in-2smr}

\subsubsection{Mendelian randomisation methods and their
assumptions}\label{mendelian-randomisation-methods-and-their-assumptions}

In this paper we consider three main classes of MR estimation. Full
details for each approach have been described extensively elsewhere.

\textbf{Mean-based methods:} The inverse variance weighted (IVW)
meta-analysis approach assumes that variant exhibits no HP (fixed
effects meta-analysis) or that HP is present but balanced (random
effects meta-analysis). Egger regression relaxes the HP assumption
further by allowing the horizontal pleiotropy to systematically occur in
a specific direction, known as directional horizontal pleiotropy. The
Rucker framework uses estimates of heterogeneity to navigate between
these nested models. A jackknife approach (random selection with
replacement of instruments) can be used to obtain a sampling
distribution for the model estimate amongst these four variations.

The Rucker framework is a system used in meta-analysis to navigate
between IVW and Egger regression, and fixed and random effects models.
It was recently applied to MR to use heterogeneity statistics to select
the most appropriate model of pleiotropy, and therefore the most
appropriate MR method amongst the four mean-based estimators. There

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments. This has the advantage that
up to half the instruments can be invalid, and the estimate will remain
unbiased. Developing the approach further to allow stronger instruments
to contribute more towards the estimate can be obtained by obtaining the
median of the weights of each instrument. The penalised weighted median
estimator \ldots{}

\textbf{Mode-based methods:} Supposing that

\subsubsection{Instrument selection}\label{instrument-selection}

\paragraph{Top hits}\label{top-hits}

The simplest approach to selecting instruments for performing MR is to
take take SNPs that have been declared significant in the published GWAS
for the exposure. This typically involves obtaining SNPs that surpass
\(p < 5 \times 10^{-8}\), using clumping to obtain independent SNPs, and
then replicating in an independent sample. These results are often
recorded in public GWAS catalogs. Alternatively the clumping procedure
can be performed using complete summary data in MR-Base. We call this
the ``top hits'' strategy.

\paragraph{Steiger filtering}\label{steiger-filtering}

With genome-wide association studies growing ever larger, the
statistical power to detect significant associations that may be
influencing the trait downstream of many other pathways increases. For
example, if a SNP \(g_{A}\) influences trait \(A\), and trait \(A\)
influences trait \(B\), then a sufficiently powered GWAS will identify
the \(g_{A}\) as being significant for trait \(B\) (Figure 1a). Using
\(g_{A}\) as an instrument to test the causal effect of \(A\) on \(B\)
is perfectly valid. But in the (incorrectly hypothesised) MR analysis of
trait \(B\) on trait \(A\) could erroneously result in the apparent
causal association of \(B\) on \(A\). If \(g_{A}\) is only one of many
known instruments for \(B\), amongst which some are valid, it is to the
advantage of the researcher to exclude \(g_{A}\) from the analysis.

An approach to inferring the causal direction between phenotypes was
developed recently, using the following basic premise. If trait \(A\)
causes trait \(B\) then

\[
\sum^M_{i=1}{cor(g_{i}, A)^2} > \sum^M_{i=1}{cor(g_{i}, B)^2}
\]

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Steiger's Z-test of correlated correlations can be used to
formally test the extent to which the two correlations are statistically
different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). In this case the Steiger test applied to each
variant in turn will identify \(g_{A}\) as being unlikely to primarily
associate with \(B\) relative to \(A\). Similarly, for SNPs that
influence confounders of \(A\) and \(B\) or those variants that exhibit
horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
\(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
SNP being excluded because the Steiger Z-test is less likely to be
significant.

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then our
objective is to estimate the variance of the risk liability explained by
the SNP, \(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here,
\(V_e = \pi^2/3\), and \(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the
log odds ratio and \(p\) is the allele frequency of the SNP in the
population. \(p\) can be estimated using the allele frequency of the SNP
in an ascertained sample by deriving the \(2 \times 2\) contingency
table from the odds ratio \(e^\beta\), allele frequency in the
ascertained sample \$p\_\{cc\}, and number of cases \(N_1\) and controls
\(N_0\).

\subsubsection{Competitive mixture of
experts}\label{competitive-mixture-of-experts}

We consider 14 MR methods, for which instruments can be supplied using
two instrument selection strategies, leading to 28 methods in total. In
the context of this analysis, each method is considered to be an
`expert', taking a set of SNP-exposure and SNP-outcome effect sizes and
their standard errors as inputs. Our objective is to select the expert
most likely to be correct for a specific MR analysis.

\paragraph{Mixture of experts}\label{mixture-of-experts}

The mixture of experts (MoE) method is a machine learning approach which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations.

\textbf{Training and testing simulations}

The MoE is trained using datasets generated from simulations. A
\emph{dataset} is the minimum data required to perform 2SMR analysis -
four columns comprising the SNP-exposure and SNP-outcome effects and
standard errors, and rows corresponding to each SNP that is used as an
instrument for the exposure. This \emph{dataset} can be fed into any of
the 28 experts to obtain MR causal effects. The simulations used to
generate these datasets seek to cover a range of pleiotropic scenarios,
including where some proportion of SNPs exhibit directional or balanced
horizontal pleiotropy, or where SNPs influencing confounding variables.

We simulate two individual level datasets for which there are \(N_x\)
and \(N_y\) samples, and \(M\) SNPs, where each SNP has effect allele
frequency of \(p_m \sim U(0.05, 0.95)\). These datasets are used to
obtain the SNP effects for the exposure trait \(x\) and the outcome
trait \(y\), respectively, using the following sampling criteria:

\[
\begin{aligned}
N_x & = \{20000, ..., 500000\} \\
N_y & = \{20000, ..., 500000\} \\
K & = \{0, ..., 10\} \\
M_x & = \{1, ..., 200\} \\
M_y & = \{1, ..., 200\} \\
M_{u_k} & = \{5,..., 30\} \\
\end{aligned}
\]

The \(M = M_x + M_y + \sum{M_{u_k}}\) SNPs can influence \(x\) directly,
\(y\) directly, or some number of confounders \(u_{k}\) directly.
Phenotypes for \(x\) and \(y\) are constructed using

\[
x = \sum^{M_x}_{i}{\beta_{gx,x,i}g_{x,i}} + \sum^{M_y}_{j}{\beta_{gy,x,j}g_{y,j}} + \sum^{K}_{k}{\beta_{ux,k} u_{k}} + e_{x}
\]

where \(\beta_{gx,x}\) is the vector of effects of each of the \(M_x\)
SNPs that influence \(x\) primarily, \(\beta_{gy,x}\) is the vector of
effects for the \(M_y\) SNPs on \(x\), where the \(M_y\) SNPs influence
\(y\) primarily but exhibit horizontal pleiotropic effects on \(x\). We
allow some proportion of these effects to be 0. \(\beta_{ux}\) is the
vector of effects of each of the \(K\) confounders on \(x\). Each
\(u_{k}\) variable is constructed using

\[
u = \sum^{M_u}_{l}{\beta_{gu,l}g_{l}} + e_{l}
\]

and finally \(y\) is constructed using

\[
y = \beta_{x,y}x + \sum^{M_y}_{i}{\beta_{gy,y,j}g_{y,j}} + \sum^{M_x}_{j}{\beta_{gx,y,i}g_{x,i}} + \sum^{K}_{k}{\beta_{uy,k} u_{k}} + e_{y}
\]

where \(\beta_{x,y}\) is the causal effect of \(x\) on \(y\). We sample
the distribution of direct SNP effects using

\[
\begin{aligned}
\beta_{gx,x,i} & \sim N(0, \sigma^2_{gx,x}) \\
\sigma^2_{gx,x,i} & \sim U(0.01, 0.1) \\
\beta_{gy,y,j} ~ N(0, \sigma^2_{gy,y}) \\
\sigma^2_{gy,y,j} & \sim U(0.01, 0.1) \\
\end{aligned}
\]

Some proportion \(s_x \sim U(0,1)\) of \(g_x\) SNPs and some proportion
\(s_y \sim U(0,1)\) of \(g_y\) SNPs exhibit horizontal pleiotropy with
effects sampled using

\[
\begin{aligned}
\beta_{gx,y,i*} & \sim N(\mu_{gx,y}, \sigma^2_{gx,y}) \times  \\
\mu_{gx,y,i*} & \sim U(-0.005, 0.005) \\
\sigma^2_{gx,y,i*} & \sim U(0.001, 0.01) \\
\beta_{gy,x,j*} & \sim N(\mu_{gy,x}, \sigma^2_{gy,x}) \\
\mu_{gy,x,j*} & \sim U(-0.005, 0.005) \\
\sigma^2_{gy,x,j*} & \sim U(0.001, 0.01) \\
\end{aligned}
\]

The genetic influences on each of the confounders are sampled using

\[
\begin{aligned}
\beta_{gu,u,l} & \sim N(0, \sigma^2_{gu,u}) \\
\sigma^2_{gu,u,l} & \sim U(0.01, 0.1) \\
\end{aligned}
\]

The influence of each confounder on \(x\) and \(y\) is obtained using

\[
\begin{aligned}
\beta_{u,x} & \sim N(0, \sigma^{2}_{u,x}) \\
\beta_{u,y} & \sim N(0, \sigma^{2}_{u,y}) \\
\end{aligned}
\]

Finally, 20\% of the simulations have a null effect of
\(\beta_{x,y} = 0\), while the other remaining 80\% have a true effect
sampled from

\[
\begin{aligned}.
\beta_{x,y} & \sim N(0, \sigma^2_{x,y}) \\
\sigma^2_{x,y} & \sim U(0.001, 0.1) \\
\end{aligned}
\]

For each simulation we used linear regression to estimate the genetic
effect of each SNP \(M\) on \(x\) in sample 1, and each SNP \(M\) on
\(y\) in sample 2. We then perform MR analysis in both directions,
retaining SNPs that have \(p < 5e-8\) in sample 1 to perform MR of \(x\)
on \(y\) (the true causal direction for non-null simulations), and
retaining SNPs that have \(p < 5e-8\) in sample 2 to perform MR of \(y\)
on \(x\) (the reverse causal direction for non-null simulations). We
treat the summary data (effect sizes and standard errors) used for
estimating \(x \rightarrow y\) the summary data used for estimating
\(y \rightarrow x\) as two separate datasets Hence, for each simulation
two datasets are generated which are analysed to produce 28 MR estimates
each. We performed 100,000 simulations using these parameters, resulting
in 200,000 datasets.

\textbf{Strategy}

Figure 2a outlines the general strategy behind the MoE implementation.
Briefly, we record 53 metrics about each dataset, hypothesising that we
can use these metrics to predict the performance of each expert, given
some optimisation criteria. For each method we train random forest
decision trees to predict that method's performance using the dataset's
metrics.

Having generated random forest decision trees for each of the 28 methods
using 133,000 of the simulations, we then applied them to the remaining
67,000 datasets to predict which method would have the highest
performance for each of the remaining datasets. Finally we compare the
performance of the method selected by the MoE against all remaining
methods. The default settings for the \texttt{randomForest} package in R
was used to train the models.

\textbf{Optimisation function}

We aim to maximise statistical power for datasets where
\(\beta_{x,y} \neq 0\) and minimise false discovery rates for datasets
where \(\beta_{x,y} = 0\). The random forest for a particular method
\(h(O_{w,d}, \textbf{z}_{d}\) is generated where the training set of
input metrics for dataset \(d\) is \(\textbf{z}_{d}\) and the response
is

\[
    O_{w,d} = 
\begin{cases}
    1,              & \text{if } \beta_{x,y} \neq 0 \text{and } p_{m,d} < 0.01\\
    1,              & \text{if } \beta_{x,y} = 0 \text{and } p_{m,d} > 0.1 \\
    0,              & \text{otherwise}
\end{cases}
\]

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-DaveySmith2003}{}
1. Davey Smith G, Ebrahim S. 'Mendelian randomization': can genetic
epidemiology contribute to understanding environmental determinants of
disease? International Journal of Epidemiology {[}Internet{]}. 2003
Feb;32(1):1--22. Available from:
\url{http://www.ije.oxfordjournals.org/cgi/doi/10.1093/ije/dyg070}

\hypertarget{ref-DaveySmithHemani2014}{}
2. Davey Smith G, Hemani G. Mendelian randomization: genetic anchors for
causal inference in epidemiological studies. Human molecular genetics.
2014 Jul;23(R1):R89-----R98.

\end{document}

### 2017-06-26 docs/manuscript.rmd
updated

---
title: Towards the construction of a causal map of the human phenome
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: no
    keep_tex: true
  word_document: default
header-includes:
   - \usepackage{amsmath}
bibliography: manuscript.bib
csl: vancouver.csl
---

## Introduction

Mendelian randomisation (MR) [@DaveySmith2003; @DaveySmithHemani2014] exploits genetic pleiotropy to infer the causal relationships between phenotypes. Suppose that one trait (the exposure) causally influences another (the outcome). If a SNP influences the outcome through the exposure then the SNP is exhibiting vertical pleiotropy. Such a genetic variant is considered to be an instrumental variable for the exposure, and can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not. Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant is an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) some violations of the assumptions of MR (most notably that the SNP influences the outcome through no pathways other than the exposure) can be evaluated through sensitivity analyses.

Genome-wide association studies (GWAS) have identified genetic instrumental variables for thousands of phenotypes. Recent developments in Mendelian randomisation have enabled knowledge of instrumental variables to be applied using only summary level data (known as two-sample MR, 2SMR). Here, in order to infer the causal effect of an exposure on an outcome all that is required is an estimate of the genetic effects of the instrumenting SNP on the exposure, and the corresponding estimate of the effect on the outcome. This has two major advantages. First, GWAS summary data is non-disclosive and often publicly available. Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting the breadth of possible causal estimates only to the availability of GWAS summary data for the traits in question.

Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure. The extent of this problem is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP. It is considered best practice to report estimates from all available methods as sensitivity analyses when presenting causal estimates, however this strategy is not necessarily optimal for several reasons. First, if different methods disagree it is not possible to know which is correct because the appropriate model of HP is not known. Second, though the IVW approach is most statistically powerful under no HP, it can have high false negative or low true positive rates in the presence of HP compared to other methods. Given that pleiotropy has been hypothesised to be universal, defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate. Third, the available methods do not cover all possible models of HP, and therefore an automated method for instrument selection may be necessary. Fourth, it could be of interest to make causal effect estimates for thousands of traits, in which case a discerning evaluation of each causal effect of interest may not be possible or convenient.

In this paper we introduce two innovations towards improving the reliability of MR estimates. First, we introduce an approach to automatically discard genetic variants that are likely to be invalid. Second, we hypothesised that characteristics of the summary data could indicate which method would be most reliable, and we introduce new machine learning approaches that attempt to automate both instrument and method selection. Using curated GWAS summary data for thousands of phenotypes, we use these new methods to construct a graph of millions of causal estimates. We consider this to be only a 'first draft' of the causal map of the human phenome.

## Methods

### GWAS summary data and their use in 2SMR

The use of summary data in two-sample MR is described in detail elsewhere. A brief outline of the procedure is as follows. First, genetic instruments for the exposure trait need to be identified - those SNPs with $p < 10^{-8}$ are retained, collecting their effect sizes and standard errors and effect alleles for the association with the exposure trait. These can be obtanied manually from complete summary data or from curated lists of significant GWAS associations. Next, the effects of those SNPs on the outcome need to be obtained, typically necessitating complete summary data because it is unlikely that these SNPs will have reached genome-wide significance (and therefore be present in curated catalogues). At its most simple implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome.

Given summary data for a large number of traits, it is straightforward to exhaustively analyse the causal relationships of every trait against every other trait for which there is sufficient summary data available. Supplementary table X provides a list of all traits that have available GWAS summary data, indicating if they have complete summary data (in which case they can be used as both exposure traits and outcome traits), or if they only have significant associations (in which they can only be used as exposure traits). Supplementary table X provides a list of the genetic instruments used for each trait.

### Mendelian randomisation methods and their assumptions

In this paper we consider three main classes of MR estimation. Full details for each approach have been described extensively elsewhere.

**Mean-based methods:** Here we consider four nested models. The inverse variance weighted (IVW) fixed effects meta-analysis approach assumes that variants exhibit no HP. IVW random effects meta-analysis relaxes the HP assumption, allowing it to be present but balanced - such that it only leads to increased heterogeneity around the regression and not introducing bias. Fixed effects Egger regression relaxes the HP assumption further by allowing a non-zero intercept which essentially allows horizontal pleiotropy be directional, where it systematically occurs in a specific direction. Random effects Egger regression allows heterogeneity around the directional HP, as long as the HP effects are not correlated with the SNP-exposure effects (this is the INSIDE assumption).

The Rucker framework uses estimates of heterogeneity in the IVW and Egger frameworks to navigate between these nested models. A jackknife approach (random selection with replacement of instruments) can be used to obtain a sampling distribution for the model estimate amongst these four variations. Using 1000 rounds of jackknife estimates, we can obtain a final estimate using the mean or the median of the distribution. We only use the jackknife approach for associations where there are 15 or more variants to avoid combinatorial saturation.

**Median-based methods:** An alternative approach is to take the median effect of all available instruments. This has the advantage that up to half the instruments can be invalid, and the estimate will remain unbiased. Developing the approach further to allow stronger instruments to contribute more towards the estimate can be obtained by obtaining the median of the weights of each instrument. The penalised weighted median estimator ...

**Mode-based methods:** Similar to the median, the mode based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments. This can be extended in two ways, first by weighting the strength of each cluster by the inverse of the variance of the SNP-outcome associations in the cluster; second by weighting the strength of each cluster by the inverse of the variance of the SNP-exposure associations.

### Instrument selection

#### Top hits

The simplest approach to selecting instruments for performing MR is to take take SNPs that have been declared significant in the published GWAS for the exposure. This typically involves obtaining SNPs that surpass $p < 5 \times 10^{-8}$, using clumping to obtain independent SNPs, and then replicating in an independent sample. These results are often recorded in public GWAS catalogs. Alternatively the clumping procedure can be performed using complete summary data in MR-Base. We call this the "top hits" strategy.

#### Steiger filtering

With genome-wide association studies growing ever larger, the statistical power to detect significant associations that may be influencing the trait downstream of many other pathways increases. For example, if a SNP $g_{A}$ influences trait $A$, and trait $A$ influences trait $B$, then a sufficiently powered GWAS will identify the $g_{A}$ as being significant for trait $B$ (Figure 1a). Using $g_{A}$ as an instrument to test the causal effect of $A$ on $B$ is perfectly valid. But in the (incorrectly hypothesised) MR analysis of trait $B$ on trait $A$ could erroneously result in the apparent causal association of $B$ on $A$. If $g_{A}$ is only one of many known instruments for $B$, amongst which some are valid, it is to the advantage of the researcher to exclude $g_{A}$ from the analysis.

An approach to inferring the causal direction between phenotypes was developed recently, using the following basic premise. If trait $A$ causes trait $B$ then

$$
\sum^M_{i=1}{cor(g_{i}, A)^2} > \sum^M_{i=1}{cor(g_{i}, B)^2}
$$

because the $cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}$. This simple inequality will not hold in some cases, for example $\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}$ where $\rho_{x, x_o}$ and $\rho_{y, y_o}$ are the precision of the measurements of the $x$ and $y$. Steiger's Z-test of correlated correlations can be used to formally test the extent to which the two correlations are statistically different.

Here we adapt this approach to automatically filter SNPs that are liable to be invalid (Figure 1a). In this case the Steiger test applied to each variant in turn will identify $g_{A}$ as being unlikely to primarily associate with $B$ relative to $A$. Similarly, for SNPs that influence confounders of $A$ and $B$ or those variants that exhibit horizontal pleiotropy, the difference in $cor(g_{i}, A)^2$ and $cor(g_{i}, B)^2$ will be reduced, increasing the likelihood of the SNP being excluded because the Steiger Z-test is less likely to be significant.

To estimate $cor(g, x)^2$, if $x$ is continuous we obtain the F-statistic from the reported p-value and sample size and then $cor(g, x)^2 = \frac{F}{N - 2 - F}$. If $x$ is binary then our objective is to estimate the variance of the risk liability explained by the SNP, $cor(g, x)^2 = \frac{V_a}{V_a + V_e}$. Here, $V_e = \pi^2/3$, and $V_a = 2\beta^2p(1-p)$, where $\beta$ is the log odds ratio and $p$ is the allele frequency of the SNP in the population. $p$ can be estimated using the allele frequency of the SNP in an ascertained sample by deriving the $2 \times 2$ contingency table from the odds ratio $e^\beta$, allele frequency in the ascertained sample $p_{cc}, and number of cases $N_1$ and controls $N_0$.

### Competitive mixture of experts

We consider 14 MR methods, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total. In the context of this analysis, each method is considered to be an 'expert', taking a set of SNP-exposure and SNP-outcome effect sizes and their standard errors as inputs. Our objective is to select the expert most likely to be correct for a specific MR analysis.

#### Mixture of experts

The mixture of experts (MoE) method is a machine learning approach which seeks to divide a parameter space into subdomains, such that a particular expert is used primarily for problems that reside in a subdomain most suited to that expert. In this case our objective is to identify characteristics of the SNP-exposure and SNP-outcome associations for which one specific MR method is most likely to yield highest statistical power for non-null associations, and lowest false discovery rates for null associations.

**Training and testing simulations**

The MoE is trained using datasets generated from simulations. A *dataset* is the minimum data required to perform 2SMR analysis - four columns comprising the SNP-exposure and SNP-outcome effects and standard errors, and rows corresponding to each SNP that is used as an instrument for the exposure. This *dataset* can be fed into any of the 28 experts to obtain MR causal effects. The simulations used to generate these datasets seek to cover a range of pleiotropic scenarios, including where some proportion of SNPs exhibit directional or balanced horizontal pleiotropy, or where SNPs influencing confounding variables.

We simulate two individual level datasets for which there are $N_x$ and $N_y$ samples, and $M$ SNPs, where each SNP has effect allele frequency of $p_m \sim U(0.05, 0.95)$. These datasets are used to obtain the SNP effects for the exposure trait $x$ and the outcome trait $y$, respectively, using the following sampling criteria:

$$
\begin{aligned}
N_x & = \{20000, ..., 500000\} \\
N_y & = \{20000, ..., 500000\} \\
K & = \{0, ..., 10\} \\
M_x & = \{1, ..., 200\} \\
M_y & = \{1, ..., 200\} \\
M_{u_k} & = \{5,..., 30\} \\
\end{aligned}
$$

The $M = M_x + M_y + \sum{M_{u_k}}$ SNPs can influence $x$ directly, $y$ directly, or some number of confounders $u_{k}$ directly. Phenotypes for $x$ and $y$ are constructed using

$$
x = \sum^{M_x}_{i}{\beta_{gx,x,i}g_{x,i}} + \sum^{M_y}_{j}{\beta_{gy,x,j}g_{y,j}} + \sum^{K}_{k}{\beta_{ux,k} u_{k}} + e_{x}
$$

where $\beta_{gx,x}$ is the vector of effects of each of the $M_x$ SNPs that influence $x$ primarily, $\beta_{gy,x}$ is the vector of effects for the $M_y$ SNPs on $x$, where the $M_y$ SNPs influence $y$ primarily but exhibit horizontal pleiotropic effects on $x$. We allow some proportion of these effects to be 0. $\beta_{ux}$ is the vector of effects of each of the $K$ confounders on $x$. Each $u_{k}$ variable is constructed using

$$
u = \sum^{M_u}_{l}{\beta_{gu,l}g_{l}} + e_{l}
$$

and finally $y$ is constructed using

$$
y = \beta_{x,y}x + \sum^{M_y}_{i}{\beta_{gy,y,j}g_{y,j}} + \sum^{M_x}_{j}{\beta_{gx,y,i}g_{x,i}} + \sum^{K}_{k}{\beta_{uy,k} u_{k}} + e_{y}
$$

where $\beta_{x,y}$ is the causal effect of $x$ on $y$. We sample the distribution of direct SNP effects using

$$
\begin{aligned}
\beta_{gx,x,i} & \sim N(0, \sigma^2_{gx,x}) \\
\sigma^2_{gx,x,i} & \sim U(0.01, 0.1) \\
\beta_{gy,y,j} ~ N(0, \sigma^2_{gy,y}) \\
\sigma^2_{gy,y,j} & \sim U(0.01, 0.1) \\
\end{aligned}
$$

Some proportion $s_x \sim U(0,1)$ of $g_x$ SNPs and some proportion $s_y \sim U(0,1)$ of $g_y$ SNPs exhibit horizontal pleiotropy with effects sampled using

$$
\begin{aligned}
\beta_{gx,y,i*} & \sim N(\mu_{gx,y}, \sigma^2_{gx,y})  \\
\mu_{gx,y,i*} & \sim U(-0.005, 0.005) \\
\sigma^2_{gx,y,i*} & \sim U(0.001, 0.01) \\
\beta_{gy,x,j*} & \sim N(\mu_{gy,x}, \sigma^2_{gy,x}) \\
\mu_{gy,x,j*} & \sim U(-0.005, 0.005) \\
\sigma^2_{gy,x,j*} & \sim U(0.001, 0.01) \\
\end{aligned}
$$

The genetic influences on each of the confounders are sampled using

$$
\begin{aligned}
\beta_{gu,u,l} & \sim N(0, \sigma^2_{gu,u}) \\
\sigma^2_{gu,u,l} & \sim U(0.01, 0.1) \\
\end{aligned}
$$

The influence of each confounder on $x$ and $y$ is obtained using

$$
\begin{aligned}
\beta_{u,x} & \sim N(0, \sigma^{2}_{u,x}) \\
\beta_{u,y} & \sim N(0, \sigma^{2}_{u,y}) \\
\end{aligned}
$$

Finally, 20% of the simulations have a null effect of $\beta_{x,y} = 0$, while the other remaining 80% have a true effect sampled from

$$
\begin{aligned}.
\beta_{x,y} & \sim N(0, \sigma^2_{x,y}) \\
\sigma^2_{x,y} & \sim U(0.001, 0.1) \\
\end{aligned}
$$

For each simulation we used linear regression to estimate the genetic effect of each SNP $M$ on $x$ in sample 1, and each SNP $M$ on $y$ in sample 2. We then perform MR analysis in both directions, retaining SNPs that have $p < 5e-8$ in sample 1 to perform MR of $x$ on $y$ (the true causal direction for non-null simulations), and retaining SNPs that have $p < 5e-8$ in sample 2 to perform MR of $y$ on $x$ (the reverse causal direction for non-null simulations). We treat the summary data (effect sizes and standard errors) used for estimating $x \rightarrow y$ the summary data used for estimating $y \rightarrow x$ as two separate datasets Hence, for each simulation two datasets are generated which are analysed to produce 28 MR estimates each. We performed 100,000 simulations using these parameters, resulting in 200,000 datasets.

**Strategy**

Figure 2a outlines the general strategy behind the MoE implementation. Briefly, we record 53 metrics about each dataset, hypothesising that we can use these metrics to predict the performance of each expert, given some optimisation criteria. For each method we train random forest decision trees  to predict that method's performance using the dataset's metrics.

Having generated random forest decision trees for each of the 28 methods using 133,000 of the simulations, we then applied them to the remaining 67,000 datasets to predict which method would have the highest performance for each of the remaining datasets. Finally we compare the performance of the method selected by the MoE against all remaining methods. The default settings for the `randomForest` package in R was used to train the models.

**Optimisation function**

We aim to maximise statistical power for datasets where $\beta_{x,y} \neq 0$ and minimise false discovery rates for datasets where $\beta_{x,y} = 0$. The random forest for a particular method $h(O_{w,d}, \textbf{z}_{d}$ is generated where the training set of input metrics for dataset $d$ is $\textbf{z}_{d}$ and the response is

$$
    O_{w,d} = 
\begin{cases}
    1,   & \text{if } \beta_{x,y} \neq 0 \text{and } p_{m,d} < 0.01\\
    1,   & \text{if } \beta_{x,y} = 0 \text{and } p_{m,d} > 0.1 \\
    0,   & \text{otherwise}
\end{cases}
$$

### Graph database of MR estimates

The set of MR estimates obtained from this analysis are recorded in a Neo4j graph database. Because each association has up to 28 different estimates, for simplicity we distill this down to a single 'best estimate' for each association using the following rules:

1. If the number of variants after Steiger filtering is greater than 5 then apply the MoE to obtain the best method
2. If the number of variants after Steiger filtering is less than or equal to 5 but greater than 1 then use the IVW random effects approach on the filtered set of variants
3. If there is 1 variant retained in the Steiger approach then use the Wald ratio on the remaining variant
4. If there are no variants remaining after Steiger filtering then declare no causal association.

For specific hypotheses we strongly recommend that estimates from all sensitivity analyses are scrutinised and reported.

## Results

### Steiger filtering improves reliability

We evaluated the efficacy of the Steiger filtering approach for improving MR reliability using 200,000 simulated datasets for in which there were varying null and non-null causal relationships, as well as varying degrees of instrument strength, sample size and forms of horizontal pleiotropy. Compared to

### Mixture of experts method selection improves over any single method

The gating function for the mixture of experts model was trained by generating a different random forest independently for each of the 28 MR approaches, using a random two-thirds of the simulated datasets. The prediction accuracy of the performance for each of these methods is shown in table X, estimated using the remaining third of datasets. The dataset characteristics with the most importance for each of the predictors is also shown, as well as the frequency for which each of the methods was selected in the testing datasets.

### A first draft of the causal map of the human phenome

It is important to note that though the gating function should be applicable to datasets of any distribution of characteristics, the frequencies of chosen methods are relative to the characteristics of the simulated datasets, and would not necessarily be the same frequencies observed empirically. Applying the MoE model to the X number of MR associations for which there was sufficient data, we found that

How many associations reach nominal or FDR corrected significance
How many bi-directional associations
How many associations could be declared for a causal chain of 2

## Discussion

- Limitations of MR that cannot be addressed without biological information
    + Impossible causal associations
    + Temporal effects
    + Risk liability etc
    + Single variants
- Further developments required to improve causal estimates
- The use of causal chains to improve power
- Repository will continually update - need to time stamp and provide access to older associations

## References

### 2017-06-26 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{\texttt{r\ format(Sys.time(),\ \textquotesingle{}\%d\ \%B\ \%Y\textquotesingle{})}}
+  \date{26 June 2017}
 
@@
 the outcome through the exposure then the SNP is exhibiting vertical
-pleiotropy. Such a genetic variant is known as an instrumental variable
-for the exposure, and can be exploited to mimic a randomised controlled
-trial, making causal inference by comparing the outcome phenotypes
-between those individuals that have the exposure-increasing allele
-against those who do not. Multiple independent genetic variants for a
-particular exposure can be used jointly to improve causal inference, the
-premise being that each variant is an independent natural experiment,
-and an overall causal estimate can be obtained by meta-analysing the
-single estimates from each instrument.
+pleiotropy. Such a genetic variant is considered to be an instrumental
+variable for the exposure, and can be exploited to mimic a randomised
+controlled trial, enabling a causal estimate to be made by comparing the
+outcome phenotypes between those individuals that have the
+exposure-increasing allele against those who do not. Multiple
+independent genetic variants for a particular exposure can be used
+jointly to improve causal inference, because a) each variant is an
+independent natural experiment, and an overall causal estimate can be
+obtained by meta-analysing the single estimates from each instrument;
+and b) some violations of the assumptions of MR (most notably that the
+SNP influences the outcome through no pathways other than the exposure)
+can be evaluated through sensitivity analyses.
 
@@
 
-In this paper we introduce new machine learning approaches that attempt
-to automate both instrument and method selection. Using curated GWAS
-summary data for thousands of phenotypes, we use these new methods to
-construct a graph of millions of causal estimates.
+In this paper we introduce two innovations towards improving the
+reliability of MR estimates. First, we introduce an approach to
+automatically discard genetic variants that are likely to be invalid.
+Second, we hypothesised that characteristics of the summary data could
+indicate which method would be most reliable, and we introduce new
+machine learning approaches that attempt to automate both instrument and
+method selection. Using curated GWAS summary data for thousands of
+phenotypes, we use these new methods to construct a graph of millions of
+causal estimates. We consider this to be only a `first draft' of the
+causal map of the human phenome.
 
@@
 2SMR}\label{gwas-summary-data-and-their-use-in-2smr}
+
+The use of summary data in two-sample MR is described in detail
+elsewhere. A brief outline of the procedure is as follows. First,
+genetic instruments for the exposure trait need to be identified - those
+SNPs with \(p < 10^{-8}\) are retained, collecting their effect sizes
+and standard errors and effect alleles for the association with the
+exposure trait. These can be obtanied manually from complete summary
+data or from curated lists of significant GWAS associations. Next, the
+effects of those SNPs on the outcome need to be obtained, typically
+necessitating complete summary data because it is unlikely that these
+SNPs will have reached genome-wide significance (and therefore be
+present in curated catalogues). At its most simple implementation, the
+regression of the SNP-exposure effect sizes against the SNP-outcome
+effect sizes, with greater weight afforded to those SNPs with smaller
+SNP-outcome standard errors, provides the estimate of the causal effect
+of the exposure on the outcome.
+
+Given summary data for a large number of traits, it is straightforward
+to exhaustively analyse the causal relationships of every trait against
+every other trait for which there is sufficient summary data available.
+Supplementary table X provides a list of all traits that have available
+GWAS summary data, indicating if they have complete summary data (in
+which case they can be used as both exposure traits and outcome traits),
+or if they only have significant associations (in which they can only be
+used as exposure traits). Supplementary table X provides a list of the
+genetic instruments used for each trait.
 
@@
 
-\textbf{Mean-based methods:} The inverse variance weighted (IVW)
-meta-analysis approach assumes that variant exhibits no HP (fixed
-effects meta-analysis) or that HP is present but balanced (random
-effects meta-analysis). Egger regression relaxes the HP assumption
-further by allowing the horizontal pleiotropy to systematically occur in
-a specific direction, known as directional horizontal pleiotropy. The
-Rucker framework uses estimates of heterogeneity to navigate between
-these nested models. A jackknife approach (random selection with
-replacement of instruments) can be used to obtain a sampling
-distribution for the model estimate amongst these four variations.
+\textbf{Mean-based methods:} Here we consider four nested models. The
+inverse variance weighted (IVW) fixed effects meta-analysis approach
+assumes that variants exhibit no HP. IVW random effects meta-analysis
+relaxes the HP assumption, allowing it to be present but balanced - such
+that it only leads to increased heterogeneity around the regression and
+not introducing bias. Fixed effects Egger regression relaxes the HP
+assumption further by allowing a non-zero intercept which essentially
+allows horizontal pleiotropy be directional, where it systematically
+occurs in a specific direction. Random effects Egger regression allows
+heterogeneity around the directional HP, as long as the HP effects are
+not correlated with the SNP-exposure effects (this is the INSIDE
+assumption).
 
-The Rucker framework is a system used in meta-analysis to navigate
-between IVW and Egger regression, and fixed and random effects models.
-It was recently applied to MR to use heterogeneity statistics to select
-the most appropriate model of pleiotropy, and therefore the most
-appropriate MR method amongst the four mean-based estimators. There
+The Rucker framework uses estimates of heterogeneity in the IVW and
+Egger frameworks to navigate between these nested models. A jackknife
+approach (random selection with replacement of instruments) can be used
+to obtain a sampling distribution for the model estimate amongst these
+four variations. Using 1000 rounds of jackknife estimates, we can obtain
+a final estimate using the mean or the median of the distribution. We
+only use the jackknife approach for associations where there are 15 or
+more variants to avoid combinatorial saturation.
 
@@
 
-\textbf{Mode-based methods:} Supposing that
+\textbf{Mode-based methods:} Similar to the median, the mode based
+estimator clusters the instruments into groups based on similarity of
+causal effects, and returns the final causal effect estimate based on
+the cluster that has the largest number of instruments. This can be
+extended in two ways, first by weighting the strength of each cluster by
+the inverse of the variance of the SNP-outcome associations in the
+cluster; second by weighting the strength of each cluster by the inverse
+of the variance of the SNP-exposure associations.
 
@@
 \begin{aligned}
-\beta_{gx,y,i*} & \sim N(\mu_{gx,y}, \sigma^2_{gx,y}) \times  \\
+\beta_{gx,y,i*} & \sim N(\mu_{gx,y}, \sigma^2_{gx,y})  \\
 \mu_{gx,y,i*} & \sim U(-0.005, 0.005) \\
@@
 
+\subsubsection{Graph database}\label{graph-database}
+
+The set of MR estimates obtained from this analysis are recorded in a
+Neo4j graph database. Because each association has up to 28 different
+estimates, for simplicity we distill this down to a single `best
+estimate' for each association using the following rules:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  If the number of variants after Steiger filtering is greater than 5
+  then apply the MoE to obtain the best method
+\item
+  If the number of variants after Steiger filtering is less than or
+  equal to 5 but greater than 1 then use the IVW random effects approach
+  on the filtered set of variants
+\item
+  If there is 1 variant retained in the Steiger approach then use the
+  Wald ratio on the remaining variant
+\item
+  If there are no variants remaining after Steiger filtering then
+  declare no causal association.
+\end{enumerate}
+
+For specific hypotheses we strongly recommend that estimates from all
+sensitivity analyses are scrutinised and reported.
+
+\subsection{Results}\label{results}
+
+\subsubsection{Steiger filtering improves
+reliability}\label{steiger-filtering-improves-reliability}
+
+\subsubsection{Mixture of experts method selection improves over any
+single
+method}\label{mixture-of-experts-method-selection-improves-over-any-single-method}
+
+Number of associations using different approaches Which methods were
+preferentially chosen
+
+\subsubsection{A first draft of the causal map of the human
+phenome}\label{a-first-draft-of-the-causal-map-of-the-human-phenome}
+
+How many associations reach nominal or FDR corrected significance How
+many bi-directional associations How many associations could be declared
+for a causal chain of 2
+
+\subsection{Discussion}\label{discussion}
+
+\begin{itemize}
+\tightlist
+\item
+  Limitations of MR that cannot be addressed without biological
+  information
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Impossible causal associations
+  \item
+    Temporal effects
+  \item
+    Risk liability etc
+  \item
+    Single variants
+  \end{itemize}
+\item
+  Further developments required to improve causal estimates
+\item
+  The use of causal chains to improve power
+\end{itemize}
+
 \subsection*{References}\label{references}

### 2017-06-27 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{26 June 2017}
+  \date{27 June 2017}
 
@@
 
-\subsubsection{Graph database}\label{graph-database}
+\subsubsection{Graph database of MR
+estimates}\label{graph-database-of-mr-estimates}
 
@@
 
+We evaluated the efficacy of the Steiger filtering approach for
+improving MR reliability using 200,000 simulated datasets for in which
+there were varying null and non-null causal relationships, as well as
+varying degrees of instrument strength, sample size and forms of
+horizontal pleiotropy. Compared to
+
 \subsubsection{Mixture of experts method selection improves over any
@@
 
-Number of associations using different approaches Which methods were
-preferentially chosen
+The gating function for the mixture of experts model was trained by
+generating a different random forest independently for each of the 28 MR
+approaches, using a random two-thirds of the simulated datasets. The
+prediction accuracy of the performance for each of these methods is
+shown in table X, estimated using the remaining third of datasets. The
+dataset characteristics with the most importance for each of the
+predictors is also shown, as well as the frequency for which each of the
+methods was selected in the testing datasets.
 
@@
 phenome}\label{a-first-draft-of-the-causal-map-of-the-human-phenome}
+
+It is important to note that though the gating function should be
+applicable to datasets of any distribution of characteristics, the
+frequencies of chosen methods are relative to the characteristics of the
+simulated datasets, and would not necessarily be the same frequencies
+observed empirically. Applying the MoE model to the X number of MR
+associations for which there was sufficient data, we found that
 
@@
   The use of causal chains to improve power
+\item
+  Repository will continually update - need to time stamp and provide
+  access to older associations
 \end{itemize}
+
+\subsection{Figures}\label{figures}
+
+\newpage
+
+\includegraphics{images/fig1.pdf}
+
+\newpage
+
+\includegraphics{images/fig2.pdf}
 

### 2017-06-27 scripts/05-create_graphdb.r
updated

@@ -95,7 +95,12 @@ ucsc_get_position <- function(snp)
 # - SNP-Trait
 # - Trait-Trait
 
-load("../results/01/nodes.rdata")
+load("../results/01/outcome_nodes.rdata")
+load("../results/01/exposure_dat.rdata")
+
+exposure_nodes <- subset(exposure_dat, !duplicated(id.exposure))
+table(exposure_nodes$id.exposure %in% outcome_nodes)
+
 load("../results/01/extract.rdata")
 load("../results/01/mr.rdata")
 

### 2017-06-27 docs/manuscript.rmd
updated

@@
 title: Towards the construction of a causal map of the human phenome
+author: Gibran Hemani, Jie Zheng, Jack Bowden, Oliver Davies, Philip Haycock, George Davey Smith, Tom Gaunt
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 Such a genetic variant is considered to be an instrumental variable for the exposure, and can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not.
-Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant is an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) some violations of the assumptions of MR (most notably that the SNP influences the outcome through no pathways other than the exposure) can be evaluated through sensitivity analyses.
+Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) some violations of the assumptions of MR (most notably that the SNP influences the outcome through no pathways other than the exposure) can be evaluated through sensitivity analyses.
 
@@
 In this paper we introduce two innovations towards improving the reliability of MR estimates.
-First, we introduce an approach to automatically discard genetic variants that are likely to be invalid.
+First, we introduce an approach to discard genetic variants that are likely to be invalid.
 Second, we hypothesised that characteristics of the summary data could indicate which method would be most reliable, and we introduce new machine learning approaches that attempt to automate both instrument and method selection.
@@
 Our objective is to select the expert most likely to be correct for a specific MR analysis.
+This involves creating a 'gating function', whose purpose is to decide which expert to use for a specific MR analysis, given the parameter space that is occupied by that dataset.
 
@@
 
-We evaluated the efficacy of the Steiger filtering approach for improving MR reliability using 200,000 simulated datasets for in which there were varying null and non-null causal relationships, as well as varying degrees of instrument strength, sample size and forms of horizontal pleiotropy.
-Compared to
+As statistical power for GWAS studies improves the likelihood of a significant association being discovered that doesn't act primarily on the trait of interest increases (Figure 1a).
+We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated datasets comprising both null and non-null causal models.
+For each dataset instruments were selected based on the tophits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.
+
+Figure 1b shows that the tophits strategy led to 49% of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype.
+Applying Steiger filtering reduced this to 25%.
+Consequently, the false discovery rates (FDR) for 12 of the 14 methods reduced substantially when applied using Steiger filtered instruments (Figure 1c).
+The true positive rates for the methods based on Steiger filtering did however reduce slightly for 10 of the 14 methods.
 
@@
 
-The gating function for the mixture of experts model was trained by generating a different random forest independently for each of the 28 MR approaches, using a random two-thirds of the simulated datasets.
-The prediction accuracy of the performance for each of these methods is shown in table X, estimated using the remaining third of datasets.
-The dataset characteristics with the most importance for each of the predictors is also shown, as well as the frequency for which each of the methods was selected in the testing datasets.
+Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to use dataset characteristics to predict the most appropriate of the 28 MR methods and instrument selection strategies (Figure 2a).
+
+The ability to predict the performance for each of these methods is shown in table 1, with prediction $R^2$ ranging between 0.04 and 0.24.
+The dataset characteristics with the most importance for each of the predictors differed substantially between each dataset, as well as the frequency for which each of the methods was selected in the testing datasets.
+The FDR of each method when chosen, compared to their averages across all datasets, reduced; and likewise the true positive rates of each method when chosen increased compared to their simulation-wide averages.
+
+We compared the MoE performance in the simulations against each of the 28 methods, testing to see if it outperformed all other single methods.
+Figures 2b and 2c show that the MoE approach had the best general performance, obtaining an AUC of 0.84 in terms of classifying the simulations as being null or non-null.
+Notably, the next best methods were median and mode based estimators using Steiger filtering.
+Under the assumption of universal pleiotropy it is clear that all methods suffer from very high false discovery rates.
 
@@
 
+## Figures
+
+\newpage
+
+![](images/fig1.pdf)
+
+\newpage
+
+![](images/fig2.pdf)
+
 ## References

### 2017-06-28 scripts/01-choose_nodes-01.r
updated

@@ -1,13 +1,0 @@
-library(tidyverse)
-
-ao <- read_csv("../data/ao.csv")
-ids <- scan("../results/01/ids.txt")
-
-outcome_nodes <- subset(ao, 
-	id %in% ids |
-	author == "Shin" |
-	author == "Kettunen"
-	) %>% 
-	dplyr::select(id, trait, pmid, author, consortium, category, subcategory, EFO, EFO_id, match, sample_size, ncase, ncontrol, unit, sd) 
-
-save(outcome_nodes, file="../results/01/outcome_nodes.rdata")

### 2017-06-28 scripts/02-extract_data-safe.r

@@ -1,59 +1,0 @@
-library(tidyverse)
-library(TwoSampleMR)
-
-#toggle_dev("test")
-
-try_extract_outcome <- function(SNP, id, tries=3)
-{
-	i <- 1
-	out <- 1
-	class(out) <- 'try-error'
-	while(class(out) == 'try-error' & i <= tries)
-	{
-		message("trying iteration ", i)
-		out <- try(extract_outcome_data(SNP, id, proxies=FALSE))
-		i <- i + 1
-	}
-	return(out)
-}
-
-###
-
-ds <- "01"
-
-load(paste0("../results/", ds, "/outcome_nodes.rdata"))
-load(paste0("../results/", ds, "/exposure_dat.rdata"))
-
-filt <- function(temp)
-{
-	temp <- filter(temp, mr_keep.outcome) %>%
-                select(id.outcome, SNP, effect_allele.outcome, other_allele.outcome, eaf.outcome, beta.outcome, se.outcome, pval.outcome, samplesize.outcome, ncase.outcome, ncontrol.outcome, proxy.outcome, proxy_snp.outcome)
-	return(temp)
-}
-
-snps <- unique(exposure_dat$SNP)
-for(i in 50)
-{
-	j <- outcome_nodes$id[i]
-	message(i," of ", nrow(outcome_nodes), ": ", j)
-	snpsneed <- snps
-	out <- paste0("../data/results/01/interim-", j, ".rdata")
-	if(file.exists(out))
-	{
-		message("Loading previous version")
-		load(out)
-		snpsgot <- unique(temp$SNP)
-		snpsneed <- snps[!snps %in% snpsgot]
-		temp2 <- try_extract_outcome(snpsneed, outcome_nodes$id[i])
-		temp2 <- filt(temp2)
-		temp <- rbind(temp, temp2)
-	} else {
-		message("No previous record")
-		temp <- try_extract_outcome(snps, outcome_nodes$id[i])
-		temp <- filter(temp, mr_keep.outcome) %>%
-			select(id.outcome, SNP, effect_allele.outcome, other_allele.outcome, eaf.outcome, beta.outcome, se.outcome, pval.outcome, samplesize.outcome, ncase.outcome, ncontrol.outcome)
-	}
-	save(temp, file=out)
-}
-
-

### 2017-06-28 scripts/02-extract_data.r

@@ -1,39 +1,0 @@
-library(tidyverse)
-library(TwoSampleMR)
-
-#toggle_dev("test")
-
-try_extract_outcome <- function(SNP, id, tries=3)
-{
-	i <- 1
-	out <- 1
-	class(out) <- 'try-error'
-	while(class(out) == 'try-error' & i <= tries)
-	{
-		message("trying iteration ", i)
-		out <- try(extract_outcome_data(SNP, id))
-		i <- i + 1
-	}
-	return(out)
-}
-
-###
-
-ds <- "01"
-
-load(paste0("../results/", ds, "/outcome_nodes.rdata"))
-load(paste0("../results/", ds, "/exposure_dat.rdata"))
-
-snps <- unique(exposure_dat$SNP)
-
-m <- list()
-for(i in 1:nrow(outcome_nodes))
-{
-	temp <- try_extract_outcome(snps, outcome_nodes$id[i])
-	m[[i]] <- filter(temp, mr_keep.outcome) %>%
-		select(id.outcome, SNP, effect_allele.outcome, other_allele.outcome, eaf.outcome, beta.outcome, se.outcome, pval.outcome, samplesize.outcome, ncase.outcome, ncontrol.outcome, proxy.outcome, proxy_snp.outcome)
-}
-
-outcome_datl <- m
-save(outcome_datl, file=paste0("../data/results/", ds, "/outcome_extract_", sec, ".rdata"))
-

### 2017-06-28 scripts/02-extract_data.sh

@@ -1,3 +1,0 @@
-#!/bin/bash
-
-Rscript 02-extract_data.r ${1}

### 2017-06-28 scripts/02-get_instruments.r

@@ -5,10 +5,6 @@
 library(tidyverse)
 library(TwoSampleMR)
 library(MRInstruments)
-data(metab_qtls)
-
-metab_qtls <- format_metab_qtls(metab_qtls)
-
 
 ## Soma scan proteins
 
@@ -29,29 +25,7 @@ soma <- format_data(
 )
 
 
-## Extract Shin QTLs
-
-shin <- read_csv("../data/shin_metab.csv")
-shin <- shin %>% separate("Effect (SE)", c("beta", "se"), sep=" \\(")
-shin$se <- gsub(")", "", shin$se) %>% as.numeric
-shin$beta <- as.numeric(shin$beta)
-shin <- shin %>% separate("EA/OA1", c("ea", "nea"), sep="/")
-
-shin <- format_data(
-	shin,
-	snp_col = "SNP",
-	effect_allele_col = "ea",
-	other_allele_col = "nea",
-	eaf_col = "EAF2",
-	phenotype_col = "Biochemical(s)",
-	beta_col = "beta",
-	se_col = "se",
-	pval_col = "P-value",
-	samplesize_col = "N"
-)
-
-
-load("../results/01/outcome_nodes.rdata")
+load("../results/02/nodes.rdata")
 
 try_extract_instrument <- function(id, pval, tries=3)
 {
@@ -67,24 +41,35 @@ try_extract_instrument <- function(id, pval, tries=3)
 	return(out)
 }
 
+nodes$type <- NA
+ao <- available_outcomes()
 
 l <- list()
 for(i in 1:nrow(nodes))
 {
-	temp <- try_extract_instrument(nodes$id[i])
-	if(!class(temp) == 'try-error' & !is.null(temp))
-	{
-		temp$id.exposure <- nodes$id[i]
-		l[[i]] <- filter(temp, mr_keep.exposure) %>%
-			select(id.exposure, SNP, effect_allele.exposure, other_allele.exposure, eaf.exposure, beta.exposure, se.exposure, pval.exposure, samplesize.exposure, ncase.exposure, ncontrol.exposure)
+	message(i)
+	if(nodes$id[i] %in% ao$id)
+	{	
+		temp <- try_extract_instrument(nodes$id[i], 5e-8)
+		if(!class(temp) == 'try-error' & !is.null(temp))
+		{
+			temp$id.exposure <- nodes$id[i]
+			temp$exposure <- nodes$trait[i]
+			nodes$type[i] <- "eo"
[…]

### 2017-06-28 scripts/03-extract_data.r

@@ -19,12 +19,10 @@ try_extract_outcome <- function(SNP, id, tries=3)
 
 ###
 
-ds <- "01"
+ds <- "02"
 
-load(paste0("../results/", ds, "/outcome_nodes.rdata"))
-# load(paste0("../results/", ds, "/exposure_dat.rdata"))
-
-load("../data/westra.rdata")
+load(paste0("../results/", ds, "/nodes.rdata"))
+load(paste0("../results/", ds, "/exposure_dat.rdata"))
 
 filt <- function(temp)
 {
@@ -33,13 +31,17 @@ filt <- function(temp)
 	return(temp)
 }
 
-snps <- unique(westra$SNP)
-for(i in 1:nrow(outcome_nodes))
+# load("../results/01/outcome_nodes.rdata")
+# load("../results/01/exposure_dat.rdata")
+
+snps <- unique(exposure_dat$SNP)
+# for(i in 1:nrow(nodes))
+for(i in 32)
 {
-	j <- outcome_nodes$id[i]
+	j <- nodes$id[i]
 	message(i," of ", nrow(outcome_nodes), ": ", j)
 	snpsneed <- snps
-	out <- paste0("../data/results/01/westra-", j, ".rdata")
+	out <- paste0("../data/results/02/interim/interim-", j, ".rdata")
 	if(file.exists(out))
 	{
 		message("Loading previous version")
@@ -57,5 +59,4 @@ for(i in 1:nrow(outcome_nodes))
 	}
 	save(temp, file=out)
 }
-
 

### 2017-06-28 scripts/03-extract_data.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+Rscript 03-extract_data.r ${1}

### 2017-06-28 scripts/04-harmonise_data.r

@@ -79,8 +79,10 @@ for(i in 1:nrow(outcome_nodes))
 {
 	message(i, ": ", outcome_nodes$id[i])
 	load(paste0("../data/results/01/interim-", outcome_nodes$id[i], ".rdata"))
-	d <- harmonise_data(exposure_dat, temp)
-	d <- merge(d, subset(exposure_dat, select=c(id.exposure, SNP, rsq.exposure)), by=c("id.exposure", "SNP"))
+	temp$other_allele.outcome[temp$other_allele.outcome == ""] <- NA
+	act <- ifelse(all(is.na(temp$other_allele.outcome)), 1, 2)
+	d <- suppressMessages(harmonise_data(exposure_dat, temp, action=act))
+	# d <- merge(d, subset(exposure_dat, select=c(id.exposure, SNP, rsq.exposure)), by=c("id.exposure", "SNP"))
 	d <- get_rsq(d, outcome_nodes$unit[i] == "log odds")
 
 	sexp <- d$samplesize.exposure

### 2017-06-28 scripts/04-perform_mr.r

@@ -1,19 +1,0 @@
-# devtools::install_github("MRCIEU/TwoSampleMR@mr_structure")
-library(TwoSampleMR)
-library(tidyverse)
-
-ds <- "01"
-
-load(paste0("../results/", ds, "/outcome_nodes.rdata"))
-load(paste0("../results/", ds, "/exposure_dat.rdata"))
-
-i <- as.numeric(commandArgs(T)[1])
-
-load(paste0("../results/01/dat/dat-", outcome_nodes$id[i], ".rdata"))
-d$outcome <- outcome_nodes$trait[i]
-
-m1 <- run_mr(subset(d, id.exposure != id.outcome))
-save(m1, file=paste0("../results/01/mr/m1-", outcome_nodes$id[i], ".rdata"))
-m2 <- run_mr(subset(d, id.exposure != id.outcome & steiger_dir & steiger_pval < 0.05))
-save(m2, file=paste0("../results/01/mr/m2-", outcome_nodes$id[i], ".rdata"))
-

### 2017-06-28 scripts/05-create_graphdb_readcsv.r

@@ -1,176 +1,0 @@
-library(tidyverse)
-
-modify_node_headers_for_neo4j <- function(x, id, idname)
-{
-	id_col <- which(names(x) == id)
-	cl <- sapply(x, class)
-	for(i in 1:length(cl))
-	{
-		if(cl[i] == "integer")
-		{
-			names(x)[i] <- paste0(names(x)[i], ":INT")
-		}
-		if(cl[i] == "numeric")
-		{
-			names(x)[i] <- paste0(names(x)[i], ":FLOAT")
-		}
-	}
-	names(x)[id_col] <- paste0(idname, "Id:ID(", idname, ")")
-	return(x)
-}
-
-modify_rel_headers_for_neo4j <- function(x, id1, id1name, id2, id2name)
-{
-	id1_col <- which(names(x) == id1)
-	id2_col <- which(names(x) == id2)
-	cl <- sapply(x, class)
-	for(i in 1:length(cl))
-	{
-		if(cl[i] == "integer")
-		{
-			names(x)[i] <- paste0(names(x)[i], ":INT")
-		}
-		if(cl[i] == "numeric")
-		{
-			names(x)[i] <- paste0(names(x)[i], ":FLOAT")
-		}
-	}
-	names(x)[id1_col] <- paste0(":START_ID(", id1name, ")")
-	names(x)[id2_col] <- paste0(":END_ID(", id2name, ")")
-	return(x)
-}
-
-get_gene <- function(dat, chrname, posname, radius=10000)
-{
-	# https://www.biostars.org/p/167818/
-	# source("https://bioconductor.org/biocLite.R")
-	# biocLite("Homo.sapiens")
-	require(Homo.sapiens)
-	require(dplyr)
-	
-	mycoords.gr <- data_frame(chrom=dat[[chrname]], start=dat[[posname]]-10000, end=dat[[posname]]+10000) %>% 
-	makeGRangesFromDataFrame
-
-	mycoords.gr <- mergeByOverlaps(genes(TxDb.Hsapiens.UCSC.hg19.knownGene), mycoords.gr)
-	mycoords.gr <- data.frame(gene_id=mycoords.gr$gene_id, gr=mycoords.gr$mycoords.gr, stringsAsFactors=FALSE) %>% as_data_frame
-
-	mycoords.gr <- inner_join(mycoords.gr, as.data.frame(org.Hs.egSYMBOL), by="gene_id")
-	mycoords.gr$pos <- mycoords.gr$gr.start + 10000
-
-	dat <- right_join(
[…]

### 2017-06-28 scripts/05-experts.sh

@@ -1,25 +1,0 @@
-#!/bin/bash
-
-#PBS -N mr2
-#PBS -o job_reports/mr2-output
-#PBS -e job_reports/mr2-error
-#PBS -l walltime=00:30:00
-#PBS -t 1-724
-#PBS -l nodes=1:ppn=1
-#PBS -S /bin/bash
-
-set -e
-
-echo "Running on ${HOSTNAME}"
-
-if [ -n "${1}" ]; then
-	echo "${1}"
-	PBS_ARRAYID=${1}
-fi
-
-i=${PBS_ARRAYID}
-
-cd ${HOME}/repo/mr-eve/scripts
-
-Rscript 05-experts.r ${i}
-

### 2017-06-28 scripts/05-perform_mr.r

@@ -1,3 +1,4 @@
+# devtools::install_github("MRCIEU/TwoSampleMR@mr_structure")
 library(TwoSampleMR)
 library(tidyverse)
 library(randomForest)
@@ -85,13 +86,11 @@ system_metrics <- function(dat)
 	return(metrics)
 }
 
-
-
 get_metrics <- function(dat)
 {
 	metrics <- system_metrics(dat)
 	# Steiger
-	steiger_keep <- dat$steiger_dir & dat$steiger_pval < 0.05
+	steiger_keep <- dat$steiger_dir
 	metrics$st_correct <- sum(steiger_keep) / nrow(dat)
 	metrics$st_unknown <- sum(dat$steiger_pval < 0.05) / nrow(dat)
 	metrics$st_incorrect <-  sum(!dat$steiger_dir & dat$steiger_pval < 0.05) / nrow(dat)
@@ -142,7 +141,7 @@ get_rf_method <- function(m1, m2, d, rf)
 			}
 			pr <- tibble(pr=pr, Method=nom)
 			res <- left_join(res, pr, by="Method")
-			ress <- res[which.max(res$pr), ]
+			ress <- res[which.max(res$pr)[1], ]
 			n[[j]] <- tibble(id.exposure=id.e, id.outcome=id.o, exposure=e, outcome=o, selected_method=ress$Method)
 			ress$Method <- "RF"
 			l[[j]] <- rbind(res, ress)
@@ -153,18 +152,28 @@ get_rf_method <- function(m1, m2, d, rf)
 	return(l)
 }
 
-load("~/repo/instrument-directionality/results/rf.rdata")
+
 
 ds <- "01"
 
 load(paste0("../results/", ds, "/outcome_nodes.rdata"))
 load(paste0("../results/", ds, "/exposure_dat.rdata"))
+load("~/repo/instrument-directionality/results/rf.rdata")
+
 
 i <- as.numeric(commandArgs(T)[1])
 
 load(paste0("../results/01/dat/dat-", outcome_nodes$id[i], ".rdata"))
-load(paste0("../results/01/mr/m1-", outcome_nodes$id[i], ".rdata"))
-load(paste0("../results/01/mr/m2-", outcome_nodes$id[i], ".rdata"))
+d$outcome <- outcome_nodes$trait[i]
 
+
+a <- subset(d, id.exposure == 2)
+run_mr(a)
+
[…]

### 2017-06-28 scripts/06-create_graphdb.r

@@ -99,26 +99,87 @@ load("../results/01/outcome_nodes.rdata")
 load("../results/01/exposure_dat.rdata")
 
 exposure_nodes <- subset(exposure_dat, !duplicated(id.exposure))
-table(exposure_nodes$id.exposure %in% outcome_nodes)
 
-load("../results/01/extract.rdata")
-load("../results/01/mr.rdata")
+exposure_nodes <- dplyr::select(exposure_nodes,
+	trait=exposure,
+	id=id.exposure,
+	unit=units.exposure,
+	sample_size=samplesize.exposure,
+	ncase=ncase.exposure,
+	ncontrol=ncontrol.exposure
+)
+
+exposure_nodes <- subset(exposure_nodes, ! id %in% outcome_nodes$id)
+# ind1 <- exposure_nodes$trait %in% outcome_nodes$trait
+# ind <- match(exposure_nodes$trait[ind1], outcome_nodes$trait)
+# ind <- ind[!is.na(ind)]
+# exposure_nodes$id[ind1] <- outcome_nodes$id[ind]
+# exposure_nodes$trait[ind1] == outcome_nodes$trait[ind]
+# exposure_nodes$id[ind1] == outcome_nodes$id[ind]
+
+outcome_nodes$id <- as.character(outcome_nodes$id)
+nodes <- bind_rows(exposure_nodes, outcome_nodes)
+
+
+# table(exposure_nodes$id.exposure %in% outcome_nodes$id)
+# table(outcome_nodes$id %in% exposure_nodes$id.exposure)
+
+# temp <- subset(exposure_nodes, data_source.exposure == "Shin metabolites")
+# table(temp$exposure %in% outcome_nodes$trait)
+# subset(temp, ! exposure %in% outcome_nodes$trait)$exposure
+# subset(outcome_nodes, ! trait %in% exposure_nodes$exposure)$trait
+
+# a <- subset(outcome_nodes, author == "Shin")$id
+
+# r <- list()
+# for(i in 1:length(a))
+# {
+# 	message(i)
+# 	load(paste0("../results/01/mr/m1-", a[i], ".rdata"))
+# 	r[[i]] <- which(sapply(m1, function(x) nrow(subset(x$out, P < 1e-3 & round(Estimate, 1) == 1))) != 0)
+# }
+
+
+# b <- subset(exposure_nodes, data_source.exposure == "Shin metabolites")$exposure
[…]

### 2017-06-28 scripts/westrasnps.r

@@ -1,53 +1,0 @@
-# westra
-# simple clumping
-
-library(tidyverse)
-library(TwoSampleMR)
-library(data.table)
-
-a <- fread("../data/2012-12-21-CisAssociationsProbeLevelFDR0.5.txt")
-b <- filter(a, PValue < 5e-5) %>%
-	arrange(PValue) %>%
-	filter(!duplicated(ProbeName))
-
-write.table(unique(b$SNPName), file="westrasnps.txt", row=F, col=F, qu=F)
-system(
-	"plink1.90 --bfile ~/data/alspac_1kg/data/derived/filtered/bestguess/maf0.01_info0.8/combined/data --extract westrasnps.txt --freq --out westrasnps.txt"
-)
-
-fr <- fread("westrasnps.txt.frq")
-
-b <- inner_join(b, dplyr::select(fr, SNP, MAF), by=c("SNPName"="SNP"))
-
-b$n <- 5300
-b$r <- get_r_from_pn(b$PValue, b$n)
-b$rsq <- b$r^2
-
-b$eff <- sqrt(b$rsq /(2 * b$MAF * (1-b$MAF)))
-b$se <- b$eff / b$OverallZScore
-
-temp <- do.call(rbind, strsplit( b$SNPType, "/"))
-index <- temp[,1] == b$AlleleAssessed
-b$otherallele <- NA
-b$otherallele[index] <- temp[index, 2]
-b$otherallele[!index] <- temp[!index, 1]
-
-
-westra <- with(b,
-	tibble(
-		SNP=SNPName,
-		exposure=ProbeName,
-		id.exposure=ProbeName,
-		beta.exposure=eff,
-		se.exposure=se,
-		pval.exposure=PValue,
-		effect_allele.exposure=AlleleAssessed,
-		other_allele.exposure=otherallele,
-		genename=HUGO,
-		probechr=ProbeChr,
-		probecentre=ProbeCenterChrPos
-	)
-)
-
-save(westra, file="../data/westra.rdata")
-

### 2017-06-29 scripts/05-perform_mr.r
updated

@@ -158,7 +158,6 @@ ds <- "01"
 
 load(paste0("../results/", ds, "/outcome_nodes.rdata"))
 load(paste0("../results/", ds, "/exposure_dat.rdata"))
-load("~/repo/instrument-directionality/results/rf.rdata")
 
 
 i <- as.numeric(commandArgs(T)[1])
@@ -166,14 +165,30 @@ i <- as.numeric(commandArgs(T)[1])
 load(paste0("../results/01/dat/dat-", outcome_nodes$id[i], ".rdata"))
 d$outcome <- outcome_nodes$trait[i]
 
+message(i, ": ", outcome_nodes$id[i])
 
-a <- subset(d, id.exposure == 2)
-run_mr(a)
+if(file.exists(paste0("../results/01/mr/m1-", outcome_nodes$id[i], ".rdata")))
+{
+	load(paste0("../results/01/mr/m1-", outcome_nodes$id[i], ".rdata"))
+} else {
+	m1 <- run_mr(subset(d, id.exposure != id.outcome))
+	save(m1, file=paste0("../results/01/mr/m1-", outcome_nodes$id[i], ".rdata"))
+}
 
-m1 <- run_mr(subset(d, id.exposure != id.outcome))
-save(m1, file=paste0("../results/01/mr/m1-", outcome_nodes$id[i], ".rdata"))
-m2 <- run_mr(subset(d, id.exposure != id.outcome & steiger_dir))
-save(m2, file=paste0("../results/01/mr/m2-", outcome_nodes$id[i], ".rdata"))
-m3 <- get_rf_method(m1, m2, d, rf)
-save(m3, file=paste0("../results/01/mr/m3-", outcome_nodes$id[i], ".rdata"))
+if(file.exists(paste0("../results/01/mr/m2-", outcome_nodes$id[i], ".rdata")))
+{
+	load(paste0("../results/01/mr/m2-", outcome_nodes$id[i], ".rdata"))
+} else {
+	m2 <- run_mr(subset(d, id.exposure != id.outcome & steiger_dir))
+	save(m2, file=paste0("../results/01/mr/m2-", outcome_nodes$id[i], ".rdata"))
+}
+
+if(file.exists(paste0("../results/01/mr/m3-", outcome_nodes$id[i], ".rdata")))
+{
+	load(paste0("../results/01/mr/m3-", outcome_nodes$id[i], ".rdata"))
+} else {
[…]

### 2017-06-29 scripts/05-perform_mr.sh

@@ -5,7 +5,7 @@
 #PBS -e job_reports/mr-error
 #PBS -l walltime=12:00:00
 #PBS -t 1-724
-#PBS -l nodes=1:ppn=2
+#PBS -l nodes=1:ppn=4
 #PBS -S /bin/bash
 
 set -e

### 2017-06-29 scripts/06-collate_mr.r
updated collation

@@ -1,0 +1,131 @@
+library(tidyverse)
+
+# best causal estimate
+# each other causal estimate
+# other metrics
+
+
+# best
+
+# m2 = 1 SNP: wald ratio
+# m2 = 5 SNP: ivw
+# m2 = >5 SNP: RF
+# m2 not in m1: 0 effect
+
+
+
+# 1. Get the results
+
+# For each SNP
+# - rsq
+# - steiger
+# - evidence of being outlier
+# - 
+
+
+fn <- function(oid)
+{
+	require(tidyverse)
+	m <- list()
+	mb <- list()
+	se <- list()
+	so <- list()
+	k <- 1
+	for(i in 1:length(oid))
+	{
+		message(i)
+		fn1 <- paste0("../results/01/mr/m1-", oid[i], ".rdata")
+		fn2 <- paste0("../results/01/mr/m2-", oid[i], ".rdata")
+		fn3 <- paste0("../results/01/mr/m3-", oid[i], ".rdata")
+		if(file.exists(fn1) & file.exists(fn2) & file.exists(fn3))
+		{
+			load(fn1)
+			load(fn2)
+			load(fn3)
+			a1 <- attributes(m1)$id.exposure
+			a2 <- attributes(m2)$id.exposure
+			a3 <- attributes(m3)$id.exposure
+
+			for(j in 1:length(m1))
+			{
+				so[[k]] <- tibble(
+					SNP=m1[[j]]$dat$SNP,
+					beta=m1[[j]]$dat$beta.outcome,
+					se=m1[[j]]$dat$se.outcome,
+					pval=m1[[j]]$dat$pval.outcome,
+					n=m1[[j]]$dat$samplesize.outcome,
+					ea=m1[[j]]$dat$a1,
+					oa=m1[[j]]$dat$a2,
+					id=oid[i]
+				)
+
+				if(a1[j] %in% a3)
+				{
+					msel <- attributes(m3)$selected_method[which(a3 == a1[j])]
+
+					if(!is.na(msel))
+					{
+						mb[[k]] <- subset(m3[[which(a3 == a1[j])]], Method == "RF")
+						mb[[k]]$Method <- msel
+					} else {
+						mb[[k]] <- subset(m3[[which(a3 == a1[j])]], Method == "Simple mode - steiger")
+					}
+					m[[k]] <- m3[[which(a3 == a1[j])]]
[…]

### 2017-07-01 scripts/07-create_graphdb.r
updated

@@ -176,9 +176,9 @@ snp_trait <- exposure_dat
 
 snps <- ucsc_get_position(unique(snp_trait$SNP))
 snps <- subset(snps, !duplicated(name)) %>% 
-	dplyr::select(name=name, chr=chrom, pos=chromStart, alleles=alleles, alleleFreqs=alleleFreqs, ucsc_func=func) %>% as_data_frame
+	dplyr::select(name=name, chr=chrom, pos=chromStart, ucsc_func=func) %>% as_data_frame
 snps <- bind_rows(snps, data.frame(name=unique(subset(snp_trait, !SNP %in% snps$name)$SNP)))
-snps$id <- 1:nrow(snps)
+snps$id <- snps$name
 snp_trait <- left_join(snp_trait, dplyr::select(snps, SNP=name, snp_id=id), by="SNP")
 
 
@@ -212,13 +212,25 @@ write.csv(snp_trait, file="../results/01/upload/snp_trait.csv", row.names=FALSE,
 
 ## trait-trait
 
-for(i in 1:8)
+ntt <- 8
+
+for(i in 5:ntt)
 {
 	message(i)
 	load(paste0("../results/01/mr/mc-", i, ".rdata"))
+
 	m <- bind_rows(a$m)
 	m <- modify_rel_headers_for_neo4j(m, "id.exposure", "trait", "id.outcome", "trait")
-	write.csv(m, paste0("../results/01/upload/trait_trait-", i, ".csv"), row.names=FALSE, na="")
+	write.csv(m, paste0("../results/01/upload/trait_trait-", i, ".csv"), row.names=FALSE, na="", col.names=i == 1)
+
+	mb <- bind_rows(a$mb)
+	mb <- modify_rel_headers_for_neo4j(mb, "id.exposure", "trait", "id.outcome", "trait")
+	write.csv(mb, paste0("../results/01/upload/trait_trait_sel-", i, ".csv"), row.names=FALSE, na="", col.names=i == 1)
+
+	# so <- bind_rows(a$so)
+	# names(so) <- c("SNP", "beta", "se", "pval", "samplesize", "effect_allele", "other_allele", "id")
+	# so <- modify_rel_headers_for_neo4j(so, "SNP", "snp", "id", "trait")
+	# write.csv(so, paste0("../results/01/upload/so-", i, ".csv"), row.names=FALSE, na=")
 }
 
 cmd <- paste0(

### 2017-07-01 scripts/07-create_graphdb.sh

@@ -1,0 +1,95 @@
+#/bin/bash
+
+module add languages/java-jdk-1.8.0-66
+
+neo4j_dir=""
+cat ../results/01/upload/trait_trait* > ../results/01/upload/trait_trait.csv
+cat ../results/01/upload/trait_trait_sel* ../results/01/upload/trait_trait_sel.csv
+
+
+${neo4j_dir}/bin/neo4j-import \
+--into ../results/01/upload/mr-eve.db \
+--id-type string \
+--nodes:gene ../data/upload/genes.csv \
+--nodes:snp ../data/upload/snps.csv \
+--nodes:trait ../data/upload/traits.csv \
+--relationships:GS ../data/upload/gene_snp.csv \
+--relationships:GA ../data/upload/snp_trait.csv \
+--relationships:MR ../data/upload/trait_trait.csv \
+--relationships:MRB ../data/upload/trait_trait_sel.csv
+
+scp -rv mr-eve.db gh13047@shark.epi.bris.ac.uk:neo4j/data/mreve.db
+
+
+
+
+
+#####
+
+
+#/bin/bash
+
+module add languages/java-jdk-1.8.0-66
+
+neo4j_dir="~/bin/neo4j-community-3.2.1"
+cat ../results/01/upload/trait_trait* > ../results/01/upload/trait_trait.csv
+cat ../results/01/upload/trait_trait_sel* > ../results/01/upload/trait_trait_sel.csv
+
+
+bin/neo4j-import \
+--into mr-eve.db \
+--id-type string \
+--nodes:gene data/upload/genes.csv \
+--nodes:snp data/upload/snps.csv \
+--nodes:trait data/upload/traits.csv \
+--relationships:GS data/upload/gene_snp.csv \
+--relationships:GA data/upload/snp_trait.csv \
+--relationships:MRB data/upload/trait_trait_sel.csv \
+--relationships:MR data/upload/trait_trait.csv
+
+scp -prv ../results/01/upload/mr-eve.db/. gh13047@shark.epi.bris.ac.uk:neo4j/data/mr-eve.db/
+
+
+
+rm -r ../results/01/upload/trait_trait.csv
+rm -r ../results/01/upload/trait_trait_sel.csv
+for i in {1..8}
+do
[…]

### 2017-07-03 shiny/server.r
created basic interface to neo4j

@@ -1,0 +1,150 @@
+library(RNeo4j)
+library(shiny)
+
+escape_quote <- function(text)
+{
+	gsub("'", "\\\\'", text)
+}
+
+
+get_query_trait <- function(graph, trait)
+{
+	query <- paste0("match 
+		(n:trait {name: '", escape_quote(trait), "'})
+		-[r:MRB]-
+		(m:trait)
+
+		return n,r,m order by r.P limit 200"
+	)
+	return(query)
+}
+
+get_query_exposure_outcome <- function(graph, exposure, outcome)
+{
+	query <- paste0("match 
+		(g:gene)-[rg:GS]->(s:snp)-[rs:GA]->(n:trait {name: '", escape_quote(exposure), "'})
+		-[r:MR]->
+		(m:trait {name: '", escape_quote(outcome), "'}) 
+		return n,r,m,rg,g,s,rs"
+	)
+	return(query)
+}
+
+get_table_exposure_outcome <- function(graph, exposure, outcome)
+{
+	query <- paste0("match 
+		(n:trait {name: '", escape_quote(exposure), "'})
+		-[r:MR]->
+		(m:trait {name: '", escape_quote(outcome), "'}) 
+		return
+		r.Method as Method, 
+		r.instruments as instruments,
+		r.nsnp as nsnp,
+		r.Estimate as Estimate, 
+		r.SE as SE,
+		r.CI_low as CI_low,
+		r.CI_upp as CI_upp,
+		r.P as P"
+	)
+	return(cypher(graph, query))
+}
+
+# get_table_exposure_outcome(graph, "Body mass index", "Parkinson's disease")
+
+
+get_trait_scan <- function(graph, trait)
+{
+	query <- paste0("match 
+		(n:trait)
+		-[r:MRB]->
+		(m:trait)
+
+		where (n.name = '", escape_quote(trait), "' and m.name <> '", escape_quote(trait), "') or (m.name = '", escape_quote(trait), "' and n.name <> '", escape_quote(trait), "')
+
+		return
+
+		n.name as Exposure,
+		m.name as Outcome,
+		r.Method as Method, 
+		r.nsnp as nsnp,
+		r.Estimate as Estimate, 
+		r.SE as SE,
+		r.CI_low as CI_low,
+		r.CI_upp as CI_upp,
+		r.P as P
+		order by P"
+	)
+	o <- cypher(graph, query)
+	o$FDR <- p.adjust(o$P, "fdr")
[…]

### 2017-07-03 shiny/ui.r

@@ -1,0 +1,124 @@
+library(RNeo4j)
+library(shiny)
+library(shinydashboard)
+
+
+graph <- startGraph("http://shark.epi.bris.ac.uk:7474/db/data", username="neo4j", password="123qwe")
+
+
+browser_url <- "http://shark.epi.bris.ac.uk:7474/browser/"
+
+# a <- getLabeledNodes(graph, "trait") # this is so slow
+# b <- getNodes(graph, "match(n:trait) return n")
+# nodes <- sapply(a, function(x) x$name)
+nodes <- cypher(graph, "match (n:trait) return n.name")$n.name
+snps <- cypher(graph, "match(n:snp) return n.name")$n.name
+consortia <- cypher(graph, "match (n:trait) return n.consortium")$n.consortium
+mrest <- cypher(graph, "match (n)-[r:MRB]-(m) return count(r) as count")$count
+
+
+load("outcome_nodes.rdata")
+outcome_nodes <- iconv(outcome_nodes$trait, sub='')
+
+dashboardPage(
+
+	dashboardHeader(title="MR-EvE"),
+
+	dashboardSidebar(
+		sidebarMenu(
+			menuItem("About", tabName = "about", icon=icon("smile-o")),
+			menuItem("Basic lookup", tabName = "basic", icon=icon("cogs")),
+			menuItem("Trait scan", tabName = "trait", icon=icon("cogs"))
+		)
+	),
+
+	dashboardBody(
+		tabItems(
+			tabItem(tabName = "about",
+				column(6,
+					box(title="MR of everything vs everything", width = 12,
+						p("This is a repository of Mendelian randomisation estimates that have been generated using the GWAS summary data in the MR-Base database. It can be used to quickly browse specific phenotypes or lookup specific causal hypotheses."),
+						p("You can find out more about MR-Base at ", tags$a("http://mrbase.org", "http://www.mrbase.org")),
+						p("Important: Mendelian randomization is not a panacea for causal inference. Methodology is still required to reduce sensitivity to violations of assumptions, and to incorporate information about the biological feasibility of causal associations.")
+					),
+					box(width=12,
+						p(tags$strong("Alpha phase release: "), "strictly experimental"),
+						p(tags$strong("Last update: "), "3rd July 2017")
+					)
+				),
+				column(6,
+					infoBox(width=12, "Phenotypes", length(nodes), icon=icon("database")),
+					infoBox(width=12, "MR estimates", mrest, icon = icon("random")),
[…]

### 2017-07-03 scripts/04-harmonise_data.r
fixed steiger pval

@@ -89,7 +89,7 @@ for(i in 1:nrow(outcome_nodes))
 	sexp[is.na(d$samplesize.exposure)] <- 1000000
 	sout <- d$samplesize.outcome
 	sout[is.na(d$samplesize.outcome)] <- 1000000
-	st <- psych::r.test(n = sexp, n2 = sout, r12 = d$rsq.exposure, r34 = d$rsq.outcome)
+	st <- psych::r.test(n = sexp, n2 = sout, r12 = sqrt(d$rsq.exposure) * sign(d$beta.exposure), r34 = sqrt(d$rsq.outcome) * sign(d$beta.outcome))
 
 	d$steiger_dir <- as.logical(sign(st$z))
 	d$steiger_pval <- st$p

### 2017-07-03 docs/manuscript.rmd
updated

@@
 title: Towards the construction of a causal map of the human phenome
-author: Gibran Hemani, Jie Zheng, Jack Bowden, Oliver Davies, Philip Haycock, George Davey Smith, Tom Gaunt
+author: Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George Davey Smith
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
   pdf_document:
-    fig_caption: no
+    fig_caption: yes
     keep_tex: true
@@
 
+## Abstract
+
+A major application for genome-wide association studies (GWAS) has been the emerging field of causal inference using Mendelian randomisation (MR), where the causal effect between a pair of traits can be estimated using only summary level data.
+MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype.
+Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy, and many methods have been developed in an attempt to address this.
+Here we show that the mechanisms that underlie horizontal pleiotropy are numerous, and that instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow.
+We have developed a mixture of experts machine learning approach (MR-MoE 1.0) that improves on both power and false discovery rates over all existing methods.
+Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a draft of the causal map of the human phenome.
+But we offer caution that its interpretation is far from straightforward.
+
 ## Introduction
@@
 Such a genetic variant is considered to be an instrumental variable for the exposure, and can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not.
-Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) some violations of the assumptions of MR (most notably that the SNP influences the outcome through no pathways other than the exposure) can be evaluated through sensitivity analyses.
+Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) some violations of the assumptions of MR (most notably that the SNP influences the outcome through no pathways other than the exposure) can be evaluated through sensitivity analyses [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Hemani2017].
 
-Genome-wide association studies (GWAS) have identified genetic instrumental variables for thousands of phenotypes.
-Recent developments in Mendelian randomisation have enabled knowledge of instrumental variables to be applied using only summary level data (known as two-sample MR, 2SMR).
+Genome-wide association studies (GWAS) have identified genetic instrumental variables for thousands of phenotypes [@Hindorff2010].
+Recent developments in Mendelian randomisation have enabled knowledge of instrumental variables to be applied using only summary level data (known as two-sample MR, 2SMR) [@Pierce2013].
 Here, in order to infer the causal effect of an exposure on an outcome all that is required is an estimate of the genetic effects of the instrumenting SNP on the exposure, and the corresponding estimate of the effect on the outcome.
@@
 First, GWAS summary data is non-disclosive and often publicly available.
-Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting the breadth of possible causal estimates only to the availability of GWAS summary data for the traits in question.
+Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting the breadth of possible causal estimates only to the availability of GWAS summary data for the traits in question [@Hemani2016].
 
 Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure.
-The extent of this problem is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP.
+The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017].
 It is considered best practice to report estimates from all available methods as sensitivity analyses when presenting causal estimates, however this strategy is not necessarily optimal for several reasons.
-First, if different methods disagree it is not possible to know which is correct because the appropriate model of HP is not known.
+First, if different methods disagree it is not possible to know which is correct because the true nature of HP exhibited by the instruments is not known.
 Second, though the IVW approach is most statistically powerful under no HP, it can have high false negative or low true positive rates in the presence of HP compared to other methods.
-Given that pleiotropy has been hypothesised to be universal, defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate.
+Given that pleiotropy has been hypothesised to be universal [@Wagner2011; @Hill2012a], defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate.
 Third, the available methods do not cover all possible models of HP, and therefore an automated method for instrument selection may be necessary.
@@
 Second, we hypothesised that characteristics of the summary data could indicate which method would be most reliable, and we introduce new machine learning approaches that attempt to automate both instrument and method selection.
-Using curated GWAS summary data for thousands of phenotypes, we use these new methods to construct a graph of millions of causal estimates.
-We consider this to be only a 'first draft' of the causal map of the human phenome.
+Using curated GWAS summary data for thousands of phenotypes [@Hemani2016], we use these new methods to construct a graph of millions of causal estimates.
+We consider this to be a 'first draft' of the causal map of the human phenome, but raise caution throughout that its interpretation is far from straightforward
 
@@
 
-The use of summary data in two-sample MR is described in detail elsewhere.
+The use of summary data in two-sample MR is described in detail elsewhere [@Hemani2016].
 A brief outline of the procedure is as follows.
-First, genetic instruments for the exposure trait need to be identified - those SNPs with $p < 10^{-8}$ are retained, collecting their effect sizes and standard errors and effect alleles for the association with the exposure trait.
+First, genetic instruments for the exposure trait need to be identified - those SNPs with $p < 10^{-8}$ are retained in order to ensure that the first assumption of Mendelian randomisation (that the instrument associates with the exposure) is satisfied.
+We collect their effect sizes, standard errors and effect alleles for the association with the exposure trait.
 These can be obtanied manually from complete summary data or from curated lists of significant GWAS associations.
@@
 At its most simple implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome.
+This is known as the inverse-variance weighted (IVW) method.
 
 Given summary data for a large number of traits, it is straightforward to exhaustively analyse the causal relationships of every trait against every other trait for which there is sufficient summary data available.
-Supplementary table X provides a list of all traits that have available GWAS summary data, indicating if they have complete summary data (in which case they can be used as both exposure traits and outcome traits), or if they only have significant associations (in which they can only be used as exposure traits).
-Supplementary table X provides a list of the genetic instruments used for each trait.
+Supplementary table 1 provides a list of all traits that have available GWAS summary data, indicating if they have complete summary data (in which case they can be used as both exposure traits and outcome traits), or if they only have significant associations (in which they can only be used as exposure traits).
 
@@
 
-**Mean-based methods:** Here we consider four nested models.
+**Mean-based methods:** Here we consider four nested models [@Bowden2017].
 The inverse variance weighted (IVW) fixed effects meta-analysis approach assumes that variants exhibit no HP.
 IVW random effects meta-analysis relaxes the HP assumption, allowing it to be present but balanced - such that it only leads to increased heterogeneity around the regression and not introducing bias.
-Fixed effects Egger regression relaxes the HP assumption further by allowing a non-zero intercept which essentially allows horizontal pleiotropy be directional, where it systematically occurs in a specific direction.
-Random effects Egger regression allows heterogeneity around the directional HP, as long as the HP effects are not correlated with the SNP-exposure effects (this is the INSIDE assumption).
+Fixed effects Egger regression [@Bowden2015] relaxes the HP assumption further by allowing a non-zero intercept which essentially allows horizontal pleiotropy to be directional, where it systematically influences the outcome in a specific direction.
+Random effects Egger regression allows heterogeneity around the directional HP, as long as the HP effects are not correlated with the SNP-exposure effects (also known as the INSIDE assumption).
 
-The Rucker framework uses estimates of heterogeneity in the IVW and Egger frameworks to navigate between these nested models.
+The Rucker framework [@Rucker2011], adapted to MR [@Bowden2017] uses estimates of heterogeneity in the IVW and Egger frameworks to navigate between these nested models.
 A jackknife approach (random selection with replacement of instruments) can be used to obtain a sampling distribution for the model estimate amongst these four variations.
@@
 We only use the jackknife approach for associations where there are 15 or more variants to avoid combinatorial saturation.
+The four nested models plus the three Rucker estimates provide seven mean based estimators.
 
-**Median-based methods:** An alternative approach is to take the median effect of all available instruments.
+**Median-based methods:** An alternative approach is to take the median effect of all available instruments [@Bowden2016b].
 This has the advantage that up to half the instruments can be invalid, and the estimate will remain unbiased.
 Developing the approach further to allow stronger instruments to contribute more towards the estimate can be obtained by obtaining the median of the weights of each instrument.
-The penalised weighted median estimator ...
+The penalised weighted median estimator introduces a further weight to the instruments, penalising any instrument that contributes substantially towards the heterogeneity statistic.
+Together, this provides three median based estimators.
 
-**Mode-based methods:** Similar to the median, the mode based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments.
+**Mode-based methods:** Similar to the median, the mode based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017].
 This can be extended in two ways, first by weighting the strength of each cluster by the inverse of the variance of the SNP-outcome associations in the cluster; second by weighting the strength of each cluster by the inverse of the variance of the SNP-exposure associations.
+When weighting by the SNP-outcome standard errors we term this a weighted analysis.
+When *not* weighting by the exposure standard errors we term this the 'no measuremenent error in the exposure (NOME)' assumption.
+Together this provides four mode-based estimators.
 
@@
 
-An approach to inferring the causal direction between phenotypes was developed recently, using the following basic premise. If trait $A$ causes trait $B$ then
+An approach to inferring the causal direction between phenotypes was developed recently [@Hemani2017], using the following basic premise. If trait $A$ causes trait $B$ then
 
@@
 This simple inequality will not hold in some cases, for example $\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}$ where $\rho_{x, x_o}$ and $\rho_{y, y_o}$ are the precision of the measurements of the $x$ and $y$.
-Steiger's Z-test of correlated correlations can be used to formally test the extent to which the two correlations are statistically different.
+Some combinations of confounding effects can also distort the $\rho_{g,x}$ and $\rho_{g,y}$ parameters, as has been discussed in detail previously.
+However, we use it here as an inexpensive and approximate method to identify variants that are likely to be invalid.
+Steiger's Z-test of correlated correlations [@Steiger1980] can be used to formally test the extent to which the two correlations are statistically different.
 
 Here we adapt this approach to automatically filter SNPs that are liable to be invalid (Figure 1a).
-In this case the Steiger test applied to each variant in turn will identify $g_{A}$ as being unlikely to primarily associate with $B$ relative to $A$.
+In this case the Steiger test applied to each variant in turn and we exclude any $g_{A}$ for which the Steiger Z-test has $p > 0.05$, indicating that it is unlikely to primarily associate with $B$ relative to $A$.
 Similarly, for SNPs that influence confounders of $A$ and $B$ or those variants that exhibit horizontal pleiotropy, the difference in $cor(g_{i}, A)^2$ and $cor(g_{i}, B)^2$ will be reduced, increasing the likelihood of the SNP being excluded because the Steiger Z-test is less likely to be significant.
@@
 To estimate $cor(g, x)^2$, if $x$ is continuous we obtain the F-statistic from the reported p-value and sample size and then $cor(g, x)^2 = \frac{F}{N - 2 - F}$.
-If $x$ is binary then our objective is to estimate the variance of the risk liability explained by the SNP, $cor(g, x)^2 = \frac{V_a}{V_a + V_e}$.
-Here, $V_e = \pi^2/3$, and $V_a = 2\beta^2p(1-p)$, where $\beta$ is the log odds ratio and $p$ is the allele frequency of the SNP in the population.
+If $x$ is binary then we estimate the variance of the underlying liability explained by the SNP, $cor(g, x)^2 = \frac{V_a}{V_a + V_e}$.
+Here, $V_e = \pi^2/3$, and $V_a = 2\beta^2p(1-p)$, where $\beta$ is the log odds ratio and $p$ is the allele frequency of the SNP in the population [@Lee2013c].
 $p$ can be estimated using the allele frequency of the SNP in an ascertained sample by deriving the $2 \times 2$ contingency table from the odds ratio $e^\beta$, allele frequency in the ascertained sample $p_{cc}, and number of cases $N_1$ and controls $N_0$.
@@
 
-We consider 14 MR methods, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total.
+We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total.
 In the context of this analysis, each method is considered to be an 'expert', taking a set of SNP-exposure and SNP-outcome effect sizes and their standard errors as inputs.
 Our objective is to select the expert most likely to be correct for a specific MR analysis.
+
+The mixture of experts method is a machine learning approach [@Jordan1994] which seeks to divide a parameter space into subdomains, such that a particular expert is used primarily for problems that reside in a subdomain most suited to that expert.
+In this case our objective is to identify characteristics of the SNP-exposure and SNP-outcome associations for which one specific MR method is most likely to yield highest statistical power for non-null associations, and lowest false discovery rates for null associations.
 This involves creating a 'gating function', whose purpose is to decide which expert to use for a specific MR analysis, given the parameter space that is occupied by that dataset.
+The metrics are a mixture of regression diagnostics and are described in Supplementary table 2.
 
-#### Mixture of experts
-
-The mixture of experts (MoE) method is a machine learning approach which seeks to divide a parameter space into subdomains, such that a particular expert is used primarily for problems that reside in a subdomain most suited to that expert.
-In this case our objective is to identify characteristics of the SNP-exposure and SNP-outcome associations for which one specific MR method is most likely to yield highest statistical power for non-null associations, and lowest false discovery rates for null associations.
+The gating function needs to be trained using data for which the true causal effect is known, and to this end we generated a large number of simulated datasets.
+We trained the gating function using random forest learning algorithms that seek to identify sectors of the parameter space that are most likely to return accurate estimates for a particular expert.
+Figure 2a illustrates how the gating function is trained using simulated data.
+New datasets can then be applied to the trained mixture of experts (MoE) model (Figure 2b).
+There are countless ways to implement this approach, we call this implementation MR-MoE 1.0.
 
@@
 
-The MoE is trained using datasets generated from simulations.
+The MoE is trained using datasets generated from simulations (Figure 2a).
 A *dataset* is the minimum data required to perform 2SMR analysis - four columns comprising the SNP-exposure and SNP-outcome effects and standard errors, and rows corresponding to each SNP that is used as an instrument for the exposure.
@@
 
-**Strategy**
-
-Figure 2a outlines the general strategy behind the MoE implementation.
-Briefly, we record 53 metrics about each dataset, hypothesising that we can use these metrics to predict the performance of each expert, given some optimisation criteria.
-For each method we train random forest decision trees  to predict that method's performance using the dataset's metrics.
-
-Having generated random forest decision trees for each of the 28 methods using 133,000 of the simulations, we then applied them to the remaining 67,000 datasets to predict which method would have the highest performance for each of the remaining datasets.
-Finally we compare the performance of the method selected by the MoE against all remaining methods.
-The default settings for the `randomForest` package in R was used to train the models.
-
 **Optimisation function**
@@
 We aim to maximise statistical power for datasets where $\beta_{x,y} \neq 0$ and minimise false discovery rates for datasets where $\beta_{x,y} = 0$.
-The random forest for a particular method $h(O_{w,d}, \textbf{z}_{d}$ is generated where the training set of input metrics for dataset $d$ is $\textbf{z}_{d}$ and the response is
+To train random forest decision trees to predict performance for a particular method $h(O_{w,d}, \textbf{z}_{d}$ is generated where the training set of input metrics for dataset $d$ is $\textbf{z}_{d}$ and the response (optimisation function) is
 
@@
 $$
+
+**Strategy**
+
+For each training dataset we record a set of 53 metrics $\textbf{z}_{d}$ (Supplementary table 2), and an outcome $O_{w,d}$, which is a measure of how well that method performed for each particular simulated dataset.
+For each of our 28 methods, we need to create a model that predicts the performance of the method based on metrics generated from a dataset.
+To do this, for each method we train random forest decision trees to predict that method's performance using the dataset's metrics.
+The random forest approach is well suited to this problem because there are likely to be non-continuous combinations of different metrics that improve on prediction over, for example, a simple linear model that does not learn about interactions.
+
+A simple hypothetical example - if a dataset has a single outlier but is otherwise exhibiting no heterogeneity then the following methods could arguably perform well:
+
+- an IVW fixed effects analysis with the outlier removed, should the Steiger test be able to detect the outlier
+- a median based approach
+- a mode based approach
+
+deciding between these methods requires finding, in general, which will minimise false discovery rates and maximise true positive rates for that particular scenario.
+
+Having generated random forest decision trees for each of the 28 methods using 133,000 of the simulations, we then applied them to the remaining 67,000 datasets to predict which method would have the highest performance for each of the remaining datasets.
+Finally we compare the performance of the method selected by the MoE against all remaining methods.
+The default settings for the `randomForest` package in R [@Liaw2002] was used to train the models.
+MR-MoE 1.0 is implemented in the TwoSampleMR R package available at [github.com/MRCIEU/TwoSampleMR](https://github.com/MRCIEU/TwoSampleMR).
 
@@
 For specific hypotheses we strongly recommend that estimates from all sensitivity analyses are scrutinised and reported.
+The graph can be queried directly using the cypher language at [http://shark.epi.bris.ac.uk:7474/browser/](http://shark.epi.bris.ac.uk:7474/browser/) or through a basic web interface at [http://shark.epi.bris.ac.uk:3838](http://shark.epi.bris.ac.uk:3838).
 
@@
 We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated datasets comprising both null and non-null causal models.
-For each dataset instruments were selected based on the tophits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.
+For each dataset, instruments were selected based on the tophits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.
 
-Figure 1b shows that the tophits strategy led to 49% of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype.
+Figure 1b shows that the tophits strategy led to over half of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype.
+The proportion of invalid instruments due to reverse cause increased as GWAS discovery sample size increased.
 Applying Steiger filtering reduced this to 25%.
@@
 
-Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to use dataset characteristics to predict the most appropriate of the 28 MR methods and instrument selection strategies (Figure 2a).
+Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to predict the most appropriate of the 28 MR methods and instrument selection strategies to apply to a particular dataset based on its characteristics (Figure 2).
 
-The ability to predict the performance for each of these methods is shown in table 1, with prediction $R^2$ ranging between 0.04 and 0.24.
+The ability to predict the performance for each of these methods is shown in Supplementary table 3, with prediction $R^2$ ranging between 0.04 and 0.24.
 The dataset characteristics with the most importance for each of the predictors differed substantially between each dataset, as well as the frequency for which each of the methods was selected in the testing datasets.
@@
 We compared the MoE performance in the simulations against each of the 28 methods, testing to see if it outperformed all other single methods.
-Figures 2b and 2c show that the MoE approach had the best general performance, obtaining an AUC of 0.84 in terms of classifying the simulations as being null or non-null.
+Figures 3a and 3b show that the MoE approach had the best general performance, obtaining an AUC of 0.84 in terms of classifying the simulations as being null or non-null.
 Notably, the next best methods were median and mode based estimators using Steiger filtering.
-Under the assumption of universal pleiotropy it is clear that all methods suffer from very high false discovery rates.
+Under the assumption of universal pleiotropy it is clear that all methods suffer from high false discovery rates.
 
-### A first draft of the causal map of the human phenome
+### Automated MR analysis of 2407 phenotypes
 
-It is important to note that though the gating function should be applicable to datasets of any distribution of characteristics, the frequencies of chosen methods are relative to the characteristics of the simulated datasets, and would not necessarily be the same frequencies observed empirically.
-Applying the MoE model to the X number of MR associations for which there was sufficient data, we found that
+We applied our analysis using summary data for 2407 phenotypes, including 149 complex traits and diseases, 575 metabolites [@Shin2014; @Kettunen2016] and 1683 plasma protein levels [@Sun2017].
+For the protein levels only the instruments were available, so they could only be evaluated as exposure phenotypes.
+The complex traits and diseases and metabolite levels had complete summary data available in MR-Base, thus they be evaluated as both exposures (if they had significant instruments) and outcomes.
+Together, we evaluated 715681 relationships.
+The majority of these associations could only be evaluated using fewer than 5 SNPs, and so the Wald ratio or IVW fixed effects methods were used, but for 61029 associations the MR-MoE approach was applied.
 
-How many associations reach nominal or FDR corrected significance
-How many bi-directional associations
-How many associations could be declared for a causal chain of 2
+There were 5660 associations following Bonferroni correction ($p < 7.0\times^{-8}$).
+Of these 2918 were obtained from the MR-MoE analysis, while the remainder were estimated using fewer than 5 SNPs using the Wald ratio or IVW fixed effects methods.
+
+The frequencies of the methods chosen by the MR-MoE analysis are shown in Supplementary table 2.
+Amongst those deemed 'significant', the IVW fixed effects analysis using tophits instruments (the only method applicable when there is no evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases.
+
+### Causal associations involving years of schooling
+
+As an example of the database, we performed a look up of associations that causally influence years of schooling [@Okbay2016], or that years of schooling itself causally influences.
+Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4).
+Though many of these putative relationships appear to be plausible, it is clear that there are serious limitations to using MR as a panacea for causal inference.
+For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g.
+childhood intelligence and birth weight) which appear to be causally downstream of years of schooling despite this being temporally impossible.
+The facility to perform these scans has been automated at the MR of everything vs everything (MR-EvE) website, [http://shark.epi.bris.ac.uk:3838](http://shark.epi.bris.ac.uk:3838), though we urge caution on interpretation of these findings.
 
@@
 
-- Limitations of MR that cannot be addressed without biological information
-    + Impossible causal associations
-    + Temporal effects
-    + Risk liability etc
-    + Single variants
-- Further developments required to improve causal estimates
-- The use of causal chains to improve power
-- Repository will continually update - need to time stamp and provide access to older associations
+The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones.
+Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception.
+Following a recent flurry in development of MR methodology, we have devised a machine learning approach, MR-MoE, that seeks to predict the performance of each MR method, selecting the one which is most likely to maximise power whilst minimising FDR for a specific dataset.
+The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy.
+
+Crucially however, we observed in our simulations that none of the available methods were conservative and all, including MR-MoE, had FDR above 0.2.
+Instrumental variables in Mendelian randomisation are typically chosen blind, with GWAS significance being the only criteria.
+We illustrated how confounders and reverse causal associations can easily lead to invalid instruments being used in MR, and that even if those instruments that only influence the exposure directly are used, most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.
+
+We applied the method to the curated set of GWAS summary data present in MR-Base [@hemani2016mr], generating several million MR estimates.
+MR-MoE selected a method that indicated some pattern of horizontal pleiotropy for almost 90% of cases, reinforcing the notion that it is the rule rather than the exception.
+Based on our simulations we expect that those associations that we reported to be 'significant' are liable to high type 1 error rates.
+
+In addition to correctly handling horizontal pleiotropy, there are many other limitations that prevent MR from being a panacea for causal inference.
+Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes.
+Though these associations can be informative, their interpretations as causal relationships are far from clear.
+Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease.
+Again, interpretation of such associations can be problematic.
+
+A large proportion of the associations that were estimated used only a single instrumental variable.
+Methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017], but separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework.
+Other problems can also manifest, for example frailty effects could induce associations for late-onset traits [@Noyce2017]; genetic variants could strongly relate to several phenotypes making it difficult to ascertain which of them is the causal exposure [@Burgess2014a]; and the measured feature for which genetic associations are known could relate in complicated ways to the biological entity that are truly causal [@DaveySmithHemani2014].
+
+We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular dataset, namely, that the data is choosing the model.
+This is liable to increase type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance.
+Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to all other MR methods that do not suffer from this issue.
+
+Despite these limitations, the construction of a causal graph of 'everything versus everything' does have appeal.
+First, though causality is not guaranteed by MR, it can still be highly informative for confirming or negating specific hypotheses.
+Second, it offers new approaches to searching for novel putative associations.
+The potential to exploit GWAS summary data within the properties of graph databases to aid with both of these endeavours could be transformative in biological research.
 
@@
 
-![](images/fig1.pdf)
+![Simulations.
+a) Schematic of how GWAS with sufficient power can lead to the selection of instrumental variables that are invalid.
+We used arbitrary numbers of SNPs and confounders to simulate GWAS summary datasets.
+b) Any SNP that has a direct influence on the exposure, or an influence on a non-confounding intermediate variable, is considered a 'direct' effect.
+The y-axis shows the proportion of instruments selected for analysis due that are either direct associations with the exposure, instruments for the outcome (reverse), or instruments for confounding traits (confounder).
+The proportions are compared over a range of different exposure discovery sample sizes (x-axis) and using either the tophits approach (left) or the Steiger approach (right) for instrument selection.
+c) Top: The false discovery rates from null simulations for each of the 14 methods using either tophits, Steiger filtered variants, or variants that are known to be directly associated with the exposure (oracle, note that direct effects can still exhibit horizontal pleiotropy in these simulations).
+Bottom: The statistical power to detect true causal associations in the non-null simulations.](images/fig1.pdf)
 
@@
 
-![](images/fig2.pdf)
+![Mixture of experts.
+a) Training.
+Datasets are simulated that have either null or non-null causal relationships, and MR is performed by each of the 28 available MR methods.
+In the toy datasets, the columns in black represent 53 metrics about each of the 67,000 training simulations.
+The columns in red are specific to each method, they represent how well that method performed in obtaining the correct answer for each of the datasets.
+Random forests are used to learn the parameter space of the 53 metrics in which a particular dataset is likely to perform well.
+Together, this creates 28 random forest decision trees, one for each method.
+b) Application.
+For a GWAS summary dataset, our objective is to choose the method most likely to return the correct causal estimate.
+Metrics are generated from the dataset and fed into each of the 28 random forest decision trees.
+This provides us with 28 performance predictions.
+Finally, we use the method for which the performance prediction is highest.
+](images/fig2.pdf)
+
+\newpage
+
+![Performance of MoE against all other methods.
+a) The power for non-null datasets is plotted against the FDR for null datasets for each of the 28 methods, plus MoE.
+No single method achieved nominal FDR for these simulations.
+b) Calculating the area under the ROC curve from the values in (a) we plotted the performance in order from lowest to highest.
+Under the assumption of pervasive horizontal pleiotropy, the MoE approach is likely most effective than any other single method.](images/fig3.pdf)
+
+\newpage
+
+![Lookup of causal associations involving Years of Schooling.
+The arrows denote causal direction and the values on the arrows denote the causal effect estimate.
+Only those relationships are shown for which the $FDR < 0.05$.](images/fig4.png)
+
+\newpage
 

### 2017-07-03 docs/manuscript.tex
updated

--- 2017-06-26 docs/manuscript.tex
@@
             pdftitle={Towards the construction of a causal map of the human phenome},
+            pdfauthor={Gibran Hemani, Jie Zheng, Philip Haycock, Oliver Davies, Jack Bowden, Tom Gaunt, George Davey Smith},
             pdfborder={0 0 0},
@@
 \fi
-\usepackage{amsmath}
 
@@
   \posttitle{\par}
-  \author{}
-  \preauthor{}\postauthor{}
+  \author{Gibran Hemani, Jie Zheng, Philip Haycock, Oliver Davies, Jack Bowden,
+Tom Gaunt, George Davey Smith}
+  \preauthor{\centering\large\emph}
+  \postauthor{\par}
   \predate{\centering\large\emph}
   \postdate{\par}
-  \date{26 June 2017}
+  \date{03 July 2017}
+
+\usepackage{amsmath}
 
@@
 \maketitle
+
+\subsection{Abstract}\label{abstract}
+
+A major application for genome-wide association studies (GWAS) has been
+the emerging field of causal inference using Mendelian randomisation
+(MR), where the causal effect between a pair of traits can be estimated
+using only summary level data. MR depends on SNPs exhibiting vertical
+pleiotropy, where the SNP influences an outcome phenotype only through
+an exposure phenotype. Issues arise when this assumption is violated due
+to SNPs exhibiting horizontal pleiotropy, and many methods have been
+developed in an attempt to address this. Here we show that the
+mechanisms that underlie horizontal pleiotropy are numerous, and that
+instrument selection will be increasingly liable to selecting invalid
+instruments as GWAS sample sizes continue to grow. We have developed a
+mixture of experts machine learning approach (MR-MoE 1.0) that improves
+on both power and false discovery rates over all existing methods. Using
+the approach, we systematically estimated the causal effects amongst
+2407 phenotypes, generating a draft of the causal map of the human
+phenome. But we offer caution that its interpretation is far from
+straightforward.
 
@@
 independent genetic variants for a particular exposure can be used
-jointly to improve causal inference, because a) each variant is an
-independent natural experiment, and an overall causal estimate can be
+jointly to improve causal inference, because a) each variant represents
+an independent natural experiment, and an overall causal estimate can be
 obtained by meta-analysing the single estimates from each instrument;
@@
 SNP influences the outcome through no pathways other than the exposure)
-can be evaluated through sensitivity analyses.
+can be evaluated through sensitivity analyses (3--7).
 
 Genome-wide association studies (GWAS) have identified genetic
-instrumental variables for thousands of phenotypes. Recent developments
-in Mendelian randomisation have enabled knowledge of instrumental
-variables to be applied using only summary level data (known as
-two-sample MR, 2SMR). Here, in order to infer the causal effect of an
-exposure on an outcome all that is required is an estimate of the
-genetic effects of the instrumenting SNP on the exposure, and the
+instrumental variables for thousands of phenotypes (8). Recent
+developments in Mendelian randomisation have enabled knowledge of
+instrumental variables to be applied using only summary level data
+(known as two-sample MR, 2SMR) (9). Here, in order to infer the causal
+effect of an exposure on an outcome all that is required is an estimate
+of the genetic effects of the instrumenting SNP on the exposure, and the
 corresponding estimate of the effect on the outcome. This has two major
@@
 limiting the breadth of possible causal estimates only to the
-availability of GWAS summary data for the traits in question.
+availability of GWAS summary data for the traits in question (10).
 
@@
 influence the outcome through a pathway other than the exposure. The
-extent of this problem is not to be understated, and many methods have
-been developed that attempt to reliably obtain unbiased causal estimates
-under specific models of HP. It is considered best practice to report
-estimates from all available methods as sensitivity analyses when
-presenting causal estimates, however this strategy is not necessarily
-optimal for several reasons. First, if different methods disagree it is
-not possible to know which is correct because the appropriate model of
-HP is not known. Second, though the IVW approach is most statistically
-powerful under no HP, it can have high false negative or low true
-positive rates in the presence of HP compared to other methods. Given
-that pleiotropy has been hypothesised to be universal, defaulting to the
-IVW method in the first instance and using other methods as sensitivity
-analyses may not be appropriate. Third, the available methods do not
-cover all possible models of HP, and therefore an automated method for
-instrument selection may be necessary. Fourth, it could be of interest
-to make causal effect estimates for thousands of traits, in which case a
-discerning evaluation of each causal effect of interest may not be
-possible or convenient.
+extent of this phenomenon is not to be understated, and many methods
+have been developed that attempt to reliably obtain unbiased causal
+estimates under specific models of HP (3--6). It is considered best
+practice to report estimates from all available methods as sensitivity
+analyses when presenting causal estimates, however this strategy is not
+necessarily optimal for several reasons. First, if different methods
+disagree it is not possible to know which is correct because the true
+nature of HP exhibited by the instruments is not known. Second, though
+the IVW approach is most statistically powerful under no HP, it can have
+high false negative or low true positive rates in the presence of HP
+compared to other methods. Given that pleiotropy has been hypothesised
+to be universal (11,12), defaulting to the IVW method in the first
+instance and using other methods as sensitivity analyses may not be
+appropriate. Third, the available methods do not cover all possible
+models of HP, and therefore an automated method for instrument selection
+may be necessary. Fourth, it could be of interest to make causal effect
+estimates for thousands of traits, in which case a discerning evaluation
+of each causal effect of interest may not be possible or convenient.
 
 In this paper we introduce two innovations towards improving the
-reliability of MR estimates. First, we introduce an approach to
-automatically discard genetic variants that are likely to be invalid.
-Second, we hypothesised that characteristics of the summary data could
-indicate which method would be most reliable, and we introduce new
-machine learning approaches that attempt to automate both instrument and
-method selection. Using curated GWAS summary data for thousands of
-phenotypes, we use these new methods to construct a graph of millions of
-causal estimates. We consider this to be only a `first draft' of the
-causal map of the human phenome.
+reliability of MR estimates. First, we introduce an approach to discard
+genetic variants that are likely to be invalid. Second, we hypothesised
+that characteristics of the summary data could indicate which method
+would be most reliable, and we introduce new machine learning approaches
+that attempt to automate both instrument and method selection. Using
+curated GWAS summary data for thousands of phenotypes (10), we use these
+new methods to construct a graph of millions of causal estimates. We
+consider this to be a `first draft' of the causal map of the human
+phenome, but raise caution throughout that its interpretation is far
+from straightforward
 
@@
 The use of summary data in two-sample MR is described in detail
-elsewhere. A brief outline of the procedure is as follows. First,
+elsewhere (10). A brief outline of the procedure is as follows. First,
 genetic instruments for the exposure trait need to be identified - those
-SNPs with \(p < 10^{-8}\) are retained, collecting their effect sizes
-and standard errors and effect alleles for the association with the
-exposure trait. These can be obtanied manually from complete summary
-data or from curated lists of significant GWAS associations. Next, the
-effects of those SNPs on the outcome need to be obtained, typically
-necessitating complete summary data because it is unlikely that these
-SNPs will have reached genome-wide significance (and therefore be
-present in curated catalogues). At its most simple implementation, the
-regression of the SNP-exposure effect sizes against the SNP-outcome
-effect sizes, with greater weight afforded to those SNPs with smaller
-SNP-outcome standard errors, provides the estimate of the causal effect
-of the exposure on the outcome.
+SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
+assumption of Mendelian randomisation (that the instrument associates
+with the exposure) is satisfied. We collect their effect sizes, standard
+errors and effect alleles for the association with the exposure trait.
+These can be obtanied manually from complete summary data or from
+curated lists of significant GWAS associations. Next, the effects of
+those SNPs on the outcome need to be obtained, typically necessitating
+complete summary data because it is unlikely that these SNPs will have
+reached genome-wide significance (and therefore be present in curated
+catalogues). At its most simple implementation, the regression of the
+SNP-exposure effect sizes against the SNP-outcome effect sizes, with
+greater weight afforded to those SNPs with smaller SNP-outcome standard
+errors, provides the estimate of the causal effect of the exposure on
+the outcome. This is known as the inverse-variance weighted (IVW)
+method.
 
@@
 every other trait for which there is sufficient summary data available.
-Supplementary table X provides a list of all traits that have available
+Supplementary table 1 provides a list of all traits that have available
 GWAS summary data, indicating if they have complete summary data (in
@@
 or if they only have significant associations (in which they can only be
-used as exposure traits). Supplementary table X provides a list of the
-genetic instruments used for each trait.
+used as exposure traits).
 
@@
 
-\textbf{Mean-based methods:} Here we consider four nested models. The
-inverse variance weighted (IVW) fixed effects meta-analysis approach
+\textbf{Mean-based methods:} Here we consider four nested models (5).
+The inverse variance weighted (IVW) fixed effects meta-analysis approach
 assumes that variants exhibit no HP. IVW random effects meta-analysis
@@
 that it only leads to increased heterogeneity around the regression and
-not introducing bias. Fixed effects Egger regression relaxes the HP
+not introducing bias. Fixed effects Egger regression (3) relaxes the HP
 assumption further by allowing a non-zero intercept which essentially
-allows horizontal pleiotropy be directional, where it systematically
-occurs in a specific direction. Random effects Egger regression allows
-heterogeneity around the directional HP, as long as the HP effects are
-not correlated with the SNP-exposure effects (this is the INSIDE
-assumption).
+allows horizontal pleiotropy to be directional, where it systematically
+influences the outcome in a specific direction. Random effects Egger
+regression allows heterogeneity around the directional HP, as long as
+the HP effects are not correlated with the SNP-exposure effects (also
+known as the INSIDE assumption).
 
-The Rucker framework uses estimates of heterogeneity in the IVW and
-Egger frameworks to navigate between these nested models. A jackknife
-approach (random selection with replacement of instruments) can be used
-to obtain a sampling distribution for the model estimate amongst these
-four variations. Using 1000 rounds of jackknife estimates, we can obtain
-a final estimate using the mean or the median of the distribution. We
-only use the jackknife approach for associations where there are 15 or
-more variants to avoid combinatorial saturation.
+The Rucker framework (13), adapted to MR (5) uses estimates of
+heterogeneity in the IVW and Egger frameworks to navigate between these
+nested models. A jackknife approach (random selection with replacement
+of instruments) can be used to obtain a sampling distribution for the
+model estimate amongst these four variations. Using 1000 rounds of
+jackknife estimates, we can obtain a final estimate using the mean or
+the median of the distribution. We only use the jackknife approach for
+associations where there are 15 or more variants to avoid combinatorial
+saturation. The four nested models plus the three Rucker estimates
+provide seven mean based estimators.
 
 \textbf{Median-based methods:} An alternative approach is to take the
-median effect of all available instruments. This has the advantage that
-up to half the instruments can be invalid, and the estimate will remain
-unbiased. Developing the approach further to allow stronger instruments
-to contribute more towards the estimate can be obtained by obtaining the
-median of the weights of each instrument. The penalised weighted median
-estimator \ldots{}
+median effect of all available instruments (4). This has the advantage
+that up to half the instruments can be invalid, and the estimate will
+remain unbiased. Developing the approach further to allow stronger
+instruments to contribute more towards the estimate can be obtained by
+obtaining the median of the weights of each instrument. The penalised
+weighted median estimator introduces a further weight to the
+instruments, penalising any instrument that contributes substantially
+towards the heterogeneity statistic. Together, this provides three
+median based estimators.
 
@@
 causal effects, and returns the final causal effect estimate based on
-the cluster that has the largest number of instruments. This can be
+the cluster that has the largest number of instruments (6). This can be
 extended in two ways, first by weighting the strength of each cluster by
@@
 cluster; second by weighting the strength of each cluster by the inverse
-of the variance of the SNP-exposure associations.
+of the variance of the SNP-exposure associations. When weighting by the
+SNP-outcome standard errors we term this a weighted analysis. When
+\emph{not} weighting by the exposure standard errors we term this the
+`no measuremenent error in the exposure (NOME)' assumption. Together
+this provides four mode-based estimators.
 
@@
 An approach to inferring the causal direction between phenotypes was
-developed recently, using the following basic premise. If trait \(A\)
-causes trait \(B\) then
+developed recently (7), using the following basic premise. If trait
+\(A\) causes trait \(B\) then
 
@@
 \(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
-\(y\). Steiger's Z-test of correlated correlations can be used to
-formally test the extent to which the two correlations are statistically
+\(y\). Some combinations of confounding effects can also distort the
+\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
+detail previously. However, we use it here as an inexpensive and
+approximate method to identify variants that are likely to be invalid.
+Steiger's Z-test of correlated correlations (14) can be used to formally
+test the extent to which the two correlations are statistically
 different.
@@
 to be invalid (Figure 1a). In this case the Steiger test applied to each
-variant in turn will identify \(g_{A}\) as being unlikely to primarily
+variant in turn and we exclude any \(g_{A}\) for which the Steiger
+Z-test has \(p > 0.05\), indicating that it is unlikely to primarily
 associate with \(B\) relative to \(A\). Similarly, for SNPs that
@@
 F-statistic from the reported p-value and sample size and then
-\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then our
-objective is to estimate the variance of the risk liability explained by
-the SNP, \(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here,
-\(V_e = \pi^2/3\), and \(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the
-log odds ratio and \(p\) is the allele frequency of the SNP in the
-population. \(p\) can be estimated using the allele frequency of the SNP
-in an ascertained sample by deriving the \(2 \times 2\) contingency
-table from the odds ratio \(e^\beta\), allele frequency in the
-ascertained sample \$p\_\{cc\}, and number of cases \(N_1\) and controls
-\(N_0\).
+\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
+estimate the variance of the underlying liability explained by the SNP,
+\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
+\(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the log odds ratio and
+\(p\) is the allele frequency of the SNP in the population (15). \(p\)
+can be estimated using the allele frequency of the SNP in an ascertained
+sample by deriving the \(2 \times 2\) contingency table from the odds
+ratio \(e^\beta\), allele frequency in the ascertained sample
+\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\).
 
@@
 
-We consider 14 MR methods, for which instruments can be supplied using
-two instrument selection strategies, leading to 28 methods in total. In
-the context of this analysis, each method is considered to be an
-`expert', taking a set of SNP-exposure and SNP-outcome effect sizes and
-their standard errors as inputs. Our objective is to select the expert
-most likely to be correct for a specific MR analysis.
+We consider the 14 MR methods described above, for which instruments can
+be supplied using two instrument selection strategies, leading to 28
+methods in total. In the context of this analysis, each method is
+considered to be an `expert', taking a set of SNP-exposure and
+SNP-outcome effect sizes and their standard errors as inputs. Our
+objective is to select the expert most likely to be correct for a
+specific MR analysis.
 
-\paragraph{Mixture of experts}\label{mixture-of-experts}
-
-The mixture of experts (MoE) method is a machine learning approach which
+The mixture of experts method is a machine learning approach (16) which
 seeks to divide a parameter space into subdomains, such that a
@@
 highest statistical power for non-null associations, and lowest false
-discovery rates for null associations.
+discovery rates for null associations. This involves creating a `gating
+function', whose purpose is to decide which expert to use for a specific
+MR analysis, given the parameter space that is occupied by that dataset.
+The metrics are a mixture of regression diagnostics and are described in
+Supplementary table 2.
+
+The gating function needs to be trained using data for which the true
+causal effect is known, and to this end we generated a large number of
+simulated datasets. We trained the gating function using random forest
+learning algorithms that seek to identify sectors of the parameter space
+that are most likely to return accurate estimates for a particular
+expert. Figure 2a illustrates how the gating function is trained using
+simulated data. New datasets can then be applied to the trained mixture
+of experts (MoE) model (Figure 2b). There are countless ways to
+implement this approach, we call this implementation MR-MoE 1.0.
 
@@
 
-The MoE is trained using datasets generated from simulations. A
-\emph{dataset} is the minimum data required to perform 2SMR analysis -
-four columns comprising the SNP-exposure and SNP-outcome effects and
-standard errors, and rows corresponding to each SNP that is used as an
-instrument for the exposure. This \emph{dataset} can be fed into any of
-the 28 experts to obtain MR causal effects. The simulations used to
-generate these datasets seek to cover a range of pleiotropic scenarios,
-including where some proportion of SNPs exhibit directional or balanced
-horizontal pleiotropy, or where SNPs influencing confounding variables.
+The MoE is trained using datasets generated from simulations (Figure
+2a). A \emph{dataset} is the minimum data required to perform 2SMR
+analysis - four columns comprising the SNP-exposure and SNP-outcome
+effects and standard errors, and rows corresponding to each SNP that is
+used as an instrument for the exposure. This \emph{dataset} can be fed
+into any of the 28 experts to obtain MR causal effects. The simulations
+used to generate these datasets seek to cover a range of pleiotropic
+scenarios, including where some proportion of SNPs exhibit directional
+or balanced horizontal pleiotropy, or where SNPs influencing confounding
+variables.
 
@@
 
-\textbf{Strategy}
-
-Figure 2a outlines the general strategy behind the MoE implementation.
-Briefly, we record 53 metrics about each dataset, hypothesising that we
-can use these metrics to predict the performance of each expert, given
-some optimisation criteria. For each method we train random forest
-decision trees to predict that method's performance using the dataset's
-metrics.
-
-Having generated random forest decision trees for each of the 28 methods
-using 133,000 of the simulations, we then applied them to the remaining
-67,000 datasets to predict which method would have the highest
-performance for each of the remaining datasets. Finally we compare the
-performance of the method selected by the MoE against all remaining
-methods. The default settings for the \texttt{randomForest} package in R
-was used to train the models.
-
 \textbf{Optimisation function}
@@
 \(\beta_{x,y} \neq 0\) and minimise false discovery rates for datasets
-where \(\beta_{x,y} = 0\). The random forest for a particular method
+where \(\beta_{x,y} = 0\). To train random forest decision trees to
+predict performance for a particular method
 \(h(O_{w,d}, \textbf{z}_{d}\) is generated where the training set of
 input metrics for dataset \(d\) is \(\textbf{z}_{d}\) and the response
-is
+(optimisation function) is
 
@@
 
-\subsubsection{Graph database}\label{graph-database}
+\textbf{Strategy}
+
+For each training dataset we record a set of 53 metrics
+\(\textbf{z}_{d}\) (Supplementary table 2), and an outcome \(O_{w,d}\),
+which is a measure of how well that method performed for each particular
+simulated dataset. For each of our 28 methods, we need to create a model
+that predicts the performance of the method based on metrics generated
+from a dataset. To do this, for each method we train random forest
+decision trees to predict that method's performance using the dataset's
+metrics. The random forest approach is well suited to this problem
+because there are likely to be non-continuous combinations of different
+metrics that improve on prediction over, for example, a simple linear
+model that does not learn about interactions.
+
+A simple hypothetical example - if a dataset has a single outlier but is
+otherwise exhibiting no heterogeneity then the following methods could
+arguably perform well:
+
+\begin{itemize}
+\tightlist
+\item
+  an IVW fixed effects analysis with the outlier removed, should the
+  Steiger test be able to detect the outlier
+\item
+  a median based approach
+\item
+  a mode based approach
+\end{itemize}
+
+deciding between these methods requires finding, in general, which will
+minimise false discovery rates and maximise true positive rates for that
+particular scenario.
+
+Having generated random forest decision trees for each of the 28 methods
+using 133,000 of the simulations, we then applied them to the remaining
+67,000 datasets to predict which method would have the highest
+performance for each of the remaining datasets. Finally we compare the
+performance of the method selected by the MoE against all remaining
+methods. The default settings for the \texttt{randomForest} package in R
+(17) was used to train the models. MR-MoE 1.0 is implemented in the
+TwoSampleMR R package available at
+\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}.
+
+\subsubsection{Graph database of MR
+estimates}\label{graph-database-of-mr-estimates}
 
@@
 For specific hypotheses we strongly recommend that estimates from all
-sensitivity analyses are scrutinised and reported.
+sensitivity analyses are scrutinised and reported. The graph can be
+queried directly using the cypher language at
+\url{http://shark.epi.bris.ac.uk:7474/browser/} or through a basic web
+interface at \url{http://shark.epi.bris.ac.uk:3838}.
 
@@
 
+As statistical power for GWAS studies improves the likelihood of a
+significant association being discovered that doesn't act primarily on
+the trait of interest increases (Figure 1a). We evaluated the efficacy
+of the Steiger filtering approach for improving instrument selection
+using 100,000 simulated datasets comprising both null and non-null
+causal models. For each dataset, instruments were selected based on the
+tophits strategy and the Steiger filtering strategy, and MR was
+performed using 14 different methods based on instruments selected from
+each of these strategies.
+
+Figure 1b shows that the tophits strategy led to 49\% of the instruments
+being primarily associated with either confounders or the outcome
+phenotype, not the exposure phenotype. Applying Steiger filtering
+reduced this to 25\%. Consequently, the false discovery rates (FDR) for
+12 of the 14 methods reduced substantially when applied using Steiger
+filtered instruments (Figure 1c). The true positive rates for the
+methods based on Steiger filtering did however reduce slightly for 10 of
+the 14 methods.
+
 \subsubsection{Mixture of experts method selection improves over any
@@
 
-Number of associations using different approaches Which methods were
-preferentially chosen
+Following evidence that the Steiger filtering approach can improve on
+existing methods, we next hypothesised that a mixture of experts (MoE)
+model would be able to use dataset characteristics to predict the most
+appropriate of the 28 MR methods and instrument selection strategies
+(Figure 2).
 
-\subsubsection{A first draft of the causal map of the human
-phenome}\label{a-first-draft-of-the-causal-map-of-the-human-phenome}
+The ability to predict the performance for each of these methods is
+shown in table 1, with prediction \(R^2\) ranging between 0.04 and 0.24.
+The dataset characteristics with the most importance for each of the
+predictors differed substantially between each dataset, as well as the
+frequency for which each of the methods was selected in the testing
+datasets. The FDR of each method when chosen, compared to their averages
+across all datasets, reduced; and likewise the true positive rates of
+each method when chosen increased compared to their simulation-wide
+averages.
 
-How many associations reach nominal or FDR corrected significance How
-many bi-directional associations How many associations could be declared
-for a causal chain of 2
+We compared the MoE performance in the simulations against each of the
+28 methods, testing to see if it outperformed all other single methods.
+Figures 3a and 3b show that the MoE approach had the best general
+performance, obtaining an AUC of 0.84 in terms of classifying the
+simulations as being null or non-null. Notably, the next best methods
+were median and mode based estimators using Steiger filtering. Under the
+assumption of universal pleiotropy it is clear that all methods suffer
+from high false discovery rates.
+
+\subsubsection{Automated MR analysis of 2407
+phenotypes}\label{automated-mr-analysis-of-2407-phenotypes}
+
+We applied our analysis using summary data for 2407 phenotypes,
+including 149 complex traits and diseases, 575 metabolites (18,19) and
+1683 plasma protein levels (20). For the protein levels only the
+instruments were available, so they could only be evaluated as exposure
+phenotypes. The complex traits and diseases and metabolite levels had
+complete summary data available in MR-Base, thus they be evaluated as
+both exposures (if they had significant instruments) and outcomes.
+Together, we evaluated 715681 relationships. The majority of these
+associations could only be evaluated using fewer than 5 SNPs, and so the
+Wald ratio or IVW fixed effects methods were used, but for 61029
+associations the MR-MoE approach was applied.
+
+There were 5660 associations following Bonferroni correction
+(\(p < 7.0\times^{-8}\)). Of these 2918 were obtained from the MR-MoE
+analysis, while the remainder were estimated using fewer than 5 SNPs
+using the Wald ratio or IVW fixed effects methods.
+
+The frequencies of the methods chosen by the MR-MoE analysis are shown
+in Supplementary table 2. Amongst those deemed `significant', the IVW
+fixed effects analysis using tophits instruments (the only method
+applicable when there is no evidence of horizontal pleiotropy of any
+sort) was selected in only 10.4\% of cases.
+
+\subsubsection{Causal associations involving years of
+schooling}\label{causal-associations-involving-years-of-schooling}
+
+As an example of the database, we performed a look up of associations
+that causally influence years of schooling (21), or that years of
+schooling itself causally influences. Using a false-discovery rate of
+0.05, 45 traits were returned as having some direction of causality with
+years of schooling (Figure 4). Though many of these putative
+relationships appear to be plausible, it is clear that there are serious
+limitations to using MR as a panacea for causal inference. For example,
+bi-directional causal relationships with college completion exist
+because of the similarity in trait definitions; and there are several
+traits (e.g.~childhood intelligence and birth weight) which appear to be
+causally downstream of years of schooling despite this being temporally
+impossible. The facility to perform these scans has been automated at
+the MR of everything vs everything (MR-EvE) website,
+\url{http://shark.epi.bris.ac.uk:3838}, though we urge caution on
+interpretation of these findings.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Limitations of MR that cannot be addressed without biological
-  information
+Following a recent flurry in development of MR methodology, we have
+devised a machine learning approach, MR-MoE, that seeks to predict the
+performance of each MR method, selecting the one which is most likely to
+maximise power whilst minimising FDR for a specific dataset. The mixture
+of experts approach that we present here is trained using random forest
+decision trees applied to extensive and diverse simulations, and we
+demonstrate that it makes substantial performance improvements over any
+other single method.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Impossible causal associations
-  \item
-    Temporal effects
-  \item
-    Risk liability etc
-  \item
-    Single variants
-  \end{itemize}
-\item
-  Further developments required to improve causal estimates
-\item
-  The use of causal chains to improve power
-\end{itemize}
+Crucially, we observed in our simulations that none of the available
+methods were conservative and all, including MR-MoE, had FDR above 0.2.
+Instrumental variables in Mendelian randomisation are typically chosen
+blind, with GWAS significance being the only criteria. We illustrated
+how confounders and reverse causal associations can easily lead to
+invalid instruments being used in MR, and that even if those instruments
+that only influence the exposure directly, most randomly generated
+patterns of horizontal pleiotropy cannot be adequately accounted for by
+any single MR method.
+
+We applied the method to a curated set of GWAS summary data that was
+present in MR-Base (22), generating hundreds of thousands of MR
+estimates. MR-MoE selected a method that indicated some pattern of
+horizontal pleiotropy for almost 90\% of cases, suggesting that it is
+the rule rather than the exception. Based on our simulations we expect
+that those associations that we reported to be `significant' are liable
+to high type 1 error rates.
+
+In addition to correctly handling horizontal pleiotropy, there are many
+other limitations that prevent MR from being a panacea for causal
+inference. Many of the associations are biologically impossible, for
+example where early life phenotypes appear to be influenced by later
+stage phenotypes. Though these associations can be informative, their
+interpretations as causal relationships are far from clear. Similarly,
+often disease traits appear to causally relate to other phenotypes, but
+GWAS is typically performed on the liability scale, hence the causal
+estimate reflects not the presence or absence of disease, but the
+underlying risk of disease. Again, interpretation of such associations
+can be problematic.
+
+A large proportion of the associations that were estimated used only a
+single instrumental variable. Methods are emerging to attempt to
+delineate between models of reverse cause, pleiotropy and multiple
+causal variants in the same region (23,24), but separating vertical from
+horizontal pleiotropy with a single instrument is not yet possible in
+the two-sample MR framework. Other problems can also manifest, for
+example frailty effects could induce associations for late-onset traits
+(25). We also encounter a potential new problem when it comes to using
+machine learning approaches to infer the correct model for a particular
+dataset, namely, that the data is choosing the model. This is liable to
+increase type 1 error rates, even though we have attempted to separate
+the information used for optimisation from the information used to
+predict performance. Though MR-MoE has high type 1 error rates, they
+still remain amongst the lowest compared to all other methods that do
+not suffer from this issue.
+
+Despite these limitations, the construction of a causal graph of
+`everything versus everything' does have appeal. First, though causality
+is not guaranteed by MR, it can still be highly informative for
+confirming or negating specific hypotheses, or to search for novel
+putative associations. The potential to exploit GWAS summary data within
+the properties of graph databases to aid with both of these endeavours
+could be transformative in biological research.
+
+\subsection{Figures}\label{figures}
+
+\newpage
+
+\begin{figure}
+\centering
+\includegraphics{images/fig1.pdf}
+\caption{Simulations. a) Schematic of how GWAS with sufficient power can
+lead to the selection of instrumental variables that are invalid. We
+used arbitrary numbers of SNPs and confounders to simulate GWAS summary
+datasets. b) Any SNP that has a direct influence on the exposure, or an
+influence on a non-confounding intermediate variable, is considered a
+`direct' effect. The bar-chart shows the number of direct effect
+instruments retained for analysis using either the tophits approach or
+the Steiger approach for instrument selection. c) Top: The false
+discovery rates from null simulations for each of the 14 methods using
+either tophits, Steiger filtered variants, or variants that are known to
+be directly associated with the exposure (oracle, note that direct
+effects can still exhibit horizontal pleiotropy in these simulations).
+Bottom: The statistical power to detect true causal associations in the
+non-null simulations.}
+\end{figure}
+
+\newpage
+
+\begin{figure}
+\centering
+\includegraphics{images/fig2.pdf}
+\caption{Mixture of experts. a) Training. Datasets are simulated that
+have either null or non-null causal relationships, and MR is performed
+by each of the 28 available MR methods. In the toy datasets, the columns
+in black represent 53 metrics about each of the 67,000 training
+simulations. The columns in red are specific to each method, they
+represent how well that method performed in obtaining the correct answer
+for each of the datasets. Random forests are used to learn the parameter
+space of the 53 metrics in which a particular dataset is likely to
+perform well. Together, this creates 28 random forest decision trees,
+one for each method. b) Application. For a GWAS summary dataset, our
+objective is to choose the method most likely to return the correct
+causal estimate. Metrics are generated from the dataset and fed into
+each of the 28 random forest decision trees. This provides us with 28
+performance predictions. Finally, we use the method for which the
+performance prediction is highest.}
+\end{figure}
+
+\newpage
+
+\begin{figure}
+\centering
+\includegraphics{images/fig3.pdf}
+\caption{Performance of MoE against all other methods. a) The power for
+non-null datasets is plotted against the FDR for null datasets for each
+of the 28 methods, plus MoE. No single method achieved nominal FDR for
+these simulations. b) Calculating the area under the ROC curve from the
+values in (a) we plotted the performance in order from lowest to
+highest. Under the assumption of pervasive horizontal pleiotropy, the
+MoE approach is likely most effective than any other single method.}
+\end{figure}
+
+\newpage
+
+\begin{figure}
+\centering
+\includegraphics{images/fig4.png}
+\caption{Lookup of causal associations involving Years of Schooling. The
+arrows denote causal direction and the values on the arrows denote the
+causal effect estimate. Only those relationships are shown for which the
+\(FDR < 0.05\).}
+\end{figure}
 
@@
 
+\hypertarget{ref-Bowden2015}{}
+3. Bowden J, Davey Smith G, Burgess S. Mendelian randomization with
+invalid instruments: effect estimation and bias detection through Egger
+regression. International Journal of Epidemiology. 2015;44(2):512--25.
+
+\hypertarget{ref-Bowden2016b}{}
+4. Bowden J, Davey Smith G, Haycock PC, Burgess S. Consistent Estimation
+in Mendelian Randomization with Some Invalid Instruments Using a
+Weighted Median Estimator. Genetic Epidemiology {[}Internet{]}. 2016
+May;40(4):304--14. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/27061298\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4849733\%20http://doi.wiley.com/10.1002/gepi.21965}{http://www.ncbi.nlm.nih.gov/pubmed/27061298 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4849733 http://doi.wiley.com/10.1002/gepi.21965}
+
+\hypertarget{ref-Bowden2017}{}
+5. Bowden J, Del~Greco~M F, Minelli C, Davey Smith G, Sheehan N,
+Thompson J. A framework for the investigation of pleiotropy in
+two-sample summary data Mendelian randomization. Statistics in Medicine
+{[}Internet{]}. 2017; Available from:
+\url{http://doi.wiley.com/10.1002/sim.7221}
+
+\hypertarget{ref-Hartwig2017}{}
+6. Hartwig FP, Davey Smith G, Bowden J. Robust Inference In Two-Sample
+Mendelian Randomisation Via The Zero Modal Pleiotropy Assumption.
+bioRxiv {[}Internet{]}. 2017; Available from:
+\url{http://biorxiv.org/content/early/2017/04/10/126102}
+
+\hypertarget{ref-Hemani2017}{}
+7. Hemani G, Tilling K, Davey Smith G. Orienting The Causal Relationship
+Between Imprecisely Measured Traits Using Genetic Instruments. bioRxiv
+{[}Internet{]}. 2017; Available from:
+\url{http://biorxiv.org/content/early/2017/03/15/117101}
+
+\hypertarget{ref-Hindorff2010}{}
+8. Hindorff LA, Junkins HA, Hall PN, Mehta JP, Manolio TA. A Catalog of
+Published Genome-Wide Association Studies, available at
+http://www.genome.gov/gwastudies. Accessed 12/10/2010. 2010.
+
+\hypertarget{ref-Pierce2013}{}
+9. Pierce BL, Burgess S. Efficient design for Mendelian randomization
+studies: subsample and 2-sample instrumental variable estimators.
+American journal of epidemiology {[}Internet{]}. 2013
+Oct;178(7):1177--84. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+
+\hypertarget{ref-Hemani2016}{}
+10. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
+MR-Base: a platform for systematic causal inference across the phenome
+using billions of genetic associations. BioRxiv. 2016;10.1101/07.
+
+\hypertarget{ref-Wagner2011}{}
+11. Wagner GP, Zhang J. The pleiotropic structure of the
+genotype--phenotype map: the evolvability of complex organisms. Nature
+Reviews Genetics {[}Internet{]}. 2011 Mar;12(3):204--13. Available from:
+\url{http://www.nature.com/doifinder/10.1038/nrg2949}
+
+\hypertarget{ref-Hill2012a}{}
+12. Hill WG, Zhang X-S. Assessing pleiotropy and its evolutionary
+consequences: pleiotropy is not necessarily limited, nor need it hinder
+the evolution of complexity. Nature Reviews Genetics {[}Internet{]}.
+2012 Feb;13(4):296. Available from:
+\url{http://www.ncbi.nlm.nih.gov/pubmed/22349131}
+
+\hypertarget{ref-Rucker2011}{}
+13. Rucker G, Schwarzer G, Carpenter JR, Binder H, Schumacher M.
+Treatment-effect estimates adjusted for small-study effects via a limit
+meta-analysis. Biostatistics {[}Internet{]}. 2011 Jan;12(1):122--42.
+Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/20656692\%20https://academic.oup.com/biostatistics/article-lookup/doi/10.1093/biostatistics/kxq046}{http://www.ncbi.nlm.nih.gov/pubmed/20656692 https://academic.oup.com/biostatistics/article-lookup/doi/10.1093/biostatistics/kxq046}
+
+\hypertarget{ref-Steiger1980}{}
+14. Steiger JH. Tests for comparing elements of a correlation matrix.
+Psychological Bulletin. 1980;87(2):245--51.
+
+\hypertarget{ref-Lee2013c}{}
+15. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+association studies: quantification of power and genomic prediction
+accuracy. PLoS One. 2013;8(8):e71494.
+
+\hypertarget{ref-Jordan1994}{}
+16. Jordan MI, Jacobs RA. Hierarchical Mixtures of Experts and the EM
+Algorithm. Neural Computation {[}Internet{]}. 1994 Mar;6(2):181--214.
+Available from:
+\url{http://www.mitpressjournals.org/doi/10.1162/neco.1994.6.2.181}
+
+\hypertarget{ref-Liaw2002}{}
+17. Liaw A, Wiener M. Classification and Regression by randomForest. R
+News {[}Internet{]}. 2002;2(3):18--22. Available from:
+\url{http://cran.r-project.org/doc/Rnews/}
+
+\hypertarget{ref-Shin2014}{}
+18. Shin S-Y, Fauman EB, Petersen A-K, Krumsiek J, Santos R, Huang J, et
+al. An atlas of genetic influences on human blood metabolites. Nature
+genetics {[}Internet{]}. 2014 Jun;46(6):543--50. Available from:
+\url{http://dx.doi.org/10.1038/ng.2982}
+
+\hypertarget{ref-Kettunen2016}{}
+19. Kettunen J, Demirkan A, Wurtz P, Draisma HHM, Haller T, Rawal R, et
+al. Genome-wide study for circulating metabolites identifies 62 loci and
+reveals novel systemic effects of LPA. Nat Commun {[}Internet{]}. 2016
+Mar;7. Available from:
+\href{http://dx.doi.org/10.1038/ncomms11122\%20http://10.0.4.14/ncomms11122}{http://dx.doi.org/10.1038/ncomms11122 http://10.0.4.14/ncomms11122}
+
+\hypertarget{ref-Sun2017}{}
+20. Sun BB, Maranville JC, Peters JE, Stacey D, Staley JR, Blackshaw J,
+et al. Consequences Of Natural Perturbations In The Human Plasma
+Proteome. bioRxiv {[}Internet{]}. 2017; Available from:
+\url{http://biorxiv.org/content/early/2017/05/05/134551}
+
+\hypertarget{ref-Okbay2016}{}
+21. Okbay A, Beauchamp JP, Fontana MA, Lee JJ, Pers TH, Rietveld CA, et
+al. Genome-wide association study identifies 74 loci associated with
+educational attainment. Nature {[}Internet{]}. 2016
+May;533(7604):539--42. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/27225129\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4883595\%20http://www.nature.com/doifinder/10.1038/nature17671}{http://www.ncbi.nlm.nih.gov/pubmed/27225129 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4883595 http://www.nature.com/doifinder/10.1038/nature17671}
+
+\hypertarget{ref-hemani2016mr}{}
+22. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
+MR-Base: a platform for systematic causal inference across the phenome
+using billions of genetic associations. bioRxiv. 2016;78972.
+
+\hypertarget{ref-Zhu2016}{}
+23. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+Integration of summary data from GWAS and eQTL studies predicts complex
+trait gene targets. Nature Genetics {[}Internet{]}. 2016
+Mar;48(5):481--7. Available from:
+\url{http://www.nature.com/doifinder/10.1038/ng.3538}
+
+\hypertarget{ref-Richardson2017}{}
+24. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
+CL, et al. Causal epigenome-wide association study identifies CpG sites
+that influence cardiovascular disease risk. bioRxiv {[}Internet{]}.
+2017; Available from:
+\url{http://biorxiv.org/content/early/2017/04/29/132019}
+
+\hypertarget{ref-Noyce2017}{}
+25. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
+E, et al. Estimating the causal influence of body mass index on risk of
+Parkinson disease: A Mendelian randomisation study. Brayne C, editor.
+PLOS Medicine {[}Internet{]}. 2017 Jun;14(6):e1002314. Available from:
+\url{http://dx.plos.org/10.1371/journal.pmed.1002314}
+
 \end{document}

### 2017-07-03 docs/manuscript.rmd
updated

@@
 ---
-title: Towards the construction of a causal map of the human phenome
-author: Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George Davey Smith
+title: Learning pleiotropic signatures to construct a causal map of the human phenome
+author: Gibran Hemani, Jack Bowden, Philip Haycock, Jie Zheng, Tom Gaunt, George Davey Smith
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 
-A major application for genome-wide association studies (GWAS) has been the emerging field of causal inference using Mendelian randomisation (MR), where the causal effect between a pair of traits can be estimated using only summary level data.
+A major application for genome-wide association studies (GWAS) has been the emerging field of causal inference using Mendelian randomization (MR), where the causal effect between a pair of traits can be estimated using only summary level data.
 MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype.
@@
 We have developed a mixture of experts machine learning approach (MR-MoE 1.0) that improves on both power and false discovery rates over all existing methods.
-Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a draft of the causal map of the human phenome.
-But we offer caution that its interpretation is far from straightforward.
+Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a draft of the causal map of the human phenome, but we offer caution that its interpretation is far from straightforward.
 
@@
 
-Mendelian randomisation (MR) [@DaveySmith2003; @DaveySmithHemani2014] exploits genetic pleiotropy to infer the causal relationships between phenotypes.
+Mendelian randomization (MR) [@DaveySmith2003; @DaveySmithHemani2014] exploits genetic pleiotropy to infer the causal relationships between phenotypes.
 Suppose that one trait (the exposure) causally influences another (the outcome).
@@
 Genome-wide association studies (GWAS) have identified genetic instrumental variables for thousands of phenotypes [@Hindorff2010].
-Recent developments in Mendelian randomisation have enabled knowledge of instrumental variables to be applied using only summary level data (known as two-sample MR, 2SMR) [@Pierce2013].
+Recent developments in MR have enabled knowledge of instrumental variables to be applied using only summary level data (known as two-sample MR, 2SMR) [@Pierce2013].
 Here, in order to infer the causal effect of an exposure on an outcome all that is required is an estimate of the genetic effects of the instrumenting SNP on the exposure, and the corresponding estimate of the effect on the outcome.
@@
 Third, the available methods do not cover all possible models of HP, and therefore an automated method for instrument selection may be necessary.
-Fourth, it could be of interest to make causal effect estimates for thousands of traits, in which case a discerning evaluation of each causal effect of interest may not be possible or convenient.
+Fourth, it could be of interest to make causal effect estimates for thousands of traits, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient.
 
@@
 Using curated GWAS summary data for thousands of phenotypes [@Hemani2016], we use these new methods to construct a graph of millions of causal estimates.
-We consider this to be a 'first draft' of the causal map of the human phenome, but raise caution throughout that its interpretation is far from straightforward
+We consider this to be a 'draft' of the causal map of the human phenome, but raise caution throughout that its interpretation is far from straightforward
 
@@
 A brief outline of the procedure is as follows.
-First, genetic instruments for the exposure trait need to be identified - those SNPs with $p < 10^{-8}$ are retained in order to ensure that the first assumption of Mendelian randomisation (that the instrument associates with the exposure) is satisfied.
+First, genetic instruments for the exposure trait need to be identified - those SNPs with $p < 10^{-8}$ are retained in order to ensure that the first assumption of MR (that the instrument associates with the exposure) is satisfied.
 We collect their effect sizes, standard errors and effect alleles for the association with the exposure trait.
@@
 
-### Mendelian randomisation methods and their assumptions
+### MR methods and their assumptions
 
-In this paper we consider three main classes of MR estimation. Full details for each approach have been described extensively elsewhere.
+In this paper we consider three main classes of MR estimation. Full details for each approach have been described previously. A summary of the methods is given in Supplementary table 2.
 
@@
 $p$ can be estimated using the allele frequency of the SNP in an ascertained sample by deriving the $2 \times 2$ contingency table from the odds ratio $e^\beta$, allele frequency in the ascertained sample $p_{cc}, and number of cases $N_1$ and controls $N_0$.
+We exclude any SNPs that do not have the correct causal direction for the hypothesis, and do not have Steiger test $p < 0.05$.
 
@@
 
-We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total.
+We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total (Supplementary table 2).
 In the context of this analysis, each method is considered to be an 'expert', taking a set of SNP-exposure and SNP-outcome effect sizes and their standard errors as inputs.
@@
 This involves creating a 'gating function', whose purpose is to decide which expert to use for a specific MR analysis, given the parameter space that is occupied by that dataset.
-The metrics are a mixture of regression diagnostics and are described in Supplementary table 2.
+The metrics are a mixture of regression diagnostics and are described in Supplementary table 3.
 
@@
 For each simulation we used linear regression to estimate the genetic effect of each SNP $M$ on $x$ in sample 1, and each SNP $M$ on $y$ in sample 2.
-We then perform MR analysis in both directions, retaining SNPs that have $p < 5e-8$ in sample 1 to perform MR of $x$ on $y$ (the true causal direction for non-null simulations), and retaining SNPs that have $p < 5e-8$ in sample 2 to perform MR of $y$ on $x$ (the reverse causal direction for non-null simulations).
+We then perform MR analysis in both directions, mimicking GWAS by retaining SNPs that have $p < 5e-8$ in sample 1 to perform MR of $x$ on $y$ (the true causal direction for non-null simulations), and retaining SNPs that have $p < 5e-8$ in sample 2 to perform MR of $y$ on $x$ (the reverse causal direction for non-null simulations).
 We treat the summary data (effect sizes and standard errors) used for estimating $x \rightarrow y$ the summary data used for estimating $y \rightarrow x$ as two separate datasets Hence, for each simulation two datasets are generated which are analysed to produce 28 MR estimates each.
@@
 We aim to maximise statistical power for datasets where $\beta_{x,y} \neq 0$ and minimise false discovery rates for datasets where $\beta_{x,y} = 0$.
-To train random forest decision trees to predict performance for a particular method $h(O_{w,d}, \textbf{z}_{d}$ is generated where the training set of input metrics for dataset $d$ is $\textbf{z}_{d}$ and the response (optimisation function) is
+To train random forest decision trees to predict performance for a particular method $h(O_{w,d}, \textbf{z}_{d})$ is generated where the training set of input metrics for dataset $d$ is $\textbf{z}_{d}$ and the response (optimisation function) is
 
@@
 
-For each training dataset we record a set of 53 metrics $\textbf{z}_{d}$ (Supplementary table 2), and an outcome $O_{w,d}$, which is a measure of how well that method performed for each particular simulated dataset.
+For each training dataset we record a set of 53 metrics $\textbf{z}_{d}$ (Supplementary table 3), and an outcome $O_{w,d}$, which is a measure of how well that method performed for each particular simulated dataset.
 For each of our 28 methods, we need to create a model that predicts the performance of the method based on metrics generated from a dataset.
@@
 
-The ability to predict the performance for each of these methods is shown in Supplementary table 3, with prediction $R^2$ ranging between 0.04 and 0.24.
+The ability to predict the performance for each of these methods is shown in Supplementary table 2, with prediction $R^2$ ranging between 0.04 and 0.24.
 The dataset characteristics with the most importance for each of the predictors differed substantially between each dataset, as well as the frequency for which each of the methods was selected in the testing datasets.
@@
 
-There were 5660 associations following Bonferroni correction ($p < 7.0\times^{-8}$).
+There were 5660 associations following Bonferroni correction ($p < 7.0 \times 10^{-8}$).
 Of these 2918 were obtained from the MR-MoE analysis, while the remainder were estimated using fewer than 5 SNPs using the Wald ratio or IVW fixed effects methods.
 
-The frequencies of the methods chosen by the MR-MoE analysis are shown in Supplementary table 2.
+The frequencies of the methods chosen by the MR-MoE analysis are shown in Supplementary table 3.
 Amongst those deemed 'significant', the IVW fixed effects analysis using tophits instruments (the only method applicable when there is no evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases.
@@
 Crucially however, we observed in our simulations that none of the available methods were conservative and all, including MR-MoE, had FDR above 0.2.
-Instrumental variables in Mendelian randomisation are typically chosen blind, with GWAS significance being the only criteria.
+Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criteria.
 We illustrated how confounders and reverse causal associations can easily lead to invalid instruments being used in MR, and that even if those instruments that only influence the exposure directly are used, most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.
@@
 The potential to exploit GWAS summary data within the properties of graph databases to aid with both of these endeavours could be transformative in biological research.
-
-## Figures
 

### 2017-07-03 docs/manuscript.tex
updated

@@
 \hypersetup{unicode=true,
-            pdftitle={Towards the construction of a causal map of the human phenome},
-            pdfauthor={Gibran Hemani, Jie Zheng, Philip Haycock, Oliver Davies, Jack Bowden, Tom Gaunt, George Davey Smith},
+            pdftitle={Learning pleiotropic signatures to construct a causal map of the human phenome},
+            pdfauthor={Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George Davey Smith},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{Towards the construction of a causal map of the human phenome}
+  \title{Learning pleiotropic signatures to construct a causal map of the human
+phenome}
   \pretitle{\vspace{\droptitle}\centering\huge}
   \posttitle{\par}
-  \author{Gibran Hemani, Jie Zheng, Philip Haycock, Oliver Davies, Jack Bowden,
-Tom Gaunt, George Davey Smith}
+  \author{Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George
+Davey Smith}
   \preauthor{\centering\large\emph}
@@
 A major application for genome-wide association studies (GWAS) has been
-the emerging field of causal inference using Mendelian randomisation
+the emerging field of causal inference using Mendelian randomization
 (MR), where the causal effect between a pair of traits can be estimated
@@
 2407 phenotypes, generating a draft of the causal map of the human
-phenome. But we offer caution that its interpretation is far from
+phenome, but we offer caution that its interpretation is far from
 straightforward.
@@
 
-Mendelian randomisation (MR) (1,2) exploits genetic pleiotropy to infer
+Mendelian randomization (MR) (1,2) exploits genetic pleiotropy to infer
 the causal relationships between phenotypes. Suppose that one trait (the
@@
 instrumental variables for thousands of phenotypes (8). Recent
-developments in Mendelian randomisation have enabled knowledge of
-instrumental variables to be applied using only summary level data
-(known as two-sample MR, 2SMR) (9). Here, in order to infer the causal
-effect of an exposure on an outcome all that is required is an estimate
-of the genetic effects of the instrumenting SNP on the exposure, and the
-corresponding estimate of the effect on the outcome. This has two major
-advantages. First, GWAS summary data is non-disclosive and often
-publicly available. Second, causal inference can be made between
-phenotypes even if they have not been measured in the same samples,
-limiting the breadth of possible causal estimates only to the
-availability of GWAS summary data for the traits in question (10).
+developments in MR have enabled knowledge of instrumental variables to
+be applied using only summary level data (known as two-sample MR, 2SMR)
+(9). Here, in order to infer the causal effect of an exposure on an
+outcome all that is required is an estimate of the genetic effects of
+the instrumenting SNP on the exposure, and the corresponding estimate of
+the effect on the outcome. This has two major advantages. First, GWAS
+summary data is non-disclosive and often publicly available. Second,
+causal inference can be made between phenotypes even if they have not
+been measured in the same samples, limiting the breadth of possible
+causal estimates only to the availability of GWAS summary data for the
+traits in question (10).
 
@@
 may be necessary. Fourth, it could be of interest to make causal effect
-estimates for thousands of traits, in which case a discerning evaluation
-of each causal effect of interest may not be possible or convenient.
+estimates for thousands of traits, in which case a critical evaluation
+of each causal estimate of interest may not be possible or convenient.
 
@@
 new methods to construct a graph of millions of causal estimates. We
-consider this to be a `first draft' of the causal map of the human
-phenome, but raise caution throughout that its interpretation is far
-from straightforward
+consider this to be a `draft' of the causal map of the human phenome,
+but raise caution throughout that its interpretation is far from
+straightforward
 
@@
 SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
-assumption of Mendelian randomisation (that the instrument associates
-with the exposure) is satisfied. We collect their effect sizes, standard
-errors and effect alleles for the association with the exposure trait.
-These can be obtanied manually from complete summary data or from
-curated lists of significant GWAS associations. Next, the effects of
-those SNPs on the outcome need to be obtained, typically necessitating
-complete summary data because it is unlikely that these SNPs will have
-reached genome-wide significance (and therefore be present in curated
+assumption of MR (that the instrument associates with the exposure) is
+satisfied. We collect their effect sizes, standard errors and effect
+alleles for the association with the exposure trait. These can be
+obtanied manually from complete summary data or from curated lists of
+significant GWAS associations. Next, the effects of those SNPs on the
+outcome need to be obtained, typically necessitating complete summary
+data because it is unlikely that these SNPs will have reached
+genome-wide significance (and therefore be present in curated
 catalogues). At its most simple implementation, the regression of the
@@
 
-\subsubsection{Mendelian randomisation methods and their
-assumptions}\label{mendelian-randomisation-methods-and-their-assumptions}
+\subsubsection{MR methods and their
+assumptions}\label{mr-methods-and-their-assumptions}
 
 In this paper we consider three main classes of MR estimation. Full
-details for each approach have been described extensively elsewhere.
+details for each approach have been described previously. A summary of
+the methods is given in Supplementary table 2.
 
@@
 ratio \(e^\beta\), allele frequency in the ascertained sample
-\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\).
+\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\). We
+exclude any SNPs that do not have the correct causal direction for the
+hypothesis, and do not have Steiger test \(p < 0.05\).
 
@@
 be supplied using two instrument selection strategies, leading to 28
-methods in total. In the context of this analysis, each method is
-considered to be an `expert', taking a set of SNP-exposure and
-SNP-outcome effect sizes and their standard errors as inputs. Our
-objective is to select the expert most likely to be correct for a
-specific MR analysis.
+methods in total (Supplementary table 2). In the context of this
+analysis, each method is considered to be an `expert', taking a set of
+SNP-exposure and SNP-outcome effect sizes and their standard errors as
+inputs. Our objective is to select the expert most likely to be correct
+for a specific MR analysis.
 
@@
 The metrics are a mixture of regression diagnostics and are described in
-Supplementary table 2.
+Supplementary table 3.
 
@@
 \(y\) in sample 2. We then perform MR analysis in both directions,
-retaining SNPs that have \(p < 5e-8\) in sample 1 to perform MR of \(x\)
-on \(y\) (the true causal direction for non-null simulations), and
-retaining SNPs that have \(p < 5e-8\) in sample 2 to perform MR of \(y\)
-on \(x\) (the reverse causal direction for non-null simulations). We
-treat the summary data (effect sizes and standard errors) used for
-estimating \(x \rightarrow y\) the summary data used for estimating
-\(y \rightarrow x\) as two separate datasets Hence, for each simulation
-two datasets are generated which are analysed to produce 28 MR estimates
-each. We performed 100,000 simulations using these parameters, resulting
-in 200,000 datasets.
+mimicking GWAS by retaining SNPs that have \(p < 5e-8\) in sample 1 to
+perform MR of \(x\) on \(y\) (the true causal direction for non-null
+simulations), and retaining SNPs that have \(p < 5e-8\) in sample 2 to
+perform MR of \(y\) on \(x\) (the reverse causal direction for non-null
+simulations). We treat the summary data (effect sizes and standard
+errors) used for estimating \(x \rightarrow y\) the summary data used
+for estimating \(y \rightarrow x\) as two separate datasets Hence, for
+each simulation two datasets are generated which are analysed to produce
+28 MR estimates each. We performed 100,000 simulations using these
+parameters, resulting in 200,000 datasets.
 
@@
 predict performance for a particular method
-\(h(O_{w,d}, \textbf{z}_{d}\) is generated where the training set of
+\(h(O_{w,d}, \textbf{z}_{d})\) is generated where the training set of
 input metrics for dataset \(d\) is \(\textbf{z}_{d}\) and the response
@@
 For each training dataset we record a set of 53 metrics
-\(\textbf{z}_{d}\) (Supplementary table 2), and an outcome \(O_{w,d}\),
+\(\textbf{z}_{d}\) (Supplementary table 3), and an outcome \(O_{w,d}\),
 which is a measure of how well that method performed for each particular
@@
 
-Figure 1b shows that the tophits strategy led to 49\% of the instruments
-being primarily associated with either confounders or the outcome
-phenotype, not the exposure phenotype. Applying Steiger filtering
-reduced this to 25\%. Consequently, the false discovery rates (FDR) for
-12 of the 14 methods reduced substantially when applied using Steiger
-filtered instruments (Figure 1c). The true positive rates for the
-methods based on Steiger filtering did however reduce slightly for 10 of
-the 14 methods.
+Figure 1b shows that the tophits strategy led to over half of the
+instruments being primarily associated with either confounders or the
+outcome phenotype, not the exposure phenotype. The proportion of invalid
+instruments due to reverse cause increased as GWAS discovery sample size
+increased. Applying Steiger filtering reduced this to 25\%.
+Consequently, the false discovery rates (FDR) for 12 of the 14 methods
+reduced substantially when applied using Steiger filtered instruments
+(Figure 1c). The true positive rates for the methods based on Steiger
+filtering did however reduce slightly for 10 of the 14 methods.
 
@@
 existing methods, we next hypothesised that a mixture of experts (MoE)
-model would be able to use dataset characteristics to predict the most
-appropriate of the 28 MR methods and instrument selection strategies
-(Figure 2).
+model would be able to predict the most appropriate of the 28 MR methods
+and instrument selection strategies to apply to a particular dataset
+based on its characteristics (Figure 2).
 
 The ability to predict the performance for each of these methods is
-shown in table 1, with prediction \(R^2\) ranging between 0.04 and 0.24.
-The dataset characteristics with the most importance for each of the
-predictors differed substantially between each dataset, as well as the
-frequency for which each of the methods was selected in the testing
-datasets. The FDR of each method when chosen, compared to their averages
-across all datasets, reduced; and likewise the true positive rates of
-each method when chosen increased compared to their simulation-wide
-averages.
+shown in Supplementary table 2, with prediction \(R^2\) ranging between
+0.04 and 0.24. The dataset characteristics with the most importance for
+each of the predictors differed substantially between each dataset, as
+well as the frequency for which each of the methods was selected in the
+testing datasets. The FDR of each method when chosen, compared to their
+averages across all datasets, reduced; and likewise the true positive
+rates of each method when chosen increased compared to their
+simulation-wide averages.
 
@@
 The frequencies of the methods chosen by the MR-MoE analysis are shown
-in Supplementary table 2. Amongst those deemed `significant', the IVW
+in Supplementary table 3. Amongst those deemed `significant', the IVW
 fixed effects analysis using tophits instruments (the only method
@@
 
-Following a recent flurry in development of MR methodology, we have
-devised a machine learning approach, MR-MoE, that seeks to predict the
-performance of each MR method, selecting the one which is most likely to
-maximise power whilst minimising FDR for a specific dataset. The mixture
-of experts approach that we present here is trained using random forest
-decision trees applied to extensive and diverse simulations, and we
-demonstrate that it makes substantial performance improvements over any
-other single method.
+The trend of increasing GWAS sample sizes continues, but while the
+opportunity that this affords MR to be furnished with more instruments
+is typically welcomed, here we have shown that it is invalid
+instruments, exhibiting horizontal pleiotropy or reverse causation, that
+are more likely to be identified than valid ones. Strategising how MR is
+to be used in practice, therefore, must consider horizontal pleiotropy
+to be the rule rather than the exception. Following a recent flurry in
+development of MR methodology, we have devised a machine learning
+approach, MR-MoE, that seeks to predict the performance of each MR
+method, selecting the one which is most likely to maximise power whilst
+minimising FDR for a specific dataset. The mixture of experts approach
+that we present here is trained using random forest decision trees
+applied to extensive and diverse simulations, and we demonstrate that it
+makes substantial performance improvements over any other single method
+in the presence or absence of extensive horizontal pleiotropy.
 
-Crucially, we observed in our simulations that none of the available
-methods were conservative and all, including MR-MoE, had FDR above 0.2.
-Instrumental variables in Mendelian randomisation are typically chosen
-blind, with GWAS significance being the only criteria. We illustrated
-how confounders and reverse causal associations can easily lead to
-invalid instruments being used in MR, and that even if those instruments
-that only influence the exposure directly, most randomly generated
+Crucially however, we observed in our simulations that none of the
+available methods were conservative and all, including MR-MoE, had FDR
+above 0.2. Instrumental variables in MR are typically chosen blind, with
+GWAS significance being the only criteria. We illustrated how
+confounders and reverse causal associations can easily lead to invalid
+instruments being used in MR, and that even if those instruments that
+only influence the exposure directly are used, most randomly generated
 patterns of horizontal pleiotropy cannot be adequately accounted for by
@@
 
-We applied the method to a curated set of GWAS summary data that was
-present in MR-Base (22), generating hundreds of thousands of MR
-estimates. MR-MoE selected a method that indicated some pattern of
-horizontal pleiotropy for almost 90\% of cases, suggesting that it is
-the rule rather than the exception. Based on our simulations we expect
-that those associations that we reported to be `significant' are liable
-to high type 1 error rates.
+We applied the method to the curated set of GWAS summary data present in
+MR-Base (22), generating several million MR estimates. MR-MoE selected a
+method that indicated some pattern of horizontal pleiotropy for almost
+90\% of cases, reinforcing the notion that it is the rule rather than
+the exception. Based on our simulations we expect that those
+associations that we reported to be `significant' are liable to high
+type 1 error rates.
 
@@
 example frailty effects could induce associations for late-onset traits
-(25). We also encounter a potential new problem when it comes to using
-machine learning approaches to infer the correct model for a particular
-dataset, namely, that the data is choosing the model. This is liable to
-increase type 1 error rates, even though we have attempted to separate
-the information used for optimisation from the information used to
-predict performance. Though MR-MoE has high type 1 error rates, they
-still remain amongst the lowest compared to all other methods that do
-not suffer from this issue.
+(25); genetic variants could strongly relate to several phenotypes
+making it difficult to ascertain which of them is the causal exposure
+(26); and the measured feature for which genetic associations are known
+could relate in complicated ways to the biological entity that are truly
+causal (2).
+
+We also encounter a potential new problem in using machine learning
+approaches to infer the correct model for a particular dataset, namely,
+that the data is choosing the model. This is liable to increase type 1
+error rates, even though we have attempted to separate the information
+used for optimisation from the information used to predict performance.
+Though MR-MoE does exhibit higher type 1 error rates than are desirable,
+they still remain amongst the lowest compared to all other MR methods
+that do not suffer from this issue.
 
@@
 is not guaranteed by MR, it can still be highly informative for
-confirming or negating specific hypotheses, or to search for novel
-putative associations. The potential to exploit GWAS summary data within
-the properties of graph databases to aid with both of these endeavours
-could be transformative in biological research.
-
-\subsection{Figures}\label{figures}
+confirming or negating specific hypotheses. Second, it offers new
+approaches to searching for novel putative associations. The potential
+to exploit GWAS summary data within the properties of graph databases to
+aid with both of these endeavours could be transformative in biological
+research.
 
@@
 influence on a non-confounding intermediate variable, is considered a
-`direct' effect. The bar-chart shows the number of direct effect
-instruments retained for analysis using either the tophits approach or
-the Steiger approach for instrument selection. c) Top: The false
-discovery rates from null simulations for each of the 14 methods using
-either tophits, Steiger filtered variants, or variants that are known to
-be directly associated with the exposure (oracle, note that direct
-effects can still exhibit horizontal pleiotropy in these simulations).
-Bottom: The statistical power to detect true causal associations in the
-non-null simulations.}
+`direct' effect. The y-axis shows the proportion of instruments selected
+for analysis due that are either direct associations with the exposure,
+instruments for the outcome (reverse), or instruments for confounding
+traits (confounder). The proportions are compared over a range of
+different exposure discovery sample sizes (x-axis) and using either the
+tophits approach (left) or the Steiger approach (right) for instrument
+selection. c) Top: The false discovery rates from null simulations for
+each of the 14 methods using either tophits, Steiger filtered variants,
+or variants that are known to be directly associated with the exposure
+(oracle, note that direct effects can still exhibit horizontal
+pleiotropy in these simulations). Bottom: The statistical power to
+detect true causal associations in the non-null simulations.}
 \end{figure}
@@
 \end{figure}
+
+\newpage
 
@@
 
+\hypertarget{ref-Burgess2014a}{}
+26. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+multivariable Mendelian randomization to disentangle the causal effects
+of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
+Available from:
+\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}
+
 \end{document}

### 2017-07-29 docs/manuscript.rmd
updated

@@
 title: Learning pleiotropic signatures to construct a causal map of the human phenome
-author: Gibran Hemani, Jack Bowden, Philip Haycock, Jie Zheng, Tom Gaunt, George Davey Smith
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
   pdf_document:
-    fig_caption: yes
+    fig_caption: no
     keep_tex: true
@@
 
+Gibran Hemani$^1$, Jack Bowden$^1$, Philip Haycock$^1$, Jie Zheng$^1$, Oliver Davis$^1$, Peter Flach, Tom Gaunt$^{1,*}$, George Davey Smith$^{1,*}$
+
+1. MRC Integrative Epidemiology Unit (IEU) at the University of Bristol, School of Social and Community Medicine, Bristol, UK
+
+Correspondence to: g.hemani@bristol.ac.uk
+
+\* Equal contribution
+
 ## Abstract
@@
 Here we show that the mechanisms that underlie horizontal pleiotropy are numerous, and that instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow.
-We have developed a mixture of experts machine learning approach (MR-MoE 1.0) that improves on both power and false discovery rates over all existing methods.
-Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a draft of the causal map of the human phenome, but we offer caution that its interpretation is far from straightforward.
+We have developed a mixture of experts machine learning approach (MR-MoE 1.0) that improves on both power and false discovery rates over unselective use of any existing methods.
+Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a working draft of the causal map of the human phenome.
+Almost 90% of causal estimates indicated some level of horizontal pleiotropy.
+All results are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)).
 
@@
 If a SNP influences the outcome through the exposure then the SNP is exhibiting vertical pleiotropy.
-Such a genetic variant is considered to be an instrumental variable for the exposure, and can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not.
-Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) some violations of the assumptions of MR (most notably that the SNP influences the outcome through no pathways other than the exposure) can be evaluated through sensitivity analyses [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Hemani2017].
+Such a genetic variant is considered to be an instrumental variable for the if it only influences the outcome through the exposure (the exclusion restriction).
+It can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not.
+Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) potential bias arising from the exclusion restriction principal can be detected or corrected by evaluating the consistency of effects across instruments [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Hemani2017; @Verbanck2017].
 
@@
 Here, in order to infer the causal effect of an exposure on an outcome all that is required is an estimate of the genetic effects of the instrumenting SNP on the exposure, and the corresponding estimate of the effect on the outcome.
-This has two major advantages.
-First, GWAS summary data is non-disclosive and often publicly available.
-Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting the breadth of possible causal estimates only to the availability of GWAS summary data for the traits in question [@Hemani2016].
+This has three major advantages.
+First, GWAS summary data are non-disclosive and often publicly available.
+Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting possible causal estimates only by the availability of GWAS summary data for the traits in question [@Hemani2016].
+Third, we can view the causal inference problem through the simple and widely understood prism of a meta-analysis.
 
-Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure.
-The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017].
-It is considered best practice to report estimates from all available methods as sensitivity analyses when presenting causal estimates, however this strategy is not necessarily optimal for several reasons.
-First, if different methods disagree it is not possible to know which is correct because the true nature of HP exhibited by the instruments is not known.
+Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure [@DaveySmithHemani2014].
+The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017].
+It is considered best practice to report estimates from all available methods as sensitivity analyses when presenting causal estimates because different pleiotropic models can then be evaluated.
+However it is also desirable to consolidate across methods under some circumstances.
+First, it could be of interest to make causal effect estimates for thousands of traits, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient.
 Second, though the IVW approach is most statistically powerful under no HP, it can have high false negative or low true positive rates in the presence of HP compared to other methods.
-Given that pleiotropy has been hypothesised to be universal [@Wagner2011; @Hill2012a], defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate.
-Third, the available methods do not cover all possible models of HP, and therefore an automated method for instrument selection may be necessary.
-Fourth, it could be of interest to make causal effect estimates for thousands of traits, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient.
+Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate.
+Third, if different methods disagree it is not possible to know which is correct because the true nature of HP exhibited by the instruments is not known.
+Fourth, the available methods do not cover all possible models of HP, and therefore an automated method for instrument selection may be necessary.
 
@@
 Using curated GWAS summary data for thousands of phenotypes [@Hemani2016], we use these new methods to construct a graph of millions of causal estimates.
-We consider this to be a 'draft' of the causal map of the human phenome, but raise caution throughout that its interpretation is far from straightforward
+Motivated by the recent avalanche [@Hartwig2016] of 2SMR publications, often with repeating [@Xu2017; @Larsson2017] or contradicting [@Prins2016; @Inoshita2016] conclusions despite using the same data, we developed a graph database to represent these estimates in a consistent manner.
+We consider this to be a 'working draft' of the causal map of the human phenome, but raise caution throughout that its interpretation is far from straightforward, and that future corrections and refinements will inevitably follow as data grows and 2SMR methods evolve.
 
@@
 A brief outline of the procedure is as follows.
-First, genetic instruments for the exposure trait need to be identified - those SNPs with $p < 10^{-8}$ are retained in order to ensure that the first assumption of MR (that the instrument associates with the exposure) is satisfied.
+First, genetic instruments for the exposure trait need to be identified - those SNPs with $p < 10^{-8}$ are retained in order to ensure that the first assumption of MR (that the instrument associates with the exposure) is generally satisfied.
 We collect their effect sizes, standard errors and effect alleles for the association with the exposure trait.
-These can be obtanied manually from complete summary data or from curated lists of significant GWAS associations.
-Next, the effects of those SNPs on the outcome need to be obtained, typically necessitating complete summary data because it is unlikely that these SNPs will have reached genome-wide significance (and therefore be present in curated catalogues).
+These can be obtained manually from complete summary data or from curated lists of significant GWAS associations.
+Next, the effects of those SNPs on the outcome need to be obtained, typically necessitating complete summary data because these SNPs will generally have not have reached genome-wide significance for other traits (and therefore be present in curated catalogues).
 At its most simple implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome.
@@
 
-Given summary data for a large number of traits, it is straightforward to exhaustively analyse the causal relationships of every trait against every other trait for which there is sufficient summary data available.
+Given summary data for a large number of traits, it is straightforward to exhaustively analyse the causal relationships of every trait against every other trait for which there are sufficient summary data available.
 Supplementary table 1 provides a list of all traits that have available GWAS summary data, indicating if they have complete summary data (in which case they can be used as both exposure traits and outcome traits), or if they only have significant associations (in which they can only be used as exposure traits).
@@
 Fixed effects Egger regression [@Bowden2015] relaxes the HP assumption further by allowing a non-zero intercept which essentially allows horizontal pleiotropy to be directional, where it systematically influences the outcome in a specific direction.
-Random effects Egger regression allows heterogeneity around the directional HP, as long as the HP effects are not correlated with the SNP-exposure effects (also known as the INSIDE assumption).
+Random effects Egger regression allows heterogeneity around the directional HP [@Bowden2017], as long as the HP effects are not correlated with the SNP-exposure effects (also known as the INSIDE assumption)[@Bowden2015].
 
 The Rucker framework [@Rucker2011], adapted to MR [@Bowden2017] uses estimates of heterogeneity in the IVW and Egger frameworks to navigate between these nested models.
-A jackknife approach (random selection with replacement of instruments) can be used to obtain a sampling distribution for the model estimate amongst these four variations.
+A jackknife approach (random selection with replacement of the complete set of instruments) can be used to obtain a sampling distribution for the model estimate amongst these four variations.
 Using 1000 rounds of jackknife estimates, we can obtain a final estimate using the mean or the median of the distribution.
-We only use the jackknife approach for associations where there are 15 or more variants to avoid combinatorial saturation.
-The four nested models plus the three Rucker estimates provide seven mean based estimators.
+We only use the jackknife approach for associations where there are 15 or more variants in order to avoid saturating the possible number of instrument combinations.
 
-**Median-based methods:** An alternative approach is to take the median effect of all available instruments [@Bowden2016b].
+The four nested models (IVW fixed effects, IVW random effects, Egger fixed effects, Egger random effects) plus the three Rucker estimates (point estimate, mean of the jackknife, median of the jackknife) provide seven mean based estimators.
+
+**Median-based methods:** An alternative approach is to take the median effect of all available instruments [@Han2008; @Bowden2016b].
 This has the advantage that up to half the instruments can be invalid, and the estimate will remain unbiased.
@@
 Together, this provides three median based estimators.
+Other estimation strategies not considered here, such as LASSO regression, have also been developed for the situation where at least half of the instruments are valid [@Kang2016].
 
 **Mode-based methods:** Similar to the median, the mode based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017].
-This can be extended in two ways, first by weighting the strength of each cluster by the inverse of the variance of the SNP-outcome associations in the cluster; second by weighting the strength of each cluster by the inverse of the variance of the SNP-exposure associations.
-When weighting by the SNP-outcome standard errors we term this a weighted analysis.
-When *not* weighting by the exposure standard errors we term this the 'no measuremenent error in the exposure (NOME)' assumption.
-Together this provides four mode-based estimators.
+There are four implementations of this method: the simple and the weighted mode, each weighted with or without the assumption of no measurement error in the exposure estimates (NOME).
+The simple mode is the unweighted mode of the empirical density function of causal estimates, whereas the weighted mode is weighted by the inverse variance of the outcome effect.
+The bandwidth parameter was set to 1 by default.
 
@@
 
-The simplest approach to selecting instruments for performing MR is to take take SNPs that have been declared significant in the published GWAS for the exposure.
+The simplest approach to selecting instruments for performing MR is to take SNPs that have been declared significant in the published GWAS for the exposure.
 This typically involves obtaining SNPs that surpass $p < 5 \times 10^{-8}$, using clumping to obtain independent SNPs, and then replicating in an independent sample.
 These results are often recorded in public GWAS catalogs.
-Alternatively the clumping procedure can be performed using complete summary data in MR-Base.
+Alternatively the clumping procedure can be performed using complete summary data in MR-Base [@Hemani2016].
 We call this the "top hits" strategy.
@@
 
-An approach to inferring the causal direction between phenotypes was developed recently [@Hemani2017], using the following basic premise. If trait $A$ causes trait $B$ then
+An approach to inferring the causal direction between phenotypes recently developed [@Hemani2017] uses the following basic premise. If trait $A$ causes trait $B$ then
 
@@
 Some combinations of confounding effects can also distort the $\rho_{g,x}$ and $\rho_{g,y}$ parameters, as has been discussed in detail previously.
-However, we use it here as an inexpensive and approximate method to identify variants that are likely to be invalid.
+However, we use it here as a computationally inexpensive and approximate method to identify variants that are likely to be invalid.
 Steiger's Z-test of correlated correlations [@Steiger1980] can be used to formally test the extent to which the two correlations are statistically different.
@@
 Here we adapt this approach to automatically filter SNPs that are liable to be invalid (Figure 1a).
-In this case the Steiger test applied to each variant in turn and we exclude any $g_{A}$ for which the Steiger Z-test has $p > 0.05$, indicating that it is unlikely to primarily associate with $B$ relative to $A$.
-Similarly, for SNPs that influence confounders of $A$ and $B$ or those variants that exhibit horizontal pleiotropy, the difference in $cor(g_{i}, A)^2$ and $cor(g_{i}, B)^2$ will be reduced, increasing the likelihood of the SNP being excluded because the Steiger Z-test is less likely to be significant.
+Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are likely to be exhibiting horizontal pleiotropy.
+Conversely, we primarily developed Steiger filtering to identify those instruments that are likely to be arising due to reverse cause.
+
+The Steiger test is applied to each variant in turn and we exclude any $g_{A}$ for which $cor(g_{i}, A)^2 > cor(g_{i}, B)^2$, indicating that it is unlikely to primarily associate with $B$ relative to $A$.
+Similarly, for SNPs that influence confounders of $A$ and $B$ or exhibit horizontal pleiotropy, the difference in $cor(g_{i}, A)^2$ and $cor(g_{i}, B)^2$ will be reduced, increasing the likelihood of the SNP being excluded because the Steiger Z-test is less likely to be significant.
+Hence we also exclude any $g_{A}$ for which the Steiger Z-test has $p > 0.05$.
 
@@
 Our objective is to select the expert most likely to be correct for a specific MR analysis.
+A overview of the approach is shown in Figure 2.
 
@@
 New datasets can then be applied to the trained mixture of experts (MoE) model (Figure 2b).
-There are countless ways to implement this approach, we call this implementation MR-MoE 1.0.
+We call this implementation MR-MoE 1.0.
+
+It should be noted that in the original hierarchical mixture of experts approach of [@Jordan1994] the gating function and the experts share the same input space and are trained simultaneously using expectation-maximisation.
+In our approach the gating function is defined over a separate input space consisting of the parameter space of the experts, and is trained separately in a supervised setting using simulated data.
+As such our MR-MoE approach fits the framework of meta-learning [@brazdil2008metalearning; @smith2009cross; @lemke2015metalearning] where machine learning methods are applied at the 'meta-level' to analyse the results of machine learning experiments at the 'base-level', in order to be able to understand and model the capabilities of each expert and recommend which expert should be applied to a given problem.
+From this perspective the 28 MR methods are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the datasets.
 
@@
 This *dataset* can be fed into any of the 28 experts to obtain MR causal effects.
-The simulations used to generate these datasets seek to cover a range of pleiotropic scenarios, including where some proportion of SNPs exhibit directional or balanced horizontal pleiotropy, or where SNPs influencing confounding variables.
+The simulations used to generate these datasets seek to cover a range of pleiotropic scenarios, including where some proportion of SNPs exhibit directional or balanced horizontal pleiotropy, or where SNPs influence confounding variables.
 
@@
 
-Some proportion $s_x \sim U(0,1)$ of $g_x$ SNPs and some proportion $s_y \sim U(0,1)$ of $g_y$ SNPs exhibit horizontal pleiotropy with effects sampled using
+Some proportion $floor(MS_{x}/M)$ of $g_x$ SNPs and $floor(MS_{y}/M)$ of $g_y$ SNPs, where $s_x$ and $s_y \sim U(0,1)$, exhibit horizontal pleiotropy with effects sampled using
 
@@
 \begin{cases}
-    1,   & \text{if } \beta_{x,y} \neq 0 \text{and } p_{m,d} < 0.01\\
-    1,   & \text{if } \beta_{x,y} = 0 \text{and } p_{m,d} > 0.1 \\
+    1,   & \text{if } \beta_{x,y} \neq 0 \text{ and } p_{m,d} < 0.01\\
+    1,   & \text{if } \beta_{x,y} = 0 \text{ and } p_{m,d} > 0.1 \\
     0,   & \text{otherwise}
@@
 $$
+
+where $p_{m,d}$ is the p-value for method $m$ on dataset $d$.
 
@@
 For each of our 28 methods, we need to create a model that predicts the performance of the method based on metrics generated from a dataset.
-To do this, for each method we train random forest decision trees to predict that method's performance using the dataset's metrics.
+To do this, for each method we train a random decision forest to predict that method's performance using the dataset's metrics.
 The random forest approach is well suited to this problem because there are likely to be non-continuous combinations of different metrics that improve on prediction over, for example, a simple linear model that does not learn about interactions.
 
-A simple hypothetical example - if a dataset has a single outlier but is otherwise exhibiting no heterogeneity then the following methods could arguably perform well:
+As a simple hypothetical example - if a dataset has a single outlier but is otherwise exhibiting no heterogeneity then the following methods could arguably perform well:
 
@@
 deciding between these methods requires finding, in general, which will minimise false discovery rates and maximise true positive rates for that particular scenario.
+In this example the IVW with Steiger filtering would likely be the clear winner because of its superior statistical power.
+Countless other scenarios could arise.
+For example if 100% of instruments are invalid but the InSIDE assumptio is met then the MR-Egger is likely to be most effective; if 40% of instruments are invalid then a median-based approach is likely most effective; and if 80% of instruments are invalid then a mode-based approach is likely most effective.
 
@@
 Finally we compare the performance of the method selected by the MoE against all remaining methods.
-The default settings for the `randomForest` package in R [@Liaw2002] was used to train the models.
-MR-MoE 1.0 is implemented in the TwoSampleMR R package available at [github.com/MRCIEU/TwoSampleMR](https://github.com/MRCIEU/TwoSampleMR).
+The default settings for the `randomForest` package in R [@Liaw2002] were used to train the models.
+MR-MoE 1.0 is implemented in the TwoSampleMR R package available at [github.com/MRCIEU/TwoSampleMR](https://github.com/MRCIEU/TwoSampleMR) [@Hemani2016].
 
@@
 For specific hypotheses we strongly recommend that estimates from all sensitivity analyses are scrutinised and reported.
-The graph can be queried directly using the cypher language at [http://shark.epi.bris.ac.uk:7474/browser/](http://shark.epi.bris.ac.uk:7474/browser/) or through a basic web interface at [http://shark.epi.bris.ac.uk:3838](http://shark.epi.bris.ac.uk:3838).
+The graph can be queried directly using the cypher language at [http://eve-neo4j.mrbase.org](http://eve-neo4j.mrbase.org) or through a basic web interface at [http://eve.mrbase.org](http://eve.mrbase.org).
 
@@
 
-The ability to predict the performance for each of these methods is shown in Supplementary table 2, with prediction $R^2$ ranging between 0.04 and 0.24.
+The ability to predict the performance for each of these methods is shown in Supplementary table 2.
+The prediction $R^2$ of whether a dataset's status was truly null ($\beta_{xy}=0$) or non-null ($\beta_{xy} \neq 0$) against the method's prediction of the dataset's status, ranged between 0.04 and 0.24.
+The method performance prediction was most effective for the Egger random effects model with Steiger filtering.
 The dataset characteristics with the most importance for each of the predictors differed substantially between each dataset, as well as the frequency for which each of the methods was selected in the testing datasets.
@@
 Notably, the next best methods were median and mode based estimators using Steiger filtering.
-Under the assumption of universal pleiotropy it is clear that all methods suffer from high false discovery rates.
+Under the assumption of widespread pleiotropy it is clear that all methods suffer from high false discovery rates.
 
@@
 Amongst those deemed 'significant', the IVW fixed effects analysis using tophits instruments (the only method applicable when there is no evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases.
+This indicates that horizontal pleiotropy is likely to be pervasive.
 
-### Causal associations involving years of schooling
+### Putative associations involving years of schooling
 
-As an example of the database, we performed a look up of associations that causally influence years of schooling [@Okbay2016], or that years of schooling itself causally influences.
-Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4).
-Though many of these putative relationships appear to be plausible, it is clear that there are serious limitations to using MR as a panacea for causal inference.
+As an example of the database we performed a look up of associations where LDL cholesterol as measured in the GLGC consortium [@Willer2013] had a false discovery rate of 0.05 using the MoE approach.
+This returned 287 putative associations, amongst which 111 involved LDL cholesterol influencing other traits and 176 involved other traits influencing LDL cholesterol.
+A large proportion of these traits were metabolites, which are dominated by lipid fractions.
+Filtering to exclude the metabolomic studies ([@Shin2014; @Kettunen2016]) returned 27 associations with 23 traits (Figure 4a).
+There was a strong association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52, $p=2.7\times10^{-12}$).
+Here the IVW random effects method was chosen after Steiger filtering, indicating a prediction of HP being present amongst the instruments.
+
+We also performed a look up of associations that causally influence years of schooling [@Okbay2016], or that years of schooling itself causally influences.
+Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4b).
+Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease has recently been examined [@Anderson2017].
+However it is clear that there are serious limitations in relying on MR as a panacea for causal inference.
 For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g.
 childhood intelligence and birth weight) which appear to be causally downstream of years of schooling despite this being temporally impossible.
-The facility to perform these scans has been automated at the MR of everything vs everything (MR-EvE) website, [http://shark.epi.bris.ac.uk:3838](http://shark.epi.bris.ac.uk:3838), though we urge caution on interpretation of these findings.
+Though explanations could be conjured for this association, for example parents' schooling influences childhood intelligence, or the genetic instruments are shared across the two traits, an intuitive interpretation of causality cannot be applied.
+
+The facility to perform these scans has been automated in the MR of everything vs everything (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).
 
@@
 The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy.
+As such, our method contributes to the field of meta-learning but also to the rapidly growing field of automatic model selection and hyper-parameter optimization in machine learning [@kotthoff2016auto].
 
@@
 
-We applied the method to the curated set of GWAS summary data present in MR-Base [@hemani2016mr], generating several million MR estimates.
+We applied the method to the curated set of GWAS summary data present in MR-Base [@Hemani2016], generating several million MR estimates.
 MR-MoE selected a method that indicated some pattern of horizontal pleiotropy for almost 90% of cases, reinforcing the notion that it is the rule rather than the exception.
@@
 A large proportion of the associations that were estimated used only a single instrumental variable.
-Methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017], but separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework.
+Methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017].
+Separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework though the use of mediation-based approaches may be informative in the one-sample setting [@Wang2017; @Hemani2017].
 Other problems can also manifest, for example frailty effects could induce associations for late-onset traits [@Noyce2017]; genetic variants could strongly relate to several phenotypes making it difficult to ascertain which of them is the causal exposure [@Burgess2014a]; and the measured feature for which genetic associations are known could relate in complicated ways to the biological entity that are truly causal [@DaveySmithHemani2014].
@@
 We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular dataset, namely, that the data is choosing the model.
-This is liable to increase type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance.
+This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance.
+The extent to which a machine based method selector suffers from this more than a human manually inferring the result most likely to be true is not clear.
 Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to all other MR methods that do not suffer from this issue.
 
-Despite these limitations, the construction of a causal graph of 'everything versus everything' does have appeal.
+The field of MR is evolving.
+The advent of GWAS databases [@Hindorff2010; @Staley2016] and automated 2SMR [@Hemani2016] has trivialised the analytical aspect of investigating specific causal hypotheses.
+Despite the limitations described above, the construction of a causal graph of 'everything versus everything' does have appeal.
 First, though causality is not guaranteed by MR, it can still be highly informative for confirming or negating specific hypotheses.
-Second, it offers new approaches to searching for novel putative associations.
+Second, it paves the way for new approaches to search for novel putative associations and improve reliability in single analyses.
 The potential to exploit GWAS summary data within the properties of graph databases to aid with both of these endeavours could be transformative in biological research.
@@
 
-![Simulations.
+![](images/fig1.pdf)
+
+Figure 1: Simulations.
 a) Schematic of how GWAS with sufficient power can lead to the selection of instrumental variables that are invalid.
@@
 c) Top: The false discovery rates from null simulations for each of the 14 methods using either tophits, Steiger filtered variants, or variants that are known to be directly associated with the exposure (oracle, note that direct effects can still exhibit horizontal pleiotropy in these simulations).
-Bottom: The statistical power to detect true causal associations in the non-null simulations.](images/fig1.pdf)
+Bottom: The statistical power to detect true causal associations in the non-null simulations.
 
@@
 
-![Mixture of experts.
+![](images/fig2.pdf)
+
+Figure 2: Mixture of experts.
 a) Training.
@@
 Finally, we use the method for which the performance prediction is highest.
-](images/fig2.pdf)
 
@@
 
-![Performance of MoE against all other methods.
+![](images/fig3.pdf)
+
+Figure 3: Performance of MoE against all other methods.
 a) The power for non-null datasets is plotted against the FDR for null datasets for each of the 28 methods, plus MoE.
@@
 b) Calculating the area under the ROC curve from the values in (a) we plotted the performance in order from lowest to highest.
-Under the assumption of pervasive horizontal pleiotropy, the MoE approach is likely most effective than any other single method.](images/fig3.pdf)
+Under the assumption of pervasive horizontal pleiotropy, the MoE approach is likely most effective than any other single method.
 
@@
 
-![Lookup of causal associations involving Years of Schooling.
+![](images/fig4-01.png)
+
+Figure 4: Lookup of causal associations involving a) LDL cholesterol and b) Years of Schooling.
+In a) the nodes are filtered to not include those obtained from the Shin et al or Kettunen et al metabolomic studies.
 The arrows denote causal direction and the values on the arrows denote the causal effect estimate.
-Only those relationships are shown for which the $FDR < 0.05$.](images/fig4.png)
+Only those relationships are shown for which the $FDR < 0.05$.
 
@@
 ## References
+
+\raggedright

### 2017-07-29 docs/manuscript.tex
updated

@@
             pdftitle={Learning pleiotropic signatures to construct a causal map of the human phenome},
-            pdfauthor={Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George Davey Smith},
             pdfborder={0 0 0},
@@
 \fi
+\usepackage{amsmath}
 
@@
   \posttitle{\par}
-  \author{Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George
-Davey Smith}
-  \preauthor{\centering\large\emph}
-  \postauthor{\par}
+  \author{}
+  \preauthor{}\postauthor{}
   \predate{\centering\large\emph}
   \postdate{\par}
-  \date{03 July 2017}
-
-\usepackage{amsmath}
+  \date{29 July 2017}
 
@@
 \maketitle
+
+Gibran Hemani\(^1\), Jack Bowden\(^1\), Philip Haycock\(^1\), Jie
+Zheng\(^1\), Oliver Davis\(^1\), Peter Flach, Tom Gaunt\(^{1,*}\),
+George Davey Smith\(^{1,*}\)
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  MRC Integrative Epidemiology Unit (IEU) at the University of Bristol,
+  School of Social and Community Medicine, Bristol, UK
+\end{enumerate}
+
+Correspondence to:
+\href{mailto:g.hemani@bristol.ac.uk}{\nolinkurl{g.hemani@bristol.ac.uk}}
+
+* Equal contribution
 
@@
 mixture of experts machine learning approach (MR-MoE 1.0) that improves
-on both power and false discovery rates over all existing methods. Using
-the approach, we systematically estimated the causal effects amongst
-2407 phenotypes, generating a draft of the causal map of the human
-phenome, but we offer caution that its interpretation is far from
-straightforward.
+on both power and false discovery rates over unselective use of any
+existing methods. Using the approach, we systematically estimated the
+causal effects amongst 2407 phenotypes, generating a working draft of
+the causal map of the human phenome. Almost 90\% of causal estimates
+indicated some level of horizontal pleiotropy. All results are organised
+into a publicly available graph database (\url{http://eve.mrbase.org}).
 
@@
 pleiotropy. Such a genetic variant is considered to be an instrumental
-variable for the exposure, and can be exploited to mimic a randomised
+variable for the if it only influences the outcome through the exposure
+(the exclusion restriction). It can be exploited to mimic a randomised
 controlled trial, enabling a causal estimate to be made by comparing the
@@
 obtained by meta-analysing the single estimates from each instrument;
-and b) some violations of the assumptions of MR (most notably that the
-SNP influences the outcome through no pathways other than the exposure)
-can be evaluated through sensitivity analyses (3--7).
+and b) potential bias arising from the exclusion restriction principal
+can be detected or corrected by evaluating the consistency of effects
+across instruments (3--8).
 
 Genome-wide association studies (GWAS) have identified genetic
-instrumental variables for thousands of phenotypes (8). Recent
+instrumental variables for thousands of phenotypes (9). Recent
 developments in MR have enabled knowledge of instrumental variables to
 be applied using only summary level data (known as two-sample MR, 2SMR)
-(9). Here, in order to infer the causal effect of an exposure on an
+(10). Here, in order to infer the causal effect of an exposure on an
 outcome all that is required is an estimate of the genetic effects of
 the instrumenting SNP on the exposure, and the corresponding estimate of
-the effect on the outcome. This has two major advantages. First, GWAS
-summary data is non-disclosive and often publicly available. Second,
+the effect on the outcome. This has three major advantages. First, GWAS
+summary data are non-disclosive and often publicly available. Second,
 causal inference can be made between phenotypes even if they have not
-been measured in the same samples, limiting the breadth of possible
-causal estimates only to the availability of GWAS summary data for the
-traits in question (10).
+been measured in the same samples, limiting possible causal estimates
+only by the availability of GWAS summary data for the traits in question
+(11). Third, we can view the causal inference problem through the simple
+and widely understood prism of a meta-analysis.
 
@@
 the genetic instruments exhibit horizontal pleiotropy (HP), where they
-influence the outcome through a pathway other than the exposure. The
+influence the outcome through a pathway other than the exposure (2). The
 extent of this phenomenon is not to be understated, and many methods
 have been developed that attempt to reliably obtain unbiased causal
-estimates under specific models of HP (3--6). It is considered best
+estimates under specific models of HP (3--6,8,12). It is considered best
 practice to report estimates from all available methods as sensitivity
-analyses when presenting causal estimates, however this strategy is not
-necessarily optimal for several reasons. First, if different methods
+analyses when presenting causal estimates because different pleiotropic
+models can then be evaluated. However it is also desirable to
+consolidate across methods under some circumstances. First, it could be
+of interest to make causal effect estimates for thousands of traits, in
+which case a critical evaluation of each causal estimate of interest may
+not be possible or convenient. Second, though the IVW approach is most
+statistically powerful under no HP, it can have high false negative or
+low true positive rates in the presence of HP compared to other methods.
+Pleiotropy has been hypothesised to be universal (13), though the degree
+to which this may be the case is contested (14,15), hence defaulting to
+the IVW method in the first instance and using other methods as
+sensitivity analyses may not be appropriate. Third, if different methods
 disagree it is not possible to know which is correct because the true
-nature of HP exhibited by the instruments is not known. Second, though
-the IVW approach is most statistically powerful under no HP, it can have
-high false negative or low true positive rates in the presence of HP
-compared to other methods. Given that pleiotropy has been hypothesised
-to be universal (11,12), defaulting to the IVW method in the first
-instance and using other methods as sensitivity analyses may not be
-appropriate. Third, the available methods do not cover all possible
-models of HP, and therefore an automated method for instrument selection
-may be necessary. Fourth, it could be of interest to make causal effect
-estimates for thousands of traits, in which case a critical evaluation
-of each causal estimate of interest may not be possible or convenient.
+nature of HP exhibited by the instruments is not known. Fourth, the
+available methods do not cover all possible models of HP, and therefore
+an automated method for instrument selection may be necessary.
 
@@
 that attempt to automate both instrument and method selection. Using
-curated GWAS summary data for thousands of phenotypes (10), we use these
-new methods to construct a graph of millions of causal estimates. We
-consider this to be a `draft' of the causal map of the human phenome,
-but raise caution throughout that its interpretation is far from
-straightforward
+curated GWAS summary data for thousands of phenotypes (11), we use these
+new methods to construct a graph of millions of causal estimates.
+Motivated by the recent avalanche (16) of 2SMR publications, often with
+repeating (17,18) or contradicting (19,20) conclusions despite using the
+same data, we developed a graph database to represent these estimates in
+a consistent manner. We consider this to be a `working draft' of the
+causal map of the human phenome, but raise caution throughout that its
+interpretation is far from straightforward, and that future corrections
+and refinements will inevitably follow as data grows and 2SMR methods
+evolve.
 
@@
 The use of summary data in two-sample MR is described in detail
-elsewhere (10). A brief outline of the procedure is as follows. First,
+elsewhere (11). A brief outline of the procedure is as follows. First,
 genetic instruments for the exposure trait need to be identified - those
@@
 assumption of MR (that the instrument associates with the exposure) is
-satisfied. We collect their effect sizes, standard errors and effect
-alleles for the association with the exposure trait. These can be
-obtanied manually from complete summary data or from curated lists of
+generally satisfied. We collect their effect sizes, standard errors and
+effect alleles for the association with the exposure trait. These can be
+obtained manually from complete summary data or from curated lists of
 significant GWAS associations. Next, the effects of those SNPs on the
 outcome need to be obtained, typically necessitating complete summary
-data because it is unlikely that these SNPs will have reached
-genome-wide significance (and therefore be present in curated
+data because these SNPs will generally have not have reached genome-wide
+significance for other traits (and therefore be present in curated
 catalogues). At its most simple implementation, the regression of the
@@
 to exhaustively analyse the causal relationships of every trait against
-every other trait for which there is sufficient summary data available.
+every other trait for which there are sufficient summary data available.
 Supplementary table 1 provides a list of all traits that have available
@@
 influences the outcome in a specific direction. Random effects Egger
-regression allows heterogeneity around the directional HP, as long as
-the HP effects are not correlated with the SNP-exposure effects (also
-known as the INSIDE assumption).
-
-The Rucker framework (13), adapted to MR (5) uses estimates of
+regression allows heterogeneity around the directional HP (5), as long
+as the HP effects are not correlated with the SNP-exposure effects (also
+known as the INSIDE assumption)(3).
+
+The Rucker framework (21), adapted to MR (5) uses estimates of
 heterogeneity in the IVW and Egger frameworks to navigate between these
 nested models. A jackknife approach (random selection with replacement
-of instruments) can be used to obtain a sampling distribution for the
-model estimate amongst these four variations. Using 1000 rounds of
-jackknife estimates, we can obtain a final estimate using the mean or
-the median of the distribution. We only use the jackknife approach for
-associations where there are 15 or more variants to avoid combinatorial
-saturation. The four nested models plus the three Rucker estimates
-provide seven mean based estimators.
+of the complete set of instruments) can be used to obtain a sampling
+distribution for the model estimate amongst these four variations. Using
+1000 rounds of jackknife estimates, we can obtain a final estimate using
+the mean or the median of the distribution. We only use the jackknife
+approach for associations where there are 15 or more variants in order
+to avoid saturating the possible number of instrument combinations.
+
+The four nested models (IVW fixed effects, IVW random effects, Egger
+fixed effects, Egger random effects) plus the three Rucker estimates
+(point estimate, mean of the jackknife, median of the jackknife) provide
+seven mean based estimators.
 
 \textbf{Median-based methods:} An alternative approach is to take the
-median effect of all available instruments (4). This has the advantage
-that up to half the instruments can be invalid, and the estimate will
-remain unbiased. Developing the approach further to allow stronger
-instruments to contribute more towards the estimate can be obtained by
-obtaining the median of the weights of each instrument. The penalised
-weighted median estimator introduces a further weight to the
+median effect of all available instruments (4,22). This has the
+advantage that up to half the instruments can be invalid, and the
+estimate will remain unbiased. Developing the approach further to allow
+stronger instruments to contribute more towards the estimate can be
+obtained by obtaining the median of the weights of each instrument. The
+penalised weighted median estimator introduces a further weight to the
 instruments, penalising any instrument that contributes substantially
 towards the heterogeneity statistic. Together, this provides three
-median based estimators.
+median based estimators. Other estimation strategies not considered
+here, such as LASSO regression, have also been developed for the
+situation where at least half of the instruments are valid (23).
 
@@
 causal effects, and returns the final causal effect estimate based on
-the cluster that has the largest number of instruments (6). This can be
-extended in two ways, first by weighting the strength of each cluster by
-the inverse of the variance of the SNP-outcome associations in the
-cluster; second by weighting the strength of each cluster by the inverse
-of the variance of the SNP-exposure associations. When weighting by the
-SNP-outcome standard errors we term this a weighted analysis. When
-\emph{not} weighting by the exposure standard errors we term this the
-`no measuremenent error in the exposure (NOME)' assumption. Together
-this provides four mode-based estimators.
+the cluster that has the largest number of instruments (6). There are
+four implementations of this method: the simple and the weighted mode,
+each weighted with or without the assumption of no measurement error in
+the exposure estimates (NOME). The simple mode is the unweighted mode of
+the empirical density function of causal estimates, whereas the weighted
+mode is weighted by the inverse variance of the outcome effect. The
+bandwidth parameter was set to 1 by default.
 
@@
 The simplest approach to selecting instruments for performing MR is to
-take take SNPs that have been declared significant in the published GWAS
-for the exposure. This typically involves obtaining SNPs that surpass
+take SNPs that have been declared significant in the published GWAS for
+the exposure. This typically involves obtaining SNPs that surpass
 \(p < 5 \times 10^{-8}\), using clumping to obtain independent SNPs, and
@@
 recorded in public GWAS catalogs. Alternatively the clumping procedure
-can be performed using complete summary data in MR-Base. We call this
-the ``top hits'' strategy.
+can be performed using complete summary data in MR-Base (11). We call
+this the ``top hits'' strategy.
 
@@
 
-An approach to inferring the causal direction between phenotypes was
-developed recently (7), using the following basic premise. If trait
-\(A\) causes trait \(B\) then
+An approach to inferring the causal direction between phenotypes
+recently developed (7) uses the following basic premise. If trait \(A\)
+causes trait \(B\) then
 
@@
 \(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
-detail previously. However, we use it here as an inexpensive and
-approximate method to identify variants that are likely to be invalid.
-Steiger's Z-test of correlated correlations (14) can be used to formally
-test the extent to which the two correlations are statistically
-different.
+detail previously. However, we use it here as a computationally
+inexpensive and approximate method to identify variants that are likely
+to be invalid. Steiger's Z-test of correlated correlations (24) can be
+used to formally test the extent to which the two correlations are
+statistically different.
 
 Here we adapt this approach to automatically filter SNPs that are liable
-to be invalid (Figure 1a). In this case the Steiger test applied to each
-variant in turn and we exclude any \(g_{A}\) for which the Steiger
-Z-test has \(p > 0.05\), indicating that it is unlikely to primarily
-associate with \(B\) relative to \(A\). Similarly, for SNPs that
-influence confounders of \(A\) and \(B\) or those variants that exhibit
-horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
+to be invalid (Figure 1a). Other methods have been developed to identify
+invalid instruments for the purposes of exclusion from MR analyses
+(8,12,25), based on the notion that outlying instruments are likely to
+be exhibiting horizontal pleiotropy. Conversely, we primarily developed
+Steiger filtering to identify those instruments that are likely to be
+arising due to reverse cause.
+
+The Steiger test is applied to each variant in turn and we exclude any
+\(g_{A}\) for which \(cor(g_{i}, A)^2 > cor(g_{i}, B)^2\), indicating
+that it is unlikely to primarily associate with \(B\) relative to \(A\).
+Similarly, for SNPs that influence confounders of \(A\) and \(B\) or
+exhibit horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
 \(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
 SNP being excluded because the Steiger Z-test is less likely to be
-significant.
+significant. Hence we also exclude any \(g_{A}\) for which the Steiger
+Z-test has \(p > 0.05\).
 
@@
 \(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the log odds ratio and
-\(p\) is the allele frequency of the SNP in the population (15). \(p\)
+\(p\) is the allele frequency of the SNP in the population (26). \(p\)
 can be estimated using the allele frequency of the SNP in an ascertained
@@
 inputs. Our objective is to select the expert most likely to be correct
-for a specific MR analysis.
-
-The mixture of experts method is a machine learning approach (16) which
+for a specific MR analysis. A overview of the approach is shown in
+Figure 2.
+
+The mixture of experts method is a machine learning approach (27) which
 seeks to divide a parameter space into subdomains, such that a
@@
 simulated data. New datasets can then be applied to the trained mixture
-of experts (MoE) model (Figure 2b). There are countless ways to
-implement this approach, we call this implementation MR-MoE 1.0.
+of experts (MoE) model (Figure 2b). We call this implementation MR-MoE
+1.0.
+
+It should be noted that in the original hierarchical mixture of experts
+approach of (27) the gating function and the experts share the same
+input space and are trained simultaneously using
+expectation-maximisation. In our approach the gating function is defined
+over a separate input space consisting of the parameter space of the
+experts, and is trained separately in a supervised setting using
+simulated data. As such our MR-MoE approach fits the framework of
+meta-learning (28--30) where machine learning methods are applied at the
+`meta-level' to analyse the results of machine learning experiments at
+the `base-level', in order to be able to understand and model the
+capabilities of each expert and recommend which expert should be applied
+to a given problem. From this perspective the 28 MR methods are the
+base-models, the gating function is the meta-model, and its
+meta-features are the parameters of the datasets.
 
@@
 scenarios, including where some proportion of SNPs exhibit directional
-or balanced horizontal pleiotropy, or where SNPs influencing confounding
+or balanced horizontal pleiotropy, or where SNPs influence confounding
 variables.
@@
 
-Some proportion \(s_x \sim U(0,1)\) of \(g_x\) SNPs and some proportion
-\(s_y \sim U(0,1)\) of \(g_y\) SNPs exhibit horizontal pleiotropy with
-effects sampled using
+Some proportion \(floor(MS_{x}/M)\) of \(g_x\) SNPs and
+\(floor(MS_{y}/M)\) of \(g_y\) SNPs, where \(s_x\) and
+\(s_y \sim U(0,1)\), exhibit horizontal pleiotropy with effects sampled
+using
 
@@
 \]
+
+where \(p_{m,d}\) is the p-value for method \(m\) on dataset \(d\).
 
@@
 that predicts the performance of the method based on metrics generated
-from a dataset. To do this, for each method we train random forest
-decision trees to predict that method's performance using the dataset's
-metrics. The random forest approach is well suited to this problem
-because there are likely to be non-continuous combinations of different
-metrics that improve on prediction over, for example, a simple linear
-model that does not learn about interactions.
-
-A simple hypothetical example - if a dataset has a single outlier but is
-otherwise exhibiting no heterogeneity then the following methods could
-arguably perform well:
+from a dataset. To do this, for each method we train a random decision
+forest to predict that method's performance using the dataset's metrics.
+The random forest approach is well suited to this problem because there
+are likely to be non-continuous combinations of different metrics that
+improve on prediction over, for example, a simple linear model that does
+not learn about interactions.
+
+As a simple hypothetical example - if a dataset has a single outlier but
+is otherwise exhibiting no heterogeneity then the following methods
+could arguably perform well:
 
@@
 minimise false discovery rates and maximise true positive rates for that
-particular scenario.
+particular scenario. In this example the IVW with Steiger filtering
+would likely be the clear winner because of its superior statistical
+power. Countless other scenarios could arise. For example if 100\% of
+instruments are invalid but the InSIDE assumptio is met then the
+MR-Egger is likely to be most effective; if 40\% of instruments are
+invalid then a median-based approach is likely most effective; and if
+80\% of instruments are invalid then a mode-based approach is likely
+most effective.
 
@@
 methods. The default settings for the \texttt{randomForest} package in R
-(17) was used to train the models. MR-MoE 1.0 is implemented in the
+(31) were used to train the models. MR-MoE 1.0 is implemented in the
 TwoSampleMR R package available at
-\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}.
+\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
+(11).
 
@@
 queried directly using the cypher language at
-\url{http://shark.epi.bris.ac.uk:7474/browser/} or through a basic web
-interface at \url{http://shark.epi.bris.ac.uk:3838}.
+\url{http://eve-neo4j.mrbase.org} or through a basic web interface at
+\url{http://eve.mrbase.org}.
 
@@
 The ability to predict the performance for each of these methods is
-shown in Supplementary table 2, with prediction \(R^2\) ranging between
-0.04 and 0.24. The dataset characteristics with the most importance for
-each of the predictors differed substantially between each dataset, as
-well as the frequency for which each of the methods was selected in the
+shown in Supplementary table 2. The prediction \(R^2\) of whether a
+dataset's status was truly null (\(\beta_{xy}=0\)) or non-null
+(\(\beta_{xy} \neq 0\)) against the method's prediction of the dataset's
+status, ranged between 0.04 and 0.24. The method performance prediction
+was most effective for the Egger random effects model with Steiger
+filtering. The dataset characteristics with the most importance for each
+of the predictors differed substantially between each dataset, as well
+as the frequency for which each of the methods was selected in the
 testing datasets. The FDR of each method when chosen, compared to their
@@
 were median and mode based estimators using Steiger filtering. Under the
-assumption of universal pleiotropy it is clear that all methods suffer
+assumption of widespread pleiotropy it is clear that all methods suffer
 from high false discovery rates.
@@
 We applied our analysis using summary data for 2407 phenotypes,
-including 149 complex traits and diseases, 575 metabolites (18,19) and
-1683 plasma protein levels (20). For the protein levels only the
+including 149 complex traits and diseases, 575 metabolites (32,33) and
+1683 plasma protein levels (34). For the protein levels only the
 instruments were available, so they could only be evaluated as exposure
@@
 There were 5660 associations following Bonferroni correction
-(\(p < 7.0\times^{-8}\)). Of these 2918 were obtained from the MR-MoE
-analysis, while the remainder were estimated using fewer than 5 SNPs
-using the Wald ratio or IVW fixed effects methods.
+(\(p < 7.0 \times 10^{-8}\)). Of these 2918 were obtained from the
+MR-MoE analysis, while the remainder were estimated using fewer than 5
+SNPs using the Wald ratio or IVW fixed effects methods.
 
@@
 applicable when there is no evidence of horizontal pleiotropy of any
-sort) was selected in only 10.4\% of cases.
-
-\subsubsection{Causal associations involving years of
-schooling}\label{causal-associations-involving-years-of-schooling}
-
-As an example of the database, we performed a look up of associations
-that causally influence years of schooling (21), or that years of
-schooling itself causally influences. Using a false-discovery rate of
-0.05, 45 traits were returned as having some direction of causality with
-years of schooling (Figure 4). Though many of these putative
-relationships appear to be plausible, it is clear that there are serious
-limitations to using MR as a panacea for causal inference. For example,
-bi-directional causal relationships with college completion exist
-because of the similarity in trait definitions; and there are several
-traits (e.g.~childhood intelligence and birth weight) which appear to be
-causally downstream of years of schooling despite this being temporally
-impossible. The facility to perform these scans has been automated at
-the MR of everything vs everything (MR-EvE) website,
-\url{http://shark.epi.bris.ac.uk:3838}, though we urge caution on
-interpretation of these findings.
+sort) was selected in only 10.4\% of cases. This indicates that
+horizontal pleiotropy is likely to be pervasive.
+
+\subsubsection{Putative associations involving years of
+schooling}\label{putative-associations-involving-years-of-schooling}
+
+As an example of the database we performed a look up of associations
+where LDL cholesterol as measured in the GLGC consortium (35) had a
+false discovery rate of 0.05 using the MoE approach. This returned 287
+putative associations, amongst which 111 involved LDL cholesterol
+influencing other traits and 176 involved other traits influencing LDL
+cholesterol. A large proportion of these traits were metabolites, which
+are dominated by lipid fractions. Filtering to exclude the metabolomic
+studies ((32,33)) returned 27 associations with 23 traits (Figure 4a).
+There was a strong association with coronary heart disease (0.41 log(OR)
+per SD, 0.31-0.52, \(p=2.7\times10^{-12}\)). Here the IVW random effects
+method was chosen after Steiger filtering, indicating a prediction of HP
+being present amongst the instruments.
+
+We also performed a look up of associations that causally influence
+years of schooling (36), or that years of schooling itself causally
+influences. Using a false-discovery rate of 0.05, 45 traits were
+returned as having some direction of causality with years of schooling
+(Figure 4b). Several of these putative relationships appear to be
+plausible, for example the influence of years of schooling on lower risk
+of Alzheimer's disease has recently been examined (37). However it is
+clear that there are serious limitations in relying on MR as a panacea
+for causal inference. For example, bi-directional causal relationships
+with college completion exist because of the similarity in trait
+definitions; and there are several traits (e.g.~childhood intelligence
+and birth weight) which appear to be causally downstream of years of
+schooling despite this being temporally impossible. Though explanations
+could be conjured for this association, for example parents' schooling
+influences childhood intelligence, or the genetic instruments are shared
+across the two traits, an intuitive interpretation of causality cannot
+be applied.
+
+The facility to perform these scans has been automated in the MR of
+everything vs everything (MR-EvE) web application,
+\url{http://eve.mrbase.org}.
 
@@
 makes substantial performance improvements over any other single method
-in the presence or absence of extensive horizontal pleiotropy.
+in the presence or absence of extensive horizontal pleiotropy. As such,
+our method contributes to the field of meta-learning but also to the
+rapidly growing field of automatic model selection and hyper-parameter
+optimization in machine learning (38).
 
@@
 We applied the method to the curated set of GWAS summary data present in
-MR-Base (22), generating several million MR estimates. MR-MoE selected a
+MR-Base (11), generating several million MR estimates. MR-MoE selected a
 method that indicated some pattern of horizontal pleiotropy for almost
@@
 delineate between models of reverse cause, pleiotropy and multiple
-causal variants in the same region (23,24), but separating vertical from
+causal variants in the same region (39,40). Separating vertical from
 horizontal pleiotropy with a single instrument is not yet possible in
-the two-sample MR framework. Other problems can also manifest, for
-example frailty effects could induce associations for late-onset traits
-(25); genetic variants could strongly relate to several phenotypes
-making it difficult to ascertain which of them is the causal exposure
-(26); and the measured feature for which genetic associations are known
-could relate in complicated ways to the biological entity that are truly
-causal (2).
+the two-sample MR framework though the use of mediation-based approaches
+may be informative in the one-sample setting (7,41). Other problems can
+also manifest, for example frailty effects could induce associations for
+late-onset traits (42); genetic variants could strongly relate to
+several phenotypes making it difficult to ascertain which of them is the
+causal exposure (43); and the measured feature for which genetic
+associations are known could relate in complicated ways to the
+biological entity that are truly causal (2).
 
@@
 approaches to infer the correct model for a particular dataset, namely,
-that the data is choosing the model. This is liable to increase type 1
-error rates, even though we have attempted to separate the information
-used for optimisation from the information used to predict performance.
-Though MR-MoE does exhibit higher type 1 error rates than are desirable,
-they still remain amongst the lowest compared to all other MR methods
-that do not suffer from this issue.
-
-Despite these limitations, the construction of a causal graph of
-`everything versus everything' does have appeal. First, though causality
-is not guaranteed by MR, it can still be highly informative for
-confirming or negating specific hypotheses. Second, it offers new
-approaches to searching for novel putative associations. The potential
-to exploit GWAS summary data within the properties of graph databases to
-aid with both of these endeavours could be transformative in biological
-research.
+that the data is choosing the model. This is liable to contribute to
+elevated type 1 error rates, even though we have attempted to separate
+the information used for optimisation from the information used to
+predict performance. The extent to which a machine based method selector
+suffers from this more than a human manually inferring the result most
+likely to be true is not clear. Though MR-MoE does exhibit higher type 1
+error rates than are desirable, they still remain amongst the lowest
+compared to all other MR methods that do not suffer from this issue.
+
+The field of MR is evolving. The advent of GWAS databases (9,44) and
+automated 2SMR (11) has trivialised the analytical aspect of
+investigating specific causal hypotheses. Despite the limitations
+described above, the construction of a causal graph of `everything
+versus everything' does have appeal. First, though causality is not
+guaranteed by MR, it can still be highly informative for confirming or
+negating specific hypotheses. Second, it paves the way for new
+approaches to search for novel putative associations and improve
+reliability in single analyses. The potential to exploit GWAS summary
+data within the properties of graph databases to aid with both of these
+endeavours could be transformative in biological research.
 
@@
 
-\begin{figure}
-\centering
 \includegraphics{images/fig1.pdf}
-\caption{Simulations. a) Schematic of how GWAS with sufficient power can
-lead to the selection of instrumental variables that are invalid. We
+
+Figure 1: Simulations. a) Schematic of how GWAS with sufficient power
+can lead to the selection of instrumental variables that are invalid. We
 used arbitrary numbers of SNPs and confounders to simulate GWAS summary
@@
 pleiotropy in these simulations). Bottom: The statistical power to
-detect true causal associations in the non-null simulations.}
-\end{figure}
+detect true causal associations in the non-null simulations.
 
@@
 
-\begin{figure}
-\centering
 \includegraphics{images/fig2.pdf}
-\caption{Mixture of experts. a) Training. Datasets are simulated that
+
+Figure 2: Mixture of experts. a) Training. Datasets are simulated that
 have either null or non-null causal relationships, and MR is performed
@@
 performance predictions. Finally, we use the method for which the
-performance prediction is highest.}
-\end{figure}
+performance prediction is highest.
 
@@
 
-\begin{figure}
-\centering
 \includegraphics{images/fig3.pdf}
-\caption{Performance of MoE against all other methods. a) The power for
+
+Figure 3: Performance of MoE against all other methods. a) The power for
 non-null datasets is plotted against the FDR for null datasets for each
@@
 highest. Under the assumption of pervasive horizontal pleiotropy, the
-MoE approach is likely most effective than any other single method.}
-\end{figure}
+MoE approach is likely most effective than any other single method.
 
@@
 
-\begin{figure}
-\centering
-\includegraphics{images/fig4.png}
-\caption{Lookup of causal associations involving Years of Schooling. The
+\includegraphics{images/fig4-01.png}
+
+Figure 4: Lookup of causal associations involving a) LDL cholesterol and
+b) Years of Schooling. In a) the nodes are filtered to not include those
+obtained from the Shin et al or Kettunen et al metabolomic studies. The
 arrows denote causal direction and the values on the arrows denote the
 causal effect estimate. Only those relationships are shown for which the
-\(FDR < 0.05\).}
-\end{figure}
+\(FDR < 0.05\).
 
@@
 
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
+\subsection{References}\label{references}
+
+\raggedright
 
@@
 
+\hypertarget{ref-Verbanck2017}{}
+8. Verbanck M, Chen C-Y, Neale B, Do R. Widespread pleiotropy confounds
+causal relationships between complex traits and diseases inferred from
+Mendelian randomization. bioRxiv {[}Internet{]}. 2017; Available from:
+\url{http://www.biorxiv.org/content/early/2017/06/30/157552}
+
 \hypertarget{ref-Hindorff2010}{}
-8. Hindorff LA, Junkins HA, Hall PN, Mehta JP, Manolio TA. A Catalog of
+9. Hindorff LA, Junkins HA, Hall PN, Mehta JP, Manolio TA. A Catalog of
 Published Genome-Wide Association Studies, available at
@@
 \hypertarget{ref-Pierce2013}{}
-9. Pierce BL, Burgess S. Efficient design for Mendelian randomization
+10. Pierce BL, Burgess S. Efficient design for Mendelian randomization
 studies: subsample and 2-sample instrumental variable estimators.
@@
 \hypertarget{ref-Hemani2016}{}
-10. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
+11. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
 MR-Base: a platform for systematic causal inference across the phenome
@@
 
+\hypertarget{ref-Zhu2017}{}
+12. Zhu Z, Zheng Z, Zhang F, Wu Y, Trzaskowski M, Maier R, et al. Causal
+associations between risk factors and common diseases inferred from GWAS
+summary data. bioRxiv {[}Internet{]}. 2017; Available from:
+\url{http://www.biorxiv.org/content/early/2017/07/26/168674}
+
+\hypertarget{ref-Wright1968}{}
+13. Wright S. Evolution and the Genetics of Populations. In: Evolution
+and the genetics of populations. Chicago: University of Chicago Press;
+1968.
+
 \hypertarget{ref-Wagner2011}{}
-11. Wagner GP, Zhang J. The pleiotropic structure of the
+14. Wagner GP, Zhang J. The pleiotropic structure of the
 genotype--phenotype map: the evolvability of complex organisms. Nature
@@
 \hypertarget{ref-Hill2012a}{}
-12. Hill WG, Zhang X-S. Assessing pleiotropy and its evolutionary
+15. Hill WG, Zhang X-S. Assessing pleiotropy and its evolutionary
 consequences: pleiotropy is not necessarily limited, nor need it hinder
@@
 
+\hypertarget{ref-Hartwig2016}{}
+16. Hartwig FP, Davies NM, Hemani G, Davey Smith G. Two-sample Mendelian
+randomization: avoiding the downsides of a powerful, widely applicable
+but potentially fallible technique. International Journal of
+Epidemiology {[}Internet{]}. 2016 Dec;45(6):1717--26. Available from:
+\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dyx028}
+
+\hypertarget{ref-Xu2017}{}
+17. Xu L, Lin SL, Schooling CM. A Mendelian randomization study of the
+effect of calcium on coronary artery disease, myocardial infarction and
+their risk factors. Scientific Reports {[}Internet{]}. 2017 Feb;7:42691.
+Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/28195141\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5307362\%20http://www.nature.com/articles/srep42691}{http://www.ncbi.nlm.nih.gov/pubmed/28195141 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5307362 http://www.nature.com/articles/srep42691}
+
+\hypertarget{ref-Larsson2017}{}
+18. Larsson SC, Burgess S, Michaëlsson K, TB H, WH C, Y P. Association
+of Genetic Variants Related to Serum Calcium Levels With Coronary Artery
+Disease and Myocardial Infarction. JAMA {[}Internet{]}. 2017
+Jul;318(4):371. Available from:
+\url{http://jama.jamanetwork.com/article.aspx?doi=10.1001/jama.2017.8981}
+
+\hypertarget{ref-Prins2016}{}
+19. Prins BP, Abbasi A, Wong A, Vaez A, Nolte I, Franceschini N, et al.
+Investigating the Causal Relationship of C-Reactive Protein with 32
+Complex Somatic and Psychiatric Outcomes: A Large-Scale Cross-Consortium
+Mendelian Randomization Study. Hay PJ, editor. PLOS Medicine
+{[}Internet{]}. 2016 Jun;13(6):e1001976. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/27327646\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4915710\%20http://dx.plos.org/10.1371/journal.pmed.1001976}{http://www.ncbi.nlm.nih.gov/pubmed/27327646 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4915710 http://dx.plos.org/10.1371/journal.pmed.1001976}
+
+\hypertarget{ref-Inoshita2016}{}
+20. Inoshita M, Numata S, Tajima A, Kinoshita M, Umehara H, Nakataki M,
+et al. A significant causal association between C-reactive protein
+levels and schizophrenia. Scientific Reports {[}Internet{]}. 2016
+Sep;6(1):26105. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/27193331\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4872134\%20http://www.nature.com/articles/srep26105}{http://www.ncbi.nlm.nih.gov/pubmed/27193331 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4872134 http://www.nature.com/articles/srep26105}
+
 \hypertarget{ref-Rucker2011}{}
-13. Rucker G, Schwarzer G, Carpenter JR, Binder H, Schumacher M.
+21. Rucker G, Schwarzer G, Carpenter JR, Binder H, Schumacher M.
 Treatment-effect estimates adjusted for small-study effects via a limit
@@
 
+\hypertarget{ref-Han2008}{}
+22. Han C. Detecting invalid instruments using L 1-GMM. Economics
+Letters. 2008;101(3):285--7.
+
+\hypertarget{ref-Kang2016}{}
+23. Kang H, Zhang A, Cai TT, Small DS. Instrumental Variables Estimation
+With Some Invalid Instruments and its Application to Mendelian
+Randomization. Journal of the American Statistical Association.
+2016;111(513):132--44.
+
 \hypertarget{ref-Steiger1980}{}
-14. Steiger JH. Tests for comparing elements of a correlation matrix.
+24. Steiger JH. Tests for comparing elements of a correlation matrix.
 Psychological Bulletin. 1980;87(2):245--51.
 
+\hypertarget{ref-Corbin2016}{}
+25. Corbin LJ, Richmond RC, Wade KH, Burgess S, Bowden J, Smith GD, et
+al. BMI as a Modifiable Risk Factor for Type 2 Diabetes: Refining and
+Understanding Causal Estimates Using Mendelian Randomization. Diabetes
+{[}Internet{]}. 2016 Oct;65(10):3002--7. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/27402723\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5279886}{http://www.ncbi.nlm.nih.gov/pubmed/27402723 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5279886}
+
 \hypertarget{ref-Lee2013c}{}
-15. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+26. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 \hypertarget{ref-Jordan1994}{}
-16. Jordan MI, Jacobs RA. Hierarchical Mixtures of Experts and the EM
+27. Jordan MI, Jacobs RA. Hierarchical Mixtures of Experts and the EM
 Algorithm. Neural Computation {[}Internet{]}. 1994 Mar;6(2):181--214.
@@
 
+\hypertarget{ref-brazdil2008metalearning}{}
+28. Brazdil P, Carrier CG, Soares C, Vilalta R. Metalearning:
+Applications to data mining. Springer Science \& Business Media; 2008.
+
+\hypertarget{ref-smith2009cross}{}
+29. Smith-Miles KA. Cross-disciplinary perspectives on meta-learning for
+algorithm selection. ACM Computing Surveys (CSUR). 2009;41(1):6.
+
+\hypertarget{ref-lemke2015metalearning}{}
+30. Lemke C, Budka M, Gabrys B. Metalearning: a survey of trends and
+technologies. Artificial intelligence review. 2015;44(1):117--30.
+
 \hypertarget{ref-Liaw2002}{}
-17. Liaw A, Wiener M. Classification and Regression by randomForest. R
+31. Liaw A, Wiener M. Classification and Regression by randomForest. R
 News {[}Internet{]}. 2002;2(3):18--22. Available from:
@@
 \hypertarget{ref-Shin2014}{}
-18. Shin S-Y, Fauman EB, Petersen A-K, Krumsiek J, Santos R, Huang J, et
+32. Shin S-Y, Fauman EB, Petersen A-K, Krumsiek J, Santos R, Huang J, et
 al. An atlas of genetic influences on human blood metabolites. Nature
@@
 \hypertarget{ref-Kettunen2016}{}
-19. Kettunen J, Demirkan A, Wurtz P, Draisma HHM, Haller T, Rawal R, et
+33. Kettunen J, Demirkan A, Wurtz P, Draisma HHM, Haller T, Rawal R, et
 al. Genome-wide study for circulating metabolites identifies 62 loci and
@@
 \hypertarget{ref-Sun2017}{}
-20. Sun BB, Maranville JC, Peters JE, Stacey D, Staley JR, Blackshaw J,
+34. Sun BB, Maranville JC, Peters JE, Stacey D, Staley JR, Blackshaw J,
 et al. Consequences Of Natural Perturbations In The Human Plasma
@@
 
+\hypertarget{ref-Willer2013}{}
+35. Willer CJ, Schmidt EM, Sengupta S, Peloso GM, Gustafsson S, Kanoni
+S, et al. Discovery and refinement of loci associated with lipid levels.
+Nature Genetics {[}Internet{]}. 2013;45(11):1274--83. Available from:
+\url{http://www.nature.com/doifinder/10.1038/ng.2797}
+
 \hypertarget{ref-Okbay2016}{}
-21. Okbay A, Beauchamp JP, Fontana MA, Lee JJ, Pers TH, Rietveld CA, et
+36. Okbay A, Beauchamp JP, Fontana MA, Lee JJ, Pers TH, Rietveld CA, et
 al. Genome-wide association study identifies 74 loci associated with
@@
 
-\hypertarget{ref-hemani2016mr}{}
-22. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
-MR-Base: a platform for systematic causal inference across the phenome
-using billions of genetic associations. bioRxiv. 2016;78972.
+\hypertarget{ref-Anderson2017}{}
+37. Anderson E, Wade KH, Hemani G, Bowden J, Korologou-Linden R, Davey
+Smith G, et al. The Causal Effect Of Educational Attainment On
+Alzheimer's Disease: A Two-Sample Mendelian Randomization Study. bioRxiv
+{[}Internet{]}. 2017; Available from:
+\url{http://www.biorxiv.org/content/early/2017/04/17/127993}
+
+\hypertarget{ref-kotthoff2016auto}{}
+38. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
+2.0: Automatic model selection and hyperparameter optimization in WEKA.
+Journal of Machine Learning Research. 2016;17:1--5.
 
 \hypertarget{ref-Zhu2016}{}
-23. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+39. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
 Integration of summary data from GWAS and eQTL studies predicts complex
@@
 \hypertarget{ref-Richardson2017}{}
-24. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
+40. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
 CL, et al. Causal epigenome-wide association study identifies CpG sites
@@
 
+\hypertarget{ref-Wang2017}{}
+41. Wang L, Michoel T. Efficient And Accurate Causal Inference With
+Hidden Confounders From Genome-Transcriptome Variation Data. bioRxiv
+{[}Internet{]}. 2017; Available from:
+\url{http://www.biorxiv.org/content/early/2017/04/19/128496}
+
 \hypertarget{ref-Noyce2017}{}
-25. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
+42. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
 E, et al. Estimating the causal influence of body mass index on risk of
@@
 \hypertarget{ref-Burgess2014a}{}
-26. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+43. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 
+\hypertarget{ref-Staley2016}{}
+44. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
+al. PhenoScanner: a database of human genotype--phenotype associations.
+Bioinformatics {[}Internet{]}. 2016 Oct;32(20):3207--9. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/27318201\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5048068\%20https://academic.oup.com/bioinformatics/article-lookup/doi/10.1093/bioinformatics/btw373}{http://www.ncbi.nlm.nih.gov/pubmed/27318201 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5048068 https://academic.oup.com/bioinformatics/article-lookup/doi/10.1093/bioinformatics/btw373}
+
 \end{document}

### 2017-07-29 docs/manuscript.rmd
updated manuscript

@@
 
-Gibran Hemani$^1$, Jack Bowden$^1$, Philip Haycock$^1$, Jie Zheng$^1$, Oliver Davis$^1$, Peter Flach, Tom Gaunt$^{1,*}$, George Davey Smith$^{1,*}$
+Gibran Hemani$^1$, Jack Bowden$^1$, Philip Haycock$^1$, Jie Zheng$^1$, Oliver Davis$^1$, Peter Flach$^2$, Tom Gaunt$^{1,*}$, George Davey Smith$^{1,*}$
 
 1. MRC Integrative Epidemiology Unit (IEU) at the University of Bristol, School of Social and Community Medicine, Bristol, UK
+2. Intelligent Systems Laboratory, Department of Computer Science, University of Bristol
 
@@
 We have developed a mixture of experts machine learning approach (MR-MoE 1.0) that improves on both power and false discovery rates over unselective use of any existing methods.
-Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a working draft of the causal map of the human phenome.
+Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a working draft of the causal map of the human phenome comprising the evaluation of causality between over a million pairs of traits.
 Almost 90% of causal estimates indicated some level of horizontal pleiotropy.
-All results are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)).
+The causal estimates are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)).
 
@@
 Next, the effects of those SNPs on the outcome need to be obtained, typically necessitating complete summary data because these SNPs will generally have not have reached genome-wide significance for other traits (and therefore be present in curated catalogues).
+
+We use the term *summary-set* to refer to the minimum data required to perform 2SMR analysis - four columns comprising the SNP-exposure and SNP-outcome effects and standard errors, and rows corresponding to each SNP that is used as an instrument for the exposure.
+
 At its most simple implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome.
@@
 Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are likely to be exhibiting horizontal pleiotropy.
-Conversely, we primarily developed Steiger filtering to identify those instruments that are likely to be arising due to reverse cause.
+A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones.
+This could arise, for example, for the *CRP* variant instrumenting C-reactive protein levels [@Prins2016], or the *SLC2A9* variant influencing serum urate levels [@white2016plasma].
 
-The Steiger test is applied to each variant in turn and we exclude any $g_{A}$ for which $cor(g_{i}, A)^2 > cor(g_{i}, B)^2$, indicating that it is unlikely to primarily associate with $B$ relative to $A$.
-Similarly, for SNPs that influence confounders of $A$ and $B$ or exhibit horizontal pleiotropy, the difference in $cor(g_{i}, A)^2$ and $cor(g_{i}, B)^2$ will be reduced, increasing the likelihood of the SNP being excluded because the Steiger Z-test is less likely to be significant.
-Hence we also exclude any $g_{A}$ for which the Steiger Z-test has $p > 0.05$.
-
-To estimate $cor(g, x)^2$, if $x$ is continuous we obtain the F-statistic from the reported p-value and sample size and then $cor(g, x)^2 = \frac{F}{N - 2 - F}$.
-If $x$ is binary then we estimate the variance of the underlying liability explained by the SNP, $cor(g, x)^2 = \frac{V_a}{V_a + V_e}$.
-Here, $V_e = \pi^2/3$, and $V_a = 2\beta^2p(1-p)$, where $\beta$ is the log odds ratio and $p$ is the allele frequency of the SNP in the population [@Lee2013c].
-$p$ can be estimated using the allele frequency of the SNP in an ascertained sample by deriving the $2 \times 2$ contingency table from the odds ratio $e^\beta$, allele frequency in the ascertained sample $p_{cc}, and number of cases $N_1$ and controls $N_0$.
-We exclude any SNPs that do not have the correct causal direction for the hypothesis, and do not have Steiger test $p < 0.05$.
+To avoid this, we primarily developed Steiger filtering to identify those instruments that are likely to be arising due to reverse cause, but it also has the potential to detect SNPs that are exhibiting horizontal pleiotropy or primarily influencing confounders.
+Further details are provided in the Online Methods.
 
@@
 Our objective is to select the expert most likely to be correct for a specific MR analysis.
-A overview of the approach is shown in Figure 2.
+An overview of the approach is shown in Figure 2.
 
@@
 In this case our objective is to identify characteristics of the SNP-exposure and SNP-outcome associations for which one specific MR method is most likely to yield highest statistical power for non-null associations, and lowest false discovery rates for null associations.
-This involves creating a 'gating function', whose purpose is to decide which expert to use for a specific MR analysis, given the parameter space that is occupied by that dataset.
+This involves creating a 'gating function', whose purpose is to decide which expert to use for a specific MR analysis, given the parameter space that is occupied by that summary-set.
 The metrics are a mixture of regression diagnostics and are described in Supplementary table 3.
 
-The gating function needs to be trained using data for which the true causal effect is known, and to this end we generated a large number of simulated datasets.
+The gating function needs to be trained using data for which the true causal effect is known, and to this end we generated a large number of simulated summary-sets.
 We trained the gating function using random forest learning algorithms that seek to identify sectors of the parameter space that are most likely to return accurate estimates for a particular expert.
 Figure 2a illustrates how the gating function is trained using simulated data.
-New datasets can then be applied to the trained mixture of experts (MoE) model (Figure 2b).
+New summary-sets can then be applied to the trained mixture of experts (MoE) model (Figure 2b).
 We call this implementation MR-MoE 1.0.
+Full details are provided in the Online Methods.
+MR-MoE 1.0 is implemented in the TwoSampleMR R package available at [github.com/MRCIEU/TwoSampleMR](https://github.com/MRCIEU/TwoSampleMR) [@Hemani2016].
+
+### Graph database of MR estimates
+
+The set of MR estimates obtained from this analysis, and the summary-sets used to generate them, are deposited in a Neo4j graph database.
+Here nodes representations exist for traits and SNPs; and relationship representations exist for SNP-trait associations and trait-trait MR estimates.
+Because each trait-trait association has up to 28 different MR estimates, for simplicity we also distill this down to a third relationship type comprising a single 'best estimate' for each trait-trait association using the following rules:
+
+1. If the number of variants after Steiger filtering is greater than 5 then apply the MoE to obtain the best method
+2. If the number of variants after Steiger filtering is less than or equal to 5 but greater than 1 then use the IVW random effects approach on the filtered set of variants
+3. If there is 1 variant retained in the Steiger approach then use the Wald ratio on the remaining variant
+4. If there are no variants remaining after Steiger filtering then declare no causal association.
+
+For specific hypotheses we strongly recommend that estimates from all sensitivity analyses are scrutinised and reported.
+The graph can be queried directly using the cypher language at [http://eve-neo4j.mrbase.org](http://eve-neo4j.mrbase.org) or through a basic web interface at [http://eve.mrbase.org](http://eve.mrbase.org).
+
+## Results
+
+### Steiger filtering improves reliability
+
+As statistical power for GWAS studies improves the likelihood of a significant association being discovered that doesn't act primarily on the trait of interest increases (Figure 1a).
+We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated summary-sets comprising both null and non-null causal models.
+For each summary-set, instruments were selected based on the tophits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.
+
+Figure 1b shows that the tophits strategy led to over half of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype.
+The proportion of invalid instruments due to reverse cause increased as GWAS discovery sample size increased.
+Applying Steiger filtering reduced this to 25%.
+Consequently, the false discovery rates (FDR) for 12 of the 14 methods reduced substantially when applied using Steiger filtered instruments (Figure 1c).
+The true positive rates for the methods based on Steiger filtering did however reduce slightly for 10 of the 14 methods.
+
+### Mixture of experts method selection improves over any single method
+
+Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to predict the most appropriate of the 28 MR methods and instrument selection strategies to apply to a particular summary-set based on its characteristics (Figure 2).
+
+The ability to predict the performance for each of these methods is shown in Supplementary table 2.
+The prediction $R^2$ of whether a summary-set's status was truly null ($\beta_{xy}=0$) or non-null ($\beta_{xy} \neq 0$) against the method's prediction of the summary-set's status, ranged between 0.04 and 0.24.
+The method performance prediction was most effective for the Egger random effects model with Steiger filtering.
+The summary-set characteristics with the most importance for each of the predictors differed substantially between each summary-set, as well as the frequency for which each of the methods was selected in the testing summary-sets.
+The FDR of each method when chosen, compared to their averages across all summary-sets, reduced; and likewise the true positive rates of each method when chosen increased compared to their simulation-wide averages.
+
+We compared the MoE performance in the simulations against each of the 28 methods, testing to see if it outperformed all other single methods.
+Figures 3a and 3b show that the MoE approach had the best general performance.
+Estimating the area under the receiver operator curve gave 0.84 for the MoE approach in terms of classifying the simulations as being null or non-null.
+Notably, the next best methods were median and mode based estimators using Steiger filtering.
+Under the assumption of widespread and diverse pleiotropic effects it is clear that all methods suffer from high false discovery rates (Figure 3a).
+
+### Automated MR analysis of 2407 phenotypes
+
+We applied our analysis using summary data for 2407 phenotypes, including 149 complex traits and diseases, 575 metabolites [@Shin2014; @Kettunen2016] and 1683 plasma protein levels [@Sun2017].
+For the protein levels only the instruments were available, so they could only be evaluated as exposure phenotypes.
+The complex traits and diseases and metabolite levels had complete summary data available in MR-Base, thus they be evaluated as both exposures (if they had significant instruments) and outcomes.
+Together, we evaluated 715681 relationships.
+The majority of these associations could only be evaluated using fewer than 5 SNPs, and so the Wald ratio or IVW fixed effects methods were used, but for 61029 associations the MR-MoE approach was applied.
+
+There were 5660 associations following Bonferroni correction ($p < 7.0 \times 10^{-8}$).
+Of these 2918 were obtained from the MR-MoE analysis, while the remainder were estimated using fewer than 5 SNPs using the Wald ratio or IVW fixed effects methods.
+
+The frequencies of the methods chosen by the MR-MoE analysis are shown in Supplementary table 3.
+Amongst those deemed 'significant', the IVW fixed effects analysis using tophits instruments (the only method applicable when there is no evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases.
+This indicates that horizontal pleiotropy is likely to be pervasive.
+
+### Putative associations involving years of schooling
+
+As an example of the database we performed a look up of associations where LDL cholesterol as measured in the GLGC consortium [@Willer2013] had a false discovery rate of 0.05 using the MoE approach.
+This returned 287 putative associations, amongst which 111 involved LDL cholesterol influencing other traits and 176 involved other traits influencing LDL cholesterol.
+A large proportion of these traits were metabolites, which are dominated by lipid fractions.
+Filtering to exclude the metabolomic studies ([@Shin2014; @Kettunen2016]) returned 27 associations with 23 traits (Figure 4a).
+There was a strong association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52, $p=2.7\times10^{-12}$).
+Here the IVW random effects method was chosen after Steiger filtering, indicating a prediction of HP being present amongst the instruments.
+
+We also performed a look up of associations that causally influence years of schooling [@Okbay2016], or that years of schooling itself causally influences.
+Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4b).
+Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease has recently been examined [@Anderson2017].
+However it is clear that there are serious limitations in relying on MR as a panacea for causal inference.
+For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g.
+childhood intelligence and birth weight) which appear to be causally downstream of years of schooling despite this being temporally impossible.
+Though explanations could be conjured for this association, for example parents' schooling influences childhood intelligence, or the genetic instruments are shared across the two traits, an intuitive interpretation of causality cannot be applied.
+
+The facility to perform these scans has been automated in the MR of everything vs everything (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).
+
+## Discussion
+
+The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones.
+Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception.
+Following a recent flurry in development of MR methodology, we have devised a machine learning approach, MR-MoE, that seeks to predict the performance of each MR method, selecting the one which is most likely to maximise power whilst minimising FDR for a specific summary-set.
+The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy.
+As such, our method contributes to the field of meta-learning but also to the rapidly growing field of automatic model selection and hyper-parameter optimization in machine learning [@kotthoff2016auto].
+
+Crucially however, we observed in our simulations that none of the available methods were conservative and all, including MR-MoE, had FDR above 0.2.
+Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criteria.
+We illustrated how confounders and reverse causal associations can easily lead to invalid instruments being used in MR, and that even if those instruments that only influence the exposure directly are used, most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.
+
+We applied the method to the curated set of GWAS summary data present in MR-Base [@Hemani2016], generating several million MR estimates.
+MR-MoE selected a method that indicated some pattern of horizontal pleiotropy for almost 90% of cases, reinforcing the notion that it is the rule rather than the exception.
+Based on our simulations we expect that those associations that we reported to be 'significant' are liable to high type 1 error rates.
+
+One advantage to generating a method that consolidates across many methods is that it can prevent cherry picking the most appealing method based on their results.
+But in addition to correctly handling horizontal pleiotropy, there are many other limitations that prevent MR from being a panacea for causal inference.
+Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes.
+Though these associations can be informative, their interpretations as causal relationships are far from clear.
+Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease.
+Again, interpretation of such associations can be problematic.
+
+A large proportion of the associations that were estimated used only a single instrumental variable.
+Methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017].
+Separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework though the use of mediation-based approaches may be informative in the one-sample setting [@Wang2017; @Hemani2017].
+Other problems can also manifest, for example frailty effects could induce associations for late-onset traits [@Noyce2017]; genetic variants could strongly relate to several phenotypes making it difficult to ascertain which of them is the causal exposure [@Burgess2014a]; and the measured feature for which genetic associations are known could relate in complicated ways to the biological entity that are truly causal [@DaveySmithHemani2014].
+
+We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular summary-set, namely, that the data is choosing the model.
+This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance.
+The extent to which a machine based method selector suffers from this more than a human manually inferring the result most likely to be true is not clear.
+Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to all other MR methods that do not suffer from this issue.
+
+The field of MR is evolving.
+The advent of GWAS databases [@Hindorff2010; @Staley2016] and automated 2SMR [@Hemani2016] has trivialised the analytical aspect of investigating specific causal hypotheses.
+Despite the limitations described above, the construction of a causal graph of 'everything versus everything' does have appeal.
+First, though causality is not guaranteed by MR, it can still be highly informative for confirming or negating specific hypotheses.
+Second, it paves the way for new approaches to search for novel putative associations and improve reliability in single analyses.
+The potential to exploit GWAS summary data within the properties of graph databases to aid with both of these endeavours could be transformative in biological research.
+
+## Acknowledgements
+
+This work was supported by the UK Medical Research Council (MC_UU_12013/1).
+
+\newpage
+
+![](images/fig1.pdf)
+
+Figure 1: Simulations.
+a) Schematic of how GWAS with sufficient power can lead to the selection of instrumental variables that are invalid.
+We used arbitrary numbers of SNPs and confounders to simulate GWAS summary datasets.
+b) Any SNP that has a direct influence on the exposure, or an influence on a non-confounding intermediate variable, is considered a 'direct' effect.
+The y-axis shows the proportion of instruments selected for analysis due that are either direct associations with the exposure, instruments for the outcome (reverse), or instruments for confounding traits (confounder).
+The proportions are compared over a range of different exposure discovery sample sizes (x-axis) and using either the tophits approach (left) or the Steiger approach (right) for instrument selection.
+c) Top: The false discovery rates from null simulations for each of the 14 methods using either tophits, Steiger filtered variants, or variants that are known to be directly associated with the exposure (oracle, note that direct effects can still exhibit horizontal pleiotropy in these simulations).
+Bottom: The statistical power to detect true causal associations in the non-null simulations.
+
+\newpage
+
+![](images/fig2.pdf)
+
+Figure 2: Mixture of experts.
+a) Training.
+Datasets are simulated that have either null or non-null causal relationships, and MR is performed by each of the 28 available MR methods.
+In the toy datasets, the columns in black represent 53 metrics about each of the 67,000 training simulations.
+The columns in red are specific to each method, they represent how well that method performed in obtaining the correct answer for each of the datasets.
+Random forests are used to learn the parameter space of the 53 metrics in which a particular dataset is likely to perform well.
+Together, this creates 28 random forest decision trees, one for each method.
+b) Application.
+For a GWAS summary dataset, our objective is to choose the method most likely to return the correct causal estimate.
+Metrics are generated from the dataset and fed into each of the 28 random forest decision trees.
+This provides us with 28 performance predictions.
+Finally, we use the method for which the performance prediction is highest.
+
+\newpage
+
+![](images/fig3.pdf)
+
+Figure 3: Performance of MoE against all other methods.
+a) The power for non-null datasets is plotted against the FDR for null datasets for each of the 28 methods, plus MoE.
+No single method achieved nominal FDR for these simulations.
+b) Calculating the area under the ROC curve from the values in (a) we plotted the performance in order from lowest to highest.
+Under the assumption of pervasive horizontal pleiotropy, the MoE approach is likely most effective than any other single method.
+
+\newpage
+
+![](images/fig4-01.png)
+
+Figure 4: Lookup of causal associations involving a) LDL cholesterol and b) Years of Schooling.
+In a) the nodes are filtered to not include those obtained from the Shin et al or Kettunen et al metabolomic studies.
+The arrows denote causal direction and the values on the arrows denote the causal effect estimate.
+Only those relationships are shown for which the $FDR < 0.05$.
+
+\newpage
+
+## Online methods
+
+### Steiger filtering
+
+An approach to inferring the causal direction between phenotypes recently developed [@Hemani2017] uses the following basic premise.
+If trait $A$ causes trait $B$ then $cor(g_{i}, A)^2 > cor(g_{i}, B)^2$ because the $cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}$.
+This will be true under most circumstances, but some parameters of measurement error or unmeasured confounding can lead to the inequality reversing direction, as has been explored previously [@Hemani2017].
+
+The Steiger test is applied to each variant in turn and we exclude any $g_{A}$ for which $cor(g_{i}, A)^2 > cor(g_{i}, B)^2$, indicating that it is unlikely to primarily associate with $B$ relative to $A$.
+Similarly, for SNPs that influence confounders of $A$ and $B$ or exhibit horizontal pleiotropy, the difference in $cor(g_{i}, A)^2$ and $cor(g_{i}, B)^2$ will be reduced, increasing the likelihood of the SNP being excluded because the Steiger Z-test is less likely to be significant.
+Hence we also exclude any $g_{A}$ for which the $cor(g, x)$ and $cor(g, y)$ are not significantly different at an arbitrary threshold of $p > 0.05$.
+
+To estimate $cor(g, x)^2$, if $x$ is continuous we obtain the F-statistic from the reported p-value and sample size and then $cor(g, x)^2 = \frac{F}{N - 2 - F}$.
+If $x$ is binary then we estimate the variance of the underlying liability explained by the SNP, $cor(g, x)^2 = \frac{V_a}{V_a + V_e}$.
+Here, $V_e = \pi^2/3$, and $V_a = 2\beta^2f(1-f)$, where $\beta$ is the log odds ratio and $f$ is the allele frequency of the SNP in the population [@Lee2013c].
+$f$ can be estimated using the allele frequency of the SNP in an ascertained sample by deriving the $2 \times 2$ contingency table from the odds ratio $e^\beta$, allele frequency in the ascertained sample $f_{cc}$, and number of cases $N_1$ and controls $N_0$.
+A Z-score to test the hypothesis that $cor(g, x)$ and $cor(g, y)$ are significantly different can be generated using the Steiger test for correlated correlations.
+
+### Mixture of experts implementation for MR
+
+The mixture of experts approach seeks to use simulations to learn the circumstances in which each of the 28 MR methods is likely to perform most accurately.
+Having trained the model on simulations, new summary-sets can be applied to obtain causal estimates from the method that is predicted to be the most reliable for that summary-set.
 
@@
 As such our MR-MoE approach fits the framework of meta-learning [@brazdil2008metalearning; @smith2009cross; @lemke2015metalearning] where machine learning methods are applied at the 'meta-level' to analyse the results of machine learning experiments at the 'base-level', in order to be able to understand and model the capabilities of each expert and recommend which expert should be applied to a given problem.
-From this perspective the 28 MR methods are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the datasets.
+From this perspective the 28 MR methods are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the summary-sets.
 
@@
 
-The MoE is trained using datasets generated from simulations (Figure 2a).
-A *dataset* is the minimum data required to perform 2SMR analysis - four columns comprising the SNP-exposure and SNP-outcome effects and standard errors, and rows corresponding to each SNP that is used as an instrument for the exposure.
-This *dataset* can be fed into any of the 28 experts to obtain MR causal effects.
-The simulations used to generate these datasets seek to cover a range of pleiotropic scenarios, including where some proportion of SNPs exhibit directional or balanced horizontal pleiotropy, or where SNPs influence confounding variables.
+The MoE is trained using summary-sets generated from simulations (Figure 2a).
+Each *summary-set* can be fed into any of the 28 experts to obtain MR causal effects.
+The simulations used to generate these summary-sets seek to cover a range of pleiotropic scenarios, including where some proportion of SNPs exhibit directional or balanced horizontal pleiotropy, or where SNPs influence confounding variables.
 
-We simulate two individual level datasets for which there are $N_x$ and $N_y$ samples, and $M$ SNPs, where each SNP has effect allele frequency of $p_m \sim U(0.05, 0.95)$.
-These datasets are used to obtain the SNP effects for the exposure trait $x$ and the outcome trait $y$, respectively, using the following sampling criteria:
+We simulate two individual level datasets for which there are $N_x$ and $N_y$ samples, and $M$ SNPs, where each SNP has effect allele frequency of $f_m \sim U(0.05, 0.95)$.
+These datasets are used to obtain the SNP effects for the exposure trait $x$ and the outcome trait $y$, respectively, to create summary-sets, using the following sampling criteria:
 
@@
 We then perform MR analysis in both directions, mimicking GWAS by retaining SNPs that have $p < 5e-8$ in sample 1 to perform MR of $x$ on $y$ (the true causal direction for non-null simulations), and retaining SNPs that have $p < 5e-8$ in sample 2 to perform MR of $y$ on $x$ (the reverse causal direction for non-null simulations).
-We treat the summary data (effect sizes and standard errors) used for estimating $x \rightarrow y$ the summary data used for estimating $y \rightarrow x$ as two separate datasets Hence, for each simulation two datasets are generated which are analysed to produce 28 MR estimates each.
-We performed 100,000 simulations using these parameters, resulting in 200,000 datasets.
+We treat the summary data (effect sizes and standard errors) used for estimating $x \rightarrow y$ the summary data used for estimating $y \rightarrow x$ as two separate summary-sets.
+Hence, for each simulation two summary-sets are generated which are analysed to produce 28 MR estimates each.
+We performed 100,000 simulations using these parameters, resulting in 200,000 summary-sets.
 
@@
 
-We aim to maximise statistical power for datasets where $\beta_{x,y} \neq 0$ and minimise false discovery rates for datasets where $\beta_{x,y} = 0$.
-To train random forest decision trees to predict performance for a particular method $h(O_{w,d}, \textbf{z}_{d})$ is generated where the training set of input metrics for dataset $d$ is $\textbf{z}_{d}$ and the response (optimisation function) is
+We aim to maximise statistical power for summary-sets where $\beta_{x,y} \neq 0$ and minimise false discovery rates for summary-sets where $\beta_{x,y} = 0$.
+To train random forest decision trees to predict performance for a particular method $h(O_{w,d}, \textbf{z}_{d})$ is generated where the training set of input metrics for summary-set $d$ is $\textbf{z}_{d}$ and the response (optimisation function) is
 
@@
 
-where $p_{m,d}$ is the p-value for method $m$ on dataset $d$.
+where $p_{m,d}$ is the p-value for method $m$ on summary-set $d$.
 
@@
 
-For each training dataset we record a set of 53 metrics $\textbf{z}_{d}$ (Supplementary table 3), and an outcome $O_{w,d}$, which is a measure of how well that method performed for each particular simulated dataset.
-For each of our 28 methods, we need to create a model that predicts the performance of the method based on metrics generated from a dataset.
-To do this, for each method we train a random decision forest to predict that method's performance using the dataset's metrics.
+For each training summary-set we record a set of 53 metrics $\textbf{z}_{d}$ (Supplementary table 3), and an outcome $O_{w,d}$, which is a measure of how well that method performed for each particular simulated summary-set.
+For each of our 28 methods, we need to create a model that predicts the performance of the method based on metrics generated from a summary-set.
+To do this, for each method we train a random decision forest to predict that method's performance using the summary-set's metrics.
 The random forest approach is well suited to this problem because there are likely to be non-continuous combinations of different metrics that improve on prediction over, for example, a simple linear model that does not learn about interactions.
 
-As a simple hypothetical example - if a dataset has a single outlier but is otherwise exhibiting no heterogeneity then the following methods could arguably perform well:
+As a simple hypothetical example - if a summary-set exhibits a single outlier but is otherwise exhibiting no heterogeneity then the following methods could arguably perform well:
 
@@
 
-Having generated random forest decision trees for each of the 28 methods using 133,000 of the simulations, we then applied them to the remaining 67,000 datasets to predict which method would have the highest performance for each of the remaining datasets.
+Having generated random forest decision trees for each of the 28 methods using 133,000 of the simulations, we then applied them to the remaining 67,000 summary-sets to predict which method would have the highest performance for each of the remaining summary-sets.
 Finally we compare the performance of the method selected by the MoE against all remaining methods.
@@
 
-### Graph database of MR estimates
-
-The set of MR estimates obtained from this analysis are recorded in a Neo4j graph database.
-Because each association has up to 28 different estimates, for simplicity we distill this down to a single 'best estimate' for each association using the following rules:
-
-1. If the number of variants after Steiger filtering is greater than 5 then apply the MoE to obtain the best method
-2. If the number of variants after Steiger filtering is less than or equal to 5 but greater than 1 then use the IVW random effects approach on the filtered set of variants
-3. If there is 1 variant retained in the Steiger approach then use the Wald ratio on the remaining variant
-4. If there are no variants remaining after Steiger filtering then declare no causal association.
-
-For specific hypotheses we strongly recommend that estimates from all sensitivity analyses are scrutinised and reported.
-The graph can be queried directly using the cypher language at [http://eve-neo4j.mrbase.org](http://eve-neo4j.mrbase.org) or through a basic web interface at [http://eve.mrbase.org](http://eve.mrbase.org).
-
-## Results
-
-### Steiger filtering improves reliability
-
-As statistical power for GWAS studies improves the likelihood of a significant association being discovered that doesn't act primarily on the trait of interest increases (Figure 1a).
-We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated datasets comprising both null and non-null causal models.
-For each dataset, instruments were selected based on the tophits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.
-
-Figure 1b shows that the tophits strategy led to over half of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype.
-The proportion of invalid instruments due to reverse cause increased as GWAS discovery sample size increased.
-Applying Steiger filtering reduced this to 25%.
-Consequently, the false discovery rates (FDR) for 12 of the 14 methods reduced substantially when applied using Steiger filtered instruments (Figure 1c).
-The true positive rates for the methods based on Steiger filtering did however reduce slightly for 10 of the 14 methods.
-
-### Mixture of experts method selection improves over any single method
-
-Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to predict the most appropriate of the 28 MR methods and instrument selection strategies to apply to a particular dataset based on its characteristics (Figure 2).
-
-The ability to predict the performance for each of these methods is shown in Supplementary table 2.
-The prediction $R^2$ of whether a dataset's status was truly null ($\beta_{xy}=0$) or non-null ($\beta_{xy} \neq 0$) against the method's prediction of the dataset's status, ranged between 0.04 and 0.24.
-The method performance prediction was most effective for the Egger random effects model with Steiger filtering.
-The dataset characteristics with the most importance for each of the predictors differed substantially between each dataset, as well as the frequency for which each of the methods was selected in the testing datasets.
-The FDR of each method when chosen, compared to their averages across all datasets, reduced; and likewise the true positive rates of each method when chosen increased compared to their simulation-wide averages.
-
-We compared the MoE performance in the simulations against each of the 28 methods, testing to see if it outperformed all other single methods.
-Figures 3a and 3b show that the MoE approach had the best general performance, obtaining an AUC of 0.84 in terms of classifying the simulations as being null or non-null.
-Notably, the next best methods were median and mode based estimators using Steiger filtering.
-Under the assumption of widespread pleiotropy it is clear that all methods suffer from high false discovery rates.
-
-### Automated MR analysis of 2407 phenotypes
-
-We applied our analysis using summary data for 2407 phenotypes, including 149 complex traits and diseases, 575 metabolites [@Shin2014; @Kettunen2016] and 1683 plasma protein levels [@Sun2017].
-For the protein levels only the instruments were available, so they could only be evaluated as exposure phenotypes.
-The complex traits and diseases and metabolite levels had complete summary data available in MR-Base, thus they be evaluated as both exposures (if they had significant instruments) and outcomes.
-Together, we evaluated 715681 relationships.
-The majority of these associations could only be evaluated using fewer than 5 SNPs, and so the Wald ratio or IVW fixed effects methods were used, but for 61029 associations the MR-MoE approach was applied.
-
-There were 5660 associations following Bonferroni correction ($p < 7.0 \times 10^{-8}$).
-Of these 2918 were obtained from the MR-MoE analysis, while the remainder were estimated using fewer than 5 SNPs using the Wald ratio or IVW fixed effects methods.
-
-The frequencies of the methods chosen by the MR-MoE analysis are shown in Supplementary table 3.
-Amongst those deemed 'significant', the IVW fixed effects analysis using tophits instruments (the only method applicable when there is no evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases.
-This indicates that horizontal pleiotropy is likely to be pervasive.
-
-### Putative associations involving years of schooling
-
-As an example of the database we performed a look up of associations where LDL cholesterol as measured in the GLGC consortium [@Willer2013] had a false discovery rate of 0.05 using the MoE approach.
-This returned 287 putative associations, amongst which 111 involved LDL cholesterol influencing other traits and 176 involved other traits influencing LDL cholesterol.
-A large proportion of these traits were metabolites, which are dominated by lipid fractions.
-Filtering to exclude the metabolomic studies ([@Shin2014; @Kettunen2016]) returned 27 associations with 23 traits (Figure 4a).
-There was a strong association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52, $p=2.7\times10^{-12}$).
-Here the IVW random effects method was chosen after Steiger filtering, indicating a prediction of HP being present amongst the instruments.
-
-We also performed a look up of associations that causally influence years of schooling [@Okbay2016], or that years of schooling itself causally influences.
-Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4b).
-Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease has recently been examined [@Anderson2017].
-However it is clear that there are serious limitations in relying on MR as a panacea for causal inference.
-For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g.
-childhood intelligence and birth weight) which appear to be causally downstream of years of schooling despite this being temporally impossible.
-Though explanations could be conjured for this association, for example parents' schooling influences childhood intelligence, or the genetic instruments are shared across the two traits, an intuitive interpretation of causality cannot be applied.
-
-The facility to perform these scans has been automated in the MR of everything vs everything (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).
-
-## Discussion
-
-The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones.
-Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception.
-Following a recent flurry in development of MR methodology, we have devised a machine learning approach, MR-MoE, that seeks to predict the performance of each MR method, selecting the one which is most likely to maximise power whilst minimising FDR for a specific dataset.
-The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy.
-As such, our method contributes to the field of meta-learning but also to the rapidly growing field of automatic model selection and hyper-parameter optimization in machine learning [@kotthoff2016auto].
-
-Crucially however, we observed in our simulations that none of the available methods were conservative and all, including MR-MoE, had FDR above 0.2.
-Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criteria.
-We illustrated how confounders and reverse causal associations can easily lead to invalid instruments being used in MR, and that even if those instruments that only influence the exposure directly are used, most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.
-
-We applied the method to the curated set of GWAS summary data present in MR-Base [@Hemani2016], generating several million MR estimates.
-MR-MoE selected a method that indicated some pattern of horizontal pleiotropy for almost 90% of cases, reinforcing the notion that it is the rule rather than the exception.
-Based on our simulations we expect that those associations that we reported to be 'significant' are liable to high type 1 error rates.
-
-In addition to correctly handling horizontal pleiotropy, there are many other limitations that prevent MR from being a panacea for causal inference.
-Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes.
-Though these associations can be informative, their interpretations as causal relationships are far from clear.
-Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease.
-Again, interpretation of such associations can be problematic.
-
-A large proportion of the associations that were estimated used only a single instrumental variable.
-Methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017].
-Separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework though the use of mediation-based approaches may be informative in the one-sample setting [@Wang2017; @Hemani2017].
-Other problems can also manifest, for example frailty effects could induce associations for late-onset traits [@Noyce2017]; genetic variants could strongly relate to several phenotypes making it difficult to ascertain which of them is the causal exposure [@Burgess2014a]; and the measured feature for which genetic associations are known could relate in complicated ways to the biological entity that are truly causal [@DaveySmithHemani2014].
-
-We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular dataset, namely, that the data is choosing the model.
-This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance.
-The extent to which a machine based method selector suffers from this more than a human manually inferring the result most likely to be true is not clear.
-Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to all other MR methods that do not suffer from this issue.
-
-The field of MR is evolving.
-The advent of GWAS databases [@Hindorff2010; @Staley2016] and automated 2SMR [@Hemani2016] has trivialised the analytical aspect of investigating specific causal hypotheses.
-Despite the limitations described above, the construction of a causal graph of 'everything versus everything' does have appeal.
-First, though causality is not guaranteed by MR, it can still be highly informative for confirming or negating specific hypotheses.
-Second, it paves the way for new approaches to search for novel putative associations and improve reliability in single analyses.
-The potential to exploit GWAS summary data within the properties of graph databases to aid with both of these endeavours could be transformative in biological research.
-
-\newpage
-
-![](images/fig1.pdf)
-
-Figure 1: Simulations.
-a) Schematic of how GWAS with sufficient power can lead to the selection of instrumental variables that are invalid.
-We used arbitrary numbers of SNPs and confounders to simulate GWAS summary datasets.
-b) Any SNP that has a direct influence on the exposure, or an influence on a non-confounding intermediate variable, is considered a 'direct' effect.
-The y-axis shows the proportion of instruments selected for analysis due that are either direct associations with the exposure, instruments for the outcome (reverse), or instruments for confounding traits (confounder).
-The proportions are compared over a range of different exposure discovery sample sizes (x-axis) and using either the tophits approach (left) or the Steiger approach (right) for instrument selection.
-c) Top: The false discovery rates from null simulations for each of the 14 methods using either tophits, Steiger filtered variants, or variants that are known to be directly associated with the exposure (oracle, note that direct effects can still exhibit horizontal pleiotropy in these simulations).
-Bottom: The statistical power to detect true causal associations in the non-null simulations.
-
-\newpage
-
-![](images/fig2.pdf)
-
-Figure 2: Mixture of experts.
-a) Training.
-Datasets are simulated that have either null or non-null causal relationships, and MR is performed by each of the 28 available MR methods.
-In the toy datasets, the columns in black represent 53 metrics about each of the 67,000 training simulations.
-The columns in red are specific to each method, they represent how well that method performed in obtaining the correct answer for each of the datasets.
-Random forests are used to learn the parameter space of the 53 metrics in which a particular dataset is likely to perform well.
-Together, this creates 28 random forest decision trees, one for each method.
-b) Application.
-For a GWAS summary dataset, our objective is to choose the method most likely to return the correct causal estimate.
-Metrics are generated from the dataset and fed into each of the 28 random forest decision trees.
-This provides us with 28 performance predictions.
-Finally, we use the method for which the performance prediction is highest.
-
-\newpage
-
-![](images/fig3.pdf)
-
-Figure 3: Performance of MoE against all other methods.
-a) The power for non-null datasets is plotted against the FDR for null datasets for each of the 28 methods, plus MoE.
-No single method achieved nominal FDR for these simulations.
-b) Calculating the area under the ROC curve from the values in (a) we plotted the performance in order from lowest to highest.
-Under the assumption of pervasive horizontal pleiotropy, the MoE approach is likely most effective than any other single method.
-
-\newpage
-
-![](images/fig4-01.png)
-
-Figure 4: Lookup of causal associations involving a) LDL cholesterol and b) Years of Schooling.
-In a) the nodes are filtered to not include those obtained from the Shin et al or Kettunen et al metabolomic studies.
-The arrows denote causal direction and the values on the arrows denote the causal effect estimate.
-Only those relationships are shown for which the $FDR < 0.05$.
-
-\newpage
-
 ## References

### 2017-07-29 docs/manuscript.tex
updated manuscript

@@
 Gibran Hemani\(^1\), Jack Bowden\(^1\), Philip Haycock\(^1\), Jie
-Zheng\(^1\), Oliver Davis\(^1\), Peter Flach, Tom Gaunt\(^{1,*}\),
+Zheng\(^1\), Oliver Davis\(^1\), Peter Flach\(^2\), Tom Gaunt\(^{1,*}\),
 George Davey Smith\(^{1,*}\)
@@
   School of Social and Community Medicine, Bristol, UK
+\item
+  Intelligent Systems Laboratory, Department of Computer Science,
+  University of Bristol
 \end{enumerate}
@@
 causal effects amongst 2407 phenotypes, generating a working draft of
-the causal map of the human phenome. Almost 90\% of causal estimates
-indicated some level of horizontal pleiotropy. All results are organised
-into a publicly available graph database (\url{http://eve.mrbase.org}).
+the causal map of the human phenome comprising the evaluation of
+causality between over a million pairs of traits. Almost 90\% of causal
+estimates indicated some level of horizontal pleiotropy. The causal
+estimates are organised into a publicly available graph database
+(\url{http://eve.mrbase.org}).
 
@@
 significance for other traits (and therefore be present in curated
-catalogues). At its most simple implementation, the regression of the
-SNP-exposure effect sizes against the SNP-outcome effect sizes, with
-greater weight afforded to those SNPs with smaller SNP-outcome standard
-errors, provides the estimate of the causal effect of the exposure on
-the outcome. This is known as the inverse-variance weighted (IVW)
-method.
+catalogues).
+
+We use the term \emph{summary-set} to refer to the minimum data required
+to perform 2SMR analysis - four columns comprising the SNP-exposure and
+SNP-outcome effects and standard errors, and rows corresponding to each
+SNP that is used as an instrument for the exposure.
+
+At its most simple implementation, the regression of the SNP-exposure
+effect sizes against the SNP-outcome effect sizes, with greater weight
+afforded to those SNPs with smaller SNP-outcome standard errors,
+provides the estimate of the causal effect of the exposure on the
+outcome. This is known as the inverse-variance weighted (IVW) method.
 
@@
 (8,12,25), based on the notion that outlying instruments are likely to
-be exhibiting horizontal pleiotropy. Conversely, we primarily developed
-Steiger filtering to identify those instruments that are likely to be
-arising due to reverse cause.
-
-The Steiger test is applied to each variant in turn and we exclude any
-\(g_{A}\) for which \(cor(g_{i}, A)^2 > cor(g_{i}, B)^2\), indicating
-that it is unlikely to primarily associate with \(B\) relative to \(A\).
-Similarly, for SNPs that influence confounders of \(A\) and \(B\) or
-exhibit horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
-\(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
-SNP being excluded because the Steiger Z-test is less likely to be
-significant. Hence we also exclude any \(g_{A}\) for which the Steiger
-Z-test has \(p > 0.05\).
-
-To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
-F-statistic from the reported p-value and sample size and then
-\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
-estimate the variance of the underlying liability explained by the SNP,
-\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
-\(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the log odds ratio and
-\(p\) is the allele frequency of the SNP in the population (26). \(p\)
-can be estimated using the allele frequency of the SNP in an ascertained
-sample by deriving the \(2 \times 2\) contingency table from the odds
-ratio \(e^\beta\), allele frequency in the ascertained sample
-\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\). We
-exclude any SNPs that do not have the correct causal direction for the
-hypothesis, and do not have Steiger test \(p < 0.05\).
+be exhibiting horizontal pleiotropy. A potential drawback of this
+approach is that the outlier SNPs in a summary-set might be the only
+reliable ones. This could arise, for example, for the \emph{CRP} variant
+instrumenting C-reactive protein levels (19), or the \emph{SLC2A9}
+variant influencing serum urate levels (26).
+
+To avoid this, we primarily developed Steiger filtering to identify
+those instruments that are likely to be arising due to reverse cause,
+but it also has the potential to detect SNPs that are exhibiting
+horizontal pleiotropy or primarily influencing confounders. Further
+details are provided in the Online Methods.
 
@@
 inputs. Our objective is to select the expert most likely to be correct
-for a specific MR analysis. A overview of the approach is shown in
+for a specific MR analysis. An overview of the approach is shown in
 Figure 2.
@@
 function', whose purpose is to decide which expert to use for a specific
-MR analysis, given the parameter space that is occupied by that dataset.
-The metrics are a mixture of regression diagnostics and are described in
-Supplementary table 3.
+MR analysis, given the parameter space that is occupied by that
+summary-set. The metrics are a mixture of regression diagnostics and are
+described in Supplementary table 3.
 
@@
 causal effect is known, and to this end we generated a large number of
-simulated datasets. We trained the gating function using random forest
-learning algorithms that seek to identify sectors of the parameter space
-that are most likely to return accurate estimates for a particular
-expert. Figure 2a illustrates how the gating function is trained using
-simulated data. New datasets can then be applied to the trained mixture
-of experts (MoE) model (Figure 2b). We call this implementation MR-MoE
-1.0.
-
-It should be noted that in the original hierarchical mixture of experts
-approach of (27) the gating function and the experts share the same
-input space and are trained simultaneously using
-expectation-maximisation. In our approach the gating function is defined
-over a separate input space consisting of the parameter space of the
-experts, and is trained separately in a supervised setting using
-simulated data. As such our MR-MoE approach fits the framework of
-meta-learning (28--30) where machine learning methods are applied at the
-`meta-level' to analyse the results of machine learning experiments at
-the `base-level', in order to be able to understand and model the
-capabilities of each expert and recommend which expert should be applied
-to a given problem. From this perspective the 28 MR methods are the
-base-models, the gating function is the meta-model, and its
-meta-features are the parameters of the datasets.
-
-\textbf{Training and testing simulations}
-
-The MoE is trained using datasets generated from simulations (Figure
-2a). A \emph{dataset} is the minimum data required to perform 2SMR
-analysis - four columns comprising the SNP-exposure and SNP-outcome
-effects and standard errors, and rows corresponding to each SNP that is
-used as an instrument for the exposure. This \emph{dataset} can be fed
-into any of the 28 experts to obtain MR causal effects. The simulations
-used to generate these datasets seek to cover a range of pleiotropic
-scenarios, including where some proportion of SNPs exhibit directional
-or balanced horizontal pleiotropy, or where SNPs influence confounding
-variables.
-
-We simulate two individual level datasets for which there are \(N_x\)
-and \(N_y\) samples, and \(M\) SNPs, where each SNP has effect allele
-frequency of \(p_m \sim U(0.05, 0.95)\). These datasets are used to
-obtain the SNP effects for the exposure trait \(x\) and the outcome
-trait \(y\), respectively, using the following sampling criteria:
-
-\[
-\begin{aligned}
-N_x & = \{20000, ..., 500000\} \\
-N_y & = \{20000, ..., 500000\} \\
-K & = \{0, ..., 10\} \\
-M_x & = \{1, ..., 200\} \\
-M_y & = \{1, ..., 200\} \\
-M_{u_k} & = \{5,..., 30\} \\
-\end{aligned}
-\]
-
-The \(M = M_x + M_y + \sum{M_{u_k}}\) SNPs can influence \(x\) directly,
-\(y\) directly, or some number of confounders \(u_{k}\) directly.
-Phenotypes for \(x\) and \(y\) are constructed using
-
-\[
-x = \sum^{M_x}_{i}{\beta_{gx,x,i}g_{x,i}} + \sum^{M_y}_{j}{\beta_{gy,x,j}g_{y,j}} + \sum^{K}_{k}{\beta_{ux,k} u_{k}} + e_{x}
-\]
-
-where \(\beta_{gx,x}\) is the vector of effects of each of the \(M_x\)
-SNPs that influence \(x\) primarily, \(\beta_{gy,x}\) is the vector of
-effects for the \(M_y\) SNPs on \(x\), where the \(M_y\) SNPs influence
-\(y\) primarily but exhibit horizontal pleiotropic effects on \(x\). We
-allow some proportion of these effects to be 0. \(\beta_{ux}\) is the
-vector of effects of each of the \(K\) confounders on \(x\). Each
-\(u_{k}\) variable is constructed using
-
-\[
-u = \sum^{M_u}_{l}{\beta_{gu,l}g_{l}} + e_{l}
-\]
-
-and finally \(y\) is constructed using
-
-\[
-y = \beta_{x,y}x + \sum^{M_y}_{i}{\beta_{gy,y,j}g_{y,j}} + \sum^{M_x}_{j}{\beta_{gx,y,i}g_{x,i}} + \sum^{K}_{k}{\beta_{uy,k} u_{k}} + e_{y}
-\]
-
-where \(\beta_{x,y}\) is the causal effect of \(x\) on \(y\). We sample
-the distribution of direct SNP effects using
-
-\[
-\begin{aligned}
-\beta_{gx,x,i} & \sim N(0, \sigma^2_{gx,x}) \\
-\sigma^2_{gx,x,i} & \sim U(0.01, 0.1) \\
-\beta_{gy,y,j} ~ N(0, \sigma^2_{gy,y}) \\
-\sigma^2_{gy,y,j} & \sim U(0.01, 0.1) \\
-\end{aligned}
-\]
-
-Some proportion \(floor(MS_{x}/M)\) of \(g_x\) SNPs and
-\(floor(MS_{y}/M)\) of \(g_y\) SNPs, where \(s_x\) and
-\(s_y \sim U(0,1)\), exhibit horizontal pleiotropy with effects sampled
-using
-
-\[
-\begin{aligned}
-\beta_{gx,y,i*} & \sim N(\mu_{gx,y}, \sigma^2_{gx,y})  \\
-\mu_{gx,y,i*} & \sim U(-0.005, 0.005) \\
-\sigma^2_{gx,y,i*} & \sim U(0.001, 0.01) \\
-\beta_{gy,x,j*} & \sim N(\mu_{gy,x}, \sigma^2_{gy,x}) \\
-\mu_{gy,x,j*} & \sim U(-0.005, 0.005) \\
-\sigma^2_{gy,x,j*} & \sim U(0.001, 0.01) \\
-\end{aligned}
-\]
-
-The genetic influences on each of the confounders are sampled using
-
-\[
-\begin{aligned}
-\beta_{gu,u,l} & \sim N(0, \sigma^2_{gu,u}) \\
-\sigma^2_{gu,u,l} & \sim U(0.01, 0.1) \\
-\end{aligned}
-\]
-
-The influence of each confounder on \(x\) and \(y\) is obtained using
-
-\[
-\begin{aligned}
-\beta_{u,x} & \sim N(0, \sigma^{2}_{u,x}) \\
-\beta_{u,y} & \sim N(0, \sigma^{2}_{u,y}) \\
-\end{aligned}
-\]
-
-Finally, 20\% of the simulations have a null effect of
-\(\beta_{x,y} = 0\), while the other remaining 80\% have a true effect
-sampled from
-
-\[
-\begin{aligned}.
-\beta_{x,y} & \sim N(0, \sigma^2_{x,y}) \\
-\sigma^2_{x,y} & \sim U(0.001, 0.1) \\
-\end{aligned}
-\]
-
-For each simulation we used linear regression to estimate the genetic
-effect of each SNP \(M\) on \(x\) in sample 1, and each SNP \(M\) on
-\(y\) in sample 2. We then perform MR analysis in both directions,
-mimicking GWAS by retaining SNPs that have \(p < 5e-8\) in sample 1 to
-perform MR of \(x\) on \(y\) (the true causal direction for non-null
-simulations), and retaining SNPs that have \(p < 5e-8\) in sample 2 to
-perform MR of \(y\) on \(x\) (the reverse causal direction for non-null
-simulations). We treat the summary data (effect sizes and standard
-errors) used for estimating \(x \rightarrow y\) the summary data used
-for estimating \(y \rightarrow x\) as two separate datasets Hence, for
-each simulation two datasets are generated which are analysed to produce
-28 MR estimates each. We performed 100,000 simulations using these
-parameters, resulting in 200,000 datasets.
-
-\textbf{Optimisation function}
-
-We aim to maximise statistical power for datasets where
-\(\beta_{x,y} \neq 0\) and minimise false discovery rates for datasets
-where \(\beta_{x,y} = 0\). To train random forest decision trees to
-predict performance for a particular method
-\(h(O_{w,d}, \textbf{z}_{d})\) is generated where the training set of
-input metrics for dataset \(d\) is \(\textbf{z}_{d}\) and the response
-(optimisation function) is
-
-\[
-    O_{w,d} = 
-\begin{cases}
-    1,   & \text{if } \beta_{x,y} \neq 0 \text{ and } p_{m,d} < 0.01\\
-    1,   & \text{if } \beta_{x,y} = 0 \text{ and } p_{m,d} > 0.1 \\
-    0,   & \text{otherwise}
-\end{cases}
-\]
-
-where \(p_{m,d}\) is the p-value for method \(m\) on dataset \(d\).
-
-\textbf{Strategy}
-
-For each training dataset we record a set of 53 metrics
-\(\textbf{z}_{d}\) (Supplementary table 3), and an outcome \(O_{w,d}\),
-which is a measure of how well that method performed for each particular
-simulated dataset. For each of our 28 methods, we need to create a model
-that predicts the performance of the method based on metrics generated
-from a dataset. To do this, for each method we train a random decision
-forest to predict that method's performance using the dataset's metrics.
-The random forest approach is well suited to this problem because there
-are likely to be non-continuous combinations of different metrics that
-improve on prediction over, for example, a simple linear model that does
-not learn about interactions.
-
-As a simple hypothetical example - if a dataset has a single outlier but
-is otherwise exhibiting no heterogeneity then the following methods
-could arguably perform well:
-
-\begin{itemize}
-\tightlist
-\item
-  an IVW fixed effects analysis with the outlier removed, should the
-  Steiger test be able to detect the outlier
-\item
-  a median based approach
-\item
-  a mode based approach
-\end{itemize}
-
-deciding between these methods requires finding, in general, which will
-minimise false discovery rates and maximise true positive rates for that
-particular scenario. In this example the IVW with Steiger filtering
-would likely be the clear winner because of its superior statistical
-power. Countless other scenarios could arise. For example if 100\% of
-instruments are invalid but the InSIDE assumptio is met then the
-MR-Egger is likely to be most effective; if 40\% of instruments are
-invalid then a median-based approach is likely most effective; and if
-80\% of instruments are invalid then a mode-based approach is likely
-most effective.
-
-Having generated random forest decision trees for each of the 28 methods
-using 133,000 of the simulations, we then applied them to the remaining
-67,000 datasets to predict which method would have the highest
-performance for each of the remaining datasets. Finally we compare the
-performance of the method selected by the MoE against all remaining
-methods. The default settings for the \texttt{randomForest} package in R
-(31) were used to train the models. MR-MoE 1.0 is implemented in the
-TwoSampleMR R package available at
+simulated summary-sets. We trained the gating function using random
+forest learning algorithms that seek to identify sectors of the
+parameter space that are most likely to return accurate estimates for a
+particular expert. Figure 2a illustrates how the gating function is
+trained using simulated data. New summary-sets can then be applied to
+the trained mixture of experts (MoE) model (Figure 2b). We call this
+implementation MR-MoE 1.0. Full details are provided in the Online
+Methods. MR-MoE 1.0 is implemented in the TwoSampleMR R package
+available at
 \href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
@@
 
-The set of MR estimates obtained from this analysis are recorded in a
-Neo4j graph database. Because each association has up to 28 different
-estimates, for simplicity we distill this down to a single `best
-estimate' for each association using the following rules:
+The set of MR estimates obtained from this analysis, and the
+summary-sets used to generate them, are deposited in a Neo4j graph
+database. Here nodes representations exist for traits and SNPs; and
+relationship representations exist for SNP-trait associations and
+trait-trait MR estimates. Because each trait-trait association has up to
+28 different MR estimates, for simplicity we also distill this down to a
+third relationship type comprising a single `best estimate' for each
+trait-trait association using the following rules:
 
@@
 of the Steiger filtering approach for improving instrument selection
-using 100,000 simulated datasets comprising both null and non-null
-causal models. For each dataset, instruments were selected based on the
-tophits strategy and the Steiger filtering strategy, and MR was
+using 100,000 simulated summary-sets comprising both null and non-null
+causal models. For each summary-set, instruments were selected based on
+the tophits strategy and the Steiger filtering strategy, and MR was
 performed using 14 different methods based on instruments selected from
@@
 model would be able to predict the most appropriate of the 28 MR methods
-and instrument selection strategies to apply to a particular dataset
+and instrument selection strategies to apply to a particular summary-set
 based on its characteristics (Figure 2).
@@
 shown in Supplementary table 2. The prediction \(R^2\) of whether a
-dataset's status was truly null (\(\beta_{xy}=0\)) or non-null
-(\(\beta_{xy} \neq 0\)) against the method's prediction of the dataset's
-status, ranged between 0.04 and 0.24. The method performance prediction
-was most effective for the Egger random effects model with Steiger
-filtering. The dataset characteristics with the most importance for each
-of the predictors differed substantially between each dataset, as well
-as the frequency for which each of the methods was selected in the
-testing datasets. The FDR of each method when chosen, compared to their
-averages across all datasets, reduced; and likewise the true positive
-rates of each method when chosen increased compared to their
-simulation-wide averages.
+summary-set's status was truly null (\(\beta_{xy}=0\)) or non-null
+(\(\beta_{xy} \neq 0\)) against the method's prediction of the
+summary-set's status, ranged between 0.04 and 0.24. The method
+performance prediction was most effective for the Egger random effects
+model with Steiger filtering. The summary-set characteristics with the
+most importance for each of the predictors differed substantially
+between each summary-set, as well as the frequency for which each of the
+methods was selected in the testing summary-sets. The FDR of each method
+when chosen, compared to their averages across all summary-sets,
+reduced; and likewise the true positive rates of each method when chosen
+increased compared to their simulation-wide averages.
 
@@
 Figures 3a and 3b show that the MoE approach had the best general
-performance, obtaining an AUC of 0.84 in terms of classifying the
-simulations as being null or non-null. Notably, the next best methods
-were median and mode based estimators using Steiger filtering. Under the
-assumption of widespread pleiotropy it is clear that all methods suffer
-from high false discovery rates.
+performance. Estimating the area under the receiver operator curve gave
+0.84 for the MoE approach in terms of classifying the simulations as
+being null or non-null. Notably, the next best methods were median and
+mode based estimators using Steiger filtering. Under the assumption of
+widespread and diverse pleiotropic effects it is clear that all methods
+suffer from high false discovery rates (Figure 3a).
 
@@
 We applied our analysis using summary data for 2407 phenotypes,
-including 149 complex traits and diseases, 575 metabolites (32,33) and
-1683 plasma protein levels (34). For the protein levels only the
+including 149 complex traits and diseases, 575 metabolites (28,29) and
+1683 plasma protein levels (30). For the protein levels only the
 instruments were available, so they could only be evaluated as exposure
@@
 As an example of the database we performed a look up of associations
-where LDL cholesterol as measured in the GLGC consortium (35) had a
+where LDL cholesterol as measured in the GLGC consortium (31) had a
 false discovery rate of 0.05 using the MoE approach. This returned 287
@@
 are dominated by lipid fractions. Filtering to exclude the metabolomic
-studies ((32,33)) returned 27 associations with 23 traits (Figure 4a).
+studies ((28,29)) returned 27 associations with 23 traits (Figure 4a).
 There was a strong association with coronary heart disease (0.41 log(OR)
@@
 We also performed a look up of associations that causally influence
-years of schooling (36), or that years of schooling itself causally
+years of schooling (32), or that years of schooling itself causally
 influences. Using a false-discovery rate of 0.05, 45 traits were
@@
 plausible, for example the influence of years of schooling on lower risk
-of Alzheimer's disease has recently been examined (37). However it is
+of Alzheimer's disease has recently been examined (33). However it is
 clear that there are serious limitations in relying on MR as a panacea
@@
 method, selecting the one which is most likely to maximise power whilst
-minimising FDR for a specific dataset. The mixture of experts approach
-that we present here is trained using random forest decision trees
-applied to extensive and diverse simulations, and we demonstrate that it
-makes substantial performance improvements over any other single method
-in the presence or absence of extensive horizontal pleiotropy. As such,
-our method contributes to the field of meta-learning but also to the
-rapidly growing field of automatic model selection and hyper-parameter
-optimization in machine learning (38).
+minimising FDR for a specific summary-set. The mixture of experts
+approach that we present here is trained using random forest decision
+trees applied to extensive and diverse simulations, and we demonstrate
+that it makes substantial performance improvements over any other single
+method in the presence or absence of extensive horizontal pleiotropy. As
+such, our method contributes to the field of meta-learning but also to
+the rapidly growing field of automatic model selection and
+hyper-parameter optimization in machine learning (34).
 
@@
 
-In addition to correctly handling horizontal pleiotropy, there are many
-other limitations that prevent MR from being a panacea for causal
-inference. Many of the associations are biologically impossible, for
-example where early life phenotypes appear to be influenced by later
-stage phenotypes. Though these associations can be informative, their
-interpretations as causal relationships are far from clear. Similarly,
-often disease traits appear to causally relate to other phenotypes, but
-GWAS is typically performed on the liability scale, hence the causal
-estimate reflects not the presence or absence of disease, but the
-underlying risk of disease. Again, interpretation of such associations
-can be problematic.
+One advantage to generating a method that consolidates across many
+methods is that it can prevent cherry picking the most appealing method
+based on their results. But in addition to correctly handling horizontal
+pleiotropy, there are many other limitations that prevent MR from being
+a panacea for causal inference. Many of the associations are
+biologically impossible, for example where early life phenotypes appear
+to be influenced by later stage phenotypes. Though these associations
+can be informative, their interpretations as causal relationships are
+far from clear. Similarly, often disease traits appear to causally
+relate to other phenotypes, but GWAS is typically performed on the
+liability scale, hence the causal estimate reflects not the presence or
+absence of disease, but the underlying risk of disease. Again,
+interpretation of such associations can be problematic.
 
@@
 delineate between models of reverse cause, pleiotropy and multiple
-causal variants in the same region (39,40). Separating vertical from
+causal variants in the same region (35,36). Separating vertical from
 horizontal pleiotropy with a single instrument is not yet possible in
 the two-sample MR framework though the use of mediation-based approaches
-may be informative in the one-sample setting (7,41). Other problems can
+may be informative in the one-sample setting (7,37). Other problems can
 also manifest, for example frailty effects could induce associations for
-late-onset traits (42); genetic variants could strongly relate to
+late-onset traits (38); genetic variants could strongly relate to
 several phenotypes making it difficult to ascertain which of them is the
-causal exposure (43); and the measured feature for which genetic
+causal exposure (39); and the measured feature for which genetic
 associations are known could relate in complicated ways to the
@@
 We also encounter a potential new problem in using machine learning
-approaches to infer the correct model for a particular dataset, namely,
-that the data is choosing the model. This is liable to contribute to
-elevated type 1 error rates, even though we have attempted to separate
-the information used for optimisation from the information used to
-predict performance. The extent to which a machine based method selector
-suffers from this more than a human manually inferring the result most
-likely to be true is not clear. Though MR-MoE does exhibit higher type 1
-error rates than are desirable, they still remain amongst the lowest
-compared to all other MR methods that do not suffer from this issue.
-
-The field of MR is evolving. The advent of GWAS databases (9,44) and
+approaches to infer the correct model for a particular summary-set,
+namely, that the data is choosing the model. This is liable to
+contribute to elevated type 1 error rates, even though we have attempted
+to separate the information used for optimisation from the information
+used to predict performance. The extent to which a machine based method
+selector suffers from this more than a human manually inferring the
+result most likely to be true is not clear. Though MR-MoE does exhibit
+higher type 1 error rates than are desirable, they still remain amongst
+the lowest compared to all other MR methods that do not suffer from this
+issue.
+
+The field of MR is evolving. The advent of GWAS databases (9,40) and
 automated 2SMR (11) has trivialised the analytical aspect of
@@
 endeavours could be transformative in biological research.
+
+\subsection{Acknowledgements}\label{acknowledgements}
+
+This work was supported by the UK Medical Research Council
+(MC\_UU\_12013/1).
 
@@
 
+\subsection{Online methods}\label{online-methods}
+
+\subsubsection{Steiger filtering}\label{steiger-filtering-1}
+
+An approach to inferring the causal direction between phenotypes
+recently developed (7) uses the following basic premise. If trait \(A\)
+causes trait \(B\) then \(cor(g_{i}, A)^2 > cor(g_{i}, B)^2\) because
+the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This will be
+true under most circumstances, but some parameters of measurement error
+or unmeasured confounding can lead to the inequality reversing
+direction, as has been explored previously (7).
+
+The Steiger test is applied to each variant in turn and we exclude any
+\(g_{A}\) for which \(cor(g_{i}, A)^2 > cor(g_{i}, B)^2\), indicating
+that it is unlikely to primarily associate with \(B\) relative to \(A\).
+Similarly, for SNPs that influence confounders of \(A\) and \(B\) or
+exhibit horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
+\(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
+SNP being excluded because the Steiger Z-test is less likely to be
+significant. Hence we also exclude any \(g_{A}\) for which the
+\(cor(g, x)\) and \(cor(g, y)\) are not significantly different at an
+arbitrary threshold of \(p > 0.05\).
+
+To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
+F-statistic from the reported p-value and sample size and then
+\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
+estimate the variance of the underlying liability explained by the SNP,
+\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
+\(V_a = 2\beta^2f(1-f)\), where \(\beta\) is the log odds ratio and
+\(f\) is the allele frequency of the SNP in the population (41). \(f\)
+can be estimated using the allele frequency of the SNP in an ascertained
+sample by deriving the \(2 \times 2\) contingency table from the odds
+ratio \(e^\beta\), allele frequency in the ascertained sample
+\(f_{cc}\), and number of cases \(N_1\) and controls \(N_0\). A Z-score
+to test the hypothesis that \(cor(g, x)\) and \(cor(g, y)\) are
+significantly different can be generated using the Steiger test for
+correlated correlations.
+
+\subsubsection{Mixture of experts implementation for
+MR}\label{mixture-of-experts-implementation-for-mr}
+
+The mixture of experts approach seeks to use simulations to learn the
+circumstances in which each of the 28 MR methods is likely to perform
+most accurately. Having trained the model on simulations, new
+summary-sets can be applied to obtain causal estimates from the method
+that is predicted to be the most reliable for that summary-set.
+
+It should be noted that in the original hierarchical mixture of experts
+approach of (27) the gating function and the experts share the same
+input space and are trained simultaneously using
+expectation-maximisation. In our approach the gating function is defined
+over a separate input space consisting of the parameter space of the
+experts, and is trained separately in a supervised setting using
+simulated data. As such our MR-MoE approach fits the framework of
+meta-learning (42--44) where machine learning methods are applied at the
+`meta-level' to analyse the results of machine learning experiments at
+the `base-level', in order to be able to understand and model the
+capabilities of each expert and recommend which expert should be applied
+to a given problem. From this perspective the 28 MR methods are the
+base-models, the gating function is the meta-model, and its
+meta-features are the parameters of the summary-sets.
+
+\textbf{Training and testing simulations}
+
+The MoE is trained using summary-sets generated from simulations (Figure
+2a). Each \emph{summary-set} can be fed into any of the 28 experts to
+obtain MR causal effects. The simulations used to generate these
+summary-sets seek to cover a range of pleiotropic scenarios, including
+where some proportion of SNPs exhibit directional or balanced horizontal
+pleiotropy, or where SNPs influence confounding variables.
+
+We simulate two individual level datasets for which there are \(N_x\)
+and \(N_y\) samples, and \(M\) SNPs, where each SNP has effect allele
+frequency of \(f_m \sim U(0.05, 0.95)\). These datasets are used to
+obtain the SNP effects for the exposure trait \(x\) and the outcome
+trait \(y\), respectively, to create summary-sets, using the following
+sampling criteria:
+
+\[
+\begin{aligned}
+N_x & = \{20000, ..., 500000\} \\
+N_y & = \{20000, ..., 500000\} \\
+K & = \{0, ..., 10\} \\
+M_x & = \{1, ..., 200\} \\
+M_y & = \{1, ..., 200\} \\
+M_{u_k} & = \{5,..., 30\} \\
+\end{aligned}
+\]
+
+The \(M = M_x + M_y + \sum{M_{u_k}}\) SNPs can influence \(x\) directly,
+\(y\) directly, or some number of confounders \(u_{k}\) directly.
+Phenotypes for \(x\) and \(y\) are constructed using
+
+\[
+x = \sum^{M_x}_{i}{\beta_{gx,x,i}g_{x,i}} + \sum^{M_y}_{j}{\beta_{gy,x,j}g_{y,j}} + \sum^{K}_{k}{\beta_{ux,k} u_{k}} + e_{x}
+\]
+
+where \(\beta_{gx,x}\) is the vector of effects of each of the \(M_x\)
+SNPs that influence \(x\) primarily, \(\beta_{gy,x}\) is the vector of
+effects for the \(M_y\) SNPs on \(x\), where the \(M_y\) SNPs influence
+\(y\) primarily but exhibit horizontal pleiotropic effects on \(x\). We
+allow some proportion of these effects to be 0. \(\beta_{ux}\) is the
+vector of effects of each of the \(K\) confounders on \(x\). Each
+\(u_{k}\) variable is constructed using
+
+\[
+u = \sum^{M_u}_{l}{\beta_{gu,l}g_{l}} + e_{l}
+\]
+
+and finally \(y\) is constructed using
+
+\[
+y = \beta_{x,y}x + \sum^{M_y}_{i}{\beta_{gy,y,j}g_{y,j}} + \sum^{M_x}_{j}{\beta_{gx,y,i}g_{x,i}} + \sum^{K}_{k}{\beta_{uy,k} u_{k}} + e_{y}
+\]
+
+where \(\beta_{x,y}\) is the causal effect of \(x\) on \(y\). We sample
+the distribution of direct SNP effects using
+
+\[
+\begin{aligned}
+\beta_{gx,x,i} & \sim N(0, \sigma^2_{gx,x}) \\
+\sigma^2_{gx,x,i} & \sim U(0.01, 0.1) \\
+\beta_{gy,y,j} ~ N(0, \sigma^2_{gy,y}) \\
+\sigma^2_{gy,y,j} & \sim U(0.01, 0.1) \\
+\end{aligned}
+\]
+
+Some proportion \(floor(MS_{x}/M)\) of \(g_x\) SNPs and
+\(floor(MS_{y}/M)\) of \(g_y\) SNPs, where \(s_x\) and
+\(s_y \sim U(0,1)\), exhibit horizontal pleiotropy with effects sampled
+using
+
+\[
+\begin{aligned}
+\beta_{gx,y,i*} & \sim N(\mu_{gx,y}, \sigma^2_{gx,y})  \\
+\mu_{gx,y,i*} & \sim U(-0.005, 0.005) \\
+\sigma^2_{gx,y,i*} & \sim U(0.001, 0.01) \\
+\beta_{gy,x,j*} & \sim N(\mu_{gy,x}, \sigma^2_{gy,x}) \\
+\mu_{gy,x,j*} & \sim U(-0.005, 0.005) \\
+\sigma^2_{gy,x,j*} & \sim U(0.001, 0.01) \\
+\end{aligned}
+\]
+
+The genetic influences on each of the confounders are sampled using
+
+\[
+\begin{aligned}
+\beta_{gu,u,l} & \sim N(0, \sigma^2_{gu,u}) \\
+\sigma^2_{gu,u,l} & \sim U(0.01, 0.1) \\
+\end{aligned}
+\]
+
+The influence of each confounder on \(x\) and \(y\) is obtained using
+
+\[
+\begin{aligned}
+\beta_{u,x} & \sim N(0, \sigma^{2}_{u,x}) \\
+\beta_{u,y} & \sim N(0, \sigma^{2}_{u,y}) \\
+\end{aligned}
+\]
+
+Finally, 20\% of the simulations have a null effect of
+\(\beta_{x,y} = 0\), while the other remaining 80\% have a true effect
+sampled from
+
+\[
+\begin{aligned}.
+\beta_{x,y} & \sim N(0, \sigma^2_{x,y}) \\
+\sigma^2_{x,y} & \sim U(0.001, 0.1) \\
+\end{aligned}
+\]
+
+For each simulation we used linear regression to estimate the genetic
+effect of each SNP \(M\) on \(x\) in sample 1, and each SNP \(M\) on
+\(y\) in sample 2. We then perform MR analysis in both directions,
+mimicking GWAS by retaining SNPs that have \(p < 5e-8\) in sample 1 to
+perform MR of \(x\) on \(y\) (the true causal direction for non-null
+simulations), and retaining SNPs that have \(p < 5e-8\) in sample 2 to
+perform MR of \(y\) on \(x\) (the reverse causal direction for non-null
+simulations). We treat the summary data (effect sizes and standard
+errors) used for estimating \(x \rightarrow y\) the summary data used
+for estimating \(y \rightarrow x\) as two separate summary-sets. Hence,
+for each simulation two summary-sets are generated which are analysed to
+produce 28 MR estimates each. We performed 100,000 simulations using
+these parameters, resulting in 200,000 summary-sets.
+
+\textbf{Optimisation function}
+
+We aim to maximise statistical power for summary-sets where
+\(\beta_{x,y} \neq 0\) and minimise false discovery rates for
+summary-sets where \(\beta_{x,y} = 0\). To train random forest decision
+trees to predict performance for a particular method
+\(h(O_{w,d}, \textbf{z}_{d})\) is generated where the training set of
+input metrics for summary-set \(d\) is \(\textbf{z}_{d}\) and the
+response (optimisation function) is
+
+\[
+    O_{w,d} = 
+\begin{cases}
+    1,   & \text{if } \beta_{x,y} \neq 0 \text{ and } p_{m,d} < 0.01\\
+    1,   & \text{if } \beta_{x,y} = 0 \text{ and } p_{m,d} > 0.1 \\
+    0,   & \text{otherwise}
+\end{cases}
+\]
+
+where \(p_{m,d}\) is the p-value for method \(m\) on summary-set \(d\).
+
+\textbf{Strategy}
+
+For each training summary-set we record a set of 53 metrics
+\(\textbf{z}_{d}\) (Supplementary table 3), and an outcome \(O_{w,d}\),
+which is a measure of how well that method performed for each particular
+simulated summary-set. For each of our 28 methods, we need to create a
+model that predicts the performance of the method based on metrics
+generated from a summary-set. To do this, for each method we train a
+random decision forest to predict that method's performance using the
+summary-set's metrics. The random forest approach is well suited to this
+problem because there are likely to be non-continuous combinations of
+different metrics that improve on prediction over, for example, a simple
+linear model that does not learn about interactions.
+
+As a simple hypothetical example - if a summary-set exhibits a single
+outlier but is otherwise exhibiting no heterogeneity then the following
+methods could arguably perform well:
+
+\begin{itemize}
+\tightlist
+\item
+  an IVW fixed effects analysis with the outlier removed, should the
+  Steiger test be able to detect the outlier
+\item
+  a median based approach
+\item
+  a mode based approach
+\end{itemize}
+
+deciding between these methods requires finding, in general, which will
+minimise false discovery rates and maximise true positive rates for that
+particular scenario. In this example the IVW with Steiger filtering
+would likely be the clear winner because of its superior statistical
+power. Countless other scenarios could arise. For example if 100\% of
+instruments are invalid but the InSIDE assumptio is met then the
+MR-Egger is likely to be most effective; if 40\% of instruments are
+invalid then a median-based approach is likely most effective; and if
+80\% of instruments are invalid then a mode-based approach is likely
+most effective.
+
+Having generated random forest decision trees for each of the 28 methods
+using 133,000 of the simulations, we then applied them to the remaining
+67,000 summary-sets to predict which method would have the highest
+performance for each of the remaining summary-sets. Finally we compare
+the performance of the method selected by the MoE against all remaining
+methods. The default settings for the \texttt{randomForest} package in R
+(45) were used to train the models. MR-MoE 1.0 is implemented in the
+TwoSampleMR R package available at
+\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
+(11).
+
 \subsection{References}\label{references}
@@
 
-\hypertarget{ref-Lee2013c}{}
-26. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
-association studies: quantification of power and genomic prediction
-accuracy. PLoS One. 2013;8(8):e71494.
+\hypertarget{ref-white2016plasma}{}
+26. White J, Sofat R, Hemani G, Shah T, Engmann J, Dale C, et al. Plasma
+urate concentration and risk of coronary heart disease: a Mendelian
+randomisation analysis. The Lancet Diabetes \& Endocrinology. 2016;
 
@@
 
-\hypertarget{ref-brazdil2008metalearning}{}
-28. Brazdil P, Carrier CG, Soares C, Vilalta R. Metalearning:
-Applications to data mining. Springer Science \& Business Media; 2008.
-
-\hypertarget{ref-smith2009cross}{}
-29. Smith-Miles KA. Cross-disciplinary perspectives on meta-learning for
-algorithm selection. ACM Computing Surveys (CSUR). 2009;41(1):6.
-
-\hypertarget{ref-lemke2015metalearning}{}
-30. Lemke C, Budka M, Gabrys B. Metalearning: a survey of trends and
-technologies. Artificial intelligence review. 2015;44(1):117--30.
-
-\hypertarget{ref-Liaw2002}{}
-31. Liaw A, Wiener M. Classification and Regression by randomForest. R
-News {[}Internet{]}. 2002;2(3):18--22. Available from:
-\url{http://cran.r-project.org/doc/Rnews/}
-
 \hypertarget{ref-Shin2014}{}
-32. Shin S-Y, Fauman EB, Petersen A-K, Krumsiek J, Santos R, Huang J, et
+28. Shin S-Y, Fauman EB, Petersen A-K, Krumsiek J, Santos R, Huang J, et
 al. An atlas of genetic influences on human blood metabolites. Nature
@@
 \hypertarget{ref-Kettunen2016}{}
-33. Kettunen J, Demirkan A, Wurtz P, Draisma HHM, Haller T, Rawal R, et
+29. Kettunen J, Demirkan A, Wurtz P, Draisma HHM, Haller T, Rawal R, et
 al. Genome-wide study for circulating metabolites identifies 62 loci and
@@
 \hypertarget{ref-Sun2017}{}
-34. Sun BB, Maranville JC, Peters JE, Stacey D, Staley JR, Blackshaw J,
+30. Sun BB, Maranville JC, Peters JE, Stacey D, Staley JR, Blackshaw J,
 et al. Consequences Of Natural Perturbations In The Human Plasma
@@
 \hypertarget{ref-Willer2013}{}
-35. Willer CJ, Schmidt EM, Sengupta S, Peloso GM, Gustafsson S, Kanoni
+31. Willer CJ, Schmidt EM, Sengupta S, Peloso GM, Gustafsson S, Kanoni
 S, et al. Discovery and refinement of loci associated with lipid levels.
@@
 \hypertarget{ref-Okbay2016}{}
-36. Okbay A, Beauchamp JP, Fontana MA, Lee JJ, Pers TH, Rietveld CA, et
+32. Okbay A, Beauchamp JP, Fontana MA, Lee JJ, Pers TH, Rietveld CA, et
 al. Genome-wide association study identifies 74 loci associated with
@@
 \hypertarget{ref-Anderson2017}{}
-37. Anderson E, Wade KH, Hemani G, Bowden J, Korologou-Linden R, Davey
+33. Anderson E, Wade KH, Hemani G, Bowden J, Korologou-Linden R, Davey
 Smith G, et al. The Causal Effect Of Educational Attainment On
@@
 \hypertarget{ref-kotthoff2016auto}{}
-38. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
+34. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
 2.0: Automatic model selection and hyperparameter optimization in WEKA.
@@
 \hypertarget{ref-Zhu2016}{}
-39. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+35. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
 Integration of summary data from GWAS and eQTL studies predicts complex
@@
 \hypertarget{ref-Richardson2017}{}
-40. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
+36. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
 CL, et al. Causal epigenome-wide association study identifies CpG sites
@@
 \hypertarget{ref-Wang2017}{}
-41. Wang L, Michoel T. Efficient And Accurate Causal Inference With
+37. Wang L, Michoel T. Efficient And Accurate Causal Inference With
 Hidden Confounders From Genome-Transcriptome Variation Data. bioRxiv
@@
 \hypertarget{ref-Noyce2017}{}
-42. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
+38. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
 E, et al. Estimating the causal influence of body mass index on risk of
@@
 \hypertarget{ref-Burgess2014a}{}
-43. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+39. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 \hypertarget{ref-Staley2016}{}
-44. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
+40. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
 al. PhenoScanner: a database of human genotype--phenotype associations.
@@
 
+\hypertarget{ref-Lee2013c}{}
+41. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+association studies: quantification of power and genomic prediction
+accuracy. PLoS One. 2013;8(8):e71494.
+
+\hypertarget{ref-brazdil2008metalearning}{}
+42. Brazdil P, Carrier CG, Soares C, Vilalta R. Metalearning:
+Applications to data mining. Springer Science \& Business Media; 2008.
+
+\hypertarget{ref-smith2009cross}{}
+43. Smith-Miles KA. Cross-disciplinary perspectives on meta-learning for
+algorithm selection. ACM Computing Surveys (CSUR). 2009;41(1):6.
+
+\hypertarget{ref-lemke2015metalearning}{}
+44. Lemke C, Budka M, Gabrys B. Metalearning: a survey of trends and
+technologies. Artificial intelligence review. 2015;44(1):117--30.
+
+\hypertarget{ref-Liaw2002}{}
+45. Liaw A, Wiener M. Classification and Regression by randomForest. R
+News {[}Internet{]}. 2002;2(3):18--22. Available from:
+\url{http://cran.r-project.org/doc/Rnews/}
+
 \end{document}

### 2017-08-02 docs/manuscript.rmd
updated

@@
 
-1. MRC Integrative Epidemiology Unit (IEU) at the University of Bristol, School of Social and Community Medicine, Bristol, UK
+1. MRC Integrative Epidemiology Unit (IEU) at the University of Bristol, Population Health Sciences, Bristol, UK
 2. Intelligent Systems Laboratory, Department of Computer Science, University of Bristol
@@
 MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype.
-Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy, and many methods have been developed in an attempt to address this.
-Here we show that the mechanisms that underlie horizontal pleiotropy are numerous, and that instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow.
-We have developed a mixture of experts machine learning approach (MR-MoE 1.0) that improves on both power and false discovery rates over unselective use of any existing methods.
-Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a working draft of the causal map of the human phenome comprising the evaluation of causality between over a million pairs of traits.
+Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy, and many methods have been developed in an attempt to address this, each assuming a different model of horizontal pleiotropy.
+Here we show that across a range of horizontal pleiotropy models instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow.
+We have integrated these methods into a mixture-of-experts machine learning approach (MR-MoE 1.0) that predicts the most appropriate model to use for any specific causal analysis, improving on both power and false discovery rates.
+Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a working draft of the causal map of the human phenome, comprising the evaluation of causality between over a million pairs of traits.
 Almost 90% of causal estimates indicated some level of horizontal pleiotropy.
@@
 
-Mendelian randomization (MR) [@DaveySmith2003; @DaveySmithHemani2014] exploits genetic pleiotropy to infer the causal relationships between phenotypes.
+Mendelian randomization (MR) [@DaveySmith2003; @DaveySmithHemani2014] exploits vertical pleiotropy to infer the causal relationships between phenotypes.
 Suppose that one trait (the exposure) causally influences another (the outcome).
-If a SNP influences the outcome through the exposure then the SNP is exhibiting vertical pleiotropy.
-Such a genetic variant is considered to be an instrumental variable for the if it only influences the outcome through the exposure (the exclusion restriction).
-It can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not.
-Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) potential bias arising from the exclusion restriction principal can be detected or corrected by evaluating the consistency of effects across instruments [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Hemani2017; @Verbanck2017].
+If a SNP influences the outcome through the exposure then the SNP is exhibiting vertical pleiotropy (see Box 1).
+Such a genetic variant is considered to be a valid instrumental variable if it only influences the outcome through the exposure (the exclusion restriction assumption).
+Vertical pleiotropy can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not, although not without caveats [@Holmes2017].
+Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) potential bias arising from the exclusion restriction assumption can be detected or corrected by evaluating the consistency of effects across instruments [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Hemani2017; @Verbanck2017].
 
-Genome-wide association studies (GWAS) have identified genetic instrumental variables for thousands of phenotypes [@Hindorff2010].
-Recent developments in MR have enabled knowledge of instrumental variables to be applied using only summary level data (known as two-sample MR, 2SMR) [@Pierce2013].
-Here, in order to infer the causal effect of an exposure on an outcome all that is required is an estimate of the genetic effects of the instrumenting SNP on the exposure, and the corresponding estimate of the effect on the outcome.
+Genome-wide association studies (GWAS) have identified potential instrumental variables for thousands of phenotypes [@Hindorff2010].
+Recent developments in MR have enabled knowledge of instrumental variables to be applied using either individual-level or only summary-level data (known as two-sample MR, 2SMR) [@Pierce2013].
+Here, in order to infer the causal effect of an exposure on an outcome all that is required are the genetic effects of the instrumenting SNP on the exposure and the outcome.
 This has three major advantages.
 First, GWAS summary data are non-disclosive and often publicly available.
-Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting possible causal estimates only by the availability of GWAS summary data for the traits in question [@Hemani2016].
-Third, we can view the causal inference problem through the simple and widely understood prism of a meta-analysis.
+Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting possible MR analyses only by the availability of GWAS summary data for the traits in question [@Hemani2016].
+Third, because each instrumental variable mimics an independent randomization study, we can view the causal inference problem through the simple and widely understood prism of a meta-analysis.
 
@@
 The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017].
-It is considered best practice to report estimates from all available methods as sensitivity analyses when presenting causal estimates because different pleiotropic models can then be evaluated.
+It is considered best practice to report estimates from a range of methods as a sensitivity analysis when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions.
 However it is also desirable to consolidate across methods under some circumstances.
-First, it could be of interest to make causal effect estimates for thousands of traits, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient.
-Second, though the IVW approach is most statistically powerful under no HP, it can have high false negative or low true positive rates in the presence of HP compared to other methods.
-Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate.
+First, it could be of interest to screen links between thousands of traits for being potential causal associations, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient.
+Second, though the simplest method, based on inverse-variance weighted (IVW) meta analysis [@Pierce2013], is most statistically powerful under no HP, it can have high false positive or low true positive rates in the presence of HP compared to other methods.
+Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree and nature (e.g.
+horizontal versus vertical) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate.
 Third, if different methods disagree it is not possible to know which is correct because the true nature of HP exhibited by the instruments is not known.
@@
 Using curated GWAS summary data for thousands of phenotypes [@Hemani2016], we use these new methods to construct a graph of millions of causal estimates.
-Motivated by the recent avalanche [@Hartwig2016] of 2SMR publications, often with repeating [@Xu2017; @Larsson2017] or contradicting [@Prins2016; @Inoshita2016] conclusions despite using the same data, we developed a graph database to represent these estimates in a consistent manner.
+Motivated by the recent avalanche [@Hartwig2016] of 2SMR publications, with similar [@Xu2017; @Larsson2017] or contradicting [@Prins2016; @Inoshita2016] conclusions despite using the same data [@Hartwig2016], we developed a graph database to represent these estimates in a consistent manner.
 We consider this to be a 'working draft' of the causal map of the human phenome, but raise caution throughout that its interpretation is far from straightforward, and that future corrections and refinements will inevitably follow as data grows and 2SMR methods evolve.
+
+### Box 1: Models of pleiotropy
+
+In this paper we separate pleiotropy into two forms for simplicity - vertical and horizontal.
+Vertical pleiotropy has also been known as mediated pleiotropy [@Solovieff2013], because the genetic variant influences one trait, which in turn influences a second trait.
+It has also been termed Type II pleiotropy [@Wagner2011] and secondary pleiotropy [@Hodgkin1998].
+
+By contrast, horizontal pleiotropy can manifest in a myriad of different ways, but the guiding principle is that the genetic variant influences two different phenotypes through independent pathways.
+Horizontal pleiotropy has also been termed biological pleiotropy [@Solovieff2013], Type I pleiotropy [@Wagner2011], developmental pleiotropy or selectional pleiotropy [@Paaby2013].
+From a MR statistical perspective, one can be more specific and state that horizontal pleiotropy occurs where a SNP influences an outcome trait through some pathway other than the exposure trait.
+
+For illustration, Hu et al (2016) (Figure 3) enumerate the biological mechanisms that can lead to pleiotropy: a single variant influences multiple genes; a single gene generates multiple gene products; a single gene product has multiple cellular functions; or a single cellular function influences multiple phenotypes [@Hu2016].
+Whether we consider each of these mechanisms to be vertical pleiotropy (advantageous for MR) or horizontal pleiotropy (problematic for MR) depends on which phenotypes are being interrogated.
+For example, consider a scenario where a SNP influences the expression levels of two genes, and the first gene influences trait A while the second gene influences trait B.
+Then:
+
+- The genetic variant is exhibiting a vertical pleiotropic effect on trait A through the first gene expression level.
+Here, all other assumptions being met, the genetic variant would be a valid instrument if used to estimate the causal effect of the gene expression level on trait A.
+However,
+- the SNP is exhibiting a horizontal pleiotropic effect on trait A and trait B through the two gene expression levels.
+Here the genetic variant would be invalid if used to estimate the causal effect of trait A on trait B.
 
@@
 
-The use of summary data in two-sample MR is described in detail elsewhere [@Hemani2016].
+The use of summary data in two-sample MR is described in detail elsewhere [@Burgess2015b; @Hemani2016].
 A brief outline of the procedure is as follows.
@@
 We collect their effect sizes, standard errors and effect alleles for the association with the exposure trait.
-These can be obtained manually from complete summary data or from curated lists of significant GWAS associations.
-Next, the effects of those SNPs on the outcome need to be obtained, typically necessitating complete summary data because these SNPs will generally have not have reached genome-wide significance for other traits (and therefore be present in curated catalogues).
+These can be obtained from published GWAS summary data, either from from curated lists of established GWAS associations [@Hindorff2010] or study websites comprising complete summary data (all SNP effects across the genome regardless of strength of association).
+Next, the effects of those SNPs on the outcome need to be obtained, typically necessitating access to complete summary data because these SNPs will not usually reach genome-wide significance for the outcome trait (and are therefore usually absent in curated catalogues).
 
-We use the term *summary-set* to refer to the minimum data required to perform 2SMR analysis - four columns comprising the SNP-exposure and SNP-outcome effects and standard errors, and rows corresponding to each SNP that is used as an instrument for the exposure.
+We use the term *summary-set* to refer to the minimum data required to perform 2SMR analysis - four columns comprising the SNP-exposure and SNP-outcome effects and their standard errors, with each rows corresponding to a SNP that is used as an instrument for the exposure.
 
-At its most simple implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome.
+In its simplest implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome.
 This is known as the inverse-variance weighted (IVW) method.
@@
 Given summary data for a large number of traits, it is straightforward to exhaustively analyse the causal relationships of every trait against every other trait for which there are sufficient summary data available.
-Supplementary table 1 provides a list of all traits that have available GWAS summary data, indicating if they have complete summary data (in which case they can be used as both exposure traits and outcome traits), or if they only have significant associations (in which they can only be used as exposure traits).
+Supplementary table 1 provides a list of all traits that have available GWAS summary data that were used in these analyses.
 
@@
 
-In this paper we consider three main classes of MR estimation. Full details for each approach have been described previously. A summary of the methods is given in Supplementary table 2.
+In this paper we consider three main classes of MR estimation.
+Full details for each approach have been described previously.
+A summary of the methods is given in Supplementary table 2.
+Though other methods also exist we have limited the analyses to the following 14 methods for simplicity.
 
@@
 The inverse variance weighted (IVW) fixed effects meta-analysis approach assumes that variants exhibit no HP.
-IVW random effects meta-analysis relaxes the HP assumption, allowing it to be present but balanced - such that it only leads to increased heterogeneity around the regression and not introducing bias.
-Fixed effects Egger regression [@Bowden2015] relaxes the HP assumption further by allowing a non-zero intercept which essentially allows horizontal pleiotropy to be directional, where it systematically influences the outcome in a specific direction.
-Random effects Egger regression allows heterogeneity around the directional HP [@Bowden2017], as long as the HP effects are not correlated with the SNP-exposure effects (also known as the INSIDE assumption)[@Bowden2015].
+IVW random effects meta-analysis relaxes the HP assumption, allowing it to be present but balanced - such that it only leads to increased heterogeneity around the regression line without affecting the slope (and therefore not introducing bias).
+Fixed effects Egger regression [@Bowden2015] relaxes the HP assumption further by allowing a non-zero intercept which essentially allows overall horizontal pleiotropy to be directional, where its total effect influences the outcome in a specific direction.
+Random effects Egger regression further allows heterogeneity around the slope having accounted for overall directional HP [@Bowden2017], as long as the HP effects are not correlated with the SNP-exposure effects (also known as the INSIDE assumption)[@Bowden2015].
 
@@
 Using 1000 rounds of jackknife estimates, we can obtain a final estimate using the mean or the median of the distribution.
-We only use the jackknife approach for associations where there are 15 or more variants in order to avoid saturating the possible number of instrument combinations.
+We only use the jackknife approach for MR analyses where there are 15 or more instruments, in order to avoid saturating the possible number of instrument combinations.
 
@@
 **Median-based methods:** An alternative approach is to take the median effect of all available instruments [@Han2008; @Bowden2016b].
-This has the advantage that up to half the instruments can be invalid, and the estimate will remain unbiased.
-Developing the approach further to allow stronger instruments to contribute more towards the estimate can be obtained by obtaining the median of the weights of each instrument.
+This has the advantage that only half the instruments need to be valid, and the estimate will remain unbiased.
+The weighted median estimate develops the approach further to allow stronger instruments to contribute more towards the estimate can be obtained by weighting the contribution of each instrument by the inverse of its variance.
 The penalised weighted median estimator introduces a further weight to the instruments, penalising any instrument that contributes substantially towards the heterogeneity statistic.
@@
 
-**Mode-based methods:** Similar to the median, the mode based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017].
+**Mode-based methods:** The mode based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017].
 There are four implementations of this method: the simple and the weighted mode, each weighted with or without the assumption of no measurement error in the exposure estimates (NOME).
@@
 Alternatively the clumping procedure can be performed using complete summary data in MR-Base [@Hemani2016].
-We call this the "top hits" strategy.
+Complete summary data refers to association results for all SNPs used in a GWAS i.e.
+not only those passing statistical thresholds for significance.
+We call this the "tophits" strategy.
 
@@
 Using $g_{A}$ as an instrument to test the causal effect of $A$ on $B$ is perfectly valid.
-But in the (incorrectly hypothesised) MR analysis of trait $B$ on trait $A$ could erroneously result in the apparent causal association of $B$ on $A$.
+But the (incorrectly hypothesised) MR analysis of trait $B$ on trait $A$ could erroneously result in the apparent causal effect of $B$ on $A$.
 If $g_{A}$ is only one of many known instruments for $B$, amongst which some are valid, it is to the advantage of the researcher to exclude $g_{A}$ from the analysis.
 
-An approach to inferring the causal direction between phenotypes recently developed [@Hemani2017] uses the following basic premise. If trait $A$ causes trait $B$ then
+An approach to inferring the causal direction between phenotypes [@Hemani2017] uses the following basic premise. If trait $A$ causes trait $B$ then
 
@@
 This simple inequality will not hold in some cases, for example $\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}$ where $\rho_{x, x_o}$ and $\rho_{y, y_o}$ are the precision of the measurements of the $x$ and $y$.
-Some combinations of confounding effects can also distort the $\rho_{g,x}$ and $\rho_{g,y}$ parameters, as has been discussed in detail previously.
+Some combinations of confounding effects can also distort the $\rho_{g,x}$ and $\rho_{g,y}$ parameters, as has been discussed in detail previously [@Hemani2017].
 However, we use it here as a computationally inexpensive and approximate method to identify variants that are likely to be invalid.
@@
 Here we adapt this approach to automatically filter SNPs that are liable to be invalid (Figure 1a).
-Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are likely to be exhibiting horizontal pleiotropy.
+Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are more likely to be a source of HP.
 A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones.
@@
 We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total (Supplementary table 2).
-In the context of this analysis, each method is considered to be an 'expert', taking a set of SNP-exposure and SNP-outcome effect sizes and their standard errors as inputs.
-Our objective is to select the expert most likely to be correct for a specific MR analysis.
+Each method assumes or performs best for a different model of pleiotropy.
+Our objective is to select the method most likely to be correct for a specific MR analysis by predicting the model of pleiotropy, and relating that prediction to the most appropriate model.
+To achieve this we use a "mixture of experts" machine learning approach [@Jordan1994], where each MR method is considered to be an 'expert', taking a summary-set as its input.
 An overview of the approach is shown in Figure 2.
 
-The mixture of experts method is a machine learning approach [@Jordan1994] which seeks to divide a parameter space into subdomains, such that a particular expert is used primarily for problems that reside in a subdomain most suited to that expert.
-In this case our objective is to identify characteristics of the SNP-exposure and SNP-outcome associations for which one specific MR method is most likely to yield highest statistical power for non-null associations, and lowest false discovery rates for null associations.
+The mixture of experts method seeks to divide a parameter space into subdomains, such that a particular expert is used primarily for problems that reside in a subdomain most suited to that expert.
+In this case we first identify characteristics of the SNP-exposure and SNP-outcome associations for which one specific MR method is most likely to yield highest statistical power for non-null associations, and lowest false discovery rates for null associations.
 This involves creating a 'gating function', whose purpose is to decide which expert to use for a specific MR analysis, given the parameter space that is occupied by that summary-set.
-The metrics are a mixture of regression diagnostics and are described in Supplementary table 3.
+The metrics are a collection of regression diagnostics and are described in Supplementary table 3.
 
@@
 
+It should be noted that in the original hierarchical mixture of experts approach of [@Jordan1994] the gating function and the experts share the same input space and are trained simultaneously using expectation-maximisation.
+In our approach the gating function is defined over a separate input space consisting of the parameter space of the experts, and is trained separately in a supervised setting using simulated data.
+As such our MR-MoE approach fits the framework of meta-learning [@brazdil2008metalearning; @smith2009cross; @lemke2015metalearning] where machine learning methods are applied at the 'meta-level' to analyse the results of machine learning experiments at the 'base-level', in order to be able to understand and model the capabilities of each expert and recommend which expert should be applied to a given problem.
+From this perspective the 28 MR methods are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the summary-sets.
+
 ### Graph database of MR estimates
@@
 The set of MR estimates obtained from this analysis, and the summary-sets used to generate them, are deposited in a Neo4j graph database.
-Here nodes representations exist for traits and SNPs; and relationship representations exist for SNP-trait associations and trait-trait MR estimates.
+Here node representations exist for traits and SNPs; and relationship representations exist for SNP-trait associations and trait-trait MR estimates.
 Because each trait-trait association has up to 28 different MR estimates, for simplicity we also distill this down to a third relationship type comprising a single 'best estimate' for each trait-trait association using the following rules:
 
-1. If the number of variants after Steiger filtering is greater than 5 then apply the MoE to obtain the best method
+1.
+If the number of variants after Steiger filtering is greater than 5 then apply the MoE to obtain the best method.
+This value is chosen arbitrarily as a minimum number of variants for which the different MR methods can model HP
 2. If the number of variants after Steiger filtering is less than or equal to 5 but greater than 1 then use the IVW random effects approach on the filtered set of variants
 3. If there is 1 variant retained in the Steiger approach then use the Wald ratio on the remaining variant
-4. If there are no variants remaining after Steiger filtering then declare no causal association.
+4. If there are no variants remaining after Steiger filtering then declare no estimate of a causal association possible.
 
@@
 
-As statistical power for GWAS studies improves the likelihood of a significant association being discovered that doesn't act primarily on the trait of interest increases (Figure 1a).
+As statistical power for GWAS studies improves the likelihood of a significant association being discovered for an outcome trait that acts primarily on some exposure trait for that outcome increases (Figure 1a).
 We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated summary-sets comprising both null and non-null causal models.
@@
 Figure 1b shows that the tophits strategy led to over half of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype.
+For brevity, we refer to the latter type of invalid instruments as “reverse causal” instruments.
 The proportion of invalid instruments due to reverse cause increased as GWAS discovery sample size increased.
@@
 Notably, the next best methods were median and mode based estimators using Steiger filtering.
-Under the assumption of widespread and diverse pleiotropic effects it is clear that all methods suffer from high false discovery rates (Figure 3a).
+But a crucial observation is that under the assumption of widespread and diverse pleiotropic effects it is clear that all methods suffer from high false discovery rates (Figure 3a), including the MoE approach.
 
@@
 We applied our analysis using summary data for 2407 phenotypes, including 149 complex traits and diseases, 575 metabolites [@Shin2014; @Kettunen2016] and 1683 plasma protein levels [@Sun2017].
-For the protein levels only the instruments were available, so they could only be evaluated as exposure phenotypes.
-The complex traits and diseases and metabolite levels had complete summary data available in MR-Base, thus they be evaluated as both exposures (if they had significant instruments) and outcomes.
+For the protein levels only GWAS significant hits were available, so they could only be evaluated as exposure phenotypes.
+The complex traits and diseases and metabolite levels had complete summary data available obtained via MR-Base, and thus they could be evaluated as both exposures (if they had significant instruments) and outcomes.
 Together, we evaluated 715681 relationships.
@@
 The frequencies of the methods chosen by the MR-MoE analysis are shown in Supplementary table 3.
-Amongst those deemed 'significant', the IVW fixed effects analysis using tophits instruments (the only method applicable when there is no evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases.
+Amongst those deemed 'significant', the IVW fixed effects analysis method combined with the tophits instruments selection strategy (only valid when there is no detectable evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases.
 This indicates that horizontal pleiotropy is likely to be pervasive.
 
-### Putative associations involving years of schooling
+### Putative associations involving LDL cholesterol and years of schooling
 
-As an example of the database we performed a look up of associations where LDL cholesterol as measured in the GLGC consortium [@Willer2013] had a false discovery rate of 0.05 using the MoE approach.
+As an example of the database we performed a look up of associations between all traits and LDL cholesterol (as measured in the GLGC consortium [@Willer2013]), where the MoE result had a false discovery rate of 0.05 using the MoE approach.
 This returned 287 putative associations, amongst which 111 involved LDL cholesterol influencing other traits and 176 involved other traits influencing LDL cholesterol.
@@
 There was a strong association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52, $p=2.7\times10^{-12}$).
-Here the IVW random effects method was chosen after Steiger filtering, indicating a prediction of HP being present amongst the instruments.
+Here the IVW random effects method was chosen after Steiger filtering, indicating the presence of HP being present amongst the instruments.
 
@@
 Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4b).
-Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease has recently been examined [@Anderson2017].
-However it is clear that there are serious limitations in relying on MR as a panacea for causal inference.
+Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease, which has recently been reported [@Anderson2017].
+However, it is clear that there are serious limitations in relying on MR as a panacea for causal inference.
 For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g.
 childhood intelligence and birth weight) which appear to be causally downstream of years of schooling despite this being temporally impossible.
-Though explanations could be conjured for this association, for example parents' schooling influences childhood intelligence, or the genetic instruments are shared across the two traits, an intuitive interpretation of causality cannot be applied.
+Though explanations could be conjured for this association, for example parents' schooling influences childhood intelligence, or the genetic instruments are shared across the two traits due to a third unmeasured trait, an intuitive interpretation of causality cannot be applied because the genetic instruments used proxy the liability to achieve some level of educational attainment, and not necessarily the experience of having attained a particular level of edication.
 
@@
 Crucially however, we observed in our simulations that none of the available methods were conservative and all, including MR-MoE, had FDR above 0.2.
-Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criteria.
-We illustrated how confounders and reverse causal associations can easily lead to invalid instruments being used in MR, and that even if those instruments that only influence the exposure directly are used, most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.
+Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criterion.
+We illustrated showed how confounders and reverse causal associations can easily lead to invalid instrument selection in MR and that most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.
 
@@
 
-One advantage to generating a method that consolidates across many methods is that it can prevent cherry picking the most appealing method based on their results.
-But in addition to correctly handling horizontal pleiotropy, there are many other limitations that prevent MR from being a panacea for causal inference.
+One advantage to generating a method that consolidates across many methods is that it can prevent cherry picking the most appealing method based on their results, but in general we strongly advocate the use of reporting multiple MR methods to appraise any specific the impact of HP on a causal estimate.
+In addition to HP, there are many other limitations that prevent MR from being a panacea for causal inference.
 Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes.
+Such associations can be explained statistically because the instruments proxy the liability for that particular phenotype, not that phenotype's observed manifestation.
 Though these associations can be informative, their interpretations as causal relationships are far from clear.
-Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease.
-Again, interpretation of such associations can be problematic.
+Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS of disease is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease.
+Again, interpretation of such associations can be problematic due to the instruments proxying disease liability and not diseae outcomes.
 
-A large proportion of the associations that were estimated used only a single instrumental variable.
-Methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017].
-Separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework though the use of mediation-based approaches may be informative in the one-sample setting [@Wang2017; @Hemani2017].
-Other problems can also manifest, for example frailty effects could induce associations for late-onset traits [@Noyce2017]; genetic variants could strongly relate to several phenotypes making it difficult to ascertain which of them is the causal exposure [@Burgess2014a]; and the measured feature for which genetic associations are known could relate in complicated ways to the biological entity that are truly causal [@DaveySmithHemani2014].
+A large proportion of the associations were estimated using only a single instrumental variable, a circumstance in which cause, reverse cause or horizontal pleiotropy cannot be immediately distinguished.
+Though methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017], separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework though the use of mediation-based approaches may be informative in the one-sample setting [@Wang2017; @Hemani2017].
+Other problems can also manifest.
+Frailty effects [@Aalen2015] could induce associations for late-onset traits, and often detailed modelling for how this can influence MR is warranted [@Noyce2017; @Anderson2017].
+Genetic variants could strongly relate to several phenotypes, for example SNPs that influence LDL cholesterol are also likely to influence other lipid fractions, making it difficult to ascertain which amongst them is the true causal exposure [@Burgess2014a].
+The measured feature for which genetic associations are known could relate in complicated ways to the biological entities that is truly causal [@DaveySmithHemani2014], for example higher circulating levels of the natural antioxidant extracellular superoxide dismutase (EC-SOD) appears to causally relate to higher risk of coronary heart disease [@Juul2003] because the phenotypic measure relates inversely to the EC-SOD levels in arterial walls, hence *in situ* activity is lower.
 
-We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular summary-set, namely, that the data is choosing the model.
+We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular summary-set, namely, that the model selection is data driven.
 This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance.
-The extent to which a machine based method selector suffers from this more than a human manually inferring the result most likely to be true is not clear.
-Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to all other MR methods that do not suffer from this issue.
+The extent to which a machine-based method selector incurs higher type 1 error rates more than a human selector is not clear.
+Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to other MR methods that do not suffer from the issue of data driven model selection.
 
@@
 b) Any SNP that has a direct influence on the exposure, or an influence on a non-confounding intermediate variable, is considered a 'direct' effect.
-The y-axis shows the proportion of instruments selected for analysis due that are either direct associations with the exposure, instruments for the outcome (reverse), or instruments for confounding traits (confounder).
+The y-axis shows the proportion of instruments selected for analysis that are either direct associations with the exposure, instruments for the outcome (reverse), or instruments for confounding traits (confounder).
 The proportions are compared over a range of different exposure discovery sample sizes (x-axis) and using either the tophits approach (left) or the Steiger approach (right) for instrument selection.
@@
 a) Training.
-Datasets are simulated that have either null or non-null causal relationships, and MR is performed by each of the 28 available MR methods.
+Datasets are simulated that have either null or non-null causal relationships, and causal estimates are obtained from each of the 28 MR methods considered in this study.
 In the toy datasets, the columns in black represent 53 metrics about each of the 67,000 training simulations.
@@
 Figure 4: Lookup of causal associations involving a) LDL cholesterol and b) Years of Schooling.
-In a) the nodes are filtered to not include those obtained from the Shin et al or Kettunen et al metabolomic studies.
+In a) the nodes are filtered to not include those obtained from the metabolomic studies [@Shin2014; @Kettunen2016].
 The arrows denote causal direction and the values on the arrows denote the causal effect estimate.
@@
 Having trained the model on simulations, new summary-sets can be applied to obtain causal estimates from the method that is predicted to be the most reliable for that summary-set.
-
-It should be noted that in the original hierarchical mixture of experts approach of [@Jordan1994] the gating function and the experts share the same input space and are trained simultaneously using expectation-maximisation.
-In our approach the gating function is defined over a separate input space consisting of the parameter space of the experts, and is trained separately in a supervised setting using simulated data.
-As such our MR-MoE approach fits the framework of meta-learning [@brazdil2008metalearning; @smith2009cross; @lemke2015metalearning] where machine learning methods are applied at the 'meta-level' to analyse the results of machine learning experiments at the 'base-level', in order to be able to understand and model the capabilities of each expert and recommend which expert should be applied to a given problem.
-From this perspective the 28 MR methods are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the summary-sets.
 

### 2017-08-02 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{29 July 2017}
+  \date{02 August 2017}
 
@@
   MRC Integrative Epidemiology Unit (IEU) at the University of Bristol,
-  School of Social and Community Medicine, Bristol, UK
+  Population Health Sciences, Bristol, UK
 \item
@@
 to SNPs exhibiting horizontal pleiotropy, and many methods have been
-developed in an attempt to address this. Here we show that the
-mechanisms that underlie horizontal pleiotropy are numerous, and that
-instrument selection will be increasingly liable to selecting invalid
-instruments as GWAS sample sizes continue to grow. We have developed a
-mixture of experts machine learning approach (MR-MoE 1.0) that improves
-on both power and false discovery rates over unselective use of any
-existing methods. Using the approach, we systematically estimated the
+developed in an attempt to address this, each assuming a different model
+of horizontal pleiotropy. Here we show that across a range of horizontal
+pleiotropy models instrument selection will be increasingly liable to
+selecting invalid instruments as GWAS sample sizes continue to grow. We
+have integrated these methods into a mixture-of-experts machine learning
+approach (MR-MoE 1.0) that predicts the most appropriate model to use
+for any specific causal analysis, improving on both power and false
+discovery rates. Using the approach, we systematically estimated the
 causal effects amongst 2407 phenotypes, generating a working draft of
-the causal map of the human phenome comprising the evaluation of
+the causal map of the human phenome, comprising the evaluation of
 causality between over a million pairs of traits. Almost 90\% of causal
@@
 
-Mendelian randomization (MR) (1,2) exploits genetic pleiotropy to infer
+Mendelian randomization (MR) (1,2) exploits vertical pleiotropy to infer
 the causal relationships between phenotypes. Suppose that one trait (the
@@
 the outcome through the exposure then the SNP is exhibiting vertical
-pleiotropy. Such a genetic variant is considered to be an instrumental
-variable for the if it only influences the outcome through the exposure
-(the exclusion restriction). It can be exploited to mimic a randomised
-controlled trial, enabling a causal estimate to be made by comparing the
-outcome phenotypes between those individuals that have the
-exposure-increasing allele against those who do not. Multiple
-independent genetic variants for a particular exposure can be used
-jointly to improve causal inference, because a) each variant represents
-an independent natural experiment, and an overall causal estimate can be
+pleiotropy (see Box 1). Such a genetic variant is considered to be a
+valid instrumental variable if it only influences the outcome through
+the exposure (the exclusion restriction assumption). Vertical pleiotropy
+can be exploited to mimic a randomised controlled trial, enabling a
+causal estimate to be made by comparing the outcome phenotypes between
+those individuals that have the exposure-increasing allele against those
+who do not, although not without caveats (3). Multiple independent
+genetic variants for a particular exposure can be used jointly to
+improve causal inference, because a) each variant represents an
+independent natural experiment, and an overall causal estimate can be
 obtained by meta-analysing the single estimates from each instrument;
-and b) potential bias arising from the exclusion restriction principal
+and b) potential bias arising from the exclusion restriction assumption
 can be detected or corrected by evaluating the consistency of effects
-across instruments (3--8).
-
-Genome-wide association studies (GWAS) have identified genetic
-instrumental variables for thousands of phenotypes (9). Recent
+across instruments (4--9).
+
+Genome-wide association studies (GWAS) have identified potential
+instrumental variables for thousands of phenotypes (10). Recent
 developments in MR have enabled knowledge of instrumental variables to
-be applied using only summary level data (known as two-sample MR, 2SMR)
-(10). Here, in order to infer the causal effect of an exposure on an
-outcome all that is required is an estimate of the genetic effects of
-the instrumenting SNP on the exposure, and the corresponding estimate of
-the effect on the outcome. This has three major advantages. First, GWAS
-summary data are non-disclosive and often publicly available. Second,
-causal inference can be made between phenotypes even if they have not
-been measured in the same samples, limiting possible causal estimates
-only by the availability of GWAS summary data for the traits in question
-(11). Third, we can view the causal inference problem through the simple
-and widely understood prism of a meta-analysis.
+be applied using either individual-level or only summary-level data
+(known as two-sample MR, 2SMR) (11). Here, in order to infer the causal
+effect of an exposure on an outcome all that is required are the genetic
+effects of the instrumenting SNP on the exposure and the outcome. This
+has three major advantages. First, GWAS summary data are non-disclosive
+and often publicly available. Second, causal inference can be made
+between phenotypes even if they have not been measured in the same
+samples, limiting possible MR analyses only by the availability of GWAS
+summary data for the traits in question (12). Third, because each
+instrumental variable mimics an independent randomization study, we can
+view the causal inference problem through the simple and widely
+understood prism of a meta-analysis.
 
@@
 have been developed that attempt to reliably obtain unbiased causal
-estimates under specific models of HP (3--6,8,12). It is considered best
-practice to report estimates from all available methods as sensitivity
-analyses when presenting causal estimates because different pleiotropic
-models can then be evaluated. However it is also desirable to
-consolidate across methods under some circumstances. First, it could be
-of interest to make causal effect estimates for thousands of traits, in
+estimates under specific models of HP (4--7,9,13). It is considered best
+practice to report estimates from a range of methods as a sensitivity
+analysis when presenting causal estimates so as to assess the potential
+for different horizontal pleiotropy models and other violations of MR
+assumptions. However it is also desirable to consolidate across methods
+under some circumstances. First, it could be of interest to screen links
+between thousands of traits for being potential causal associations, in
 which case a critical evaluation of each causal estimate of interest may
-not be possible or convenient. Second, though the IVW approach is most
-statistically powerful under no HP, it can have high false negative or
+not be possible or convenient. Second, though the simplest method, based
+on inverse-variance weighted (IVW) meta analysis (11), is most
+statistically powerful under no HP, it can have high false positive or
 low true positive rates in the presence of HP compared to other methods.
-Pleiotropy has been hypothesised to be universal (13), though the degree
-to which this may be the case is contested (14,15), hence defaulting to
-the IVW method in the first instance and using other methods as
-sensitivity analyses may not be appropriate. Third, if different methods
-disagree it is not possible to know which is correct because the true
-nature of HP exhibited by the instruments is not known. Fourth, the
-available methods do not cover all possible models of HP, and therefore
-an automated method for instrument selection may be necessary.
+Pleiotropy has been hypothesised to be universal (14), though the degree
+and nature (e.g.~horizontal versus vertical) to which this may be the
+case is contested (15,16), hence defaulting to the IVW method in the
+first instance and using other methods as sensitivity analyses may not
+be appropriate. Third, if different methods disagree it is not possible
+to know which is correct because the true nature of HP exhibited by the
+instruments is not known. Fourth, the available methods do not cover all
+possible models of HP, and therefore an automated method for instrument
+selection may be necessary.
 
@@
 that attempt to automate both instrument and method selection. Using
-curated GWAS summary data for thousands of phenotypes (11), we use these
+curated GWAS summary data for thousands of phenotypes (12), we use these
 new methods to construct a graph of millions of causal estimates.
-Motivated by the recent avalanche (16) of 2SMR publications, often with
-repeating (17,18) or contradicting (19,20) conclusions despite using the
-same data, we developed a graph database to represent these estimates in
-a consistent manner. We consider this to be a `working draft' of the
-causal map of the human phenome, but raise caution throughout that its
-interpretation is far from straightforward, and that future corrections
-and refinements will inevitably follow as data grows and 2SMR methods
-evolve.
+Motivated by the recent avalanche (17) of 2SMR publications, with
+similar (18,19) or contradicting (20,21) conclusions despite using the
+same data (17), we developed a graph database to represent these
+estimates in a consistent manner. We consider this to be a `working
+draft' of the causal map of the human phenome, but raise caution
+throughout that its interpretation is far from straightforward, and that
+future corrections and refinements will inevitably follow as data grows
+and 2SMR methods evolve.
+
+\subsubsection{Box 1: Models of
+pleiotropy}\label{box-1-models-of-pleiotropy}
+
+In this paper we separate pleiotropy into two forms for simplicity -
+vertical and horizontal. Vertical pleiotropy has also been known as
+mediated pleiotropy (22), because the genetic variant influences one
+trait, which in turn influences a second trait. It has also been termed
+Type II pleiotropy (15) and secondary pleiotropy (23).
+
+By contrast, horizontal pleiotropy can manifest in a myriad of different
+ways, but the guiding principle is that the genetic variant influences
+two different phenotypes through independent pathways. Horizontal
+pleiotropy has also been termed biological pleiotropy (22), Type I
+pleiotropy (15), developmental pleiotropy or selectional pleiotropy
+(24). From a MR statistical perspective, one can be more specific and
+state that horizontal pleiotropy occurs where a SNP influences an
+outcome trait through some pathway other than the exposure trait.
+
+For illustration, Hu et al (2016) (Figure 3) enumerate the biological
+mechanisms that can lead to pleiotropy: a single variant influences
+multiple genes; a single gene generates multiple gene products; a single
+gene product has multiple cellular functions; or a single cellular
+function influences multiple phenotypes (25). Whether we consider each
+of these mechanisms to be vertical pleiotropy (advantageous for MR) or
+horizontal pleiotropy (problematic for MR) depends on which phenotypes
+are being interrogated. For example, consider a scenario where a SNP
+influences the expression levels of two genes, and the first gene
+influences trait A while the second gene influences trait B. Then:
+
+\begin{itemize}
+\tightlist
+\item
+  The genetic variant is exhibiting a vertical pleiotropic effect on
+  trait A through the first gene expression level. Here, all other
+  assumptions being met, the genetic variant would be a valid instrument
+  if used to estimate the causal effect of the gene expression level on
+  trait A. However,
+\item
+  the SNP is exhibiting a horizontal pleiotropic effect on trait A and
+  trait B through the two gene expression levels. Here the genetic
+  variant would be invalid if used to estimate the causal effect of
+  trait A on trait B.
+\end{itemize}
 
@@
 The use of summary data in two-sample MR is described in detail
-elsewhere (11). A brief outline of the procedure is as follows. First,
-genetic instruments for the exposure trait need to be identified - those
-SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
-assumption of MR (that the instrument associates with the exposure) is
-generally satisfied. We collect their effect sizes, standard errors and
-effect alleles for the association with the exposure trait. These can be
-obtained manually from complete summary data or from curated lists of
-significant GWAS associations. Next, the effects of those SNPs on the
-outcome need to be obtained, typically necessitating complete summary
-data because these SNPs will generally have not have reached genome-wide
-significance for other traits (and therefore be present in curated
-catalogues).
+elsewhere (12,26). A brief outline of the procedure is as follows.
+First, genetic instruments for the exposure trait need to be identified
+- those SNPs with \(p < 10^{-8}\) are retained in order to ensure that
+the first assumption of MR (that the instrument associates with the
+exposure) is generally satisfied. We collect their effect sizes,
+standard errors and effect alleles for the association with the exposure
+trait. These can be obtained from published GWAS summary data, either
+from from curated lists of established GWAS associations (10) or study
+websites comprising complete summary data (all SNP effects across the
+genome regardless of strength of association). Next, the effects of
+those SNPs on the outcome need to be obtained, typically necessitating
+access to complete summary data because these SNPs will not usually
+reach genome-wide significance for the outcome trait (and are therefore
+usually absent in curated catalogues).
 
@@
 to perform 2SMR analysis - four columns comprising the SNP-exposure and
-SNP-outcome effects and standard errors, and rows corresponding to each
-SNP that is used as an instrument for the exposure.
-
-At its most simple implementation, the regression of the SNP-exposure
+SNP-outcome effects and their standard errors, with each rows
+corresponding to a SNP that is used as an instrument for the exposure.
+
+In its simplest implementation, the regression of the SNP-exposure
 effect sizes against the SNP-outcome effect sizes, with greater weight
@@
 Supplementary table 1 provides a list of all traits that have available
-GWAS summary data, indicating if they have complete summary data (in
-which case they can be used as both exposure traits and outcome traits),
-or if they only have significant associations (in which they can only be
-used as exposure traits).
+GWAS summary data that were used in these analyses.
 
@@
 details for each approach have been described previously. A summary of
-the methods is given in Supplementary table 2.
-
-\textbf{Mean-based methods:} Here we consider four nested models (5).
+the methods is given in Supplementary table 2. Though other methods also
+exist we have limited the analyses to the following 14 methods for
+simplicity.
+
+\textbf{Mean-based methods:} Here we consider four nested models (6).
 The inverse variance weighted (IVW) fixed effects meta-analysis approach
@@
 relaxes the HP assumption, allowing it to be present but balanced - such
-that it only leads to increased heterogeneity around the regression and
-not introducing bias. Fixed effects Egger regression (3) relaxes the HP
-assumption further by allowing a non-zero intercept which essentially
-allows horizontal pleiotropy to be directional, where it systematically
+that it only leads to increased heterogeneity around the regression line
+without affecting the slope (and therefore not introducing bias). Fixed
+effects Egger regression (4) relaxes the HP assumption further by
+allowing a non-zero intercept which essentially allows overall
+horizontal pleiotropy to be directional, where its total effect
 influences the outcome in a specific direction. Random effects Egger
-regression allows heterogeneity around the directional HP (5), as long
-as the HP effects are not correlated with the SNP-exposure effects (also
-known as the INSIDE assumption)(3).
-
-The Rucker framework (21), adapted to MR (5) uses estimates of
+regression further allows heterogeneity around the slope having
+accounted for overall directional HP (6), as long as the HP effects are
+not correlated with the SNP-exposure effects (also known as the INSIDE
+assumption)(4).
+
+The Rucker framework (27), adapted to MR (6) uses estimates of
 heterogeneity in the IVW and Egger frameworks to navigate between these
@@
 the mean or the median of the distribution. We only use the jackknife
-approach for associations where there are 15 or more variants in order
-to avoid saturating the possible number of instrument combinations.
+approach for MR analyses where there are 15 or more instruments, in
+order to avoid saturating the possible number of instrument
+combinations.
 
@@
 \textbf{Median-based methods:} An alternative approach is to take the
-median effect of all available instruments (4,22). This has the
-advantage that up to half the instruments can be invalid, and the
-estimate will remain unbiased. Developing the approach further to allow
-stronger instruments to contribute more towards the estimate can be
-obtained by obtaining the median of the weights of each instrument. The
-penalised weighted median estimator introduces a further weight to the
-instruments, penalising any instrument that contributes substantially
-towards the heterogeneity statistic. Together, this provides three
-median based estimators. Other estimation strategies not considered
-here, such as LASSO regression, have also been developed for the
-situation where at least half of the instruments are valid (23).
-
-\textbf{Mode-based methods:} Similar to the median, the mode based
-estimator clusters the instruments into groups based on similarity of
-causal effects, and returns the final causal effect estimate based on
-the cluster that has the largest number of instruments (6). There are
-four implementations of this method: the simple and the weighted mode,
-each weighted with or without the assumption of no measurement error in
-the exposure estimates (NOME). The simple mode is the unweighted mode of
-the empirical density function of causal estimates, whereas the weighted
-mode is weighted by the inverse variance of the outcome effect. The
-bandwidth parameter was set to 1 by default.
+median effect of all available instruments (5,28). This has the
+advantage that only half the instruments need to be valid, and the
+estimate will remain unbiased. The weighted median estimate develops the
+approach further to allow stronger instruments to contribute more
+towards the estimate can be obtained by weighting the contribution of
+each instrument by the inverse of its variance. The penalised weighted
+median estimator introduces a further weight to the instruments,
+penalising any instrument that contributes substantially towards the
+heterogeneity statistic. Together, this provides three median based
+estimators. Other estimation strategies not considered here, such as
+LASSO regression, have also been developed for the situation where at
+least half of the instruments are valid (29).
+
+\textbf{Mode-based methods:} The mode based estimator clusters the
+instruments into groups based on similarity of causal effects, and
+returns the final causal effect estimate based on the cluster that has
+the largest number of instruments (7). There are four implementations of
+this method: the simple and the weighted mode, each weighted with or
+without the assumption of no measurement error in the exposure estimates
+(NOME). The simple mode is the unweighted mode of the empirical density
+function of causal estimates, whereas the weighted mode is weighted by
+the inverse variance of the outcome effect. The bandwidth parameter was
+set to 1 by default.
 
@@
 recorded in public GWAS catalogs. Alternatively the clumping procedure
-can be performed using complete summary data in MR-Base (11). We call
-this the ``top hits'' strategy.
+can be performed using complete summary data in MR-Base (12). Complete
+summary data refers to association results for all SNPs used in a GWAS
+i.e.~not only those passing statistical thresholds for significance. We
+call this the ``tophits'' strategy.
 
@@
 \(g_{A}\) as an instrument to test the causal effect of \(A\) on \(B\)
-is perfectly valid. But in the (incorrectly hypothesised) MR analysis of
+is perfectly valid. But the (incorrectly hypothesised) MR analysis of
 trait \(B\) on trait \(A\) could erroneously result in the apparent
-causal association of \(B\) on \(A\). If \(g_{A}\) is only one of many
-known instruments for \(B\), amongst which some are valid, it is to the
+causal effect of \(B\) on \(A\). If \(g_{A}\) is only one of many known
+instruments for \(B\), amongst which some are valid, it is to the
 advantage of the researcher to exclude \(g_{A}\) from the analysis.
 
-An approach to inferring the causal direction between phenotypes
-recently developed (7) uses the following basic premise. If trait \(A\)
-causes trait \(B\) then
+An approach to inferring the causal direction between phenotypes (8)
+uses the following basic premise. If trait \(A\) causes trait \(B\) then
 
@@
 \(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
-detail previously. However, we use it here as a computationally
+detail previously (8). However, we use it here as a computationally
 inexpensive and approximate method to identify variants that are likely
-to be invalid. Steiger's Z-test of correlated correlations (24) can be
+to be invalid. Steiger's Z-test of correlated correlations (30) can be
 used to formally test the extent to which the two correlations are
@@
 invalid instruments for the purposes of exclusion from MR analyses
-(8,12,25), based on the notion that outlying instruments are likely to
-be exhibiting horizontal pleiotropy. A potential drawback of this
-approach is that the outlier SNPs in a summary-set might be the only
-reliable ones. This could arise, for example, for the \emph{CRP} variant
-instrumenting C-reactive protein levels (19), or the \emph{SLC2A9}
-variant influencing serum urate levels (26).
+(9,13,31), based on the notion that outlying instruments are more likely
+to be a source of HP. A potential drawback of this approach is that the
+outlier SNPs in a summary-set might be the only reliable ones. This
+could arise, for example, for the \emph{CRP} variant instrumenting
+C-reactive protein levels (20), or the \emph{SLC2A9} variant influencing
+serum urate levels (32).
 
@@
 be supplied using two instrument selection strategies, leading to 28
-methods in total (Supplementary table 2). In the context of this
-analysis, each method is considered to be an `expert', taking a set of
-SNP-exposure and SNP-outcome effect sizes and their standard errors as
-inputs. Our objective is to select the expert most likely to be correct
-for a specific MR analysis. An overview of the approach is shown in
-Figure 2.
-
-The mixture of experts method is a machine learning approach (27) which
-seeks to divide a parameter space into subdomains, such that a
-particular expert is used primarily for problems that reside in a
-subdomain most suited to that expert. In this case our objective is to
-identify characteristics of the SNP-exposure and SNP-outcome
+methods in total (Supplementary table 2). Each method assumes or
+performs best for a different model of pleiotropy. Our objective is to
+select the method most likely to be correct for a specific MR analysis
+by predicting the model of pleiotropy, and relating that prediction to
+the most appropriate model. To achieve this we use a ``mixture of
+experts'' machine learning approach (33), where each MR method is
+considered to be an `expert', taking a summary-set as its input. An
+overview of the approach is shown in Figure 2.
+
+The mixture of experts method seeks to divide a parameter space into
+subdomains, such that a particular expert is used primarily for problems
+that reside in a subdomain most suited to that expert. In this case we
+first identify characteristics of the SNP-exposure and SNP-outcome
 associations for which one specific MR method is most likely to yield
@@
 MR analysis, given the parameter space that is occupied by that
-summary-set. The metrics are a mixture of regression diagnostics and are
-described in Supplementary table 3.
+summary-set. The metrics are a collection of regression diagnostics and
+are described in Supplementary table 3.
 
@@
 \href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
-(11).
+(12).
+
+It should be noted that in the original hierarchical mixture of experts
+approach of (33) the gating function and the experts share the same
+input space and are trained simultaneously using
+expectation-maximisation. In our approach the gating function is defined
+over a separate input space consisting of the parameter space of the
+experts, and is trained separately in a supervised setting using
+simulated data. As such our MR-MoE approach fits the framework of
+meta-learning (34--36) where machine learning methods are applied at the
+`meta-level' to analyse the results of machine learning experiments at
+the `base-level', in order to be able to understand and model the
+capabilities of each expert and recommend which expert should be applied
+to a given problem. From this perspective the 28 MR methods are the
+base-models, the gating function is the meta-model, and its
+meta-features are the parameters of the summary-sets.
 
@@
 summary-sets used to generate them, are deposited in a Neo4j graph
-database. Here nodes representations exist for traits and SNPs; and
+database. Here node representations exist for traits and SNPs; and
 relationship representations exist for SNP-trait associations and
@@
   If the number of variants after Steiger filtering is greater than 5
-  then apply the MoE to obtain the best method
+  then apply the MoE to obtain the best method. This value is chosen
+  arbitrarily as a minimum number of variants for which the different MR
+  methods can model HP
 \item
@@
   If there are no variants remaining after Steiger filtering then
-  declare no causal association.
+  declare no estimate of a causal association possible.
 \end{enumerate}
@@
 As statistical power for GWAS studies improves the likelihood of a
-significant association being discovered that doesn't act primarily on
-the trait of interest increases (Figure 1a). We evaluated the efficacy
-of the Steiger filtering approach for improving instrument selection
-using 100,000 simulated summary-sets comprising both null and non-null
-causal models. For each summary-set, instruments were selected based on
-the tophits strategy and the Steiger filtering strategy, and MR was
-performed using 14 different methods based on instruments selected from
-each of these strategies.
+significant association being discovered for an outcome trait that acts
+primarily on some exposure trait for that outcome increases (Figure 1a).
+We evaluated the efficacy of the Steiger filtering approach for
+improving instrument selection using 100,000 simulated summary-sets
+comprising both null and non-null causal models. For each summary-set,
+instruments were selected based on the tophits strategy and the Steiger
+filtering strategy, and MR was performed using 14 different methods
+based on instruments selected from each of these strategies.
 
@@
 instruments being primarily associated with either confounders or the
-outcome phenotype, not the exposure phenotype. The proportion of invalid
-instruments due to reverse cause increased as GWAS discovery sample size
-increased. Applying Steiger filtering reduced this to 25\%.
-Consequently, the false discovery rates (FDR) for 12 of the 14 methods
-reduced substantially when applied using Steiger filtered instruments
-(Figure 1c). The true positive rates for the methods based on Steiger
-filtering did however reduce slightly for 10 of the 14 methods.
+outcome phenotype, not the exposure phenotype. For brevity, we refer to
+the latter type of invalid instruments as ``reverse causal''
+instruments. The proportion of invalid instruments due to reverse cause
+increased as GWAS discovery sample size increased. Applying Steiger
+filtering reduced this to 25\%. Consequently, the false discovery rates
+(FDR) for 12 of the 14 methods reduced substantially when applied using
+Steiger filtered instruments (Figure 1c). The true positive rates for
+the methods based on Steiger filtering did however reduce slightly for
+10 of the 14 methods.
 
@@
 being null or non-null. Notably, the next best methods were median and
-mode based estimators using Steiger filtering. Under the assumption of
-widespread and diverse pleiotropic effects it is clear that all methods
-suffer from high false discovery rates (Figure 3a).
+mode based estimators using Steiger filtering. But a crucial observation
+is that under the assumption of widespread and diverse pleiotropic
+effects it is clear that all methods suffer from high false discovery
+rates (Figure 3a), including the MoE approach.
 
@@
 We applied our analysis using summary data for 2407 phenotypes,
-including 149 complex traits and diseases, 575 metabolites (28,29) and
-1683 plasma protein levels (30). For the protein levels only the
-instruments were available, so they could only be evaluated as exposure
-phenotypes. The complex traits and diseases and metabolite levels had
-complete summary data available in MR-Base, thus they be evaluated as
-both exposures (if they had significant instruments) and outcomes.
-Together, we evaluated 715681 relationships. The majority of these
-associations could only be evaluated using fewer than 5 SNPs, and so the
-Wald ratio or IVW fixed effects methods were used, but for 61029
-associations the MR-MoE approach was applied.
+including 149 complex traits and diseases, 575 metabolites (37,38) and
+1683 plasma protein levels (39). For the protein levels only GWAS
+significant hits were available, so they could only be evaluated as
+exposure phenotypes. The complex traits and diseases and metabolite
+levels had complete summary data available obtained via MR-Base, and
+thus they could be evaluated as both exposures (if they had significant
+instruments) and outcomes. Together, we evaluated 715681 relationships.
+The majority of these associations could only be evaluated using fewer
+than 5 SNPs, and so the Wald ratio or IVW fixed effects methods were
+used, but for 61029 associations the MR-MoE approach was applied.
 
@@
 in Supplementary table 3. Amongst those deemed `significant', the IVW
-fixed effects analysis using tophits instruments (the only method
-applicable when there is no evidence of horizontal pleiotropy of any
-sort) was selected in only 10.4\% of cases. This indicates that
-horizontal pleiotropy is likely to be pervasive.
-
-\subsubsection{Putative associations involving years of
-schooling}\label{putative-associations-involving-years-of-schooling}
+fixed effects analysis method combined with the tophits instruments
+selection strategy (only valid when there is no detectable evidence of
+horizontal pleiotropy of any sort) was selected in only 10.4\% of cases.
+This indicates that horizontal pleiotropy is likely to be pervasive.
+
+\subsubsection{Putative associations involving LDL cholesterol and years
+of
+schooling}\label{putative-associations-involving-ldl-cholesterol-and-years-of-schooling}
 
 As an example of the database we performed a look up of associations
-where LDL cholesterol as measured in the GLGC consortium (31) had a
-false discovery rate of 0.05 using the MoE approach. This returned 287
-putative associations, amongst which 111 involved LDL cholesterol
-influencing other traits and 176 involved other traits influencing LDL
-cholesterol. A large proportion of these traits were metabolites, which
-are dominated by lipid fractions. Filtering to exclude the metabolomic
-studies ((28,29)) returned 27 associations with 23 traits (Figure 4a).
-There was a strong association with coronary heart disease (0.41 log(OR)
-per SD, 0.31-0.52, \(p=2.7\times10^{-12}\)). Here the IVW random effects
-method was chosen after Steiger filtering, indicating a prediction of HP
-being present amongst the instruments.
+between all traits and LDL cholesterol (as measured in the GLGC
+consortium (40)), where the MoE result had a false discovery rate of
+0.05 using the MoE approach. This returned 287 putative associations,
+amongst which 111 involved LDL cholesterol influencing other traits and
+176 involved other traits influencing LDL cholesterol. A large
+proportion of these traits were metabolites, which are dominated by
+lipid fractions. Filtering to exclude the metabolomic studies ((37,38))
+returned 27 associations with 23 traits (Figure 4a). There was a strong
+association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52,
+\(p=2.7\times10^{-12}\)). Here the IVW random effects method was chosen
+after Steiger filtering, indicating the presence of HP being present
+amongst the instruments.
 
 We also performed a look up of associations that causally influence
-years of schooling (32), or that years of schooling itself causally
+years of schooling (41), or that years of schooling itself causally
 influences. Using a false-discovery rate of 0.05, 45 traits were
@@
 plausible, for example the influence of years of schooling on lower risk
-of Alzheimer's disease has recently been examined (33). However it is
-clear that there are serious limitations in relying on MR as a panacea
-for causal inference. For example, bi-directional causal relationships
-with college completion exist because of the similarity in trait
-definitions; and there are several traits (e.g.~childhood intelligence
-and birth weight) which appear to be causally downstream of years of
-schooling despite this being temporally impossible. Though explanations
-could be conjured for this association, for example parents' schooling
-influences childhood intelligence, or the genetic instruments are shared
-across the two traits, an intuitive interpretation of causality cannot
-be applied.
+of Alzheimer's disease, which has recently been reported (42). However,
+it is clear that there are serious limitations in relying on MR as a
+panacea for causal inference. For example, bi-directional causal
+relationships with college completion exist because of the similarity in
+trait definitions; and there are several traits (e.g.~childhood
+intelligence and birth weight) which appear to be causally downstream of
+years of schooling despite this being temporally impossible. Though
+explanations could be conjured for this association, for example
+parents' schooling influences childhood intelligence, or the genetic
+instruments are shared across the two traits due to a third unmeasured
+trait, an intuitive interpretation of causality cannot be applied
+because the genetic instruments used proxy the liability to achieve some
+level of educational attainment, and not necessarily the experience of
+having attained a particular level of edication.
 
@@
 the rapidly growing field of automatic model selection and
-hyper-parameter optimization in machine learning (34).
+hyper-parameter optimization in machine learning (43).
 
@@
 above 0.2. Instrumental variables in MR are typically chosen blind, with
-GWAS significance being the only criteria. We illustrated how
+GWAS significance being the only criterion. We illustrated showed how
 confounders and reverse causal associations can easily lead to invalid
-instruments being used in MR, and that even if those instruments that
-only influence the exposure directly are used, most randomly generated
-patterns of horizontal pleiotropy cannot be adequately accounted for by
-any single MR method.
+instrument selection in MR and that most randomly generated patterns of
+horizontal pleiotropy cannot be adequately accounted for by any single
+MR method.
 
 We applied the method to the curated set of GWAS summary data present in
-MR-Base (11), generating several million MR estimates. MR-MoE selected a
+MR-Base (12), generating several million MR estimates. MR-MoE selected a
 method that indicated some pattern of horizontal pleiotropy for almost
@@
 methods is that it can prevent cherry picking the most appealing method
-based on their results. But in addition to correctly handling horizontal
-pleiotropy, there are many other limitations that prevent MR from being
-a panacea for causal inference. Many of the associations are
-biologically impossible, for example where early life phenotypes appear
-to be influenced by later stage phenotypes. Though these associations
-can be informative, their interpretations as causal relationships are
-far from clear. Similarly, often disease traits appear to causally
-relate to other phenotypes, but GWAS is typically performed on the
-liability scale, hence the causal estimate reflects not the presence or
-absence of disease, but the underlying risk of disease. Again,
-interpretation of such associations can be problematic.
-
-A large proportion of the associations that were estimated used only a
-single instrumental variable. Methods are emerging to attempt to
-delineate between models of reverse cause, pleiotropy and multiple
-causal variants in the same region (35,36). Separating vertical from
-horizontal pleiotropy with a single instrument is not yet possible in
-the two-sample MR framework though the use of mediation-based approaches
-may be informative in the one-sample setting (7,37). Other problems can
-also manifest, for example frailty effects could induce associations for
-late-onset traits (38); genetic variants could strongly relate to
-several phenotypes making it difficult to ascertain which of them is the
-causal exposure (39); and the measured feature for which genetic
-associations are known could relate in complicated ways to the
-biological entity that are truly causal (2).
+based on their results, but in general we strongly advocate the use of
+reporting multiple MR methods to appraise any specific the impact of HP
+on a causal estimate. In addition to HP, there are many other
+limitations that prevent MR from being a panacea for causal inference.
+Many of the associations are biologically impossible, for example where
+early life phenotypes appear to be influenced by later stage phenotypes.
+Such associations can be explained statistically because the instruments
+proxy the liability for that particular phenotype, not that phenotype's
+observed manifestation. Though these associations can be informative,
+their interpretations as causal relationships are far from clear.
+Similarly, often disease traits appear to causally relate to other
+phenotypes, but GWAS of disease is typically performed on the liability
+scale, hence the causal estimate reflects not the presence or absence of
+disease, but the underlying risk of disease. Again, interpretation of
+such associations can be problematic due to the instruments proxying
+disease liability and not diseae outcomes.
+
+A large proportion of the associations were estimated using only a
+single instrumental variable, a circumstance in which cause, reverse
+cause or horizontal pleiotropy cannot be immediately distinguished.
+Though methods are emerging to attempt to delineate between models of
+reverse cause, pleiotropy and multiple causal variants in the same
+region (44,45), separating vertical from horizontal pleiotropy with a
+single instrument is not yet possible in the two-sample MR framework
+though the use of mediation-based approaches may be informative in the
+one-sample setting (8,46). Other problems can also manifest. Frailty
+effects (47) could induce associations for late-onset traits, and often
+detailed modelling for how this can influence MR is warranted (42,48).
+Genetic variants could strongly relate to several phenotypes, for
+example SNPs that influence LDL cholesterol are also likely to influence
+other lipid fractions, making it difficult to ascertain which amongst
+them is the true causal exposure (49). The measured feature for which
+genetic associations are known could relate in complicated ways to the
+biological entities that is truly causal (2), for example higher
+circulating levels of the natural antioxidant extracellular superoxide
+dismutase (EC-SOD) appears to causally relate to higher risk of coronary
+heart disease (50) because the phenotypic measure relates inversely to
+the EC-SOD levels in arterial walls, hence \emph{in situ} activity is
+lower.
 
@@
 approaches to infer the correct model for a particular summary-set,
-namely, that the data is choosing the model. This is liable to
+namely, that the model selection is data driven. This is liable to
 contribute to elevated type 1 error rates, even though we have attempted
 to separate the information used for optimisation from the information
-used to predict performance. The extent to which a machine based method
-selector suffers from this more than a human manually inferring the
-result most likely to be true is not clear. Though MR-MoE does exhibit
-higher type 1 error rates than are desirable, they still remain amongst
-the lowest compared to all other MR methods that do not suffer from this
-issue.
-
-The field of MR is evolving. The advent of GWAS databases (9,40) and
-automated 2SMR (11) has trivialised the analytical aspect of
+used to predict performance. The extent to which a machine-based method
+selector incurs higher type 1 error rates more than a human selector is
+not clear. Though MR-MoE does exhibit higher type 1 error rates than are
+desirable, they still remain amongst the lowest compared to other MR
+methods that do not suffer from the issue of data driven model
+selection.
+
+The field of MR is evolving. The advent of GWAS databases (10,51) and
+automated 2SMR (12) has trivialised the analytical aspect of
 investigating specific causal hypotheses. Despite the limitations
@@
 `direct' effect. The y-axis shows the proportion of instruments selected
-for analysis due that are either direct associations with the exposure,
+for analysis that are either direct associations with the exposure,
 instruments for the outcome (reverse), or instruments for confounding
@@
 Figure 2: Mixture of experts. a) Training. Datasets are simulated that
-have either null or non-null causal relationships, and MR is performed
-by each of the 28 available MR methods. In the toy datasets, the columns
-in black represent 53 metrics about each of the 67,000 training
-simulations. The columns in red are specific to each method, they
-represent how well that method performed in obtaining the correct answer
-for each of the datasets. Random forests are used to learn the parameter
-space of the 53 metrics in which a particular dataset is likely to
-perform well. Together, this creates 28 random forest decision trees,
-one for each method. b) Application. For a GWAS summary dataset, our
-objective is to choose the method most likely to return the correct
-causal estimate. Metrics are generated from the dataset and fed into
-each of the 28 random forest decision trees. This provides us with 28
-performance predictions. Finally, we use the method for which the
-performance prediction is highest.
+have either null or non-null causal relationships, and causal estimates
+are obtained from each of the 28 MR methods considered in this study. In
+the toy datasets, the columns in black represent 53 metrics about each
+of the 67,000 training simulations. The columns in red are specific to
+each method, they represent how well that method performed in obtaining
+the correct answer for each of the datasets. Random forests are used to
+learn the parameter space of the 53 metrics in which a particular
+dataset is likely to perform well. Together, this creates 28 random
+forest decision trees, one for each method. b) Application. For a GWAS
+summary dataset, our objective is to choose the method most likely to
+return the correct causal estimate. Metrics are generated from the
+dataset and fed into each of the 28 random forest decision trees. This
+provides us with 28 performance predictions. Finally, we use the method
+for which the performance prediction is highest.
 
@@
 b) Years of Schooling. In a) the nodes are filtered to not include those
-obtained from the Shin et al or Kettunen et al metabolomic studies. The
-arrows denote causal direction and the values on the arrows denote the
-causal effect estimate. Only those relationships are shown for which the
+obtained from the metabolomic studies (37,38). The arrows denote causal
+direction and the values on the arrows denote the causal effect
+estimate. Only those relationships are shown for which the
 \(FDR < 0.05\).
@@
 An approach to inferring the causal direction between phenotypes
-recently developed (7) uses the following basic premise. If trait \(A\)
+recently developed (8) uses the following basic premise. If trait \(A\)
 causes trait \(B\) then \(cor(g_{i}, A)^2 > cor(g_{i}, B)^2\) because
@@
 or unmeasured confounding can lead to the inequality reversing
-direction, as has been explored previously (7).
+direction, as has been explored previously (8).
 
@@
 \(V_a = 2\beta^2f(1-f)\), where \(\beta\) is the log odds ratio and
-\(f\) is the allele frequency of the SNP in the population (41). \(f\)
+\(f\) is the allele frequency of the SNP in the population (52). \(f\)
 can be estimated using the allele frequency of the SNP in an ascertained
@@
 that is predicted to be the most reliable for that summary-set.
-
-It should be noted that in the original hierarchical mixture of experts
-approach of (27) the gating function and the experts share the same
-input space and are trained simultaneously using
-expectation-maximisation. In our approach the gating function is defined
-over a separate input space consisting of the parameter space of the
-experts, and is trained separately in a supervised setting using
-simulated data. As such our MR-MoE approach fits the framework of
-meta-learning (42--44) where machine learning methods are applied at the
-`meta-level' to analyse the results of machine learning experiments at
-the `base-level', in order to be able to understand and model the
-capabilities of each expert and recommend which expert should be applied
-to a given problem. From this perspective the 28 MR methods are the
-base-models, the gating function is the meta-model, and its
-meta-features are the parameters of the summary-sets.
 
@@
 methods. The default settings for the \texttt{randomForest} package in R
-(45) were used to train the models. MR-MoE 1.0 is implemented in the
+(53) were used to train the models. MR-MoE 1.0 is implemented in the
 TwoSampleMR R package available at
 \href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
-(11).
+(12).
 
@@
 
+\hypertarget{ref-Holmes2017}{}
+3. Holmes MV, Ala-Korpela M, Smith GD. Mendelian randomization in
+cardiometabolic disease: challenges in evaluating causality. Nature
+Reviews Cardiology {[}Internet{]}. 2017 Jun; Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/28569269\%20http://www.nature.com/doifinder/10.1038/nrcardio.2017.78}{http://www.ncbi.nlm.nih.gov/pubmed/28569269 http://www.nature.com/doifinder/10.1038/nrcardio.2017.78}
+
 \hypertarget{ref-Bowden2015}{}
-3. Bowden J, Davey Smith G, Burgess S. Mendelian randomization with
+4. Bowden J, Davey Smith G, Burgess S. Mendelian randomization with
 invalid instruments: effect estimation and bias detection through Egger
@@
 \hypertarget{ref-Bowden2016b}{}
-4. Bowden J, Davey Smith G, Haycock PC, Burgess S. Consistent Estimation
+5. Bowden J, Davey Smith G, Haycock PC, Burgess S. Consistent Estimation
 in Mendelian Randomization with Some Invalid Instruments Using a
@@
 \hypertarget{ref-Bowden2017}{}
-5. Bowden J, Del~Greco~M F, Minelli C, Davey Smith G, Sheehan N,
+6. Bowden J, Del~Greco~M F, Minelli C, Davey Smith G, Sheehan N,
 Thompson J. A framework for the investigation of pleiotropy in
@@
 \hypertarget{ref-Hartwig2017}{}
-6. Hartwig FP, Davey Smith G, Bowden J. Robust Inference In Two-Sample
+7. Hartwig FP, Davey Smith G, Bowden J. Robust Inference In Two-Sample
 Mendelian Randomisation Via The Zero Modal Pleiotropy Assumption.
@@
 \hypertarget{ref-Hemani2017}{}
-7. Hemani G, Tilling K, Davey Smith G. Orienting The Causal Relationship
+8. Hemani G, Tilling K, Davey Smith G. Orienting The Causal Relationship
 Between Imprecisely Measured Traits Using Genetic Instruments. bioRxiv
@@
 \hypertarget{ref-Verbanck2017}{}
-8. Verbanck M, Chen C-Y, Neale B, Do R. Widespread pleiotropy confounds
+9. Verbanck M, Chen C-Y, Neale B, Do R. Widespread pleiotropy confounds
 causal relationships between complex traits and diseases inferred from
@@
 \hypertarget{ref-Hindorff2010}{}
-9. Hindorff LA, Junkins HA, Hall PN, Mehta JP, Manolio TA. A Catalog of
+10. Hindorff LA, Junkins HA, Hall PN, Mehta JP, Manolio TA. A Catalog of
 Published Genome-Wide Association Studies, available at
@@
 \hypertarget{ref-Pierce2013}{}
-10. Pierce BL, Burgess S. Efficient design for Mendelian randomization
+11. Pierce BL, Burgess S. Efficient design for Mendelian randomization
 studies: subsample and 2-sample instrumental variable estimators.
@@
 \hypertarget{ref-Hemani2016}{}
-11. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
+12. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
 MR-Base: a platform for systematic causal inference across the phenome
@@
 \hypertarget{ref-Zhu2017}{}
-12. Zhu Z, Zheng Z, Zhang F, Wu Y, Trzaskowski M, Maier R, et al. Causal
+13. Zhu Z, Zheng Z, Zhang F, Wu Y, Trzaskowski M, Maier R, et al. Causal
 associations between risk factors and common diseases inferred from GWAS
@@
 \hypertarget{ref-Wright1968}{}
-13. Wright S. Evolution and the Genetics of Populations. In: Evolution
+14. Wright S. Evolution and the Genetics of Populations. In: Evolution
 and the genetics of populations. Chicago: University of Chicago Press;
@@
 \hypertarget{ref-Wagner2011}{}
-14. Wagner GP, Zhang J. The pleiotropic structure of the
+15. Wagner GP, Zhang J. The pleiotropic structure of the
 genotype--phenotype map: the evolvability of complex organisms. Nature
@@
 \hypertarget{ref-Hill2012a}{}
-15. Hill WG, Zhang X-S. Assessing pleiotropy and its evolutionary
+16. Hill WG, Zhang X-S. Assessing pleiotropy and its evolutionary
 consequences: pleiotropy is not necessarily limited, nor need it hinder
@@
 \hypertarget{ref-Hartwig2016}{}
-16. Hartwig FP, Davies NM, Hemani G, Davey Smith G. Two-sample Mendelian
+17. Hartwig FP, Davies NM, Hemani G, Davey Smith G. Two-sample Mendelian
 randomization: avoiding the downsides of a powerful, widely applicable
@@
 \hypertarget{ref-Xu2017}{}
-17. Xu L, Lin SL, Schooling CM. A Mendelian randomization study of the
+18. Xu L, Lin SL, Schooling CM. A Mendelian randomization study of the
 effect of calcium on coronary artery disease, myocardial infarction and
@@
 \hypertarget{ref-Larsson2017}{}
-18. Larsson SC, Burgess S, Michaëlsson K, TB H, WH C, Y P. Association
+19. Larsson SC, Burgess S, Michaëlsson K, TB H, WH C, Y P. Association
 of Genetic Variants Related to Serum Calcium Levels With Coronary Artery
@@
 \hypertarget{ref-Prins2016}{}
-19. Prins BP, Abbasi A, Wong A, Vaez A, Nolte I, Franceschini N, et al.
+20. Prins BP, Abbasi A, Wong A, Vaez A, Nolte I, Franceschini N, et al.
 Investigating the Causal Relationship of C-Reactive Protein with 32
@@
 \hypertarget{ref-Inoshita2016}{}
-20. Inoshita M, Numata S, Tajima A, Kinoshita M, Umehara H, Nakataki M,
+21. Inoshita M, Numata S, Tajima A, Kinoshita M, Umehara H, Nakataki M,
 et al. A significant causal association between C-reactive protein
@@
 
+\hypertarget{ref-Solovieff2013}{}
+22. Solovieff N, Cotsapas C, Lee PH, Purcell SM, Smoller JW. Pleiotropy
+in complex traits: challenges and strategies. Nat Rev Genet
+{[}Internet{]}. 2013 Jun;14(7):483--95. Available from:
+\href{http://www.nature.com/doifinder/10.1038/nrg3461\%20http://dx.doi.org/10.1038/nrg3461\%2010.1038/nrg3461}{http://www.nature.com/doifinder/10.1038/nrg3461 http://dx.doi.org/10.1038/nrg3461 10.1038/nrg3461}
+
+\hypertarget{ref-Hodgkin1998}{}
+23. Hodgkin J. Seven types of pleiotropy. The International journal of
+developmental biology {[}Internet{]}. 1998;42(3):501--5. Available from:
+\url{http://www.ncbi.nlm.nih.gov/pubmed/9654038}
+
+\hypertarget{ref-Paaby2013}{}
+24. Paaby AB, Rockman MV. The many faces of pleiotropy {[}Internet{]}.
+Vol. 29. 2013. pp. 66--73. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/23140989\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC3558540\%20http://linkinghub.elsevier.com/retrieve/pii/S0168952512001692}{http://www.ncbi.nlm.nih.gov/pubmed/23140989 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC3558540 http://linkinghub.elsevier.com/retrieve/pii/S0168952512001692}
+
+\hypertarget{ref-Hu2016}{}
+25. Hu JX, Thomas CE, Brunak S. Network biology concepts in complex
+disease comorbidities. Nature reviews Genetics {[}Internet{]}.
+2016;17(10):615--29. Available from:
+\href{http://www.nature.com/doifinder/10.1038/nrg.2016.87\%7B/\%\%7D5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}{http://www.nature.com/doifinder/10.1038/nrg.2016.87\{\textbackslash{}\%\}5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}
+
+\hypertarget{ref-Burgess2015b}{}
+26. Burgess S, Scott RA, Timpson NJ, Davey Smith G, Thompson SG, EPIC-
+InterAct Consortium. Using published data in Mendelian randomization: a
+blueprint for efficient identification of causal risk factors. European
+Journal of Epidemiology {[}Internet{]}. 2015 Jul;30(7):543--52.
+Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/25773750\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4516908\%20http://link.springer.com/10.1007/s10654-015-0011-z}{http://www.ncbi.nlm.nih.gov/pubmed/25773750 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4516908 http://link.springer.com/10.1007/s10654-015-0011-z}
+
 \hypertarget{ref-Rucker2011}{}
-21. Rucker G, Schwarzer G, Carpenter JR, Binder H, Schumacher M.
+27. Rucker G, Schwarzer G, Carpenter JR, Binder H, Schumacher M.
 Treatment-effect estimates adjusted for small-study effects via a limit
@@
 \hypertarget{ref-Han2008}{}
-22. Han C. Detecting invalid instruments using L 1-GMM. Economics
+28. Han C. Detecting invalid instruments using L 1-GMM. Economics
 Letters. 2008;101(3):285--7.
@@
 \hypertarget{ref-Kang2016}{}
-23. Kang H, Zhang A, Cai TT, Small DS. Instrumental Variables Estimation
+29. Kang H, Zhang A, Cai TT, Small DS. Instrumental Variables Estimation
 With Some Invalid Instruments and its Application to Mendelian
@@
 \hypertarget{ref-Steiger1980}{}
-24. Steiger JH. Tests for comparing elements of a correlation matrix.
+30. Steiger JH. Tests for comparing elements of a correlation matrix.
 Psychological Bulletin. 1980;87(2):245--51.
@@
 \hypertarget{ref-Corbin2016}{}
-25. Corbin LJ, Richmond RC, Wade KH, Burgess S, Bowden J, Smith GD, et
+31. Corbin LJ, Richmond RC, Wade KH, Burgess S, Bowden J, Smith GD, et
 al. BMI as a Modifiable Risk Factor for Type 2 Diabetes: Refining and
@@
 \hypertarget{ref-white2016plasma}{}
-26. White J, Sofat R, Hemani G, Shah T, Engmann J, Dale C, et al. Plasma
+32. White J, Sofat R, Hemani G, Shah T, Engmann J, Dale C, et al. Plasma
 urate concentration and risk of coronary heart disease: a Mendelian
@@
 \hypertarget{ref-Jordan1994}{}
-27. Jordan MI, Jacobs RA. Hierarchical Mixtures of Experts and the EM
+33. Jordan MI, Jacobs RA. Hierarchical Mixtures of Experts and the EM
 Algorithm. Neural Computation {[}Internet{]}. 1994 Mar;6(2):181--214.
@@
 
+\hypertarget{ref-brazdil2008metalearning}{}
+34. Brazdil P, Carrier CG, Soares C, Vilalta R. Metalearning:
+Applications to data mining. Springer Science \& Business Media; 2008.
+
+\hypertarget{ref-smith2009cross}{}
+35. Smith-Miles KA. Cross-disciplinary perspectives on meta-learning for
+algorithm selection. ACM Computing Surveys (CSUR). 2009;41(1):6.
+
+\hypertarget{ref-lemke2015metalearning}{}
+36. Lemke C, Budka M, Gabrys B. Metalearning: a survey of trends and
+technologies. Artificial intelligence review. 2015;44(1):117--30.
+
 \hypertarget{ref-Shin2014}{}
-28. Shin S-Y, Fauman EB, Petersen A-K, Krumsiek J, Santos R, Huang J, et
+37. Shin S-Y, Fauman EB, Petersen A-K, Krumsiek J, Santos R, Huang J, et
 al. An atlas of genetic influences on human blood metabolites. Nature
@@
 \hypertarget{ref-Kettunen2016}{}
-29. Kettunen J, Demirkan A, Wurtz P, Draisma HHM, Haller T, Rawal R, et
+38. Kettunen J, Demirkan A, Wurtz P, Draisma HHM, Haller T, Rawal R, et
 al. Genome-wide study for circulating metabolites identifies 62 loci and
@@
 \hypertarget{ref-Sun2017}{}
-30. Sun BB, Maranville JC, Peters JE, Stacey D, Staley JR, Blackshaw J,
+39. Sun BB, Maranville JC, Peters JE, Stacey D, Staley JR, Blackshaw J,
 et al. Consequences Of Natural Perturbations In The Human Plasma
@@
 \hypertarget{ref-Willer2013}{}
-31. Willer CJ, Schmidt EM, Sengupta S, Peloso GM, Gustafsson S, Kanoni
+40. Willer CJ, Schmidt EM, Sengupta S, Peloso GM, Gustafsson S, Kanoni
 S, et al. Discovery and refinement of loci associated with lipid levels.
@@
 \hypertarget{ref-Okbay2016}{}
-32. Okbay A, Beauchamp JP, Fontana MA, Lee JJ, Pers TH, Rietveld CA, et
+41. Okbay A, Beauchamp JP, Fontana MA, Lee JJ, Pers TH, Rietveld CA, et
 al. Genome-wide association study identifies 74 loci associated with
@@
 \hypertarget{ref-Anderson2017}{}
-33. Anderson E, Wade KH, Hemani G, Bowden J, Korologou-Linden R, Davey
+42. Anderson E, Wade KH, Hemani G, Bowden J, Korologou-Linden R, Davey
 Smith G, et al. The Causal Effect Of Educational Attainment On
@@
 \hypertarget{ref-kotthoff2016auto}{}
-34. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
+43. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
 2.0: Automatic model selection and hyperparameter optimization in WEKA.
@@
 \hypertarget{ref-Zhu2016}{}
-35. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+44. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
 Integration of summary data from GWAS and eQTL studies predicts complex
@@
 \hypertarget{ref-Richardson2017}{}
-36. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
+45. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
 CL, et al. Causal epigenome-wide association study identifies CpG sites
@@
 \hypertarget{ref-Wang2017}{}
-37. Wang L, Michoel T. Efficient And Accurate Causal Inference With
+46. Wang L, Michoel T. Efficient And Accurate Causal Inference With
 Hidden Confounders From Genome-Transcriptome Variation Data. bioRxiv
@@
 
+\hypertarget{ref-Aalen2015}{}
+47. Aalen OO, Valberg M, Grotmol T, Tretli S. Understanding variation in
+disease risk: the elusive concept of frailty. International journal of
+epidemiology {[}Internet{]}. 2015 Aug;44(4):1408--21. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/25501685\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4588855}{http://www.ncbi.nlm.nih.gov/pubmed/25501685 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4588855}
+
 \hypertarget{ref-Noyce2017}{}
-38. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
+48. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
 E, et al. Estimating the causal influence of body mass index on risk of
@@
 \hypertarget{ref-Burgess2014a}{}
-39. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+49. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 
+\hypertarget{ref-Juul2003}{}
+50. Juul K, Tybjaerg-Hansen A, Marklund S, Heegaard NHH, Steffensen R,
+Sillesen H, et al. Genetically Reduced Antioxidative Protection and
+Increased Ischemic Heart Disease Risk: The Copenhagen City Heart Study.
+Circulation {[}Internet{]}. 2003 Dec;109(1):59--65. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/14662715\%20http://circ.ahajournals.org/cgi/doi/10.1161/01.CIR.0000105720.28086.6C}{http://www.ncbi.nlm.nih.gov/pubmed/14662715 http://circ.ahajournals.org/cgi/doi/10.1161/01.CIR.0000105720.28086.6C}
+
 \hypertarget{ref-Staley2016}{}
-40. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
+51. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
 al. PhenoScanner: a database of human genotype--phenotype associations.
@@
 \hypertarget{ref-Lee2013c}{}
-41. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+52. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 
-\hypertarget{ref-brazdil2008metalearning}{}
-42. Brazdil P, Carrier CG, Soares C, Vilalta R. Metalearning:
-Applications to data mining. Springer Science \& Business Media; 2008.
-
-\hypertarget{ref-smith2009cross}{}
-43. Smith-Miles KA. Cross-disciplinary perspectives on meta-learning for
-algorithm selection. ACM Computing Surveys (CSUR). 2009;41(1):6.
-
-\hypertarget{ref-lemke2015metalearning}{}
-44. Lemke C, Budka M, Gabrys B. Metalearning: a survey of trends and
-technologies. Artificial intelligence review. 2015;44(1):117--30.
-
 \hypertarget{ref-Liaw2002}{}
-45. Liaw A, Wiener M. Classification and Regression by randomForest. R
+53. Liaw A, Wiener M. Classification and Regression by randomForest. R
 News {[}Internet{]}. 2002;2(3):18--22. Available from:

### 2017-08-03 docs/manuscript.rmd
updated

@@
 ---
-title: Learning pleiotropic signatures to construct a causal map of the human phenome
+title: A putative causal map of natural variation across the human phenome
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype.
-Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy, and many methods have been developed in an attempt to address this, each assuming a different model of horizontal pleiotropy.
-Here we show that across a range of horizontal pleiotropy models instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow.
-We have integrated these methods into a mixture-of-experts machine learning approach (MR-MoE 1.0) that predicts the most appropriate model to use for any specific causal analysis, improving on both power and false discovery rates.
-Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a working draft of the causal map of the human phenome, comprising the evaluation of causality between over a million pairs of traits.
+Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy.
+We demonstrate that across a range of pleiotropy models, instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow.
+Methods have been developed in an attempt to protect MR from different patterns of horizontal pleiotropy, and here we have designed a mixture-of-experts machine learning framework (MR-MoE 1.0) that predicts the most appropriate model to use for any specific causal analysis, improving on both power and false discovery rates.
+Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, comprising the evaluation of causality between over a million pairs of traits.
 Almost 90% of causal estimates indicated some level of horizontal pleiotropy.
-The causal estimates are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)).
+The causal estimates are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)), and we use it here to highlight the numerous challenges that remain in automated causal inference.
 
@@
 
-Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure [@DaveySmithHemani2014].
+Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure (Box 1).
 The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017].
-It is considered best practice to report estimates from a range of methods as a sensitivity analysis when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions.
-However it is also desirable to consolidate across methods under some circumstances.
+It is considered best practice to triangulate estimates from a range of MR methods (and other experimental designs) when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions [@Lawlor2017].
+However there are circumstances in which it is desirable to consolidate across many methods to obtain a single MR estimate.
 First, it could be of interest to screen links between thousands of traits for being potential causal associations, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient.
@@
 Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree and nature (e.g.
-horizontal versus vertical) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate.
+horizontal versus vertical, see Box 1) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate.
 Third, if different methods disagree it is not possible to know which is correct because the true nature of HP exhibited by the instruments is not known.
@@
 In this paper we introduce two innovations towards improving the reliability of MR estimates.
-First, we introduce an approach to discard genetic variants that are likely to be invalid.
+First, we present an approach to discard genetic variants that are likely to be invalid.
 Second, we hypothesised that characteristics of the summary data could indicate which method would be most reliable, and we introduce new machine learning approaches that attempt to automate both instrument and method selection.
@@
 
-### Box 1: Models of pleiotropy
+### Box 1: Pleiotropic mechanisms
 
-In this paper we separate pleiotropy into two forms for simplicity - vertical and horizontal.
+Pleiotropy is the phenomenon whereby a single genetic variant influences multiple phenotypes.
+The mechanisms behind it can arise through many different mechanisms, for example a single variant influences multiple genes; a single gene generates multiple gene products; a single gene product has multiple cellular functions; or a single cellular function influences multiple phenotypes [@Hu2016].
+How we classify the pleiotropic mechanism depends on the application, and in this paper we discuss pleiotropy in two forms - vertical and horizontal.
 Vertical pleiotropy has also been known as mediated pleiotropy [@Solovieff2013], because the genetic variant influences one trait, which in turn influences a second trait.
-It has also been termed Type II pleiotropy [@Wagner2011] and secondary pleiotropy [@Hodgkin1998].
+It has also been termed, in some contexts, Type II pleiotropy [@Wagner2011] and secondary pleiotropy [@Hodgkin1998].
+It is this form of pleiotropy that MR assumes to make causal inference between two traits.
 
-By contrast, horizontal pleiotropy can manifest in a myriad of different ways, but the guiding principle is that the genetic variant influences two different phenotypes through independent pathways.
-Horizontal pleiotropy has also been termed biological pleiotropy [@Solovieff2013], Type I pleiotropy [@Wagner2011], developmental pleiotropy or selectional pleiotropy [@Paaby2013].
-From a MR statistical perspective, one can be more specific and state that horizontal pleiotropy occurs where a SNP influences an outcome trait through some pathway other than the exposure trait.
+By contrast, horizontal pleiotropy can manifest in a myriad of different ways, but the basic principle is that the genetic variant influences two different phenotypes through independent pathways.
+Depending on context, horizontal pleiotropy can also be termed biological pleiotropy [@Solovieff2013], Type I pleiotropy [@Wagner2011], developmental pleiotropy or selectional pleiotropy [@Paaby2013].
+From the perspective of MR, horizontal pleiotropy arises when a SNP influences an outcome trait through some pathway other than the exposure trait.
 
-For illustration, Hu et al (2016) (Figure 3) enumerate the biological mechanisms that can lead to pleiotropy: a single variant influences multiple genes; a single gene generates multiple gene products; a single gene product has multiple cellular functions; or a single cellular function influences multiple phenotypes [@Hu2016].
-Whether we consider each of these mechanisms to be vertical pleiotropy (advantageous for MR) or horizontal pleiotropy (problematic for MR) depends on which phenotypes are being interrogated.
-For example, consider a scenario where a SNP influences the expression levels of two genes, and the first gene influences trait A while the second gene influences trait B.
+A genetic variant can exhibit both horizontal and vertical pleiotropy in different MR analyses.
+For example, consider a scenario where a SNP influences the expression levels of two genes (X and Y), and gene X influences trait A while gene Y influences trait B.
 Then:
 
-- The genetic variant is exhibiting a vertical pleiotropic effect on trait A through the first gene expression level.
+- The genetic variant is exhibiting a vertical pleiotropic effect on trait A and the expression level of gene X.
 Here, all other assumptions being met, the genetic variant would be a valid instrument if used to estimate the causal effect of the gene expression level on trait A.
@@
 Here the genetic variant would be invalid if used to estimate the causal effect of trait A on trait B.
+
+![](images/pleiotropy.png)
+
+Adapted from Figure 3 in Hu et al (2016).
 
@@
 
-The simplest approach to selecting instruments for performing MR is to take SNPs that have been declared significant in the published GWAS for the exposure.
+The simplest approach to selecting instruments is to take SNPs that have been declared significant in the published GWAS for the exposure.
 This typically involves obtaining SNPs that surpass $p < 5 \times 10^{-8}$, using clumping to obtain independent SNPs, and then replicating in an independent sample.
@@
 
+Motivated by a recent flurry of activity in MR methodology development, we have devised a machine learning approach, MR-MoE, that seeks to predict the performance of each MR method, selecting the one which is most likely to maximise power whilst minimising FDR for a specific summary-set.
+The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy.
+As such, our method contributes to the field of meta-learning but also to the rapidly growing field of automatic model selection and hyper-parameter optimization in machine learning [@kotthoff2016auto].
+We applied the method to the curated set of GWAS summary data present in MR-Base [@Hemani2016], generating several million MR estimates.
+MR-MoE selected a method that indicated some pattern of horizontal pleiotropy for almost 90% of cases, reinforcing the notion that it is the rule rather than the exception.
+
 The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones.
 Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception.
-Following a recent flurry in development of MR methodology, we have devised a machine learning approach, MR-MoE, that seeks to predict the performance of each MR method, selecting the one which is most likely to maximise power whilst minimising FDR for a specific summary-set.
-The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy.
-As such, our method contributes to the field of meta-learning but also to the rapidly growing field of automatic model selection and hyper-parameter optimization in machine learning [@kotthoff2016auto].
+Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criterion.
+We illustrated how confounders and reverse causal associations can easily lead to invalid instrument selection in MR and that most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.
 
-Crucially however, we observed in our simulations that none of the available methods were conservative and all, including MR-MoE, had FDR above 0.2.
-Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criterion.
-We illustrated showed how confounders and reverse causal associations can easily lead to invalid instrument selection in MR and that most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.
+Methods have been introduced recently that seek to improve MR estimates through invalid instrument detection and removal [@Corbin2016; @Verbanck2017; @Zhu2017].
+These attack the problem from the perspective that variants which have a relatively large influence on the causal estimate, or provide substantially different estimates from the other variants, are likely to be exhibiting HP, and their removal from the analysis may be warranted.
+A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones.
+This could arise, for example, for the *CRP* variant instrumenting C-reactive protein levels [@Prins2016], because CRP itself is downstream of many cellular functions and most GWAS signals will not be specific for the trait.
+Steiger filtering has attempted to sidestep this issue in removing invalid instruments, seeking to detect those instruments which have associations with the outcome that are too large to be consistent with the exposure causing the outcome.
+We acknowledge that unmeasured confounding and measurement error can introduce problems with Steiger filtering [@Hemani2017], but note that our simulations indicate that all individual MR methods improve in performance following its application.
 
-We applied the method to the curated set of GWAS summary data present in MR-Base [@Hemani2016], generating several million MR estimates.
-MR-MoE selected a method that indicated some pattern of horizontal pleiotropy for almost 90% of cases, reinforcing the notion that it is the rule rather than the exception.
-Based on our simulations we expect that those associations that we reported to be 'significant' are liable to high type 1 error rates.
-
-One advantage to generating a method that consolidates across many methods is that it can prevent cherry picking the most appealing method based on their results, but in general we strongly advocate the use of reporting multiple MR methods to appraise any specific the impact of HP on a causal estimate.
 In addition to HP, there are many other limitations that prevent MR from being a panacea for causal inference.
@@
 Though methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017], separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework though the use of mediation-based approaches may be informative in the one-sample setting [@Wang2017; @Hemani2017].
-Other problems can also manifest.
+
+Other problems, in addition to HP, can also manifest.
 Frailty effects [@Aalen2015] could induce associations for late-onset traits, and often detailed modelling for how this can influence MR is warranted [@Noyce2017; @Anderson2017].
 Genetic variants could strongly relate to several phenotypes, for example SNPs that influence LDL cholesterol are also likely to influence other lipid fractions, making it difficult to ascertain which amongst them is the true causal exposure [@Burgess2014a].
-The measured feature for which genetic associations are known could relate in complicated ways to the biological entities that is truly causal [@DaveySmithHemani2014], for example higher circulating levels of the natural antioxidant extracellular superoxide dismutase (EC-SOD) appears to causally relate to higher risk of coronary heart disease [@Juul2003] because the phenotypic measure relates inversely to the EC-SOD levels in arterial walls, hence *in situ* activity is lower.
+We see this issue manifesting in Figure 4, where there appear to be causal relationships between different lipid fractions.
+The measured feature for which genetic associations are known could relate in complicated ways to the biological entities that is truly causal [@DaveySmithHemani2014], for example higher circulating levels of the natural antioxidant extracellular superoxide dismutase (EC-SOD) appears to causally relate to higher risk of coronary heart disease [@Juul2003] because the phenotypic measure relates inversely to the EC-SOD levels in arterial walls, hence *in situ* activity is lower [@DaveySmithHemani2014].
+Dynastic effects can also confound MR estimates, arising when the exposure trait in a previous generation influences the environment of the current generation.
+Here, the SNPs being used as instruments for the exposure will be associated with a potential confounder in the analysis [@VonHinke2016].
+The associations depicted in Figure 4 involving educational attainment may be susceptible to bias arising from this mechanism, and protecting against it will likely require alternative study designs, for example by using sibling pairs.
 
@@
 This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance.
+One advantage to generating a method that consolidates across many methods is that it can prevent cherry picking the most appealing method based on their results.
 The extent to which a machine-based method selector incurs higher type 1 error rates more than a human selector is not clear.
 Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to other MR methods that do not suffer from the issue of data driven model selection.
+
+In light of these issues, we emphasise that we do not consider the reporting of MR-MoE (or any other single MR method) to be seen as performing a Mendelian randomization study.
+Rather, it can be used to motivate further detailed follow up which should include sensitivity analyses [@Hemani2016], incorporating biological knowledge regarding instruments, and triangulation with other sources of evidence or experimental designs [@Lawlor2017].
+Demanding a rigorous follow up of putative associations is also essential for avoiding issues that may arise due to cherry picking of MR results.
+This is especially important when a large repository of putative associations, such as MR-EvE, is made publicly available.
 
@@
 Despite the limitations described above, the construction of a causal graph of 'everything versus everything' does have appeal.
-First, though causality is not guaranteed by MR, it can still be highly informative for confirming or negating specific hypotheses.
+First, though causality is not guaranteed by MR, it can still be highly informative for supporting or negating specific hypotheses.
+Indeed, a well-powered MR that negates a hypothesis is likely less fallible to the pitfalls described above than using MR to confirm a positive association [@Vanderweele2014].
 Second, it paves the way for new approaches to search for novel putative associations and improve reliability in single analyses.
@@
 
-This work was supported by the UK Medical Research Council (MC_UU_12013/1).
+This work was supported by the UK Medical Research Council (MC_UU_12013/1, MC_UU_12013/8, MR/N501906/1); Cancer Research UK (C52724/A20138, C18281/A19169); and the SPHERE Interdisciplinary Research Collaboration funded by the UK Engineering and Physical Sciences Research Council under grant EP/K031910/1
 

### 2017-08-03 docs/manuscript.tex
updated

@@
 \hypersetup{unicode=true,
-            pdftitle={Learning pleiotropic signatures to construct a causal map of the human phenome},
+            pdftitle={A putative causal map of natural variation across the human phenome},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{Learning pleiotropic signatures to construct a causal map of the human
-phenome}
+  \title{A putative causal map of natural variation across the human phenome}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
   \postdate{\par}
-  \date{02 August 2017}
+  \date{03 August 2017}
+
+\usepackage{amsmath}
 
@@
 an exposure phenotype. Issues arise when this assumption is violated due
-to SNPs exhibiting horizontal pleiotropy, and many methods have been
-developed in an attempt to address this, each assuming a different model
-of horizontal pleiotropy. Here we show that across a range of horizontal
-pleiotropy models instrument selection will be increasingly liable to
-selecting invalid instruments as GWAS sample sizes continue to grow. We
-have integrated these methods into a mixture-of-experts machine learning
-approach (MR-MoE 1.0) that predicts the most appropriate model to use
-for any specific causal analysis, improving on both power and false
-discovery rates. Using the approach, we systematically estimated the
-causal effects amongst 2407 phenotypes, generating a working draft of
-the causal map of the human phenome, comprising the evaluation of
-causality between over a million pairs of traits. Almost 90\% of causal
-estimates indicated some level of horizontal pleiotropy. The causal
-estimates are organised into a publicly available graph database
-(\url{http://eve.mrbase.org}).
+to SNPs exhibiting horizontal pleiotropy. We demonstrate that across a
+range of pleiotropy models, instrument selection will be increasingly
+liable to selecting invalid instruments as GWAS sample sizes continue to
+grow. Methods have been developed in an attempt to protect MR from
+different patterns of horizontal pleiotropy, and here we have designed a
+mixture-of-experts machine learning framework (MR-MoE 1.0) that predicts
+the most appropriate model to use for any specific causal analysis,
+improving on both power and false discovery rates. Using the approach,
+we systematically estimated the causal effects amongst 2407 phenotypes,
+comprising the evaluation of causality between over a million pairs of
+traits. Almost 90\% of causal estimates indicated some level of
+horizontal pleiotropy. The causal estimates are organised into a
+publicly available graph database (\url{http://eve.mrbase.org}), and we
+use it here to highlight the numerous challenges that remain in
+automated causal inference.
 
@@
 the genetic instruments exhibit horizontal pleiotropy (HP), where they
-influence the outcome through a pathway other than the exposure (2). The
-extent of this phenomenon is not to be understated, and many methods
+influence the outcome through a pathway other than the exposure (Box 1).
+The extent of this phenomenon is not to be understated, and many methods
 have been developed that attempt to reliably obtain unbiased causal
 estimates under specific models of HP (4--7,9,13). It is considered best
-practice to report estimates from a range of methods as a sensitivity
-analysis when presenting causal estimates so as to assess the potential
-for different horizontal pleiotropy models and other violations of MR
-assumptions. However it is also desirable to consolidate across methods
-under some circumstances. First, it could be of interest to screen links
+practice to triangulate estimates from a range of MR methods (and other
+experimental designs) when presenting causal estimates so as to assess
+the potential for different horizontal pleiotropy models and other
+violations of MR assumptions (14). However there are circumstances in
+which it is desirable to consolidate across many methods to obtain a
+single MR estimate. First, it could be of interest to screen links
 between thousands of traits for being potential causal associations, in
@@
 low true positive rates in the presence of HP compared to other methods.
-Pleiotropy has been hypothesised to be universal (14), though the degree
-and nature (e.g.~horizontal versus vertical) to which this may be the
-case is contested (15,16), hence defaulting to the IVW method in the
-first instance and using other methods as sensitivity analyses may not
-be appropriate. Third, if different methods disagree it is not possible
-to know which is correct because the true nature of HP exhibited by the
-instruments is not known. Fourth, the available methods do not cover all
-possible models of HP, and therefore an automated method for instrument
-selection may be necessary.
+Pleiotropy has been hypothesised to be universal (15), though the degree
+and nature (e.g.~horizontal versus vertical, see Box 1) to which this
+may be the case is contested (16,17), hence defaulting to the IVW method
+in the first instance and using other methods as sensitivity analyses
+may not be appropriate. Third, if different methods disagree it is not
+possible to know which is correct because the true nature of HP
+exhibited by the instruments is not known. Fourth, the available methods
+do not cover all possible models of HP, and therefore an automated
+method for instrument selection may be necessary.
 
 In this paper we introduce two innovations towards improving the
-reliability of MR estimates. First, we introduce an approach to discard
+reliability of MR estimates. First, we present an approach to discard
 genetic variants that are likely to be invalid. Second, we hypothesised
@@
 new methods to construct a graph of millions of causal estimates.
-Motivated by the recent avalanche (17) of 2SMR publications, with
-similar (18,19) or contradicting (20,21) conclusions despite using the
-same data (17), we developed a graph database to represent these
+Motivated by the recent avalanche (18) of 2SMR publications, with
+similar (19,20) or contradicting (21,22) conclusions despite using the
+same data (18), we developed a graph database to represent these
 estimates in a consistent manner. We consider this to be a `working
@@
 
-\subsubsection{Box 1: Models of
-pleiotropy}\label{box-1-models-of-pleiotropy}
-
-In this paper we separate pleiotropy into two forms for simplicity -
-vertical and horizontal. Vertical pleiotropy has also been known as
-mediated pleiotropy (22), because the genetic variant influences one
-trait, which in turn influences a second trait. It has also been termed
-Type II pleiotropy (15) and secondary pleiotropy (23).
+\subsubsection{Box 1: Pleiotropic
+mechanisms}\label{box-1-pleiotropic-mechanisms}
+
+Pleiotropy is the phenomenon whereby a single genetic variant influences
+multiple phenotypes. The mechanisms behind it can arise through many
+different mechanisms, for example a single variant influences multiple
+genes; a single gene generates multiple gene products; a single gene
+product has multiple cellular functions; or a single cellular function
+influences multiple phenotypes (23). How we classify the pleiotropic
+mechanism depends on the application, and in this paper we discuss
+pleiotropy in two forms - vertical and horizontal. Vertical pleiotropy
+has also been known as mediated pleiotropy (24), because the genetic
+variant influences one trait, which in turn influences a second trait.
+It has also been termed, in some contexts, Type II pleiotropy (16) and
+secondary pleiotropy (25). It is this form of pleiotropy that MR assumes
+to make causal inference between two traits.
 
 By contrast, horizontal pleiotropy can manifest in a myriad of different
-ways, but the guiding principle is that the genetic variant influences
-two different phenotypes through independent pathways. Horizontal
-pleiotropy has also been termed biological pleiotropy (22), Type I
-pleiotropy (15), developmental pleiotropy or selectional pleiotropy
-(24). From a MR statistical perspective, one can be more specific and
-state that horizontal pleiotropy occurs where a SNP influences an
-outcome trait through some pathway other than the exposure trait.
-
-For illustration, Hu et al (2016) (Figure 3) enumerate the biological
-mechanisms that can lead to pleiotropy: a single variant influences
-multiple genes; a single gene generates multiple gene products; a single
-gene product has multiple cellular functions; or a single cellular
-function influences multiple phenotypes (25). Whether we consider each
-of these mechanisms to be vertical pleiotropy (advantageous for MR) or
-horizontal pleiotropy (problematic for MR) depends on which phenotypes
-are being interrogated. For example, consider a scenario where a SNP
-influences the expression levels of two genes, and the first gene
-influences trait A while the second gene influences trait B. Then:
+ways, but the basic principle is that the genetic variant influences two
+different phenotypes through independent pathways. Depending on context,
+horizontal pleiotropy can also be termed biological pleiotropy (24),
+Type I pleiotropy (16), developmental pleiotropy or selectional
+pleiotropy (26). From the perspective of MR, horizontal pleiotropy
+arises when a SNP influences an outcome trait through some pathway other
+than the exposure trait.
+
+A genetic variant can exhibit both horizontal and vertical pleiotropy in
+different MR analyses. For example, consider a scenario where a SNP
+influences the expression levels of two genes (X and Y), and gene X
+influences trait A while gene Y influences trait B. Then:
 
@@
   The genetic variant is exhibiting a vertical pleiotropic effect on
-  trait A through the first gene expression level. Here, all other
+  trait A and the expression level of gene X. Here, all other
   assumptions being met, the genetic variant would be a valid instrument
@@
 
+\includegraphics{images/pleiotropy.png}
+
+Adapted from Figure 3 in Hu et al (2016).
+
 \subsection{Methods}\label{methods}
@@
 The use of summary data in two-sample MR is described in detail
-elsewhere (12,26). A brief outline of the procedure is as follows.
+elsewhere (12,27). A brief outline of the procedure is as follows.
 First, genetic instruments for the exposure trait need to be identified
@@
 
-The Rucker framework (27), adapted to MR (6) uses estimates of
+The Rucker framework (28), adapted to MR (6) uses estimates of
 heterogeneity in the IVW and Egger frameworks to navigate between these
@@
 \textbf{Median-based methods:} An alternative approach is to take the
-median effect of all available instruments (5,28). This has the
+median effect of all available instruments (5,29). This has the
 advantage that only half the instruments need to be valid, and the
@@
 LASSO regression, have also been developed for the situation where at
-least half of the instruments are valid (29).
+least half of the instruments are valid (30).
 
@@
 
-The simplest approach to selecting instruments for performing MR is to
-take SNPs that have been declared significant in the published GWAS for
-the exposure. This typically involves obtaining SNPs that surpass
-\(p < 5 \times 10^{-8}\), using clumping to obtain independent SNPs, and
-then replicating in an independent sample. These results are often
-recorded in public GWAS catalogs. Alternatively the clumping procedure
-can be performed using complete summary data in MR-Base (12). Complete
-summary data refers to association results for all SNPs used in a GWAS
-i.e.~not only those passing statistical thresholds for significance. We
-call this the ``tophits'' strategy.
+The simplest approach to selecting instruments is to take SNPs that have
+been declared significant in the published GWAS for the exposure. This
+typically involves obtaining SNPs that surpass \(p < 5 \times 10^{-8}\),
+using clumping to obtain independent SNPs, and then replicating in an
+independent sample. These results are often recorded in public GWAS
+catalogs. Alternatively the clumping procedure can be performed using
+complete summary data in MR-Base (12). Complete summary data refers to
+association results for all SNPs used in a GWAS i.e.~not only those
+passing statistical thresholds for significance. We call this the
+``tophits'' strategy.
 
@@
 inexpensive and approximate method to identify variants that are likely
-to be invalid. Steiger's Z-test of correlated correlations (30) can be
+to be invalid. Steiger's Z-test of correlated correlations (31) can be
 used to formally test the extent to which the two correlations are
@@
 invalid instruments for the purposes of exclusion from MR analyses
-(9,13,31), based on the notion that outlying instruments are more likely
+(9,13,32), based on the notion that outlying instruments are more likely
 to be a source of HP. A potential drawback of this approach is that the
@@
 could arise, for example, for the \emph{CRP} variant instrumenting
-C-reactive protein levels (20), or the \emph{SLC2A9} variant influencing
-serum urate levels (32).
+C-reactive protein levels (21), or the \emph{SLC2A9} variant influencing
+serum urate levels (33).
 
@@
 the most appropriate model. To achieve this we use a ``mixture of
-experts'' machine learning approach (33), where each MR method is
+experts'' machine learning approach (34), where each MR method is
 considered to be an `expert', taking a summary-set as its input. An
@@
 It should be noted that in the original hierarchical mixture of experts
-approach of (33) the gating function and the experts share the same
+approach of (34) the gating function and the experts share the same
 input space and are trained simultaneously using
@@
 simulated data. As such our MR-MoE approach fits the framework of
-meta-learning (34--36) where machine learning methods are applied at the
+meta-learning (35--37) where machine learning methods are applied at the
 `meta-level' to analyse the results of machine learning experiments at
@@
 We applied our analysis using summary data for 2407 phenotypes,
-including 149 complex traits and diseases, 575 metabolites (37,38) and
-1683 plasma protein levels (39). For the protein levels only GWAS
+including 149 complex traits and diseases, 575 metabolites (38,39) and
+1683 plasma protein levels (40). For the protein levels only GWAS
 significant hits were available, so they could only be evaluated as
@@
 between all traits and LDL cholesterol (as measured in the GLGC
-consortium (40)), where the MoE result had a false discovery rate of
+consortium (41)), where the MoE result had a false discovery rate of
 0.05 using the MoE approach. This returned 287 putative associations,
@@
 proportion of these traits were metabolites, which are dominated by
-lipid fractions. Filtering to exclude the metabolomic studies ((37,38))
+lipid fractions. Filtering to exclude the metabolomic studies ((38,39))
 returned 27 associations with 23 traits (Figure 4a). There was a strong
@@
 We also performed a look up of associations that causally influence
-years of schooling (41), or that years of schooling itself causally
+years of schooling (42), or that years of schooling itself causally
 influences. Using a false-discovery rate of 0.05, 45 traits were
@@
 plausible, for example the influence of years of schooling on lower risk
-of Alzheimer's disease, which has recently been reported (42). However,
+of Alzheimer's disease, which has recently been reported (43). However,
 it is clear that there are serious limitations in relying on MR as a
@@
 
+Motivated by a recent flurry of activity in MR methodology development,
+we have devised a machine learning approach, MR-MoE, that seeks to
+predict the performance of each MR method, selecting the one which is
+most likely to maximise power whilst minimising FDR for a specific
+summary-set. The mixture of experts approach that we present here is
+trained using random forest decision trees applied to extensive and
+diverse simulations, and we demonstrate that it makes substantial
+performance improvements over any other single method in the presence or
+absence of extensive horizontal pleiotropy. As such, our method
+contributes to the field of meta-learning but also to the rapidly
+growing field of automatic model selection and hyper-parameter
+optimization in machine learning (44). We applied the method to the
+curated set of GWAS summary data present in MR-Base (12), generating
+several million MR estimates. MR-MoE selected a method that indicated
+some pattern of horizontal pleiotropy for almost 90\% of cases,
+reinforcing the notion that it is the rule rather than the exception.
+
 The trend of increasing GWAS sample sizes continues, but while the
@@
 to be used in practice, therefore, must consider horizontal pleiotropy
-to be the rule rather than the exception. Following a recent flurry in
-development of MR methodology, we have devised a machine learning
-approach, MR-MoE, that seeks to predict the performance of each MR
-method, selecting the one which is most likely to maximise power whilst
-minimising FDR for a specific summary-set. The mixture of experts
-approach that we present here is trained using random forest decision
-trees applied to extensive and diverse simulations, and we demonstrate
-that it makes substantial performance improvements over any other single
-method in the presence or absence of extensive horizontal pleiotropy. As
-such, our method contributes to the field of meta-learning but also to
-the rapidly growing field of automatic model selection and
-hyper-parameter optimization in machine learning (43).
-
-Crucially however, we observed in our simulations that none of the
-available methods were conservative and all, including MR-MoE, had FDR
-above 0.2. Instrumental variables in MR are typically chosen blind, with
-GWAS significance being the only criterion. We illustrated showed how
-confounders and reverse causal associations can easily lead to invalid
-instrument selection in MR and that most randomly generated patterns of
-horizontal pleiotropy cannot be adequately accounted for by any single
-MR method.
-
-We applied the method to the curated set of GWAS summary data present in
-MR-Base (12), generating several million MR estimates. MR-MoE selected a
-method that indicated some pattern of horizontal pleiotropy for almost
-90\% of cases, reinforcing the notion that it is the rule rather than
-the exception. Based on our simulations we expect that those
-associations that we reported to be `significant' are liable to high
-type 1 error rates.
-
-One advantage to generating a method that consolidates across many
-methods is that it can prevent cherry picking the most appealing method
-based on their results, but in general we strongly advocate the use of
-reporting multiple MR methods to appraise any specific the impact of HP
-on a causal estimate. In addition to HP, there are many other
-limitations that prevent MR from being a panacea for causal inference.
-Many of the associations are biologically impossible, for example where
-early life phenotypes appear to be influenced by later stage phenotypes.
-Such associations can be explained statistically because the instruments
-proxy the liability for that particular phenotype, not that phenotype's
-observed manifestation. Though these associations can be informative,
-their interpretations as causal relationships are far from clear.
-Similarly, often disease traits appear to causally relate to other
-phenotypes, but GWAS of disease is typically performed on the liability
-scale, hence the causal estimate reflects not the presence or absence of
-disease, but the underlying risk of disease. Again, interpretation of
-such associations can be problematic due to the instruments proxying
-disease liability and not diseae outcomes.
+to be the rule rather than the exception. Instrumental variables in MR
+are typically chosen blind, with GWAS significance being the only
+criterion. We illustrated how confounders and reverse causal
+associations can easily lead to invalid instrument selection in MR and
+that most randomly generated patterns of horizontal pleiotropy cannot be
+adequately accounted for by any single MR method.
+
+Methods have been introduced recently that seek to improve MR estimates
+through invalid instrument detection and removal (9,13,32). These attack
+the problem from the perspective that variants which have a relatively
+large influence on the causal estimate, or provide substantially
+different estimates from the other variants, are likely to be exhibiting
+HP, and their removal from the analysis may be warranted. A potential
+drawback of this approach is that the outlier SNPs in a summary-set
+might be the only reliable ones. This could arise, for example, for the
+\emph{CRP} variant instrumenting C-reactive protein levels (21), because
+CRP itself is downstream of many cellular functions and most GWAS
+signals will not be specific for the trait. Steiger filtering has
+attempted to sidestep this issue in removing invalid instruments,
+seeking to detect those instruments which have associations with the
+outcome that are too large to be consistent with the exposure causing
+the outcome. We acknowledge that unmeasured confounding and measurement
+error can introduce problems with Steiger filtering (8), but note that
+our simulations indicate that all individual MR methods improve in
+performance following its application.
+
+In addition to HP, there are many other limitations that prevent MR from
+being a panacea for causal inference. Many of the associations are
+biologically impossible, for example where early life phenotypes appear
+to be influenced by later stage phenotypes. Such associations can be
+explained statistically because the instruments proxy the liability for
+that particular phenotype, not that phenotype's observed manifestation.
+Though these associations can be informative, their interpretations as
+causal relationships are far from clear. Similarly, often disease traits
+appear to causally relate to other phenotypes, but GWAS of disease is
+typically performed on the liability scale, hence the causal estimate
+reflects not the presence or absence of disease, but the underlying risk
+of disease. Again, interpretation of such associations can be
+problematic due to the instruments proxying disease liability and not
+diseae outcomes.
 
@@
 reverse cause, pleiotropy and multiple causal variants in the same
-region (44,45), separating vertical from horizontal pleiotropy with a
+region (45,46), separating vertical from horizontal pleiotropy with a
 single instrument is not yet possible in the two-sample MR framework
 though the use of mediation-based approaches may be informative in the
-one-sample setting (8,46). Other problems can also manifest. Frailty
-effects (47) could induce associations for late-onset traits, and often
-detailed modelling for how this can influence MR is warranted (42,48).
-Genetic variants could strongly relate to several phenotypes, for
-example SNPs that influence LDL cholesterol are also likely to influence
-other lipid fractions, making it difficult to ascertain which amongst
-them is the true causal exposure (49). The measured feature for which
-genetic associations are known could relate in complicated ways to the
-biological entities that is truly causal (2), for example higher
-circulating levels of the natural antioxidant extracellular superoxide
-dismutase (EC-SOD) appears to causally relate to higher risk of coronary
-heart disease (50) because the phenotypic measure relates inversely to
-the EC-SOD levels in arterial walls, hence \emph{in situ} activity is
-lower.
+one-sample setting (8,47).
+
+Other problems, in addition to HP, can also manifest. Frailty effects
+(48) could induce associations for late-onset traits, and often detailed
+modelling for how this can influence MR is warranted (43,49). Genetic
+variants could strongly relate to several phenotypes, for example SNPs
+that influence LDL cholesterol are also likely to influence other lipid
+fractions, making it difficult to ascertain which amongst them is the
+true causal exposure (50). We see this issue manifesting in Figure 4,
+where there appear to be causal relationships between different lipid
+fractions. The measured feature for which genetic associations are known
+could relate in complicated ways to the biological entities that is
+truly causal (2), for example higher circulating levels of the natural
+antioxidant extracellular superoxide dismutase (EC-SOD) appears to
+causally relate to higher risk of coronary heart disease (51) because
+the phenotypic measure relates inversely to the EC-SOD levels in
+arterial walls, hence \emph{in situ} activity is lower (2). Dynastic
+effects can also confound MR estimates, arising when the exposure trait
+in a previous generation influences the environment of the current
+generation. Here, the SNPs being used as instruments for the exposure
+will be associated with a potential confounder in the analysis (52). The
+associations depicted in Figure 4 involving educational attainment may
+be susceptible to bias arising from this mechanism, and protecting
+against it will likely require alternative study designs, for example by
+using sibling pairs.
 
@@
 to separate the information used for optimisation from the information
-used to predict performance. The extent to which a machine-based method
-selector incurs higher type 1 error rates more than a human selector is
-not clear. Though MR-MoE does exhibit higher type 1 error rates than are
-desirable, they still remain amongst the lowest compared to other MR
-methods that do not suffer from the issue of data driven model
-selection.
-
-The field of MR is evolving. The advent of GWAS databases (10,51) and
+used to predict performance. One advantage to generating a method that
+consolidates across many methods is that it can prevent cherry picking
+the most appealing method based on their results. The extent to which a
+machine-based method selector incurs higher type 1 error rates more than
+a human selector is not clear. Though MR-MoE does exhibit higher type 1
+error rates than are desirable, they still remain amongst the lowest
+compared to other MR methods that do not suffer from the issue of data
+driven model selection.
+
+In light of these issues, we emphasise that we do not consider the
+reporting of MR-MoE (or any other single MR method) to be seen as
+performing a Mendelian randomization study. Rather, it can be used to
+motivate further detailed follow up which should include sensitivity
+analyses (12), incorporating biological knowledge regarding instruments,
+and triangulation with other sources of evidence or experimental designs
+(14). Demanding a rigorous follow up of putative associations is also
+essential for avoiding issues that may arise due to cherry picking of MR
+results. This is especially important when a large repository of
+putative associations, such as MR-EvE, is made publicly available.
+
+The field of MR is evolving. The advent of GWAS databases (10,53) and
 automated 2SMR (12) has trivialised the analytical aspect of
@@
 versus everything' does have appeal. First, though causality is not
-guaranteed by MR, it can still be highly informative for confirming or
-negating specific hypotheses. Second, it paves the way for new
-approaches to search for novel putative associations and improve
-reliability in single analyses. The potential to exploit GWAS summary
-data within the properties of graph databases to aid with both of these
-endeavours could be transformative in biological research.
+guaranteed by MR, it can still be highly informative for supporting or
+negating specific hypotheses. Indeed, a well-powered MR that negates a
+hypothesis is likely less fallible to the pitfalls described above than
+using MR to confirm a positive association (54). Second, it paves the
+way for new approaches to search for novel putative associations and
+improve reliability in single analyses. The potential to exploit GWAS
+summary data within the properties of graph databases to aid with both
+of these endeavours could be transformative in biological research.
 
@@
 This work was supported by the UK Medical Research Council
-(MC\_UU\_12013/1).
+(MC\_UU\_12013/1, MC\_UU\_12013/8, MR/N501906/1); Cancer Research UK
+(C52724/A20138, C18281/A19169); and the SPHERE Interdisciplinary
+Research Collaboration funded by the UK Engineering and Physical
+Sciences Research Council under grant EP/K031910/1
 
@@
 b) Years of Schooling. In a) the nodes are filtered to not include those
-obtained from the metabolomic studies (37,38). The arrows denote causal
+obtained from the metabolomic studies (38,39). The arrows denote causal
 direction and the values on the arrows denote the causal effect
@@
 \(V_a = 2\beta^2f(1-f)\), where \(\beta\) is the log odds ratio and
-\(f\) is the allele frequency of the SNP in the population (52). \(f\)
+\(f\) is the allele frequency of the SNP in the population (55). \(f\)
 can be estimated using the allele frequency of the SNP in an ascertained
@@
 methods. The default settings for the \texttt{randomForest} package in R
-(53) were used to train the models. MR-MoE 1.0 is implemented in the
+(56) were used to train the models. MR-MoE 1.0 is implemented in the
 TwoSampleMR R package available at
@@
 
+\hypertarget{ref-Lawlor2017}{}
+14. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
+epidemiology. International Journal of Epidemiology {[}Internet{]}. 2017
+Jan;19(R1):dyw314. Available from:
+\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dyw314}
+
 \hypertarget{ref-Wright1968}{}
-14. Wright S. Evolution and the Genetics of Populations. In: Evolution
+15. Wright S. Evolution and the Genetics of Populations. In: Evolution
 and the genetics of populations. Chicago: University of Chicago Press;
@@
 \hypertarget{ref-Wagner2011}{}
-15. Wagner GP, Zhang J. The pleiotropic structure of the
+16. Wagner GP, Zhang J. The pleiotropic structure of the
 genotype--phenotype map: the evolvability of complex organisms. Nature
@@
 \hypertarget{ref-Hill2012a}{}
-16. Hill WG, Zhang X-S. Assessing pleiotropy and its evolutionary
+17. Hill WG, Zhang X-S. Assessing pleiotropy and its evolutionary
 consequences: pleiotropy is not necessarily limited, nor need it hinder
@@
 \hypertarget{ref-Hartwig2016}{}
-17. Hartwig FP, Davies NM, Hemani G, Davey Smith G. Two-sample Mendelian
+18. Hartwig FP, Davies NM, Hemani G, Davey Smith G. Two-sample Mendelian
 randomization: avoiding the downsides of a powerful, widely applicable
@@
 \hypertarget{ref-Xu2017}{}
-18. Xu L, Lin SL, Schooling CM. A Mendelian randomization study of the
+19. Xu L, Lin SL, Schooling CM. A Mendelian randomization study of the
 effect of calcium on coronary artery disease, myocardial infarction and
@@
 \hypertarget{ref-Larsson2017}{}
-19. Larsson SC, Burgess S, Michaëlsson K, TB H, WH C, Y P. Association
+20. Larsson SC, Burgess S, Michaëlsson K, TB H, WH C, Y P. Association
 of Genetic Variants Related to Serum Calcium Levels With Coronary Artery
@@
 \hypertarget{ref-Prins2016}{}
-20. Prins BP, Abbasi A, Wong A, Vaez A, Nolte I, Franceschini N, et al.
+21. Prins BP, Abbasi A, Wong A, Vaez A, Nolte I, Franceschini N, et al.
 Investigating the Causal Relationship of C-Reactive Protein with 32
@@
 \hypertarget{ref-Inoshita2016}{}
-21. Inoshita M, Numata S, Tajima A, Kinoshita M, Umehara H, Nakataki M,
+22. Inoshita M, Numata S, Tajima A, Kinoshita M, Umehara H, Nakataki M,
 et al. A significant causal association between C-reactive protein
@@
 
+\hypertarget{ref-Hu2016}{}
+23. Hu JX, Thomas CE, Brunak S. Network biology concepts in complex
+disease comorbidities. Nature reviews Genetics {[}Internet{]}.
+2016;17(10):615--29. Available from:
+\href{http://www.nature.com/doifinder/10.1038/nrg.2016.87\%7B/\%\%7D5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}{http://www.nature.com/doifinder/10.1038/nrg.2016.87\{\textbackslash{}\%\}5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}
+
 \hypertarget{ref-Solovieff2013}{}
-22. Solovieff N, Cotsapas C, Lee PH, Purcell SM, Smoller JW. Pleiotropy
+24. Solovieff N, Cotsapas C, Lee PH, Purcell SM, Smoller JW. Pleiotropy
 in complex traits: challenges and strategies. Nat Rev Genet
@@
 \hypertarget{ref-Hodgkin1998}{}
-23. Hodgkin J. Seven types of pleiotropy. The International journal of
+25. Hodgkin J. Seven types of pleiotropy. The International journal of
 developmental biology {[}Internet{]}. 1998;42(3):501--5. Available from:
@@
 \hypertarget{ref-Paaby2013}{}
-24. Paaby AB, Rockman MV. The many faces of pleiotropy {[}Internet{]}.
+26. Paaby AB, Rockman MV. The many faces of pleiotropy {[}Internet{]}.
 Vol. 29. 2013. pp. 66--73. Available from:
@@
 
-\hypertarget{ref-Hu2016}{}
-25. Hu JX, Thomas CE, Brunak S. Network biology concepts in complex
-disease comorbidities. Nature reviews Genetics {[}Internet{]}.
-2016;17(10):615--29. Available from:
-\href{http://www.nature.com/doifinder/10.1038/nrg.2016.87\%7B/\%\%7D5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}{http://www.nature.com/doifinder/10.1038/nrg.2016.87\{\textbackslash{}\%\}5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}
-
 \hypertarget{ref-Burgess2015b}{}
-26. Burgess S, Scott RA, Timpson NJ, Davey Smith G, Thompson SG, EPIC-
+27. Burgess S, Scott RA, Timpson NJ, Davey Smith G, Thompson SG, EPIC-
 InterAct Consortium. Using published data in Mendelian randomization: a
@@
 \hypertarget{ref-Rucker2011}{}
-27. Rucker G, Schwarzer G, Carpenter JR, Binder H, Schumacher M.
+28. Rucker G, Schwarzer G, Carpenter JR, Binder H, Schumacher M.
 Treatment-effect estimates adjusted for small-study effects via a limit
@@
 \hypertarget{ref-Han2008}{}
-28. Han C. Detecting invalid instruments using L 1-GMM. Economics
+29. Han C. Detecting invalid instruments using L 1-GMM. Economics
 Letters. 2008;101(3):285--7.
@@
 \hypertarget{ref-Kang2016}{}
-29. Kang H, Zhang A, Cai TT, Small DS. Instrumental Variables Estimation
+30. Kang H, Zhang A, Cai TT, Small DS. Instrumental Variables Estimation
 With Some Invalid Instruments and its Application to Mendelian
@@
 \hypertarget{ref-Steiger1980}{}
-30. Steiger JH. Tests for comparing elements of a correlation matrix.
+31. Steiger JH. Tests for comparing elements of a correlation matrix.
 Psychological Bulletin. 1980;87(2):245--51.
@@
 \hypertarget{ref-Corbin2016}{}
-31. Corbin LJ, Richmond RC, Wade KH, Burgess S, Bowden J, Smith GD, et
+32. Corbin LJ, Richmond RC, Wade KH, Burgess S, Bowden J, Smith GD, et
 al. BMI as a Modifiable Risk Factor for Type 2 Diabetes: Refining and
@@
 \hypertarget{ref-white2016plasma}{}
-32. White J, Sofat R, Hemani G, Shah T, Engmann J, Dale C, et al. Plasma
+33. White J, Sofat R, Hemani G, Shah T, Engmann J, Dale C, et al. Plasma
 urate concentration and risk of coronary heart disease: a Mendelian
@@
 \hypertarget{ref-Jordan1994}{}
-33. Jordan MI, Jacobs RA. Hierarchical Mixtures of Experts and the EM
+34. Jordan MI, Jacobs RA. Hierarchical Mixtures of Experts and the EM
 Algorithm. Neural Computation {[}Internet{]}. 1994 Mar;6(2):181--214.
@@
 \hypertarget{ref-brazdil2008metalearning}{}
-34. Brazdil P, Carrier CG, Soares C, Vilalta R. Metalearning:
+35. Brazdil P, Carrier CG, Soares C, Vilalta R. Metalearning:
 Applications to data mining. Springer Science \& Business Media; 2008.
@@
 \hypertarget{ref-smith2009cross}{}
-35. Smith-Miles KA. Cross-disciplinary perspectives on meta-learning for
+36. Smith-Miles KA. Cross-disciplinary perspectives on meta-learning for
 algorithm selection. ACM Computing Surveys (CSUR). 2009;41(1):6.
@@
 \hypertarget{ref-lemke2015metalearning}{}
-36. Lemke C, Budka M, Gabrys B. Metalearning: a survey of trends and
+37. Lemke C, Budka M, Gabrys B. Metalearning: a survey of trends and
 technologies. Artificial intelligence review. 2015;44(1):117--30.
@@
 \hypertarget{ref-Shin2014}{}
-37. Shin S-Y, Fauman EB, Petersen A-K, Krumsiek J, Santos R, Huang J, et
+38. Shin S-Y, Fauman EB, Petersen A-K, Krumsiek J, Santos R, Huang J, et
 al. An atlas of genetic influences on human blood metabolites. Nature
@@
 \hypertarget{ref-Kettunen2016}{}
-38. Kettunen J, Demirkan A, Wurtz P, Draisma HHM, Haller T, Rawal R, et
+39. Kettunen J, Demirkan A, Wurtz P, Draisma HHM, Haller T, Rawal R, et
 al. Genome-wide study for circulating metabolites identifies 62 loci and
@@
 \hypertarget{ref-Sun2017}{}
-39. Sun BB, Maranville JC, Peters JE, Stacey D, Staley JR, Blackshaw J,
+40. Sun BB, Maranville JC, Peters JE, Stacey D, Staley JR, Blackshaw J,
 et al. Consequences Of Natural Perturbations In The Human Plasma
@@
 \hypertarget{ref-Willer2013}{}
-40. Willer CJ, Schmidt EM, Sengupta S, Peloso GM, Gustafsson S, Kanoni
+41. Willer CJ, Schmidt EM, Sengupta S, Peloso GM, Gustafsson S, Kanoni
 S, et al. Discovery and refinement of loci associated with lipid levels.
@@
 \hypertarget{ref-Okbay2016}{}
-41. Okbay A, Beauchamp JP, Fontana MA, Lee JJ, Pers TH, Rietveld CA, et
+42. Okbay A, Beauchamp JP, Fontana MA, Lee JJ, Pers TH, Rietveld CA, et
 al. Genome-wide association study identifies 74 loci associated with
@@
 \hypertarget{ref-Anderson2017}{}
-42. Anderson E, Wade KH, Hemani G, Bowden J, Korologou-Linden R, Davey
+43. Anderson E, Wade KH, Hemani G, Bowden J, Korologou-Linden R, Davey
 Smith G, et al. The Causal Effect Of Educational Attainment On
@@
 \hypertarget{ref-kotthoff2016auto}{}
-43. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
+44. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
 2.0: Automatic model selection and hyperparameter optimization in WEKA.
@@
 \hypertarget{ref-Zhu2016}{}
-44. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+45. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
 Integration of summary data from GWAS and eQTL studies predicts complex
@@
 \hypertarget{ref-Richardson2017}{}
-45. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
+46. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
 CL, et al. Causal epigenome-wide association study identifies CpG sites
@@
 \hypertarget{ref-Wang2017}{}
-46. Wang L, Michoel T. Efficient And Accurate Causal Inference With
+47. Wang L, Michoel T. Efficient And Accurate Causal Inference With
 Hidden Confounders From Genome-Transcriptome Variation Data. bioRxiv
@@
 \hypertarget{ref-Aalen2015}{}
-47. Aalen OO, Valberg M, Grotmol T, Tretli S. Understanding variation in
+48. Aalen OO, Valberg M, Grotmol T, Tretli S. Understanding variation in
 disease risk: the elusive concept of frailty. International journal of
@@
 \hypertarget{ref-Noyce2017}{}
-48. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
+49. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
 E, et al. Estimating the causal influence of body mass index on risk of
@@
 \hypertarget{ref-Burgess2014a}{}
-49. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+50. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 \hypertarget{ref-Juul2003}{}
-50. Juul K, Tybjaerg-Hansen A, Marklund S, Heegaard NHH, Steffensen R,
+51. Juul K, Tybjaerg-Hansen A, Marklund S, Heegaard NHH, Steffensen R,
 Sillesen H, et al. Genetically Reduced Antioxidative Protection and
@@
 
+\hypertarget{ref-VonHinke2016}{}
+52. Hinke S von, Davey Smith G, Lawlor D, Propper C, Windmeijer F.
+Genetic markers as instrumental variables. J Health Econ.
+2016;45:131--48.
+
 \hypertarget{ref-Staley2016}{}
-51. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
+53. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
 al. PhenoScanner: a database of human genotype--phenotype associations.
@@
 
+\hypertarget{ref-Vanderweele2014}{}
+54. Vanderweele TJ, Tchetgen Tchetgen EJ, Cornelis M, Kraft P.
+Methodological challenges in mendelian randomization. Epidemiology
+(Cambridge, Mass) {[}Internet{]}. 2014 May;25(3):427--35. Available
+from: \url{http://www.ncbi.nlm.nih.gov/pubmed/24681576}
+
 \hypertarget{ref-Lee2013c}{}
-52. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+55. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 \hypertarget{ref-Liaw2002}{}
-53. Liaw A, Wiener M. Classification and Regression by randomForest. R
+56. Liaw A, Wiener M. Classification and Regression by randomForest. R
 News {[}Internet{]}. 2002;2(3):18--22. Available from:

### 2017-08-08 docs/manuscript.rmd
updated

@@
 ---
-title: A putative causal map of natural variation across the human phenome
+title: Automating Mendelian randomization through machine learning to construct a putative causal map of the human phenome
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 Methods have been developed in an attempt to protect MR from different patterns of horizontal pleiotropy, and here we have designed a mixture-of-experts machine learning framework (MR-MoE 1.0) that predicts the most appropriate model to use for any specific causal analysis, improving on both power and false discovery rates.
-Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, comprising the evaluation of causality between over a million pairs of traits.
+Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes.
 Almost 90% of causal estimates indicated some level of horizontal pleiotropy.
@@
 Second, causal inference can be made between phenotypes even if they have not been measured in the same samples, limiting possible MR analyses only by the availability of GWAS summary data for the traits in question [@Hemani2016].
-Third, because each instrumental variable mimics an independent randomization study, we can view the causal inference problem through the simple and widely understood prism of a meta-analysis.
+Third, because each instrumental variable mimics an independent randomization study, we can view the causal inference problem through the simple and widely understood prism of a meta-analysis [@Bowden2017].
 
-Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure (Box 1).
-The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017].
+Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy, where they influence the outcome through a pathway other than the exposure (Box 1).
+The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of horizontal pleiotropy [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017].
 It is considered best practice to triangulate estimates from a range of MR methods (and other experimental designs) when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions [@Lawlor2017].
@@
 First, it could be of interest to screen links between thousands of traits for being potential causal associations, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient.
-Second, though the simplest method, based on inverse-variance weighted (IVW) meta analysis [@Pierce2013], is most statistically powerful under no HP, it can have high false positive or low true positive rates in the presence of HP compared to other methods.
+Second, though the simplest method, based on inverse-variance weighted (IVW) meta analysis [@Pierce2013], is most statistically powerful under no horizontal pleiotropy, it can have high false positive or low true positive rates in the presence of horizontal pleiotropy compared to other methods.
 Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree and nature (e.g.
 horizontal versus vertical, see Box 1) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate.
-Third, if different methods disagree it is not possible to know which is correct because the true nature of HP exhibited by the instruments is not known.
-Fourth, the available methods do not cover all possible models of HP, and therefore an automated method for instrument selection may be necessary.
+Third, if different methods disagree it is not possible to know which is correct because the true nature of horizontal pleiotropy exhibited by the instruments is not known.
+Fourth, the available methods do not cover all possible models of horizontal pleiotropy, and therefore an automated method for instrument selection may be necessary.
 
@@
 The mechanisms behind it can arise through many different mechanisms, for example a single variant influences multiple genes; a single gene generates multiple gene products; a single gene product has multiple cellular functions; or a single cellular function influences multiple phenotypes [@Hu2016].
-How we classify the pleiotropic mechanism depends on the application, and in this paper we discuss pleiotropy in two forms - vertical and horizontal.
+How we classify the pleiotropic mechanism depends on question and context, and in this paper we discuss pleiotropy in two forms - vertical and horizontal.
 Vertical pleiotropy has also been known as mediated pleiotropy [@Solovieff2013], because the genetic variant influences one trait, which in turn influences a second trait.
@@
 In its simplest implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome.
-This is known as the inverse-variance weighted (IVW) method.
+This is known as the IVW method.
 
@@
 **Mean-based methods:** Here we consider four nested models [@Bowden2017].
-The inverse variance weighted (IVW) fixed effects meta-analysis approach assumes that variants exhibit no HP.
-IVW random effects meta-analysis relaxes the HP assumption, allowing it to be present but balanced - such that it only leads to increased heterogeneity around the regression line without affecting the slope (and therefore not introducing bias).
-Fixed effects Egger regression [@Bowden2015] relaxes the HP assumption further by allowing a non-zero intercept which essentially allows overall horizontal pleiotropy to be directional, where its total effect influences the outcome in a specific direction.
-Random effects Egger regression further allows heterogeneity around the slope having accounted for overall directional HP [@Bowden2017], as long as the HP effects are not correlated with the SNP-exposure effects (also known as the INSIDE assumption)[@Bowden2015].
+The inverse variance weighted (IVW) fixed effects meta-analysis approach assumes that variants exhibit no horizontal pleiotropy.
+IVW random effects meta-analysis relaxes the horizontal pleiotropy assumption, allowing it to be present but balanced - such that it only leads to increased heterogeneity around the regression line without affecting the slope (and therefore not introducing bias).
+Fixed effects Egger regression [@Bowden2015] relaxes the horizontal pleiotropy assumption further by allowing a non-zero intercept which essentially allows overall horizontal pleiotropy to be directional, where its total effect influences the outcome in a specific direction.
+Random effects Egger regression further allows heterogeneity around the slope having accounted for overall directional horizontal pleiotropy [@Bowden2017], as long as the horizontal pleiotropy effects are not correlated with the SNP-exposure effects (also known as the INSIDE assumption)[@Bowden2015].
 
@@
 
-The four nested models (IVW fixed effects, IVW random effects, Egger fixed effects, Egger random effects) plus the three Rucker estimates (point estimate, mean of the jackknife, median of the jackknife) provide seven mean based estimators.
+The four nested models (IVW fixed effects, IVW random effects, Egger fixed effects, Egger random effects) plus the three Rucker estimates (point estimate, mean of the jackknife, median of the jackknife) provide seven mean-based estimators.
 
@@
 The penalised weighted median estimator introduces a further weight to the instruments, penalising any instrument that contributes substantially towards the heterogeneity statistic.
-Together, this provides three median based estimators.
+Together, this provides three median-based estimators.
 Other estimation strategies not considered here, such as LASSO regression, have also been developed for the situation where at least half of the instruments are valid [@Kang2016].
 
-**Mode-based methods:** The mode based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017].
+**Mode-based methods:** The mode-based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017].
 There are four implementations of this method: the simple and the weighted mode, each weighted with or without the assumption of no measurement error in the exposure estimates (NOME).
@@
 not only those passing statistical thresholds for significance.
-We call this the "tophits" strategy.
+We call this the "top-hits" strategy.
 
@@
 For example, if a SNP $g_{A}$ influences trait $A$, and trait $A$ influences trait $B$, then a sufficiently powered GWAS will identify the $g_{A}$ as being significant for trait $B$ (Figure 1a).
-Using $g_{A}$ as an instrument to test the causal effect of $A$ on $B$ is perfectly valid.
-But the (incorrectly hypothesised) MR analysis of trait $B$ on trait $A$ could erroneously result in the apparent causal effect of $B$ on $A$.
-If $g_{A}$ is only one of many known instruments for $B$, amongst which some are valid, it is to the advantage of the researcher to exclude $g_{A}$ from the analysis.
+Using $g_{A}$ as an instrument to test the causal effect of $A$ on $B$ is perfectly valid in some cases.
+But the (incorrectly hypothesised) MR analysis of trait $B$ on trait $A$, for example, could erroneously result in the apparent causal effect of $B$ on $A$.
+It is to the advantage of the researcher to exclude $g_{A}$ from the analysis from the analysis of $B$ against $A$.
 
@@
 because the $cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}$.
-This simple inequality will not hold in some cases, for example $\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}$ where $\rho_{x, x_o}$ and $\rho_{y, y_o}$ are the precision of the measurements of the $x$ and $y$.
-Some combinations of confounding effects can also distort the $\rho_{g,x}$ and $\rho_{g,y}$ parameters, as has been discussed in detail previously [@Hemani2017].
-However, we use it here as a computationally inexpensive and approximate method to identify variants that are likely to be invalid.
+This simple inequality will not hold in some cases, for example $cor(x, x_o) < cor(x,y)cor(y,y_o)$ where $cor(x, x_o)$ and $cor(y, y_o)$ are the precision of the measurements of the $x$ and $y$.
+Some combinations of confounding effects can also distort the $cor(g,x)$ and $cor(g,y)$ parameters, as has been discussed in detail previously [@Hemani2017].
+Large differences in sample sizes between the exposure and outcome GWASs may also have a practical impact on the efficacy of this approach.
+However, we use it here as a computationally inexpensive and approximate method to identify variants that are likely to be invalid (Figure 1a).
 Steiger's Z-test of correlated correlations [@Steiger1980] can be used to formally test the extent to which the two correlations are statistically different.
 
-Here we adapt this approach to automatically filter SNPs that are liable to be invalid (Figure 1a).
-Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are more likely to be a source of HP.
+Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are more likely to be a source of horizontal pleiotropy.
 A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones.
@@
 
-We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total (Supplementary table 2).
+We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 strategies in total (Supplementary table 2).
 Each method assumes or performs best for a different model of pleiotropy.
 Our objective is to select the method most likely to be correct for a specific MR analysis by predicting the model of pleiotropy, and relating that prediction to the most appropriate model.
-To achieve this we use a "mixture of experts" machine learning approach [@Jordan1994], where each MR method is considered to be an 'expert', taking a summary-set as its input.
+To achieve this we use a "mixture of experts" machine learning approach [@Jordan1994], where each MR strategy is considered to be an 'expert', taking a summary-set as its input.
 An overview of the approach is shown in Figure 2.
@@
 As such our MR-MoE approach fits the framework of meta-learning [@brazdil2008metalearning; @smith2009cross; @lemke2015metalearning] where machine learning methods are applied at the 'meta-level' to analyse the results of machine learning experiments at the 'base-level', in order to be able to understand and model the capabilities of each expert and recommend which expert should be applied to a given problem.
-From this perspective the 28 MR methods are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the summary-sets.
+From this perspective the 28 MR strategies are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the summary-sets.
 
@@
 If the number of variants after Steiger filtering is greater than 5 then apply the MoE to obtain the best method.
-This value is chosen arbitrarily as a minimum number of variants for which the different MR methods can model HP
+This value is chosen arbitrarily as a minimum number of variants for which the different MR methods can model horizontal pleiotropy
 2. If the number of variants after Steiger filtering is less than or equal to 5 but greater than 1 then use the IVW random effects approach on the filtered set of variants
@@
 
+The graph can be queried directly using the cypher language at [http://eve-neo4j.mrbase.org](http://eve-neo4j.mrbase.org) or through a basic web interface at [http://eve.mrbase.org](http://eve.mrbase.org).
 For specific hypotheses we strongly recommend that estimates from all sensitivity analyses are scrutinised and reported.
-The graph can be queried directly using the cypher language at [http://eve-neo4j.mrbase.org](http://eve-neo4j.mrbase.org) or through a basic web interface at [http://eve.mrbase.org](http://eve.mrbase.org).
 
@@
 
-As statistical power for GWAS studies improves the likelihood of a significant association being discovered for an outcome trait that acts primarily on some exposure trait for that outcome increases (Figure 1a).
+As statistical power for GWAS studies improves the likelihood of a significant association being discovered for a trait that acts primarily on one of its precursors increases (Figure 1a).
+This presents a problem when GWAS significance is used as the sole criterion for instrument selection if the hypothesis being tested is either a) the trait causing the precursor, or b) the trait causing some other trait and the precursor causally relates to both (in which case the precursor is a confounder).
 We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated summary-sets comprising both null and non-null causal models.
-For each summary-set, instruments were selected based on the tophits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.
+For each summary-set, instruments were selected based on the top-hits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.
 
-Figure 1b shows that the tophits strategy led to over half of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype.
+Figure 1b shows that the top-hits strategy led to over half of the instruments being primarily associated with either confounders or the outcome phenotype, not the exposure phenotype.
 For brevity, we refer to the latter type of invalid instruments as “reverse causal” instruments.
@@
 
-Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to predict the most appropriate of the 28 MR methods and instrument selection strategies to apply to a particular summary-set based on its characteristics (Figure 2).
+Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to predict the most appropriate of the 14 MR methods and two instrument selection strategies (giving a total of 28 MR strategies) to apply to a particular summary-set based on its characteristics (Figure 2).
 
@@
 
-We compared the MoE performance in the simulations against each of the 28 methods, testing to see if it outperformed all other single methods.
+We compared the MoE performance in the simulations against each of the 28 strategies, testing to see if it outperformed all other single strategies.
 Figures 3a and 3b show that the MoE approach had the best general performance.
@@
 
-### Automated MR analysis of 2407 phenotypes
+### Automated MR analysis of 2407 phenotypes reveals substantial horizontal pleiotropy
 
@@
 The frequencies of the methods chosen by the MR-MoE analysis are shown in Supplementary table 3.
-Amongst those deemed 'significant', the IVW fixed effects analysis method combined with the tophits instruments selection strategy (only valid when there is no detectable evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases.
+Amongst those deemed 'significant', the IVW fixed effects analysis method combined with the top-hits instruments selection strategy (only valid when there is no detectable evidence of horizontal pleiotropy of any sort) was selected in only 10.4% of cases.
 This indicates that horizontal pleiotropy is likely to be pervasive.
 
-### Putative associations involving LDL cholesterol and years of schooling
+### Interpretation of the putative associations in a broad scan of phenotypes requires detailed follow up
 
-As an example of the database we performed a look up of associations between all traits and LDL cholesterol (as measured in the GLGC consortium [@Willer2013]), where the MoE result had a false discovery rate of 0.05 using the MoE approach.
+We performed a look up of associations between all traits and LDL cholesterol (as measured in the GLGC consortium [@Willer2013]), where the result had a false discovery rate of 0.05 using the MoE approach.
 This returned 287 putative associations, amongst which 111 involved LDL cholesterol influencing other traits and 176 involved other traits influencing LDL cholesterol.
@@
 There was a strong association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52, $p=2.7\times10^{-12}$).
-Here the IVW random effects method was chosen after Steiger filtering, indicating the presence of HP being present amongst the instruments.
+Here the IVW random effects method was chosen after Steiger filtering, indicating the presence of horizontal pleiotropy being present amongst the instruments.
 
@@
 Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease, which has recently been reported [@Anderson2017].
-However, it is clear that there are serious limitations in relying on MR as a panacea for causal inference.
+There is also a protective effect of more years of schooling on lower risk of myocardial infarction, which has been supported by detailed follow up analyses [] However, it is clear that there are limitations in relying on MR as a panacea for causal inference.
 For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g.
@@
 
-The facility to perform these scans has been automated in the MR of everything vs everything (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).
+The facility to perform these scans has been automated in the MR of 'everything vs everything' (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).
 
@@
 Methods have been introduced recently that seek to improve MR estimates through invalid instrument detection and removal [@Corbin2016; @Verbanck2017; @Zhu2017].
-These attack the problem from the perspective that variants which have a relatively large influence on the causal estimate, or provide substantially different estimates from the other variants, are likely to be exhibiting HP, and their removal from the analysis may be warranted.
+These attack the problem from the perspective that variants which have a relatively large influence on the causal estimate, or provide substantially different estimates from the other variants, are likely to be exhibiting horizontal pleiotropy, and their removal from the analysis may be warranted.
 A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones.
@@
 
-In addition to HP, there are many other limitations that prevent MR from being a panacea for causal inference.
+In addition to horizontal pleiotropy, there are many other limitations that prevent MR from being a panacea for causal inference.
 Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes.
@@
 
-Other problems, in addition to HP, can also manifest.
+Other problems, in addition to horizontal pleiotropy, can also manifest.
 Frailty effects [@Aalen2015] could induce associations for late-onset traits, and often detailed modelling for how this can influence MR is warranted [@Noyce2017; @Anderson2017].
@@
 
-We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular summary-set, namely, that the model selection is data driven.
+A new problem may arise when using machine learning approaches to infer the correct model for a particular summary-set, namely, that the model selection is data driven.
 This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance.
@@
 The y-axis shows the proportion of instruments selected for analysis that are either direct associations with the exposure, instruments for the outcome (reverse), or instruments for confounding traits (confounder).
-The proportions are compared over a range of different exposure discovery sample sizes (x-axis) and using either the tophits approach (left) or the Steiger approach (right) for instrument selection.
-c) Top: The false discovery rates from null simulations for each of the 14 methods using either tophits, Steiger filtered variants, or variants that are known to be directly associated with the exposure (oracle, note that direct effects can still exhibit horizontal pleiotropy in these simulations).
+The proportions are compared over a range of different exposure discovery sample sizes (x-axis) and using either the top-hits approach (left) or the Steiger approach (right) for instrument selection.
+c) Top: The false discovery rates from null simulations for each of the 14 methods using either top-hits, Steiger filtered variants, or variants that are known to be directly associated with the exposure (oracle, note that direct effects can still exhibit horizontal pleiotropy in these simulations).
 Bottom: The statistical power to detect true causal associations in the non-null simulations.
@@
 a) Training.
-Datasets are simulated that have either null or non-null causal relationships, and causal estimates are obtained from each of the 28 MR methods considered in this study.
+Datasets are simulated that have either null or non-null causal relationships, and causal estimates are obtained from each of the 28 MR strategies considered in this study.
 In the toy datasets, the columns in black represent 53 metrics about each of the 67,000 training simulations.
-The columns in red are specific to each method, they represent how well that method performed in obtaining the correct answer for each of the datasets.
-Random forests are used to learn the parameter space of the 53 metrics in which a particular dataset is likely to perform well.
-Together, this creates 28 random forest decision trees, one for each method.
+The columns in red are specific to each strategy, they represent how well that strategy performed in obtaining the correct answer for each of the datasets.
+Random forests are used to learn the parameter space of the 53 metrics in which a particular summary-set is likely to perform well.
+Together, this creates 28 random forest decision trees, one for each MR strategy.
 b) Application.
-For a GWAS summary dataset, our objective is to choose the method most likely to return the correct causal estimate.
+For a GWAS summary dataset, our objective is to choose the strategy most likely to return the correct causal estimate.
 Metrics are generated from the dataset and fed into each of the 28 random forest decision trees.
 This provides us with 28 performance predictions.
-Finally, we use the method for which the performance prediction is highest.
+Finally, we use the strategy for which the performance prediction is highest.
 
@@
 
-Figure 3: Performance of MoE against all other methods.
-a) The power for non-null datasets is plotted against the FDR for null datasets for each of the 28 methods, plus MoE.
-No single method achieved nominal FDR for these simulations.
+Figure 3: Performance of MoE against all other MR strategies.
+a) The power for non-null datasets is plotted against the FDR for null datasets for each of the 28 strategies, plus MoE.
+No single MR strategy achieved nominal FDR for these simulations.
 b) Calculating the area under the ROC curve from the values in (a) we plotted the performance in order from lowest to highest.
-Under the assumption of pervasive horizontal pleiotropy, the MoE approach is likely most effective than any other single method.
+Under the assumption of pervasive horizontal pleiotropy, the MoE approach is likely most effective than any other single MR strategy.
 
@@
 
-The mixture of experts approach seeks to use simulations to learn the circumstances in which each of the 28 MR methods is likely to perform most accurately.
-Having trained the model on simulations, new summary-sets can be applied to obtain causal estimates from the method that is predicted to be the most reliable for that summary-set.
+The mixture of experts approach seeks to use simulations to learn the circumstances in which each of the 28 MR strategies is likely to perform most accurately.
+Having trained the model on simulations, new summary-sets can be applied to obtain causal estimates from the strategy that is predicted to be the most reliable for that summary-set.
 
@@
 For each training summary-set we record a set of 53 metrics $\textbf{z}_{d}$ (Supplementary table 3), and an outcome $O_{w,d}$, which is a measure of how well that method performed for each particular simulated summary-set.
-For each of our 28 methods, we need to create a model that predicts the performance of the method based on metrics generated from a summary-set.
-To do this, for each method we train a random decision forest to predict that method's performance using the summary-set's metrics.
+For each of our 28 MR strategies, we need to create a model that predicts the performance of the strategy based on metrics generated from a summary-set.
+To do this, for each strategy we train a random decision forest to predict that strategy's performance using the summary-set's metrics.
 The random forest approach is well suited to this problem because there are likely to be non-continuous combinations of different metrics that improve on prediction over, for example, a simple linear model that does not learn about interactions.
@@
 
-- an IVW fixed effects analysis with the outlier removed, should the Steiger test be able to detect the outlier
-- a median based approach
-- a mode based approach
+- an IVW fixed effects analysis with Steiger filtering, should the Steiger test be able to detect the outlier
+- a median based approach without Steiger filtering
+- a mode based approach without Steiger filtering
 
-deciding between these methods requires finding, in general, which will minimise false discovery rates and maximise true positive rates for that particular scenario.
+deciding between these strategies requires finding, in general, which will minimise false discovery rates and maximise true positive rates for that particular scenario.
 In this example the IVW with Steiger filtering would likely be the clear winner because of its superior statistical power.
@@
 
-Having generated random forest decision trees for each of the 28 methods using 133,000 of the simulations, we then applied them to the remaining 67,000 summary-sets to predict which method would have the highest performance for each of the remaining summary-sets.
-Finally we compare the performance of the method selected by the MoE against all remaining methods.
+Having generated random forest decision trees for each of the 28 strategies using 133,000 of the simulations, we then applied them to the remaining 67,000 summary-sets to predict which method would have the highest performance for each of the remaining summary-sets.
+Finally we compare the performance of the method selected by the MoE against all remaining strategies.
 The default settings for the `randomForest` package in R [@Liaw2002] were used to train the models.

### 2017-08-08 docs/manuscript.tex
updated

@@
 \hypersetup{unicode=true,
-            pdftitle={A putative causal map of natural variation across the human phenome},
+            pdftitle={Automating Mendelian randomization through machine learning to construct a putative causal map of the human phenome},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{A putative causal map of natural variation across the human phenome}
+  \title{Automating Mendelian randomization through machine learning to construct
+a putative causal map of the human phenome}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
   \postdate{\par}
-  \date{03 August 2017}
+  \date{08 August 2017}
 
@@
 improving on both power and false discovery rates. Using the approach,
-we systematically estimated the causal effects amongst 2407 phenotypes,
-comprising the evaluation of causality between over a million pairs of
-traits. Almost 90\% of causal estimates indicated some level of
-horizontal pleiotropy. The causal estimates are organised into a
-publicly available graph database (\url{http://eve.mrbase.org}), and we
-use it here to highlight the numerous challenges that remain in
-automated causal inference.
+we systematically estimated the causal effects amongst 2407 phenotypes.
+Almost 90\% of causal estimates indicated some level of horizontal
+pleiotropy. The causal estimates are organised into a publicly available
+graph database (\url{http://eve.mrbase.org}), and we use it here to
+highlight the numerous challenges that remain in automated causal
+inference.
 
@@
 view the causal inference problem through the simple and widely
-understood prism of a meta-analysis.
+understood prism of a meta-analysis (6).
 
 Problems with obtaining unbiased causal effects can arise, however, if
-the genetic instruments exhibit horizontal pleiotropy (HP), where they
+the genetic instruments exhibit horizontal pleiotropy, where they
 influence the outcome through a pathway other than the exposure (Box 1).
@@
 have been developed that attempt to reliably obtain unbiased causal
-estimates under specific models of HP (4--7,9,13). It is considered best
-practice to triangulate estimates from a range of MR methods (and other
-experimental designs) when presenting causal estimates so as to assess
-the potential for different horizontal pleiotropy models and other
-violations of MR assumptions (14). However there are circumstances in
-which it is desirable to consolidate across many methods to obtain a
-single MR estimate. First, it could be of interest to screen links
-between thousands of traits for being potential causal associations, in
-which case a critical evaluation of each causal estimate of interest may
-not be possible or convenient. Second, though the simplest method, based
-on inverse-variance weighted (IVW) meta analysis (11), is most
-statistically powerful under no HP, it can have high false positive or
-low true positive rates in the presence of HP compared to other methods.
+estimates under specific models of horizontal pleiotropy (4--7,9,13). It
+is considered best practice to triangulate estimates from a range of MR
+methods (and other experimental designs) when presenting causal
+estimates so as to assess the potential for different horizontal
+pleiotropy models and other violations of MR assumptions (14). However
+there are circumstances in which it is desirable to consolidate across
+many methods to obtain a single MR estimate. First, it could be of
+interest to screen links between thousands of traits for being potential
+causal associations, in which case a critical evaluation of each causal
+estimate of interest may not be possible or convenient. Second, though
+the simplest method, based on inverse-variance weighted (IVW) meta
+analysis (11), is most statistically powerful under no horizontal
+pleiotropy, it can have high false positive or low true positive rates
+in the presence of horizontal pleiotropy compared to other methods.
 Pleiotropy has been hypothesised to be universal (15), though the degree
@@
 may not be appropriate. Third, if different methods disagree it is not
-possible to know which is correct because the true nature of HP
-exhibited by the instruments is not known. Fourth, the available methods
-do not cover all possible models of HP, and therefore an automated
-method for instrument selection may be necessary.
+possible to know which is correct because the true nature of horizontal
+pleiotropy exhibited by the instruments is not known. Fourth, the
+available methods do not cover all possible models of horizontal
+pleiotropy, and therefore an automated method for instrument selection
+may be necessary.
 
@@
 The inverse variance weighted (IVW) fixed effects meta-analysis approach
-assumes that variants exhibit no HP. IVW random effects meta-analysis
-relaxes the HP assumption, allowing it to be present but balanced - such
-that it only leads to increased heterogeneity around the regression line
-without affecting the slope (and therefore not introducing bias). Fixed
-effects Egger regression (4) relaxes the HP assumption further by
+assumes that variants exhibit no horizontal pleiotropy. IVW random
+effects meta-analysis relaxes the horizontal pleiotropy assumption,
+allowing it to be present but balanced - such that it only leads to
+increased heterogeneity around the regression line without affecting the
+slope (and therefore not introducing bias). Fixed effects Egger
+regression (4) relaxes the horizontal pleiotropy assumption further by
 allowing a non-zero intercept which essentially allows overall
@@
 regression further allows heterogeneity around the slope having
-accounted for overall directional HP (6), as long as the HP effects are
-not correlated with the SNP-exposure effects (also known as the INSIDE
-assumption)(4).
+accounted for overall directional horizontal pleiotropy (6), as long as
+the horizontal pleiotropy effects are not correlated with the
+SNP-exposure effects (also known as the INSIDE assumption)(4).
 
@@
 passing statistical thresholds for significance. We call this the
-``tophits'' strategy.
+``top-hits'' strategy.
 
@@
 simple inequality will not hold in some cases, for example
-\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
-\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
+\(cor(x, x_o) < cor(x,y)cor(y,y_o)\) where \(cor(x, x_o)\) and
+\(cor(y, y_o)\) are the precision of the measurements of the \(x\) and
 \(y\). Some combinations of confounding effects can also distort the
-\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
-detail previously (8). However, we use it here as a computationally
+\(cor(g,x)\) and \(cor(g,y)\) parameters, as has been discussed in
+detail previously (8). Large differences in sample sizes between the
+exposure and outcome GWASs may also have a practical impact on the
+efficacy of this approach. However, we use it here as a computationally
 inexpensive and approximate method to identify variants that are likely
-to be invalid. Steiger's Z-test of correlated correlations (31) can be
-used to formally test the extent to which the two correlations are
-statistically different.
-
-Here we adapt this approach to automatically filter SNPs that are liable
-to be invalid (Figure 1a). Other methods have been developed to identify
-invalid instruments for the purposes of exclusion from MR analyses
-(9,13,32), based on the notion that outlying instruments are more likely
-to be a source of HP. A potential drawback of this approach is that the
+to be invalid (Figure 1a). Steiger's Z-test of correlated correlations
+(31) can be used to formally test the extent to which the two
+correlations are statistically different.
+
+Other methods have been developed to identify invalid instruments for
+the purposes of exclusion from MR analyses (9,13,32), based on the
+notion that outlying instruments are more likely to be a source of
+horizontal pleiotropy. A potential drawback of this approach is that the
 outlier SNPs in a summary-set might be the only reliable ones. This
@@
 be supplied using two instrument selection strategies, leading to 28
-methods in total (Supplementary table 2). Each method assumes or
+strategies in total (Supplementary table 2). Each method assumes or
 performs best for a different model of pleiotropy. Our objective is to
@@
 the most appropriate model. To achieve this we use a ``mixture of
-experts'' machine learning approach (34), where each MR method is
+experts'' machine learning approach (34), where each MR strategy is
 considered to be an `expert', taking a summary-set as its input. An
@@
 capabilities of each expert and recommend which expert should be applied
-to a given problem. From this perspective the 28 MR methods are the
+to a given problem. From this perspective the 28 MR strategies are the
 base-models, the gating function is the meta-model, and its
@@
   arbitrarily as a minimum number of variants for which the different MR
-  methods can model HP
+  methods can model horizontal pleiotropy
 \item
@@
 As statistical power for GWAS studies improves the likelihood of a
-significant association being discovered for an outcome trait that acts
-primarily on some exposure trait for that outcome increases (Figure 1a).
-We evaluated the efficacy of the Steiger filtering approach for
-improving instrument selection using 100,000 simulated summary-sets
+significant association being discovered for a trait that acts primarily
+on one of its precursors increases (Figure 1a). This presents a problem
+when GWAS significance is used as the sole criterion for instrument
+selection if the hypothesis being tested is either a) the trait causing
+the precursor, or b) the trait causing some other trait and the
+precursor causally relates to both (in which case the precursor is a
+confounder). We evaluated the efficacy of the Steiger filtering approach
+for improving instrument selection using 100,000 simulated summary-sets
 comprising both null and non-null causal models. For each summary-set,
-instruments were selected based on the tophits strategy and the Steiger
+instruments were selected based on the top-hits strategy and the Steiger
 filtering strategy, and MR was performed using 14 different methods
@@
 
-Figure 1b shows that the tophits strategy led to over half of the
+Figure 1b shows that the top-hits strategy led to over half of the
 instruments being primarily associated with either confounders or the
@@
 existing methods, we next hypothesised that a mixture of experts (MoE)
-model would be able to predict the most appropriate of the 28 MR methods
-and instrument selection strategies to apply to a particular summary-set
-based on its characteristics (Figure 2).
+model would be able to predict the most appropriate of the 28 MR
+strategies and instrument selection strategies to apply to a particular
+summary-set based on its characteristics (Figure 2).
 
@@
 We compared the MoE performance in the simulations against each of the
-28 methods, testing to see if it outperformed all other single methods.
-Figures 3a and 3b show that the MoE approach had the best general
-performance. Estimating the area under the receiver operator curve gave
-0.84 for the MoE approach in terms of classifying the simulations as
-being null or non-null. Notably, the next best methods were median and
-mode based estimators using Steiger filtering. But a crucial observation
-is that under the assumption of widespread and diverse pleiotropic
-effects it is clear that all methods suffer from high false discovery
-rates (Figure 3a), including the MoE approach.
-
-\subsubsection{Automated MR analysis of 2407
-phenotypes}\label{automated-mr-analysis-of-2407-phenotypes}
+28 strategies, testing to see if it outperformed all other single
+strategies. Figures 3a and 3b show that the MoE approach had the best
+general performance. Estimating the area under the receiver operator
+curve gave 0.84 for the MoE approach in terms of classifying the
+simulations as being null or non-null. Notably, the next best methods
+were median and mode based estimators using Steiger filtering. But a
+crucial observation is that under the assumption of widespread and
+diverse pleiotropic effects it is clear that all methods suffer from
+high false discovery rates (Figure 3a), including the MoE approach.
+
+\subsubsection{Automated MR analysis of 2407 phenotypes reveals
+substantial horizontal
+pleiotropy}\label{automated-mr-analysis-of-2407-phenotypes-reveals-substantial-horizontal-pleiotropy}
 
@@
 in Supplementary table 3. Amongst those deemed `significant', the IVW
-fixed effects analysis method combined with the tophits instruments
+fixed effects analysis method combined with the top-hits instruments
 selection strategy (only valid when there is no detectable evidence of
@@
 
-\subsubsection{Putative associations involving LDL cholesterol and years
-of
-schooling}\label{putative-associations-involving-ldl-cholesterol-and-years-of-schooling}
+\subsubsection{Interpretation of the putative associations in a broad
+scan of phenotypes requires detailed follow
+up}\label{interpretation-of-the-putative-associations-in-a-broad-scan-of-phenotypes-requires-detailed-follow-up}
 
@@
 \(p=2.7\times10^{-12}\)). Here the IVW random effects method was chosen
-after Steiger filtering, indicating the presence of HP being present
-amongst the instruments.
+after Steiger filtering, indicating the presence of horizontal
+pleiotropy being present amongst the instruments.
 
@@
 The facility to perform these scans has been automated in the MR of
-everything vs everything (MR-EvE) web application,
+`everything vs everything' (MR-EvE) web application,
 \url{http://eve.mrbase.org}.
@@
 different estimates from the other variants, are likely to be exhibiting
-HP, and their removal from the analysis may be warranted. A potential
-drawback of this approach is that the outlier SNPs in a summary-set
-might be the only reliable ones. This could arise, for example, for the
-\emph{CRP} variant instrumenting C-reactive protein levels (21), because
-CRP itself is downstream of many cellular functions and most GWAS
-signals will not be specific for the trait. Steiger filtering has
-attempted to sidestep this issue in removing invalid instruments,
-seeking to detect those instruments which have associations with the
-outcome that are too large to be consistent with the exposure causing
-the outcome. We acknowledge that unmeasured confounding and measurement
-error can introduce problems with Steiger filtering (8), but note that
-our simulations indicate that all individual MR methods improve in
-performance following its application.
-
-In addition to HP, there are many other limitations that prevent MR from
-being a panacea for causal inference. Many of the associations are
-biologically impossible, for example where early life phenotypes appear
-to be influenced by later stage phenotypes. Such associations can be
-explained statistically because the instruments proxy the liability for
-that particular phenotype, not that phenotype's observed manifestation.
-Though these associations can be informative, their interpretations as
-causal relationships are far from clear. Similarly, often disease traits
-appear to causally relate to other phenotypes, but GWAS of disease is
-typically performed on the liability scale, hence the causal estimate
-reflects not the presence or absence of disease, but the underlying risk
-of disease. Again, interpretation of such associations can be
-problematic due to the instruments proxying disease liability and not
-diseae outcomes.
+horizontal pleiotropy, and their removal from the analysis may be
+warranted. A potential drawback of this approach is that the outlier
+SNPs in a summary-set might be the only reliable ones. This could arise,
+for example, for the \emph{CRP} variant instrumenting C-reactive protein
+levels (21), because CRP itself is downstream of many cellular functions
+and most GWAS signals will not be specific for the trait. Steiger
+filtering has attempted to sidestep this issue in removing invalid
+instruments, seeking to detect those instruments which have associations
+with the outcome that are too large to be consistent with the exposure
+causing the outcome. We acknowledge that unmeasured confounding and
+measurement error can introduce problems with Steiger filtering (8), but
+note that our simulations indicate that all individual MR methods
+improve in performance following its application.
+
+In addition to horizontal pleiotropy, there are many other limitations
+that prevent MR from being a panacea for causal inference. Many of the
+associations are biologically impossible, for example where early life
+phenotypes appear to be influenced by later stage phenotypes. Such
+associations can be explained statistically because the instruments
+proxy the liability for that particular phenotype, not that phenotype's
+observed manifestation. Though these associations can be informative,
+their interpretations as causal relationships are far from clear.
+Similarly, often disease traits appear to causally relate to other
+phenotypes, but GWAS of disease is typically performed on the liability
+scale, hence the causal estimate reflects not the presence or absence of
+disease, but the underlying risk of disease. Again, interpretation of
+such associations can be problematic due to the instruments proxying
+disease liability and not diseae outcomes.
 
@@
 
-Other problems, in addition to HP, can also manifest. Frailty effects
-(48) could induce associations for late-onset traits, and often detailed
-modelling for how this can influence MR is warranted (43,49). Genetic
-variants could strongly relate to several phenotypes, for example SNPs
-that influence LDL cholesterol are also likely to influence other lipid
-fractions, making it difficult to ascertain which amongst them is the
-true causal exposure (50). We see this issue manifesting in Figure 4,
-where there appear to be causal relationships between different lipid
-fractions. The measured feature for which genetic associations are known
-could relate in complicated ways to the biological entities that is
-truly causal (2), for example higher circulating levels of the natural
-antioxidant extracellular superoxide dismutase (EC-SOD) appears to
-causally relate to higher risk of coronary heart disease (51) because
-the phenotypic measure relates inversely to the EC-SOD levels in
-arterial walls, hence \emph{in situ} activity is lower (2). Dynastic
-effects can also confound MR estimates, arising when the exposure trait
-in a previous generation influences the environment of the current
-generation. Here, the SNPs being used as instruments for the exposure
-will be associated with a potential confounder in the analysis (52). The
-associations depicted in Figure 4 involving educational attainment may
-be susceptible to bias arising from this mechanism, and protecting
-against it will likely require alternative study designs, for example by
-using sibling pairs.
-
-We also encounter a potential new problem in using machine learning
-approaches to infer the correct model for a particular summary-set,
-namely, that the model selection is data driven. This is liable to
-contribute to elevated type 1 error rates, even though we have attempted
-to separate the information used for optimisation from the information
-used to predict performance. One advantage to generating a method that
-consolidates across many methods is that it can prevent cherry picking
-the most appealing method based on their results. The extent to which a
-machine-based method selector incurs higher type 1 error rates more than
-a human selector is not clear. Though MR-MoE does exhibit higher type 1
-error rates than are desirable, they still remain amongst the lowest
-compared to other MR methods that do not suffer from the issue of data
-driven model selection.
+Other problems, in addition to horizontal pleiotropy, can also manifest.
+Frailty effects (48) could induce associations for late-onset traits,
+and often detailed modelling for how this can influence MR is warranted
+(43,49). Genetic variants could strongly relate to several phenotypes,
+for example SNPs that influence LDL cholesterol are also likely to
+influence other lipid fractions, making it difficult to ascertain which
+amongst them is the true causal exposure (50). We see this issue
+manifesting in Figure 4, where there appear to be causal relationships
+between different lipid fractions. The measured feature for which
+genetic associations are known could relate in complicated ways to the
+biological entities that is truly causal (2), for example higher
+circulating levels of the natural antioxidant extracellular superoxide
+dismutase (EC-SOD) appears to causally relate to higher risk of coronary
+heart disease (51) because the phenotypic measure relates inversely to
+the EC-SOD levels in arterial walls, hence \emph{in situ} activity is
+lower (2). Dynastic effects can also confound MR estimates, arising when
+the exposure trait in a previous generation influences the environment
+of the current generation. Here, the SNPs being used as instruments for
+the exposure will be associated with a potential confounder in the
+analysis (52). The associations depicted in Figure 4 involving
+educational attainment may be susceptible to bias arising from this
+mechanism, and protecting against it will likely require alternative
+study designs, for example by using sibling pairs.
+
+A new problem may arise when using machine learning approaches to infer
+the correct model for a particular summary-set, namely, that the model
+selection is data driven. This is liable to contribute to elevated type
+1 error rates, even though we have attempted to separate the information
+used for optimisation from the information used to predict performance.
+One advantage to generating a method that consolidates across many
+methods is that it can prevent cherry picking the most appealing method
+based on their results. The extent to which a machine-based method
+selector incurs higher type 1 error rates more than a human selector is
+not clear. Though MR-MoE does exhibit higher type 1 error rates than are
+desirable, they still remain amongst the lowest compared to other MR
+methods that do not suffer from the issue of data driven model
+selection.
 
@@
 different exposure discovery sample sizes (x-axis) and using either the
-tophits approach (left) or the Steiger approach (right) for instrument
+top-hits approach (left) or the Steiger approach (right) for instrument
 selection. c) Top: The false discovery rates from null simulations for
-each of the 14 methods using either tophits, Steiger filtered variants,
+each of the 14 methods using either top-hits, Steiger filtered variants,
 or variants that are known to be directly associated with the exposure
@@
 have either null or non-null causal relationships, and causal estimates
-are obtained from each of the 28 MR methods considered in this study. In
-the toy datasets, the columns in black represent 53 metrics about each
-of the 67,000 training simulations. The columns in red are specific to
-each method, they represent how well that method performed in obtaining
-the correct answer for each of the datasets. Random forests are used to
-learn the parameter space of the 53 metrics in which a particular
-dataset is likely to perform well. Together, this creates 28 random
-forest decision trees, one for each method. b) Application. For a GWAS
-summary dataset, our objective is to choose the method most likely to
-return the correct causal estimate. Metrics are generated from the
-dataset and fed into each of the 28 random forest decision trees. This
-provides us with 28 performance predictions. Finally, we use the method
-for which the performance prediction is highest.
+are obtained from each of the 28 MR strategies considered in this study.
+In the toy datasets, the columns in black represent 53 metrics about
+each of the 67,000 training simulations. The columns in red are specific
+to each strategy, they represent how well that strategy performed in
+obtaining the correct answer for each of the datasets. Random forests
+are used to learn the parameter space of the 53 metrics in which a
+particular summary-set is likely to perform well. Together, this creates
+28 random forest decision trees, one for each MR strategy. b)
+Application. For a GWAS summary dataset, our objective is to choose the
+strategy most likely to return the correct causal estimate. Metrics are
+generated from the dataset and fed into each of the 28 random forest
+decision trees. This provides us with 28 performance predictions.
+Finally, we use the strategy for which the performance prediction is
+highest.
 
@@
 
-Figure 3: Performance of MoE against all other methods. a) The power for
-non-null datasets is plotted against the FDR for null datasets for each
-of the 28 methods, plus MoE. No single method achieved nominal FDR for
-these simulations. b) Calculating the area under the ROC curve from the
-values in (a) we plotted the performance in order from lowest to
-highest. Under the assumption of pervasive horizontal pleiotropy, the
-MoE approach is likely most effective than any other single method.
+Figure 3: Performance of MoE against all other MR strategies. a) The
+power for non-null datasets is plotted against the FDR for null datasets
+for each of the 28 strategies, plus MoE. No single MR strategy achieved
+nominal FDR for these simulations. b) Calculating the area under the ROC
+curve from the values in (a) we plotted the performance in order from
+lowest to highest. Under the assumption of pervasive horizontal
+pleiotropy, the MoE approach is likely most effective than any other
+single MR strategy.
 
@@
 The mixture of experts approach seeks to use simulations to learn the
-circumstances in which each of the 28 MR methods is likely to perform
+circumstances in which each of the 28 MR strategies is likely to perform
 most accurately. Having trained the model on simulations, new
-summary-sets can be applied to obtain causal estimates from the method
+summary-sets can be applied to obtain causal estimates from the strategy
 that is predicted to be the most reliable for that summary-set.
@@
 which is a measure of how well that method performed for each particular
-simulated summary-set. For each of our 28 methods, we need to create a
-model that predicts the performance of the method based on metrics
-generated from a summary-set. To do this, for each method we train a
-random decision forest to predict that method's performance using the
-summary-set's metrics. The random forest approach is well suited to this
-problem because there are likely to be non-continuous combinations of
-different metrics that improve on prediction over, for example, a simple
-linear model that does not learn about interactions.
+simulated summary-set. For each of our 28 MR strategies, we need to
+create a model that predicts the performance of the strategy based on
+metrics generated from a summary-set. To do this, for each strategy we
+train a random decision forest to predict that strategy's performance
+using the summary-set's metrics. The random forest approach is well
+suited to this problem because there are likely to be non-continuous
+combinations of different metrics that improve on prediction over, for
+example, a simple linear model that does not learn about interactions.
 
@@
 \item
-  an IVW fixed effects analysis with the outlier removed, should the
+  an IVW fixed effects analysis with Steiger filtering, should the
   Steiger test be able to detect the outlier
 \item
-  a median based approach
+  a median based approach without Steiger filtering
 \item
-  a mode based approach
+  a mode based approach without Steiger filtering
 \end{itemize}
 
-deciding between these methods requires finding, in general, which will
-minimise false discovery rates and maximise true positive rates for that
-particular scenario. In this example the IVW with Steiger filtering
+deciding between these strategies requires finding, in general, which
+will minimise false discovery rates and maximise true positive rates for
+that particular scenario. In this example the IVW with Steiger filtering
 would likely be the clear winner because of its superior statistical
@@
 
-Having generated random forest decision trees for each of the 28 methods
-using 133,000 of the simulations, we then applied them to the remaining
-67,000 summary-sets to predict which method would have the highest
-performance for each of the remaining summary-sets. Finally we compare
-the performance of the method selected by the MoE against all remaining
-methods. The default settings for the \texttt{randomForest} package in R
-(56) were used to train the models. MR-MoE 1.0 is implemented in the
-TwoSampleMR R package available at
+Having generated random forest decision trees for each of the 28
+strategies using 133,000 of the simulations, we then applied them to the
+remaining 67,000 summary-sets to predict which method would have the
+highest performance for each of the remaining summary-sets. Finally we
+compare the performance of the method selected by the MoE against all
+remaining strategies. The default settings for the \texttt{randomForest}
+package in R (56) were used to train the models. MR-MoE 1.0 is
+implemented in the TwoSampleMR R package available at
 \href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}

### 2017-12-11 scripts/07-create_graphdb.r
updated

@@ -214,7 +214,7 @@ write.csv(snp_trait, file="../results/01/upload/snp_trait.csv", row.names=FALSE,
 
 ntt <- 8
 
-for(i in 5:ntt)
+for(i in 1:4)
 {
 	message(i)
 	load(paste0("../results/01/mr/mc-", i, ".rdata"))
@@ -244,4 +244,4 @@ cmd <- paste0(
 " --relationships:GA ../results/01/upload/snp_trait.csv", 
 " --relationships:MR ../results/01/upload/trait_trait.csv"
 )
-system(cmd)
+# system(cmd)

### 2017-12-11 scripts/07-create_graphdb.sh

@@ -1,53 +1,81 @@
 #/bin/bash
 
+# On the server run:
+# docker run -d \
+# --name neo4j321 \
+# --publish=7475:7474  --publish=7687:7687 \
+# --volume=$HOME/neo4j/data3.2:/data \
+# --volume=$HOME/neo4j/logs3.2:/logs \
+# neo4j:3.2.1
+
+# docker exec -i -t neo4j321 /bin/bash
+# bolt://shark.epi.bris.ac.uk:7687
+
+# docker run -d \
+# --name neo4j23 \
+# --publish=7474:7474 \
+# --volume=$HOME/neo4j/data:/data \
+# --volume=$HOME/neo4j/logs:/logs \
+# neo4j:2.3
+
+# docker exec -i -t neo4j321 /bin/bash
+
+# sudo chown --recursive gh13047 neo4j/data3.2
+# mv neo4j/data3.2/databases/graph.db neo4j/data3.2/databases/graph.db_old
+
+
+
 module add languages/java-jdk-1.8.0-66
 
-neo4j_dir=""
-cat ../results/01/upload/trait_trait* > ../results/01/upload/trait_trait.csv
-cat ../results/01/upload/trait_trait_sel* ../results/01/upload/trait_trait_sel.csv
+neo4j_dir="/panfs/panasas01/sscm/gh13047/bin/neo4j-community-3.2.1"
 
+rm -f ../results/01/upload/trait_trait.csv
+rm -f ../results/01/upload/trait_trait_sel.csv
+cp ../results/01/upload/trait_trait-1.csv ../results/01/upload/trait_trait.csv
+cp ../results/01/upload/trait_trait_sel-1.csv ../results/01/upload/trait_trait_sel.csv
+
+for i in {2..8}
+do
+	echo $i
+	sed 1d ../results/01/upload/trait_trait-${i}.csv >> ../results/01/upload/trait_trait.csv
+	sed 1d ../results/01/upload/trait_trait_sel-${i}.csv >> ../results/01/upload/trait_trait_sel.csv
+done
+
+rm -rf ../results/01/upload/graph.db
 
 ${neo4j_dir}/bin/neo4j-import \
---into ../results/01/upload/mr-eve.db \
+--into ../results/01/upload/graph.db \
 --id-type string \
---nodes:gene ../data/upload/genes.csv \
---nodes:snp ../data/upload/snps.csv \
---nodes:trait ../data/upload/traits.csv \
---relationships:GS ../data/upload/gene_snp.csv \
[…]

### 2017-12-11 docs/manuscript.rmd

@@
 Vertical pleiotropy can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not, although not without caveats [@Holmes2017].
-Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) potential bias arising from the exclusion restriction assumption can be detected or corrected by evaluating the consistency of effects across instruments [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Hemani2017; @Verbanck2017].
+Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) potential bias arising from the exclusion restriction assumption can be detected or corrected by evaluating the consistency of effects across instruments [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017a; @Hemani2017; @Verbanck2017].
 
@@
 Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy, where they influence the outcome through a pathway other than the exposure (Box 1).
-The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of horizontal pleiotropy [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017].
+The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of horizontal pleiotropy [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017a; @Verbanck2017; @Zhu2017].
 It is considered best practice to triangulate estimates from a range of MR methods (and other experimental designs) when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions [@Lawlor2017].
@@
 
-**Mode-based methods:** The mode-based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017].
+**Mode-based methods:** The mode-based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017a].
 There are four implementations of this method: the simple and the weighted mode, each weighted with or without the assumption of no measurement error in the exposure estimates (NOME).
@@
 Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease, which has recently been reported [@Anderson2017].
-There is also a protective effect of more years of schooling on lower risk of myocardial infarction, which has been supported by detailed follow up analyses [] However, it is clear that there are limitations in relying on MR as a panacea for causal inference.
+There is also a protective effect of more years of schooling on lower risk of myocardial infarction, which has been supported by detailed follow up analyses [@Tillmann2017].
+
+However, it is clear that there are limitations in relying on MR as a panacea for causal inference.
 For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g.
@@
 
-The facility to perform these scans has been automated in the MR of 'everything vs everything' (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).
+The facility to perform these look ups has been automated in the MR of 'everything vs everything' (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).
 
@@
 
-The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones.
-Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception.
+The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have demonstrated that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that may be more likely to be identified than valid ones.
+Strategising how MR is to be used in practice, therefore, must consider some presence of horizontal pleiotropy to be the rule rather than the exception.
 Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criterion.

### 2017-12-11 docs/manuscript.tex

@@
 influences multiple phenotypes (23). How we classify the pleiotropic
-mechanism depends on the application, and in this paper we discuss
+mechanism depends on question and context, and in this paper we discuss
 pleiotropy in two forms - vertical and horizontal. Vertical pleiotropy
@@
 provides the estimate of the causal effect of the exposure on the
-outcome. This is known as the inverse-variance weighted (IVW) method.
+outcome. This is known as the IVW method.
 
@@
 (point estimate, mean of the jackknife, median of the jackknife) provide
-seven mean based estimators.
+seven mean-based estimators.
 
@@
 penalising any instrument that contributes substantially towards the
-heterogeneity statistic. Together, this provides three median based
+heterogeneity statistic. Together, this provides three median-based
 estimators. Other estimation strategies not considered here, such as
@@
 
-\textbf{Mode-based methods:} The mode based estimator clusters the
+\textbf{Mode-based methods:} The mode-based estimator clusters the
 instruments into groups based on similarity of causal effects, and
@@
 \(g_{A}\) as an instrument to test the causal effect of \(A\) on \(B\)
-is perfectly valid. But the (incorrectly hypothesised) MR analysis of
-trait \(B\) on trait \(A\) could erroneously result in the apparent
-causal effect of \(B\) on \(A\). If \(g_{A}\) is only one of many known
-instruments for \(B\), amongst which some are valid, it is to the
-advantage of the researcher to exclude \(g_{A}\) from the analysis.
+is perfectly valid in some cases. But the (incorrectly hypothesised) MR
+analysis of trait \(B\) on trait \(A\), for example, could erroneously
+result in the apparent causal effect of \(B\) on \(A\). It is to the
+advantage of the researcher to exclude \(g_{A}\) from the analysis from
+the analysis of \(B\) against \(A\).
 
@@
 
-For specific hypotheses we strongly recommend that estimates from all
-sensitivity analyses are scrutinised and reported. The graph can be
-queried directly using the cypher language at
+The graph can be queried directly using the cypher language at
 \url{http://eve-neo4j.mrbase.org} or through a basic web interface at
-\url{http://eve.mrbase.org}.
+\url{http://eve.mrbase.org}. For specific hypotheses we strongly
+recommend that estimates from all sensitivity analyses are scrutinised
+and reported.
 
@@
 existing methods, we next hypothesised that a mixture of experts (MoE)
-model would be able to predict the most appropriate of the 28 MR
-strategies and instrument selection strategies to apply to a particular
-summary-set based on its characteristics (Figure 2).
+model would be able to predict the most appropriate of the 14 MR methods
+and two instrument selection strategies (giving a total of 28 MR
+strategies) to apply to a particular summary-set based on its
+characteristics (Figure 2).
 
@@
 
-As an example of the database we performed a look up of associations
-between all traits and LDL cholesterol (as measured in the GLGC
-consortium (41)), where the MoE result had a false discovery rate of
-0.05 using the MoE approach. This returned 287 putative associations,
-amongst which 111 involved LDL cholesterol influencing other traits and
-176 involved other traits influencing LDL cholesterol. A large
-proportion of these traits were metabolites, which are dominated by
-lipid fractions. Filtering to exclude the metabolomic studies ((38,39))
-returned 27 associations with 23 traits (Figure 4a). There was a strong
-association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52,
-\(p=2.7\times10^{-12}\)). Here the IVW random effects method was chosen
-after Steiger filtering, indicating the presence of horizontal
-pleiotropy being present amongst the instruments.
+We performed a look up of associations between all traits and LDL
+cholesterol (as measured in the GLGC consortium (41)), where the result
+had a false discovery rate of 0.05 using the MoE approach. This returned
+287 putative associations, amongst which 111 involved LDL cholesterol
+influencing other traits and 176 involved other traits influencing LDL
+cholesterol. A large proportion of these traits were metabolites, which
+are dominated by lipid fractions. Filtering to exclude the metabolomic
+studies ((38,39)) returned 27 associations with 23 traits (Figure 4a).
+There was a strong association with coronary heart disease (0.41 log(OR)
+per SD, 0.31-0.52, \(p=2.7\times10^{-12}\)). Here the IVW random effects
+method was chosen after Steiger filtering, indicating the presence of
+horizontal pleiotropy being present amongst the instruments.
 
@@
 plausible, for example the influence of years of schooling on lower risk
-of Alzheimer's disease, which has recently been reported (43). However,
-it is clear that there are serious limitations in relying on MR as a
+of Alzheimer's disease, which has recently been reported (43). There is
+also a protective effect of more years of schooling on lower risk of
+myocardial infarction, which has been supported by detailed follow up
+analyses (44).
+
+However, it is clear that there are limitations in relying on MR as a
 panacea for causal inference. For example, bi-directional causal
@@
 
-The facility to perform these scans has been automated in the MR of
+The facility to perform these look ups has been automated in the MR of
 `everything vs everything' (MR-EvE) web application,
@@
 growing field of automatic model selection and hyper-parameter
-optimization in machine learning (44). We applied the method to the
+optimization in machine learning (45). We applied the method to the
 curated set of GWAS summary data present in MR-Base (12), generating
@@
 opportunity that this affords MR to be furnished with more instruments
-is typically welcomed, here we have shown that it is invalid
+is typically welcomed, here we have demonstrated that it is invalid
 instruments, exhibiting horizontal pleiotropy or reverse causation, that
-are more likely to be identified than valid ones. Strategising how MR is
-to be used in practice, therefore, must consider horizontal pleiotropy
-to be the rule rather than the exception. Instrumental variables in MR
-are typically chosen blind, with GWAS significance being the only
-criterion. We illustrated how confounders and reverse causal
-associations can easily lead to invalid instrument selection in MR and
-that most randomly generated patterns of horizontal pleiotropy cannot be
-adequately accounted for by any single MR method.
+may be more likely to be identified than valid ones. Strategising how MR
+is to be used in practice, therefore, must consider some presence of
+horizontal pleiotropy to be the rule rather than the exception.
+Instrumental variables in MR are typically chosen blind, with GWAS
+significance being the only criterion. We illustrated how confounders
+and reverse causal associations can easily lead to invalid instrument
+selection in MR and that most randomly generated patterns of horizontal
+pleiotropy cannot be adequately accounted for by any single MR method.
 
@@
 reverse cause, pleiotropy and multiple causal variants in the same
-region (45,46), separating vertical from horizontal pleiotropy with a
+region (46,47), separating vertical from horizontal pleiotropy with a
 single instrument is not yet possible in the two-sample MR framework
 though the use of mediation-based approaches may be informative in the
-one-sample setting (8,47).
+one-sample setting (8,48).
 
 Other problems, in addition to horizontal pleiotropy, can also manifest.
-Frailty effects (48) could induce associations for late-onset traits,
+Frailty effects (49) could induce associations for late-onset traits,
 and often detailed modelling for how this can influence MR is warranted
-(43,49). Genetic variants could strongly relate to several phenotypes,
+(43,50). Genetic variants could strongly relate to several phenotypes,
 for example SNPs that influence LDL cholesterol are also likely to
 influence other lipid fractions, making it difficult to ascertain which
-amongst them is the true causal exposure (50). We see this issue
+amongst them is the true causal exposure (51). We see this issue
 manifesting in Figure 4, where there appear to be causal relationships
@@
 dismutase (EC-SOD) appears to causally relate to higher risk of coronary
-heart disease (51) because the phenotypic measure relates inversely to
+heart disease (52) because the phenotypic measure relates inversely to
 the EC-SOD levels in arterial walls, hence \emph{in situ} activity is
@@
 the exposure will be associated with a potential confounder in the
-analysis (52). The associations depicted in Figure 4 involving
+analysis (53). The associations depicted in Figure 4 involving
 educational attainment may be susceptible to bias arising from this
@@
 
-The field of MR is evolving. The advent of GWAS databases (10,53) and
+The field of MR is evolving. The advent of GWAS databases (10,54) and
 automated 2SMR (12) has trivialised the analytical aspect of
@@
 hypothesis is likely less fallible to the pitfalls described above than
-using MR to confirm a positive association (54). Second, it paves the
+using MR to confirm a positive association (55). Second, it paves the
 way for new approaches to search for novel putative associations and
@@
 \(V_a = 2\beta^2f(1-f)\), where \(\beta\) is the log odds ratio and
-\(f\) is the allele frequency of the SNP in the population (55). \(f\)
+\(f\) is the allele frequency of the SNP in the population (56). \(f\)
 can be estimated using the allele frequency of the SNP in an ascertained
@@
 remaining strategies. The default settings for the \texttt{randomForest}
-package in R (56) were used to train the models. MR-MoE 1.0 is
+package in R (57) were used to train the models. MR-MoE 1.0 is
 implemented in the TwoSampleMR R package available at
@@
 
-\hypertarget{ref-Hartwig2017}{}
-7. Hartwig FP, Davey Smith G, Bowden J. Robust Inference In Two-Sample
-Mendelian Randomisation Via The Zero Modal Pleiotropy Assumption.
-bioRxiv {[}Internet{]}. 2017; Available from:
-\url{http://biorxiv.org/content/early/2017/04/10/126102}
+\hypertarget{ref-Hartwig2017a}{}
+7. Hartwig FP, Smith GD, Bowden J. Robust inference in summary data
+Mendelian randomization via the zero modal pleiotropy assumption.
+International Journal of Epidemiology {[}Internet{]}. 2017;dyx102:1--14.
+Available from:
+\href{https://oup.silverchair-cdn.com/oup/backfile/Content\%7B/_\%7Dpublic/Journal/ije/PAP/10.1093\%7B/_\%7Dije\%7B/_\%7Ddyx102/2/dyx102.pdf?Expires=1502303601\%7B/\&\%7DSignature=FC72PSJXMhTtysiYRAtlBTDUthqtIDgOIbjxvoO\%7B~\%7Dn0oiMIEOdXKEpXu7Wxq0EKKPev57T9LPxLzh81sX7\%7B~\%7DYzHu0s6Fd41zYJUfdISx8ev65rchmVp4AMcM}{https://oup.silverchair-cdn.com/oup/backfile/Content\{\textbackslash{}\_\}public/Journal/ije/PAP/10.1093\{\textbackslash{}\_\}ije\{\textbackslash{}\_\}dyx102/2/dyx102.pdf?Expires=1502303601\{\textbackslash{}\&\}Signature=FC72PSJXMhTtysiYRAtlBTDUthqtIDgOIbjxvoO\{\textasciitilde{}\}n0oiMIEOdXKEpXu7Wxq0EKKPev57T9LPxLzh81sX7\{\textasciitilde{}\}YzHu0s6Fd41zYJUfdISx8ev65rchmVp4AMcM}
 
@@
 23. Hu JX, Thomas CE, Brunak S. Network biology concepts in complex
-disease comorbidities. Nature reviews Genetics {[}Internet{]}.
-2016;17(10):615--29. Available from:
-\href{http://www.nature.com/doifinder/10.1038/nrg.2016.87\%7B/\%\%7D5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}{http://www.nature.com/doifinder/10.1038/nrg.2016.87\{\textbackslash{}\%\}5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}
+disease comorbidities. Nat Rev Genet {[}Internet{]}. 2016
+Oct;17(10):615--29. Available from:
+\href{http://dx.doi.org/10.1038/nrg.2016.87\%20http://10.0.4.14/nrg.2016.87}{http://dx.doi.org/10.1038/nrg.2016.87 http://10.0.4.14/nrg.2016.87}
 
@@
 \hypertarget{ref-Paaby2013}{}
-26. Paaby AB, Rockman MV. The many faces of pleiotropy {[}Internet{]}.
-Vol. 29. 2013. pp. 66--73. Available from:
+26. Paaby AB, Rockman MV. The many faces of pleiotropy. Trends in
+Genetics {[}Internet{]}. 2013 Feb;29(2):66--73. Available from:
 \href{http://www.ncbi.nlm.nih.gov/pubmed/23140989\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC3558540\%20http://linkinghub.elsevier.com/retrieve/pii/S0168952512001692}{http://www.ncbi.nlm.nih.gov/pubmed/23140989 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC3558540 http://linkinghub.elsevier.com/retrieve/pii/S0168952512001692}
@@
 
+\hypertarget{ref-Tillmann2017}{}
+44. Tillmann T, Vaucher J, Okbay A, Pikhart H, Peasey A, Kubinova R, et
+al. Education and coronary heart disease: a Mendelian randomization
+study. bioRxiv {[}Internet{]}. 2017; Available from:
+\url{http://www.biorxiv.org/content/early/2017/02/06/106237}
+
 \hypertarget{ref-kotthoff2016auto}{}
-44. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
+45. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
 2.0: Automatic model selection and hyperparameter optimization in WEKA.
@@
 \hypertarget{ref-Zhu2016}{}
-45. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+46. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
 Integration of summary data from GWAS and eQTL studies predicts complex
@@
 \hypertarget{ref-Richardson2017}{}
-46. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
+47. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
 CL, et al. Causal epigenome-wide association study identifies CpG sites
@@
 \hypertarget{ref-Wang2017}{}
-47. Wang L, Michoel T. Efficient And Accurate Causal Inference With
+48. Wang L, Michoel T. Efficient And Accurate Causal Inference With
 Hidden Confounders From Genome-Transcriptome Variation Data. bioRxiv
@@
 \hypertarget{ref-Aalen2015}{}
-48. Aalen OO, Valberg M, Grotmol T, Tretli S. Understanding variation in
+49. Aalen OO, Valberg M, Grotmol T, Tretli S. Understanding variation in
 disease risk: the elusive concept of frailty. International journal of
@@
 \hypertarget{ref-Noyce2017}{}
-49. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
+50. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
 E, et al. Estimating the causal influence of body mass index on risk of
@@
 \hypertarget{ref-Burgess2014a}{}
-50. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+51. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 \hypertarget{ref-Juul2003}{}
-51. Juul K, Tybjaerg-Hansen A, Marklund S, Heegaard NHH, Steffensen R,
+52. Juul K, Tybjaerg-Hansen A, Marklund S, Heegaard NHH, Steffensen R,
 Sillesen H, et al. Genetically Reduced Antioxidative Protection and
@@
 \hypertarget{ref-VonHinke2016}{}
-52. Hinke S von, Davey Smith G, Lawlor D, Propper C, Windmeijer F.
+53. Hinke S von, Davey Smith G, Lawlor D, Propper C, Windmeijer F.
 Genetic markers as instrumental variables. J Health Econ.
@@
 \hypertarget{ref-Staley2016}{}
-53. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
+54. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
 al. PhenoScanner: a database of human genotype--phenotype associations.
@@
 \hypertarget{ref-Vanderweele2014}{}
-54. Vanderweele TJ, Tchetgen Tchetgen EJ, Cornelis M, Kraft P.
+55. Vanderweele TJ, Tchetgen Tchetgen EJ, Cornelis M, Kraft P.
 Methodological challenges in mendelian randomization. Epidemiology
@@
 \hypertarget{ref-Lee2013c}{}
-55. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+56. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 \hypertarget{ref-Liaw2002}{}
-56. Liaw A, Wiener M. Classification and Regression by randomForest. R
+57. Liaw A, Wiener M. Classification and Regression by randomForest. R
 News {[}Internet{]}. 2002;2(3):18--22. Available from:

### 2019-02-21 R/neo4j.r
updated

@@ -73,8 +73,8 @@ modify_rel_headers_for_neo4j <- function(x, id1, id1name, id2, id2name)
 #' @return
 write_split <- function(obj, splitsize, prefix, id1, id1name, id2=NULL, id2name=NULL)
 {
-	splitnum <- round(length(mr) / splitsize)
-	splits <- split(1:length(mr), 1:splitnum)
+	splitnum <- ceiling(length(obj) / splitsize)
+	splits <- split(1:length(obj), 1:splitnum)
 	nsplit <- length(splits)
 	filenames <- paste0(prefix, 1:nsplit, ".csv.gz")
 	lapply(1:length(splits), function(x)
@@ -90,9 +90,9 @@ write_split <- function(obj, splitsize, prefix, id1, id1name, id2=NULL, id2name=
 		gz1 <- gzfile(filenames[x], "w")
 		if(x == 1)
 		{
-			write.csv(temp, file=gz1, row.names=FALSE, na="")
+			write.table(temp, file=gz1, row.names=FALSE, na="", sep=",")
 		} else {
-			write.csv(temp, file=gz1, row.names=FALSE, na="", col.names=FALSE)
+			write.table(temp, file=gz1, row.names=FALSE, na="", col.names=FALSE, sep=",")
 		}
 		close(gz1)
 	})
@@ -121,7 +121,7 @@ write_simple <- function(obj, filename, id1, id1name, id2=NULL, id2name=NULL)
 		temp <- modify_rel_headers_for_neo4j(obj, id1, id1name, id2, id2name)
 	}
 	gz1 <- gzfile(filename, "w")
-	write.csv(temp, file=gz1, row.names=FALSE, na="")
+	write.table(temp, file=gz1, row.names=FALSE, na="", sep=",")
 	close(gz1)
 	return(filename)
 }

### 2019-02-21 scripts/mr-bc4.sh

@@ -5,7 +5,7 @@
 #SBATCH --nodes=1 --cpus-per-task=10 --time=0-12:00:00
 #SBATCH --partition=mrcieu
 #SBATCH --output=job_reports/slurm-%A_%a.out
-#SBATCH --mem=20G
+#SBATCH --mem=15G
 
 echo "Running on ${HOSTNAME}"
 module add languages/r/3.4.4
@@ -18,19 +18,19 @@ fi
 i=${SLURM_ARRAY_TASK_ID}
 # i=$((i + 1000))
 
+cd ${HOME}/mr-eve/mr-eve
+
 
 GWASDIR='../gwas-files'
-OUTPUT="$GWASDIR/$id/derived/instruments/mr.rdata"
 IDLIST="resources/idlist.txt"
 IDINFO="resources/idinfo.rdata"
 RF="resources/rf.rdata"
 THREADS="10"
-
-cd ${HOME}/mr-eve/mr-eve
 ids=($(cat $IDLIST))
+id=`echo ${ids[$i]}`
+OUTPUT="$GWASDIR/$id/derived/instruments/mr.rdata"
 
 echo "Total number of ids: ${#ids[@]}"
-id=`echo ${ids[$i]}`
 echo "Analysing $id"
 if [ -z "$id" ]
 then

### 2019-02-21 scripts/mr.r

@@ -46,6 +46,7 @@ load(args[["rf"]])
 param <- mreve::determine_analyses(id, idlist, newidlist, args[["what"]])
 
 a <- try(mreve::readml(filename, idinfo))
+glimpse(a$exposure_dat)
 if(nrow(a$exposure_dat) == 0)
 {
 	param <- subset(param, exposure != id)
@@ -61,17 +62,18 @@ if(nrow(param) == 0)
 scan <- parallel::mclapply(param$id, function(i)
 {
 	p <- subset(param, id == i)
+	message(p$id)
 	res <- try({
 		if(p$exposure == id)
 		{
 			b <- mreve::readml(file.path(args[["gwasdir"]], p$outcome, "derived/instruments/ml.csv.gz"), idinfo)
-			dat <- TwoSampleMR::harmonise_data(a$exposure_dat, b$outcome_dat)
+			dat <- suppressMessages(TwoSampleMR::harmonise_data(a$exposure_dat, b$outcome_dat))
 		} else {
 			b <- mreve::readml(file.path(args[["gwasdir"]], p$exposure, "derived/instruments/ml.csv.gz"), idinfo)
 			dat <- TwoSampleMR::harmonise_data(b$exposure_dat, a$outcome_dat)
 		}
-		res <- TwoSampleMR::mr_wrapper(dat)
-		res <- TwoSampleMR::mr_moe(res, rf)
+		res <- suppressMessages(TwoSampleMR::mr_wrapper(dat))
+		res <- suppressMessages(TwoSampleMR::mr_moe(res, rf))
 		rm(b,dat)
 		res
 	})

### 2019-02-21 scripts/prepare_neo4j.r

@@ -60,6 +60,7 @@ for(i in 1:nrow(idinfo))
 	x <- idinfo$id[i]
 	message(x, " - ", i, " of ", nrow(idinfo))
 	a <- readml(file.path("../gwas-files", x, "derived/instruments/ml.csv.gz"), idinfo, format="none")
+	a <- subset(a, SNP %in% vars$rsid)
 	a$bgcid <- x
 	vt[[i]] <- subset(a, select=c(SNP, bgcid, beta, se, pval, eaf, samplesize, ncase, ncontrol))
 	vtinst[[i]] <- subset(vt[[i]], a$instrument)
@@ -68,7 +69,7 @@ for(i in 1:nrow(idinfo))
 
 vtfiles <- write_split(vt, 200, "resources/neo4j_stage/vt", "SNP", "variant", "bgcid", "bgcid")
 vtinst <- bind_rows(vtinst)
-vtinstfile <- write_split("resources/neo4j_stage/instruments.csv.gz", "SNP", "variant", "bgcid", "bgcid")
+vtinstfile <- write_simple(vtinst, "resources/neo4j_stage/instruments.csv.gz", "SNP", "variant", "bgcid", "bgcid")
 
 
 # mr
@@ -88,21 +89,27 @@ for(i in 1:nrow(idinfo))
 	message(x, " - ", i, " of ", nrow(idinfo))
 	load(file.path("../gwas-files", x, "derived/instruments/mr.rdata"))
 
-	estimates <- lapply(scan, function(x) x$estimates) %>% bind_rows
+	estimates <- lapply(scan, function(x) {
+		if(is.null(x$estimates)) return(NULL)
+		estimates <- x$estimates
+		if("MOE" %in% names(estimates))
+		{
+			estimates <- dplyr::select(estimates, -c(method2, steiger_filtered, outlier_filtered))
+		} else {
+			estimates$selection <- "Tophits"
+			estimates$selection[estimates$steiger_filtered & estimates$outlier_filtered] <- "DF + HF"
+			estimates$selection[!estimates$steiger_filtered & estimates$outlier_filtered] <- "HF"
+			estimates$selection[estimates$steiger_filtered & !estimates$outlier_filtered] <- "DF"
+			estimates <- estimates %>% dplyr::select(-c(steiger_filtered, outlier_filtered))
+			estimates$MOE <- 0
+			ind <- estimates$method %in% c("Wald ratio", "FE IVW", "Steiger null") & estimates$selection == "DF"
[…]

### 2019-02-21 scripts/mr-bc4.sh
upd

@@ -10,24 +10,25 @@
 echo "Running on ${HOSTNAME}"
 module add languages/r/3.4.4
 
-if [ -n "${1}" ]; then
-  echo "${1}"
-  SLURM_ARRAY_TASK_ID=${1}
-fi
-
-i=${SLURM_ARRAY_TASK_ID}
-# i=$((i + 1000))
-
 cd ${HOME}/mr-eve/mr-eve
 
-
 GWASDIR='../gwas-files'
-IDLIST="resources/idlist.txt"
 IDINFO="resources/idinfo.rdata"
 RF="resources/rf.rdata"
 THREADS="10"
+IDLIST="resources/idlist.txt"
 ids=($(cat $IDLIST))
-id=`echo ${ids[$i]}`
+
+
+if [ -n "${1}" ]; then
+	echo "${1}"
+	id=$1
+else
+	i=${SLURM_ARRAY_TASK_ID}
+	# i=$((i + 1000))
+	id=`echo ${ids[$i]}`
+fi
+
 OUTPUT="$GWASDIR/$id/derived/instruments/mr.rdata"
 
 echo "Total number of ids: ${#ids[@]}"

### 2019-02-21 scripts/prepare_neo4j-bc4.sh

@@ -1,0 +1,15 @@
+#!/bin/bash
+
+#SBATCH --job-name=neo4j
+#SBATCH --array=0
+#SBATCH --nodes=1 --cpus-per-task=10 --time=0-12:00:00
+#SBATCH --partition=mrcieu
+#SBATCH --output=job_reports/slurm-%A_neo4j.out
+#SBATCH --mem=15G
+
+echo "Running on ${HOSTNAME}"
+module add languages/r/3.4.4
+
+cd ${HOME}/mr-eve/mr-eve
+
+time Rscript scripts/prepare_neo4j.r

### 2019-02-21 scripts/prepare_neo4j.r

@@ -104,6 +104,12 @@ for(i in 1:nrow(idinfo))
 			estimates$MOE <- 0
 			ind <- estimates$method %in% c("Wald ratio", "FE IVW", "Steiger null") & estimates$selection == "DF"
 			estimates$MOE[ind] <- 1
+			ind <- is.infinite(estimates$b)
+			estimates$b[ind] <- 0
+			estimates$se[ind] <- NA
+			estimates$ci_low[ind] <- NA
+			estimates$ci_upp[ind] <- NA
+			estimates$pval[ind] <- 1
 		}
 		estimates
 	}) %>% bind_rows
@@ -151,18 +157,20 @@ for(i in 1:nrow(idinfo))
 	}
 }
 
+mrhetfiles <- write_simple(mrhet, 200, "resources/neo4j_stage/mrhet", "id.exposure", "bgcid", "id.outcome", "bgcid")
+mrinterceptfiles <- write_simple(mrintercept, 200, "resources/neo4j_stage/mrintercept", "id.exposure", "bgcid", "id.outcome", "bgcid")
+mrmoefiles <- write_simple(mrmoe, "resources/neo4j_stage/mrmoe", "id.exposure", "bgcid", "id.outcome", "bgcid")
+metricsfiles <- write_simple(metrics, "resources/neo4j_stage/metrics", "id.exposure", "bgcid", "id.outcome", "bgcid")
 mrfiles <- write_split(mr, 200, "resources/neo4j_stage/mr", "id.exposure", "bgcid", "id.outcome", "bgcid")
-mrhetfiles <- write_split(mrhet, 200, "resources/neo4j_stage/mrhet", "id.exposure", "bgcid", "id.outcome", "bgcid")
-mrinterceptfiles <- write_split(mrintercept, 200, "resources/neo4j_stage/mrintercept", "id.exposure", "bgcid", "id.outcome", "bgcid")
-mrmoefiles <- write_split(mrmoe, 200, "resources/neo4j_stage/mrmoe", "id.exposure", "bgcid", "id.outcome", "bgcid")
-metricsfiles <- write_split(metrics, 200, "resources/neo4j_stage/metrics", "id.exposure", "bgcid", "id.outcome", "bgcid")
 
 
 
+
+system("rm -rf ~/mr-eve/neo4j/neo4j-enterprise-3.5.3/data/databases/graph.db")
 
 cmd <- paste0(
[…]

### 2019-02-22 scripts/00initialise.r
renaming to makemreve

@@ -1,2 +1,2 @@
-library(mreve)
+library(makemreve)
 setup_inventory()

### 2019-02-22 scripts/initialise_analysis.r

@@ -1,7 +1,7 @@
 stopifnot(basename(getwd()) == "mr-eve")
 
 
-library(mreve)
+library(makemreve)
 library(argparse)
 library(data.table)
 

### 2019-02-22 scripts/mr.r

@@ -1,6 +1,6 @@
 suppressWarnings(suppressPackageStartupMessages({
 	library(TwoSampleMR)
-	library(mreve)
+	library(makemreve)
 	library(tidyverse)
 	library(car)
 	library(randomForest)

### 2019-02-22 scripts/prepare_neo4j.r

@@ -1,5 +1,5 @@
 # Make trait, snp and gene nodes
-library(mreve)
+library(makemreve)
 library(tidyverse)
 library(data.table)
 library(GenomicRanges)

### 2019-02-22 shiny/server.r

@@ -10,9 +10,9 @@ escape_quote <- function(text)
 get_query_trait <- function(graph, trait)
 {
 	query <- paste0("match 
-		(n:trait {name: '", escape_quote(trait), "'})
-		-[r:MRB]-
-		(m:trait)
+		(n:TRAIT {trait: '", escape_quote(trait), "'})
+		-[r:MR]-
+		(m:TRAIT)
 
 		return n,r,m order by r.P limit 200"
 	)
@@ -22,9 +22,9 @@ get_query_trait <- function(graph, trait)
 get_query_exposure_outcome <- function(graph, exposure, outcome)
 {
 	query <- paste0("match 
-		(g:gene)-[rg:GS]->(s:snp)-[rs:GA]->(n:trait {name: '", escape_quote(exposure), "'})
+		(g:GENE)-[rg:ANNOTATION]->(s:VARIANT)-[rs:INSTRUMENT]->(n:TRAIT {trait: '", escape_quote(exposure), "'})
 		-[r:MR]->
-		(m:trait {name: '", escape_quote(outcome), "'}) 
+		(m:TRAIT {trait: '", escape_quote(outcome), "'}) 
 		return n,r,m,rg,g,s,rs"
 	)
 	return(query)
@@ -33,18 +33,17 @@ get_query_exposure_outcome <- function(graph, exposure, outcome)
 get_table_exposure_outcome <- function(graph, exposure, outcome)
 {
 	query <- paste0("match 
-		(n:trait {name: '", escape_quote(exposure), "'})
+		(n:TRAIT {trait: '", escape_quote(exposure), "'})
 		-[r:MR]->
-		(m:trait {name: '", escape_quote(outcome), "'}) 
+		(m:TRAIT {trait: '", escape_quote(outcome), "'}) 
 		return
-		r.Method as Method, 
-		r.instruments as instruments,
+		r.method+' - '+r.selection as Method, 
 		r.nsnp as nsnp,
-		r.Estimate as Estimate, 
-		r.SE as SE,
-		r.CI_low as CI_low,
-		r.CI_upp as CI_upp,
-		r.P as P"
+		r.b as Estimate, 
+		r.se as SE,
+		r.ci_low as CI_low,
+		r.ci_upp as CI_upp,
+		r.pval as P"
 	)
 	return(cypher(graph, query))
 }
@@ -55,23 +54,23 @@ get_table_exposure_outcome <- function(graph, exposure, outcome)
 get_trait_scan <- function(graph, trait)
 {
 	query <- paste0("match 
[…]

### 2019-02-22 shiny/ui.r

@@ -3,18 +3,18 @@ library(shiny)
 library(shinydashboard)
 
 
-graph <- startGraph("http://shark.epi.bris.ac.uk:7474/db/data", username="neo4j", password="123qwe")
+graph <- startGraph("http://shark.epi.bris.ac.uk:8474/db/data", username="neo4j", password="123qwe")
 
 
-browser_url <- "http://shark.epi.bris.ac.uk:7474/browser/"
+browser_url <- "http://shark.epi.bris.ac.uk:8474/browser/"
 
 # a <- getLabeledNodes(graph, "trait") # this is so slow
-# b <- getNodes(graph, "match(n:trait) return n")
+# b <- getNodes(graph, "match(n:TRAIT) return n")
 # nodes <- sapply(a, function(x) x$name)
-nodes <- cypher(graph, "match (n:trait) return n.name")$n.name
-snps <- cypher(graph, "match(n:snp) return n.name")$n.name
-consortia <- cypher(graph, "match (n:trait) return n.consortium")$n.consortium
-mrest <- cypher(graph, "match (n)-[r:MRB]-(m) return count(r) as count")$count
+nodes <- cypher(graph, "match (n:TRAIT) return n.trait")$n.trait
+snps <- cypher(graph, "match(n:VARIANT) return n.variantId")$n.variantId
+consortia <- cypher(graph, "match (n:TRAIT) return n.consortium")$n.consortium
+mrest <- cypher(graph, "match (n)-[r:MRMOE]-(m) return count(r) as count")$count
 
 
 load("outcome_nodes.rdata")
@@ -48,9 +48,9 @@ dashboardPage(
 				),
 				column(6,
 					infoBox(width=12, "Phenotypes", length(nodes), icon=icon("database")),
-					infoBox(width=12, "MR estimates", mrest, icon = icon("random")),
+					infoBox(width=12, "MR analyses", mrest, icon = icon("random")),
 					infoBox(width=12, "Instrumental variables", length(snps), icon=icon("database")),
-					infoBox(width=12, "MR methods", "29", icon=icon("cogs")),
+					infoBox(width=12, "MR methods", "45", icon=icon("cogs")),
 					infoBox(width=12, "GWAS consortia", length(unique(consortia)), icon=icon("globe"))
 				)
 			),

### 2019-02-23 shiny/server.r
updated readme and moving shiny to separate repo

@@ -1,149 +1,0 @@
-library(RNeo4j)
-library(shiny)
-
-escape_quote <- function(text)
-{
-	gsub("'", "\\\\'", text)
-}
-
-
-get_query_trait <- function(graph, trait)
-{
-	query <- paste0("match 
-		(n:TRAIT {trait: '", escape_quote(trait), "'})
-		-[r:MR]-
-		(m:TRAIT)
-
-		return n,r,m order by r.P limit 200"
-	)
-	return(query)
-}
-
-get_query_exposure_outcome <- function(graph, exposure, outcome)
-{
-	query <- paste0("match 
-		(g:GENE)-[rg:ANNOTATION]->(s:VARIANT)-[rs:INSTRUMENT]->(n:TRAIT {trait: '", escape_quote(exposure), "'})
-		-[r:MR]->
-		(m:TRAIT {trait: '", escape_quote(outcome), "'}) 
-		return n,r,m,rg,g,s,rs"
-	)
-	return(query)
-}
-
-get_table_exposure_outcome <- function(graph, exposure, outcome)
-{
-	query <- paste0("match 
-		(n:TRAIT {trait: '", escape_quote(exposure), "'})
-		-[r:MR]->
-		(m:TRAIT {trait: '", escape_quote(outcome), "'}) 
-		return
-		r.method+' - '+r.selection as Method, 
-		r.nsnp as nsnp,
-		r.b as Estimate, 
-		r.se as SE,
-		r.ci_low as CI_low,
-		r.ci_upp as CI_upp,
-		r.pval as P"
-	)
-	return(cypher(graph, query))
-}
-
-# get_table_exposure_outcome(graph, "Body mass index", "Parkinson's disease")
-
-
-get_trait_scan <- function(graph, trait)
-{
-	query <- paste0("match 
-		(n:TRAIT)
-		-[r:MRMOE]->
-		(m:TRAIT)
-
-		where (n.trait = '", escape_quote(trait), "' and m.trait <> '", escape_quote(trait), "') or (m.trait = '", escape_quote(trait), "' and n.trait <> '", escape_quote(trait), "')
-
-		return
-
-		n.trait as Exposure,
-		m.trait as Outcome,
-		r.method+' - '+r.selection as Method, 
-		r.nsnp as nsnp,
-		r.b as Estimate, 
-		r.se as SE,
-		r.ci_low as CI_low,
-		r.ci_upp as CI_upp,
[…]

### 2019-02-23 shiny/ui.r

@@ -1,124 +1,0 @@
-library(RNeo4j)
-library(shiny)
-library(shinydashboard)
-
-
-graph <- startGraph("http://shark.epi.bris.ac.uk:8474/db/data", username="neo4j", password="123qwe")
-
-
-browser_url <- "http://shark.epi.bris.ac.uk:8474/browser/"
-
-# a <- getLabeledNodes(graph, "trait") # this is so slow
-# b <- getNodes(graph, "match(n:TRAIT) return n")
-# nodes <- sapply(a, function(x) x$name)
-nodes <- cypher(graph, "match (n:TRAIT) return n.trait")$n.trait
-snps <- cypher(graph, "match(n:VARIANT) return n.variantId")$n.variantId
-consortia <- cypher(graph, "match (n:TRAIT) return n.consortium")$n.consortium
-mrest <- cypher(graph, "match (n)-[r:MRMOE]-(m) return count(r) as count")$count
-
-
-load("outcome_nodes.rdata")
-outcome_nodes <- iconv(outcome_nodes$trait, sub='')
-
-dashboardPage(
-
-	dashboardHeader(title="MR-EvE"),
-
-	dashboardSidebar(
-		sidebarMenu(
-			menuItem("About", tabName = "about", icon=icon("smile-o")),
-			menuItem("Basic lookup", tabName = "basic", icon=icon("cogs")),
-			menuItem("Trait scan", tabName = "trait", icon=icon("cogs"))
-		)
-	),
-
-	dashboardBody(
-		tabItems(
-			tabItem(tabName = "about",
-				column(6,
-					box(title="MR of everything vs everything", width = 12,
-						p("This is a repository of Mendelian randomisation estimates that have been generated using the GWAS summary data in the MR-Base database. It can be used to quickly browse specific phenotypes or lookup specific causal hypotheses."),
-						p("You can find out more about MR-Base at ", tags$a("http://mrbase.org", "http://www.mrbase.org")),
-						p("Important: Mendelian randomization is not a panacea for causal inference. Methodology is still required to reduce sensitivity to violations of assumptions, and to incorporate information about the biological feasibility of causal associations.")
-					),
-					box(width=12,
-						p(tags$strong("Alpha phase release: "), "strictly experimental"),
-						p(tags$strong("Last update: "), "3rd July 2017")
-					)
-				),
-				column(6,
-					infoBox(width=12, "Phenotypes", length(nodes), icon=icon("database")),
-					infoBox(width=12, "MR analyses", mrest, icon = icon("random")),
[…]

### 2019-02-24 analysis/horizontal_vs_vertical.r
updated

@@ -1,0 +1,126 @@
+library(ggplot2)
+library(ggrepel)
+library(dplyr)
+library(mrever)
+
+
+# compare the magnitude of vertical vs horizontal pleiotropy
+
+simple_r2 <- function(b, se, n)
+{
+	f <- b^2 / se^2
+	r2 <- f / (n + f - 2)
+	return(r2)
+}
+
+estimate_trait_sd <- function(b, se, n, p)
+{
+	z <- b / se
+	standardised_bhat <- sqrt((z^2/(z^2+n-2)) / (2 * p * (1-p))) * sign(z)
+	estimated_sd <- b / standardised_bhat
+	return(median(estimated_sd, na.rm=T))
+}
+
+adjr2 <- function(r2, n)
+{
+	1 - (1 - sum(r2)) * (n - 1) / (n - length(r2) - 1)
+}
+
+
+calc_plei <- function(exp, exp_rep, out, adj=FALSE)
+{
+	a <- get_instruments(exp)	
+	arep <- get_genassoc(a$variantId, exp_rep)
+	b <- get_genassoc(a$variantId, out)
+	biv <- get_mr(exp, out)$mr %>% filter(!is.na(moescore)) %>% arrange(desc(moescore)) %>% slice(1) %>% .$b
+	arep$rsq <- simple_r2(arep$beta, arep$se, arep$samplesize)
+	b$rsq <- simple_r2(b$beta, b$se, b$samplesize)
+	if(all(is.na(b$eaf)))
+	{
+		temp <- merge(a, b, by="variantId")
+		vy <- estimate_trait_sd(temp$beta.y, temp$se.y, temp$samplesize.y, temp$eaf.x)^2
+	} else {
+		vy <- estimate_trait_sd(b$beta, b$se, b$samplesize, b$eaf)^2
+	}
+	ivr2 <- biv^2 * estimate_trait_sd(a$beta, a$se, a$samplesize, a$eaf)^2 / vy
+
+	if(adj)
+	{
+		totalr2 <- adjr2(b$rsq, median(b$samplesize))
+		mediatedr2 <- adjr2(arep$rsq * ivr2, median(b$samplesize))
+	} else {
+		mediatedr2 <- sum(arep$rsq * ivr2, na.rm=TRUE)
+		totalr2 <- sum(b$rsq, na.rm=TRUE)
+	}
+
[…]

### 2019-02-24 analysis/broad_characteristics.rmd
updated

---
title: Broad characteristics
author: Gibran Hemani
date: 23/02/2019
---

```{r}
library(mrever)
connect()
library(dplyr)
library(data.table)
library(knitr)
library(ggplot2)

traits <- get_traits()
o <- list()
for(i in 1:8)
{
	message(i)

	a <- fread(paste0("zcat ../resources/neo4j_stage/mr", i, ".csv.gz"), he=i == 1, sep=",")
	if(i != 1) names(a) <- names(o[[1]])[-length(names(o[[1]]))]
	names(a) <- gsub(":FLOAT", "", names(a))
	names(a) <- gsub(":INT", "", names(a))
	names(a)[1:2] <- c("exposure", "outcome")
	a$id <- paste(a$exposure, a$outcome)
	a$exposure <- as.character(a$exposure)
	a$outcome <- as.character(a$outcome)
	k <- subset(a, method == "FE IVW" & selection == "Tophits" & nsnp > 5) %>% .$id
	a <- subset(a, paste(exposure, outcome) %in% k)
	o[[i]] <- a
}
o <- bind_rows(o)

moe1 <- subset(o, method == "Steiger null") %>% subset(!duplicated(id))
moe2 <- filter(o, ! id %in% moe1$id) %>% group_by(id) %>% arrange(desc(moescore)) %>% slice(1)

moe <- bind_rows(moe1, moe2)
table(moe$method, moe$selection)
table(moe$method)

moe$z <- abs(moe$b / moe$se)
moe$z[moe$z > 10] <- 10
table(z > 100)

table(a$method, a$selection)

ggplot(moe, aes(x=z)) +
geom_histogram(aes(fill=selection), position="dodge") +
facet_grid(method ~ .) +
theme(strip.text = element_text(angle=0))

```

## 1. Reverse direction results

```{r}

b <- subset(a, id %in% subset(a, method == "Steiger null")$id)
bsig <- subset(b, method == "RE IVW" & selection == "HF" & p.adjust(pval, "fdr") < 0.05)
bsig <- merge(bsig, subset(traits, select=c(bgcidId, trait)), by.x="exposure", by.y="bgcidId")
bsig <- merge(bsig, subset(traits, select=c(bgcidId, trait)), by.x="outcome", by.y="bgcidId")
subset(bsig, select=c(trait.x, trait.y)) %>% kable
```

## 2.

## 1. Compare causal effect estimates across ivw, ivw HF.

What are notable discrepancies?

2. Compare

### 2019-02-25 analysis/broad_characteristics.rmd
upd

@@
 library(ggplot2)
+library(tidyr)
 
@@
 
-## 2.
+## 2. Instrument selection
+
+```{r}
+
+library(data.table)
+library(dplyr)
+library(tidyr)
+library(mrever)
+library(ggplot2)
+connect()
+
+b <- subset(o, method %in% c("FE IVW", "Wald ratio", "Steiger null"))
+
+traits <- get_traits() %>% dplyr::select(c(bgcidId, sample_size, trait, nsnp))
+out <- group_by(b, exposure, outcome, selection) %>% 
+summarise(instruments = first(nsnp)) %>% 
+inner_join(., traits, by=c("exposure"="bgcidId"))
+
+p <- tidyr::spread(o, key=selection, value=instruments)
+
+save(out, file="out.rdata")
+
+ggplot(out, aes(x=sample_size, y=instruments)) +
+# geom_point(aes(colour=selection)) +
+geom_smooth(aes(colour=selection))
+
+out$n2 <- out$sample_size
+out$n2[out$n2 > 500000] <- NA
+
+out$nbin <- cut(out$n2, 10)
+table(out$nbin)
+
+ggplot(subset(out, !is.na(n2)), aes(x=nbin, y=instruments)) +
+geom_boxplot(aes(fill=selection), position="dodge", outlier.shape=NA) +
+scale_fill_brewer(type="qual") +
+theme(axis.text.x=element_text(angle=90))
+
+temp <- subset(out, selection %in% c("DF", "Tophits")) %>% 
+group_by(exposure, outcome) %>%
+summarise(nrev = abs(diff(instruments)))
+
+temp <- tidyr::spread(out, key=selection, value=instruments)
+temp$nrev <- temp$Tophits - temp$DF
+summary(lm(nrev ~ I(n2/10000), temp))
+
+## Import result:
+
+temp$prev <- temp$nrev / temp$Tophits
+summary(lm(prev ~ I(n2/100000), temp))
+summary(temp$prev)
+
+# Across 758955 MR analyses, for every 100k increase in sample size the proportion of instruments that are reverse causal increases by 2.8% (s.e. 0.023, p < 2e-16)
+
+ggplot(temp, aes(x=n2, y=prev)) +
+geom_point(alpha=0.4) +
+geom_smooth() +
+geom_smooth(method="lm")
+
+ggplot(temp %>% filter(!is.na(n2)), aes(x=nbin, y=prev)) +
+geom_boxplot()
+
+temp2 <- group_by(temp, exposure, n2) %>%
+summarise(
+	mdf = mean(Tophits - DF), pmdf = mdf / max(Tophits),
+	mhf = mean(Tophits - HF), pmhf = mhf / max(Tophits)
+)
+
+summary(lm(pmdf ~ n2, temp2))
+summary(lm(pmhf ~ n2, temp2))
+
+ggplot(temp2 %>% , aes(x=))
+
+summary(lm(nrev ~ I(n2/1000), temp))
+
+```
 

### 2019-02-27 analysis/ivr2.rmd
updated

---
title: Using the graph
---

1. Trait info is problematic
2. Method is not really applicable when fewer than say 5 SNPs

1. Find instruments for x
2. See if x associates with any other X
3. Get all instruments for X and make an "exposure set"
4. For x-y estimate remove y from exposure set
5.

1. Get list of independent traits
2.

Overcoming reverse causal instruments
- how many reverse causal instruments are there??

Predicting the parameter space improves over single methods

Application to summary data reveals extensive connectedness throughout the phenome

Horizontal pleiotropy increases as more instruments are discovered

```{r}
# library(mrever)
library(igraph)
connect()

mrever_to_mr <- function(id1, id2)
{
	a <- TwoSampleMR::format_data(
		get_instruments(id1),
		snp_col="variantId",
		effect_allele_col="alt",
		other_allele_col="ref",
		id_col="bgcidId",
		units_col="unit"
	)
	b <- TwoSampleMR::format_data(
		get_genassoc(a$SNP, id2),
		type="outcome",
		snp_col="variantId",
		effect_allele_col="alt",
		other_allele_col="ref",
		id_col="bgcidId",
		units_col="unit"
	)
	return(TwoSampleMR::harmonise_data(a,b))
}

fill_in_sd <- function(traits)
{

	estimate_trait_sd <- function(b, se, n, p)
	{
		z <- b / se
		standardised_bhat <- sqrt((z^2/(z^2+n-2)) / (2 * p * (1-p))) * sign(z)
		estimated_sd <- b / standardised_bhat
		return(median(estimated_sd, na.rm=T))
	}
	# table(is.na(traits$sd), traits$unit) %>%t
	traits$unit[is.na(traits$unit)] <- "log odds"

	# For the time being just treat logor as continuous
	traits$SD <- traits$sd
	# table(is.na(traits$SD), traits$unit) %>% t
	traits$unit[traits$unit == "NA" & traits$ncase != 0] <- "risk difference"
	traits$unit[traits$unit == "NA" & traits$ncase == 0] <- "SD"
	traits$SD[grepl("SD", traits$unit)] <- 1

	ind <- traits$unit == "risk difference" & traits$ncontrol == 0
	traits$ncontrol[ind] <- traits$sample_size[ind] - traits$ncase[ind]

	ind <- traits$unit == 'risk difference'
	# traits$SD[ind] <- (traits$ncase[ind] / traits$sample_size[ind]) * (traits$ncontrol[ind] / traits$sample_size[ind])
	traits$SD[ind] <- NA

	rsidlist <- get_instruments(2)$variantId
	todo <- subset(traits, is.na(SD))$bgcidId
	for(i in 1:length(todo))
	{
		message(i)
		inst <- get_instruments(todo[i])
		if(is.null(inst)) inst <- get_genassoc(rsidlist, todo[i])
		miseaf <- inst$variantId[is.na(inst$eaf)]
		if(length(miseaf) > 0)
		{
			eaf <- get_genassoc(miseaf, "UKB-a:100") %>% dplyr::select(variantId, eaf) %>% filter(variantId %in% inst$variantId)
			inst$eaf[match(eaf$variantId, inst$variantId)] <- as.numeric(eaf$eaf)
			inst$eaf <- as.numeric(inst$eaf)
		}
		s <- estimate_trait_sd(inst$beta, inst$se, inst$samplesize, inst$eaf)
		message(s)
		traits$SD[traits$bgcidId == todo[i]] <- s
	}	
	traits$sd <- traits$SD
	return(traits)

}
```

```{r}
# l <- list()
# for(i in 1:nrow(traits))
# {
# 	inst <- get_instruments(traits$bgcidId[i])
# 	if(is.null(inst))
# 		next
# 	cands <- genassocscan(inst$variantId, 0.05 / length(traits) / nrow(inst))

# }

traits <- get_traits() %>% fill_in_sd(.)
traitst <- subset(traits, unit != "mm3")

# Do phewas of everything
l <- lapply(1:nrow(traits), function(i) phewas(traits$bgcidId[i]))
l[sapply(l, is.null)] <- NULL

# Remove traits that have fewer than 80% of results
l[sapply(l, function(x) nrow(x) < length(l) * 0.8)] <- NULL

phe <- bind_rows(l)
save(phe, file="phewas.rdata")

idlist <- unique(phe$exposure)
phe <- subset(phe, outcome %in% exposure)
table(phe$method)
phe$se[is.na(phe$se)] <- 1
phe$se <- abs(phe$se)
phe$z <- pmin(abs(phe$b / phe$se), 10)

tr <- bind_cols(tibble(bgcidId=traits$bgcidId), subset(traits, select=-c(bgcidId))) %>% filter(bgcidId %in% c(l$exposure))

gr <- graph_from_data_frame(subset(l, z > qnorm(1e-16, low=F), select=c(exposure, outcome)), directed=TRUE, vertices = tr)
E(gr)$weight <- subset(l, z > qnorm(1e-16, low=F))$z

hist(l$z)
```

estimate_ivr2 <- function(l, traits)
{
	indx <- match(l$exposure, traits$bgcidId)
	indy <- match(l$outcome, traits$bgcidId)
	l$ivr2 <- l$b^2 * traits$sd[indx]^2 / traits$sd[indy]^2
	return(l)
}

o <- estimate_ivr2(phe, traits)
o <- subset(o, pval < 1e-6)
subset(o, ivr2 > 10) %>% arrange(desc(ivr2)) %>% head

hist(o$ivr2)

o <- subset(o, exposure %in% traitst$bgcidId & outcome %in% traitst$bgcidId)

o %>% arrange(desc(ivr2)) %>% head
hist($ivr2)

wc <- cluster_walktrap(gr, steps=1, weights = E(gr)$weight)
com <- communities(wc)

hs <- hub_score(gr)$vector

plot(gr)

l <- subset(l, )

com <- communities(gr)
